### 2020-06-21 LaTeX/Ditch detection in high-resolution digital elevation models.tex
Adds work on article as well as rejoinder.

% interacttfvsample.tex
% v1.02 - September 2016

\documentclass[]{interact}

%Custom packages---------
\usepackage{hyperref}

\usepackage{floatrow}

\usepackage{booktabs}

\usepackage{float}

\usepackage{mathastext}

\floatstyle{plaintop}
\restylefloat{table}

\usepackage[parfill]{parskip}

%-------------------

\usepackage{epstopdf}% To incorporate .eps illustrations using PDFLaTeX, etc.
\usepackage{subfigure}% Support for small, `sub' figures and tables

\usepackage{natbib}% Citation support using natbib.sty
\bibpunct[, ]{(}{)}{,}{a}{}{,}% Citation support using natbib.sty
\renewcommand\bibfont{\fontsize{10}{12}\selectfont}% Bibliography support using natbib.sty

\theoremstyle{plain}% Theorem-like structures
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}

\theoremstyle{remark}
\newtheorem{remark}{Remark}
\newtheorem{notation}{Notation}

\begin{document}

\title{Ditch detection in high-resolution digital elevation models}

\author{
\name{Filip Andersson\textsuperscript{a,1}, Jonatan Flyckt\textsuperscript{a,1}, Niklas Lavesson\textsuperscript{a}\thanks{CONTACT Niklas Lavesson. Email: niklas.lavesson@ju.se}, Liselott Nilsson\textsuperscript{b}, and Anneli \AA gren\textsuperscript{c}} \affil{\textsuperscript{a}J\"onk\"oping Artificial Intelligence Laboratory, J\"onk\"oping University, Gjuterigatan 5, 551 11, J\"onk\"oping, Sweden; \textsuperscript{b}Forest Department, Swedish Forest Agency, Skeppargatan 17, 931 32 Skellefte\aa, Sweden; \textsuperscript{c}Department of Forest Ecology and Management, Swedish University of Agricultural Sciences, SLU, Skogsmarksgr\"and 17, 901 83, Ume\aa, Sweden}
\footnote{\textsuperscript{1} Filip Andersson and Jonatan Flyckt contributed equally to this work as first authors.}}

\maketitle

\begin{abstract}
In this study, a method was developed for detecting ditches in high resolution digital elevation models derived from LiDAR scans.

Thresholding methods using digital terrain indices, such as Sky View Factor, Impoundment Index, or High Pass Median Filter can be used to detect ditches. However, a single threshold generally does not capture the variability in the landscape, generating many false positives and negatives. We hypothesised that, by combining the digital terrain indices using a machine learner, we could improve the ditch detection.

In addition to the raw terrain indices, additional input variables were generated by transforming the data by including neighbouring cells to improve the predictability of the ditches. Random Forests was used to locate the ditches, and the probability output from this classifier was processed to remove noise, and binarised to produce the final ditch prediction.

The confidence interval for the Cohen's Kappa index ranged \textbf{[0.693 , 0.786]} between the evaluation plots with a confidence level of \textbf{95\%}. Input variables based on the Impoundment Index overall represented the strongest ditch predictors.

The study shows that by combining the information in several digital terrain indices using machine learning, more ditch networks could be automatically retrieved.
\end{abstract}

\begin{keywords}
Machine learning; Geographic information systems; Classification and regression trees; Supervised learning by classification
\end{keywords}

\section{Introduction} \label{introduction}

Over the last century many of the wetlands in North Europe and parts of North America have been ditched. Mapping and modelling the hydrology of a flat landscape is challenging because of the low topographic gradients. In many places both the landscape and the channel network have also been heavily modified by humans, especially in agricultural landscapes, for example in mid-western United States \citep{passalacqua}. In North Europe, ditches have also been dug to turn flat peatlands into productive forests, thereby profoundly altering the hydrology in North Europe \citep{peatlands}. Sweden is one of the North European countries with the most ditches; over five million hectares of peatland have been drained for forestry in the 20th century. This has resulted in almost one million km of ditches, which has more than doubled the length of watercourses in Sweden \citep{hasselquist}.

A comparison between a field inventory of ditches (within the National Inventory of Landscapes in Sweden (NILS)) and current maps, shows that approximately 90 \% of the ditch networks are missing on current maps ({\AA}gren, unpublished). To better prioritise the restoration of ditched wetlands, it is necessary to know where the ditches are located, but field inventories are often too costly. The research objective of this study is therefore to automatically detect ditches from high-resolution digital elevation models (DEMs) using digital terrain indices. Previous attempts \citep{uppsala} have shown that methods such as Sky View Factor \citep{zaksek}, Impoundment Index \citep{whiteboxtools} and High Pass Median Filter \citep{whiteboxtools} can be used to detect the ditches. However, all methods yield false negatives, and Sky View Factor tends to give a large amount of false positives in steep terrain, whereas Impoundment Index produces more false positives in flat terrain if volume is used as a measure, and will falsely detect natural stream channels if dam height is used as a measure. High Pass Median Filter tends to also detect small sinks as ditches. This is in part due to the Light Detection and Ranging (LiDAR) measurements sometimes being interfered by trees, bushes, or grass - leaving gaps in the ditch model, or sometimes due to a poorly set threshold. Therefore, using solely the digital terrain indices for ditch mapping tends to yield an imprecise result.

In this study, we introduce new methods for the automated mapping of ditches. From the high resolution DEM, we use machine learning to locate the ditch networks. We also examine which digital terrain indices are the most relevant for a correct ditch prediction. To accomplish this, we conduct experiments using a supervised learning algorithm with labelled ditch data to automatically extract ditches from externally supplied occurrences (data not before seen by the model) \citep{kotsiantis}. These occurrences are used as input variables to the algorithm, and these input variables consist of refined or raw digital terrain indices derived from a DEM.

There are other studies where geographical data has been used with machine learning to map ditches. \citet{roelens} used the Random Forests algorithm to detect ditches in Belgium, using LiDAR point clouds. They used several different input variables where neighbouring points of a specific LiDAR point were examined. Values of all LiDAR attributes were represented for the neighbouring area for both 0.5, 1, 1.5, 2, 3 and 4 metre radii. We reproduce these neighbouring area variables when building the model in our study. In a pilot study, we compared several different algorithms (Random Forests, XGBoost, Naive Bayes, and Support Vector Machines), and it was found that Random Forests produced the most accurate results. Random Forests also computes Mean Decrease in Impurity, or Gini importance, highlighting what input variables are the most important for a given prediction \citep{gini}. Because of these factors, and the fact that it proved to be suitable for \citet{roelens}, a Random Forests model is used to locate the ditches.

We hypothesise that by combining the information from all the digital terrain indices using a machine learner (Random Forests), we can improve the detection of the ditches. To test this, we conduct a study using a high resolution DEM of the Krycklan Catchment in Sweden \citep{krycklancatchment},  where a high precision manual mapping has been performed. This mapping was used to both train the Random Forests model and to ground-truth the models. To evaluate the performance of the model, we capture the various aspects of detection performance through the following measures: accuracy, precision, recall, Cohen’s Kappa index, and Area under Precision-Recall curve score.

\section{Materials and method}
\label{method}

\subsection{Study area}
Several other studies \citep{roelens, bailly, rapinel, kiss} that have located ditches from LiDAR data have worked on smaller areas (smaller than 2 $km^2$) where artificial drainage dominates. Here, we work on landscape scale in a mesoscale catchment. The  Krycklan Study Catchment in Sweden is 68 $km^2$, which means that we cover more landscape types. The quaternary deposits are dominated by till (51\%) and sorted sediments (30\%). The catchment ranges in elevation from 114 to 405 m above sea level. Forest covers 87\% of the catchment, mires 9\%, thin soils 7\%, and rock outcrops 1\%. The land use is dominated by forestry; approximately 25\% of the Krycklan catchment has been protected since 1922, but most of the other area is second growth forest \citep{krycklancatchment}.

A historical survey of the catchment (Gudrun Norstedt, unpublished) shows that the ditches in the catchment were dug between 1900 and 1993. The earliest ditches were used to drain mires to harvest hay. Later, many of the ditches were dug based on the notion that any mire could be turned into productive forest land by drainage \citep{paivanen}. Today we know that nutrient limitation also plays a key role, and some drained wetlands did not turn into productive forest \citep{sikstrom}. Once a canopy of trees and shrubs has been established in the forest, the increased evapotranspiration \citep{price, koivusalo}, increased interception \citep{price}, and reduced groundwater recharge will continue to naturally lower the water-table in these areas, making ditch maintenance unnecessary, at least until the clear-cut phase when the canopy is removed. Also,  lowering the water table, the oxygenated soils show increasing bulk density and decreasing hydraulic conductivity \citep{silins, price}, causing a subsidence of the peat layer near the ditch. This is likely the most important factor causing ditches to become shallow \citep{heikurainen}.

\subsection{Digitising the ground truth}
A high resolution DEM was generated from LiDAR data, see \hyperref[lidartodem]{Appendix} \ref{lidartodem} for specifications of the process. To train and evaluate the Random Forests model, a vector layer of all ditches in the Krycklan Catchment \citep{krycklancatchment} was digitised. Road ditches, forest ditches, and agricultural ditches can be found in the catchment. A multidirectional hillshade of the 0.5 m DEM was calculated in ArcGIS Pro to locate the ditches. From this layer, most of the ditches could be easily identified. To ensure that the digitised ditch was located correctly inside the ditch and not on the bank nearby, the digitisation was conducted by zooming in to a scale of 1:600. In addition to this, both current and historical records and orthophotos dating back to 1960’s were used to help locate all the ditches. In uncertain areas, we also found it useful to view the data in 3D. However, the digitising was conducted in 2D, and in uncertain areas the ditches were verified in the field. We are confident that we have captured all the ditches in the catchment according to the digitised  ditch layers in \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan}. The stream network in \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan} shows a network with a 8 hectare flow initiation threshold. Though, the length of the stream network vary substantially between season, ranging from 140 km during baseflow to 630 km during high-flow conditions \citep{mappingtemporal}. The digitisation of the ditch networks showed that there were 107 km road ditches and 208 km forest/agricultural ditches (\hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan}). The stream network in \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan}, which aims to show median flow conditions, was 197 km.

\begin{figure}
    \centering
    \includegraphics[width=1\linewidth]{./images/Krycklan_lo.jpg}
    \caption{Map of Krycklan showing ditches and stream channels, as well as our division into 21 subsections. The 10 subsections with a white border were used for developing the model before the experiment. The 11 subsections with a black border were used in the experiment for evaluation with 11-fold cross validation. Each rectangular subsection represents $2997 \cdot 2620$ pixels (roughly 196 hectare).}
    \label{fig:swedenkrycklan}
\end{figure}

From the ditch mapping, the vector layer was rasterised so it could be compared to the automatically derived ditches from the $0.5*0.5$ m DEM. Because the observed average width of ditches is larger than 0.5 metres, all pixels within a radius of three pixels (1.5 metres) of the vectors were labelled as ditch pixels. \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{a} shows the ditches rasterised from vectors and \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{b} shows the ditches after widening. The data in \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{b} is the labelled data that was used to train the Random Forests model. The ditches all have a generalised width based on a perceived average of ditch width in the area. Due to all ditches varying in width, it was impossible to produce a perfect representation of each ditch. However, this made for a good compromise for the average ditch.

Because our priority was to detect ditches, rather than accurately detecting each pixel labelled as a ditch, some adjustments were made when producing the prediction results. The dataset was divided into a lower resolution grid of $6*6$ pixels (9 $m^2$) for each grid. Each grid cell that contained at least 25\% ditch pixels was then labelled as a ditch, see \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{c}. This layer was used to ground-truth the Random Forests model.

\begin{figure} [htb!]
    \centering
    \subfigure[]{
        \resizebox*{4.5cm}{!}{\includegraphics{./images/publ_ditch_preprocess_A_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{4.5cm}{!}{\includegraphics{./images/publ_ditch_preprocess_B_lo.jpg}}}
    \subfigure[]{
        \resizebox*{4.5cm}{!}{\includegraphics{./images/publ_ditch_preprocess_C_lo.jpg}}}
    \caption{Processing of ditch labels. \textbf{a: }Rasterised ditches with a width of one pixel. \textbf{b: }Ditches after a widening process, seven pixels (3.5 metres) wide. Used as input when training the model. \textbf{c: }Ditches after a grid zone conversion. Used for evaluating the results from the model.} \label{sample-figure}
    \label{fig:ditchpreprocess}
\end{figure}

\subsection{Extracting ditches with digital terrain indices}
From the DEM, several digital terrain indices were calculated:
\begin{itemize} \label{data_attributes}

  \item \textbf{Sky View Factor} \newline
  The Sky View Factor is a digital terrain index with a value ranging from 0 to 1, representing how much of the sky that is visible from a certain point on the ground \citep{zaksek}. On an open field all of the sky would be visible, but at the bottom of a ditch much of the sky would be obscured, meaning low values would indicate ditches. This method tends to give false positives in steep terrain, where the sky can be obscured by a hillside, or in narrow valleys or ravines that also show low numbers. The Sky View Factor was calculated in SAGA GIS, using a search radius of 10 m.

  \item \textbf{Impoundment Index}\newline
  The tool Impoundment Index in Whitebox Tools \citep{whiteboxtools} is a digital terrain index designed to serve as a measure of topographic incision and flow constriction. The index effectively maps the area and average depth of the reservoir that would be created by placing a dam of a user-specified length at each grid cell in the DEM. The tool can be tuned to output the flooded area or volume (in which case it highlights areas of flow constriction), or the average flooded depth (which seems to measure topographic incision). Additionally, the tool will output a 'dam height' raster that is particularly useful for mapping channels, such as ditches. Here, we used dam height calculated with a dam length of 3 m. Typical false positives with this method are natural stream channels with a clear incised channel.

  \item \textbf{High Pass Median Filter} \newline
  High Pass Median Filters (HPMF) highlight ditches by indicating local depressions in the DEM. The algorithm operates essentially by subtracting the value at the grid cell at the centre of the window from the median value in the window. Negative values indicate depressions in the DEM, such as ditches. A typical false positive with this index are small hollows that also show negative numbers. HPMF was calculated using Whitebox Tools \citep{whiteboxtools} with a window size of $4.5 * 4.5$ m.

    \item \textbf{Slope} \newline
    The Slope index represents the degree of slope at a certain point, with a value ranging from $0^{\circ}$ to $90^{\circ}$. It was calculated on the 0.5 m grid, using ArcGIS \citep{EsriArcGisBook}. This index contains no information about the direction of the slope.
\end{itemize}

\subsubsection{Thresholding the digital terrain indices}
To benchmark our method against a non-machine learning method, we binarised the original four digital terrain indices using thresholds. Different threshold values were tested and the value that yielded the highest Cohen's Kappa score for each respective index was chosen. This test was conducted for the 11 subsections described in \ref{trainingvalidationdatasets}. The Sky View Factor index was classified to only include values below 0.955, the Impoundment Index Dam Height using values above 0.27 $m$, the HPMF index using values below -0.18 m, and the Slope index using values above $14 ^{\circ}$ (Table 2).

\subsection{Data preparation}

\subsubsection{Processing the digital terrain indices}
Developing the Random Forests model involved examining how different kinds of input variables affected the prediction. Several possible data manipulation methods could theoretically produce a better prediction. To improve the prediction, steps were taken where neighbouring pixels were included to give a representation of the area surrounding a specific pixel, similar to the approach of \citet{roelens}.

The digital terrain indices provided a satisfactory foundation for the model, but lacked in the generalisability of their predictions. More diverse input variables were extracted using simple statistical aggregates such as mean, median, min, max, and standard deviation. Using these statistical aggregations with the help of neighbouring areas around pixels aided in pruning pixels with outlier values, often smoothing out the data to represent ditches more accurately on a per-pixel basis. These variables were calculated by gathering all data points in different circular radii around the studied pixel, before computing one of the statistical aggregations, see \hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{b}, \hyperref[fig:features]{c}, \hyperref[fig:features]{h}, and \hyperref[fig:features]{j}.

In addition to the general input variables, a number of custom input variables were also developed. Several of these variables were calculated by looking at the separate digital terrain indices, attempting to amplify the ditches. This was done in several different ways for the indices by combining statistical aggregations, and by using thresholds to exclude, lower, or amplify the values of pixels that indicated ditches or non-ditches, see \hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{e} and \hyperref[fig:features]{k}.

From the Slope and Sky View Factor indices, input variables for the model were also extracted to attempt to amplify the values of pixels that lay in streams or in hilly areas, but leave pixels in ditches or other geographical areas untouched, see \hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{i}.

\newpage

Both HPMF and Sky View Factor were used with the image processing \textit{gabor filter}, which can be used to detect lines of a certain orientation in an image \citep{gabor}. 30 Gabor filters, which were rotated in different angles and with different frequencies, were combined to detect lines, amplifying ditches by utilising the fact that ditches have a linear elongated shape (\hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{d} and \hyperref[fig:features]{g}).

\label{impoundmentstreamremoval}
The raw Impoundment Index dam height was used to create a mask, attempting to mark out streams, but retain ditches. Only areas with a relatively high dam height, above 14 metres, which would indicate that these areas most likely contained streams, were marked out. After widening the resulting area, this mask was used to attempt to remove streams from all the aforementioned custom input variables, generating one new input variable from each. The features marked with \textit{streams removed} in \hyperref[featuretable]{Table} \ref{featuretable} make use of this filter (\hyperref[fig:features]{Figure} \ref{fig:features} \hyperref[fig:features]{f} and \hyperref[fig:features]{l}).

\begin{figure} [!htb]
    \centering
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_a_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_b_lo.jpg}}}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_c_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_d_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_e_lo.jpg}}}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_f_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_g_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_h_lo.jpg}}}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_i_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_j_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_k_lo.jpg}}}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_l_lo.jpg}}}\hspace{5pt}
    \caption{Example of 11 of the 81 input variables, in addition to ditch labels used by the model for a small sample area. The radii represents pixels with a 0.5 m resolution. \newline \textbf{a:} Labelled ditches, \textbf{b:} Slope standard deviation, radius 6, \textbf{c:} HPMF mean, radius 4, \textbf{d:} HPMF Gabor, \textbf{e:} HPMF ditch amplification, \textbf{f:} HPMF ditch amplification - streams removed \textbf{g:} Sky View Factor Gabor, \textbf{h:} Sky View Factor max, radius 6, \textbf{i:} Sky View Factor non-ditch amplification  \textbf{j:} Impoundment mean, radius 3, \textbf{k:} Impoundment ditch amplification, \textbf{l:} Impoundment ditch amplification - streams removed}
    \label{fig:features}
\end{figure}
\clearpage

\subsubsection{Training and validation datasets}
\label{trainingvalidationdatasets}
To develop and evaluate our model, the raster and ditch label data of Krycklan were manually divided into 21 subsections.  Each subsection represents an area of roughly 196 hectare. From this division, 11 of the subsections were put aside as hold-out data, only for use in the experiment to evaluate the performance of the predictions. The remaining 10 subsections were used solely before the experiment to develop the model to test different ways of preparing the data. This allowed the model to be evaluated on unseen data to strengthen the validity of the experiment. \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan} shows which subsections were used for development and evaluation respectively.

Using the 11 subsections in the hold-out data for the final experiment, a process called \textit{k-fold cross validation} was employed (11-fold cross validation in our case). K-fold cross validation is a method where you divide your dataset into folds of similar size and train a model on all but one of your folds (subsections). You then use that subsection to evaluate the results \citep{crossvalidation}. Using this technique, shifting which subsection to leave out from the training, allowed us to train 11 different Random Forests classifying models with a large amount of data from the remaining 10 subsections in the hold-out data, producing 11 sub-experiments to evaluate the method on.

\subsection{Building the Random Forests model}

\textit{Model ensembles} is a category of machine learning where a set of models can be combined to increase diversity and robustness of predictions \citep{flach}. Random Forests is a supervised learning algorithm, first developed by \citet{breiman}, that makes use of the model ensemble technique by using a set of decision trees to build a \textit{forest} of trees. The outputs from these trees are then examined, and a class probability estimation ranging from 0 to 1 can be produced for each sample by dividing the amount of tree outputs of a class with the total amount of trees \citep{breiman}.

The decision tree data structure is a \textit{tree} of connected nodes where the end node of each branch (leaf) produces an output value. Each step in the tree aims to minimise the entropy of the outputs \citep{kotsiantis}. Each node in a decision tree represents a variable or an abstraction of variables from the inputs and each split is made by using this variable to separate the inputs.

Random Forests uses two techniques called \textit{subspace sampling} and \textit{bootstrap aggregation} (bagging) in tandem \citep{breiman}. Subspace sampling is the process of drawing a random subset of variables from the input variable set \citet{ho}. The bagging technique takes, with replacement, a set of different random samples for constructing each tree \citep{flach}. There are several hyperparameters that can be tuned to achieve better results from the model. For example, you can adjust the amount of decision trees to use, or the maximum amount of input variables to use for each subspace sampling \citep{scikit-learn}. We performed a simple hyperparameter tuning to determine what parameter values for the Random Forests algorithm would yield the best prediction. Evaluating a maximum of 25 input variables for each node, and using 200 trees produced the best results. Setting the class weight to \textit{balanced} also improved the performance of the classifier.

\newpage

Random Forests computes Gini importance, which shows how often a variable is selected for a split for a certain classification \citep{gini}. Each input variable is given a Gini impurity rating, which measures the frequency of incorrect labelling occurring from its use. As the Gini impurity rating for a variable increases, its Gini importance decreases \citep{gini}. When the model is trained, it examines the Gini importance from all trees and produces the total Gini importance as a value between 0 and 1 \citep{gini}. The Gini importance can help when developing a model by indirectly giving advice on new input variables to introduce.

The Random Forests algorithm from the Python library \textit{scikit-learn} was used in the experiment, and the classifier was trained on all the input variables seen in \hyperref[featuretable]{Table} \ref{featuretable}. The probability predictions used in this Random Forests implementation are computed by counting the number of occurrences of a class output and dividing with the total amount of trees in the forest \citep{scikit-learn}. The testing phase showed that the classifier produced poor results when the ratio of ditch- versus non-ditch pixels in the training data was very high. A high ratio led to the model not being punished for mislabelling ditches as non-ditches, causing it to prioritise a high accuracy over a high recall. According to \citet{balanced}, an imbalanced training dataset causes a minority class to be less accurately predicted. Because the ditch class is much less common than the non-ditch class, we attempted to train the model with a roughly equal amount of ditch pixels and non-ditch pixels to balance the model. The model was, however, still evaluated on all the pixels in the 11 validation subsections.

\begin{table} [!htb]
    \tbl{The 81 input variables used when training the model.}
    {\begin{tabular}{ll} \toprule
      Variable/Algorithm\textsuperscript{a} & Circular radii\textsuperscript{b} \\ \midrule

      Sky View Factor raw & -\\
      Sky View Factor mean & 2, 3, 4, 6 \\
      Sky View Factor median &2, 4, 6 \\
      Sky View Factor standard deviation & 2, 4, 6 \\
      Sky View Factor min & 2, 4, 6 \\
      Sky View Factor max & 2, 4, 6 \\
      Sky View Factor non ditch amplification & - \\ 
      Sky View Factor conic filter - streams removed & - \\ 
      Sky View Factor Gabor & - \\
      Sky View Factor Gabor - streams removed & -\\

      Impoundment raw & - \\
      Impoundment mean & 2, 3, 4, 6 \\
      Impoundment median & 2, 4, 6 \\
      Impoundment standard deviation & 2, 4, 6 \\
      Impoundment min & 2, 4, 6 \\
      Impoundment max & 2, 4, 6 \\
      Impoundment ditch amplification & - \\
      Impoundment ditch amplification - streams removed & - \\

      HPMF raw & - \\
      HPMF mean & 2, 3, 4, 6 \\
      HPMF median & 2, 4, 6 \\
      HPMF standard deviation & 2, 4, 6 \\
      HPMF min & 2, 4, 6 \\
      HPMF max & 2, 4, 6 \\
      HPMF ditch amplification & - \\
      HPMF ditch amplification - streams removed & - \\
      HPMF Gabor & -\\
      HPMF Gabor - streams removed & -\\

      Slope raw & - \\
      Slope mean & 2, 3, 4, 6 \\
      Slope median & 2, 4, 6 \\
      Slope standard deviation & 2, 4, 6 \\
      Slope min & 2, 4, 6 \\
      Slope max & 2, 4, 6 \\
      Slope non-ditch amplification & - \\ \bottomrule
    \end{tabular}}
    \tabnote{\textsuperscript{a} Displays algorithm used to produce the variable. \newline
        \textsuperscript{b} Represents the radius of the circular mask (if one was used) to determine \newline what neighbouring pixels to use in an algorithm. The radii represents pixels \newline \mbox{with a 0.5 m resolution.\itshape\ignorespaces}}
    \label{featuretable}
\end{table}

The first step to create a more balanced training dataset was to extract all pixels labelled as ditches, as well as pixels within close proximity of ditches. Secondly, pixels were sampled randomly from the entire subsection. This allowed the training dataset to be fairly balanced while still containing most of the geographical features of each subsection (\hyperref[fig:balancedmasks]{Figure} \ref{fig:balancedmasks}).

\begin{figure} [!htb]
    \centering
    \subfigure[]{
        \resizebox*{6.85cm}{!}{\includegraphics{./images/publ_balanced_masks_A_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{6.85cm}{!}{\includegraphics{./images/publ_balanced_masks_B_lo.jpg}}}
    \caption{Black pixels here indicate the balanced masks used to determine what pixels were used when training the model. \textbf{a, b: }Examples of two subsection from the training dataset.}
    \label{fig:balancedmasks}
\end{figure}

\subsection{Post-processing}
The model outputs a ditch class probability prediction for each pixel. Values close to zero indicate a low probability of a pixel lying in a ditch, whereas values close to one equals a high ditch probability (\hyperref[fig:postprocessing1]{Figure} \ref{fig:postprocessing1} \hyperref[fig:postprocessing1]{a}).

\subsubsection{Noise reduction and gap filling}
The probability predictions contained a lot of noise in geographical locations far away from ditches, which needed to be excluded. The first step for removing noise was to use a bilateral de-noising filter on the entire prediction image. This left linear properties and pixels with a very high value intact, while lowering the value of pixels that did not contribute to an accurate prediction (\hyperref[fig:postprocessing1]{Figure} \ref{fig:postprocessing1} \hyperref[fig:postprocessing1]{b}).

\begin{figure} [!htb]
    \centering
    \subfigure[]{
        \resizebox*{6.85cm}{!}{\includegraphics{./images/publ_post_process_step_1_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{6.85cm}{!}{\includegraphics{./images/publ_post_process_step_2_lo.jpg}}}
    \caption{Step one and two of the post-process. Darker pixels indicate a higher ditch probability. \textbf{a: }Raw probability prediction produced by the Random Forests model. \textbf{b: }Prediction after bilateral de-noising.}
    \label{fig:postprocessing1}
\end{figure}

The second step for removing noise was to use a custom function to remove pixels that had a semi-high ditch probability, but that lay far away from any other high probability pixels. A threshold value was used to avoid removing pixels with a high enough ditch probability, helping to retain pixels that lay in or close to a ditch. The max ditch probability value in a circular radius of 10 pixels was then calculated. If this max value was too low, the probability value of the examined pixel was lowered (\hyperref[fig:postprocessing2]{Figure} \ref{fig:postprocessing2} \hyperref[fig:postprocessing2]{a}).

The third step involved taking measures to try to fill gaps in ditches that the model failed to correctly predict. This was done by aggregating pixels covered by cone masks expanding outwards in different directions from the examined pixel. This step also amplified some of the noise that was left, but filling the gaps in the ditches was judged to be more important to help make the next step more effective (\hyperref[fig:postprocessing2]{Figure} \ref{fig:postprocessing2} \hyperref[fig:postprocessing2]{b}).

\begin{figure} [!htb]
    \centering
    \subfigure[]{
        \resizebox*{6.85cm}{!}{\includegraphics{./images/publ_post_process_step_3_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{6.85cm}{!}{\includegraphics{./images/publ_post_process_step_4_lo.jpg}}}
    \caption{Step three and four of the post-process. Darker pixels indicate a higher ditch probability. \textbf{a: }Prediction after custom de-noising. \textbf{b: }Prediction after gap filling.}
    \label{fig:postprocessing2}
\end{figure}

\subsubsection{Binarisation with grid zones}
The model's ditch prediction is given in the original resolution raster format. To allow for comparison to the evaluation data, the same grid conversion was performed on the prediction raster as on the evaluation data seen in \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{c}. A mean probability rating was calculated for each $6*6$ grid zone, classifying the entire zone as part of a ditch if the mean probability exceeded 35 \%. This helped to fill in gaps where lone pixels in ditches had been incorrectly classified, and also helped in the next step of the post-processing (\hyperref[fig:postprocessing3]{Figure} \ref{fig:postprocessing3} \hyperref[fig:postprocessing3]{a}).

\subsubsection{Cluster removal}
To remove noise from the binary prediction, a custom cluster detection algorithm was used. By finding the number of connected pixels with a true value and removing those whose cluster size were below a given threshold, minor noise in the prediction could be removed while still retaining most of the ditch pixels. Actual ditches that have a low probability and therefore create small clusters may still be excluded by this method, but the noise removal advantages were judged to outweigh the loss in recall. A distance calculation was also performed in tandem with this method to find the largest distance of pixels inside each given cluster. This helped to remove sinks and hollows that were not removed by the initial small cluster removal, but that had a shape that indicated that they did not represent a ditch (\hyperref[fig:postprocessing3]{Figure} \ref{fig:postprocessing3} \hyperref[fig:postprocessing3]{b}).

\begin{figure} [!htb]
    \centering
    \subfigure[]{
        \resizebox*{6.85cm}{!}{\includegraphics{./images/publ_post_process_step_5_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{6.85cm}{!}{\includegraphics{./images/publ_post_process_step_6_lo.jpg}}}
    \caption{Step five and six of the post-process. Black pixels indicate a ditch prediction. \textbf{a: }Prediction after binarisation with grid zone probability. \textbf{b: }Final binary prediction after cluster removal.}
    \label{fig:postprocessing3}
\end{figure}

\subsection{Evaluation} \label{evaluation}

Using only accuracy as an evaluation metric when dealing with an imbalanced dataset (roughly 98 \% of all pixels are non-ditch) would produce a poor performance assessment \citep{balanced}; by simply classifying all pixels as non-ditches, we would by default attain 98 \% accuracy. For this reason, the results were mainly evaluated using Cohen's Kappa (Cohen's $\kappa$) index, and the Area under Precision-Recall curve. Cohen's $\kappa$ index measures how much better a prediction is compared to a completely random prediction, where random would yield a value of zero \citep{kappa123}. With our data, a $\kappa$ value close to zero would be attained by predicting 2 \% of the occurrences as ditch pixels completely at random. Values above zero are better than random, and values below zero are worse than random.

\newpage

The random rating $P_c$ of a prediction of $n$ occurrences is calculated with:\footnote{ P = Positive (ditch pixel), N = Negative (non-ditch pixel), T = true, F = false}

$$
P_c = \frac{\left(\frac{(TP + FN) \cdot (TP + FP)}{n}\right) + \left(\frac{(FN + TN) \cdot (FP + TN)}{n}\right)}{n}
$$

$$
\text{Cohen's } \kappa \text{ is then calculated as a value between -1 and 1 with:}
$$
$$\kappa = \frac{Accuracy - P_c}{1 - P_c}$$

The Precision-Recall curve and the Area under Precision-Recall curve (AUPRC) are additional metrics that can be used when evaluating datasets with a largely imbalanced class distribution \citep{precision_recall_curve}. The Precision-Recall curve has the recall value on the x-axis and the precision value on the y-axis, and the area under the curve that is defined by this point gives the AUPRC value. The area under this curve is given as a value between zero and one, where a value closer to one is better. The weighting causes the Precision-Recall curve to not place an equal value on true negatives and true positives \citep{precision_recall_curve}. For our ditch detection problem, this means that the AUPRC evaluation metric favours accurately classifying ditch pixels over accurately classifying non-ditch pixels.

To circumvent some of the performance evaluation issues arising from the use of pixel classification for ditch objects as well as not having completely accurate labels on a pixel basis (due to uncertainties in the width of the ditches), we modified the evaluation labels to allow for some error in close proximity to ditches in the prediction. False negatives and false positive grid zones that lay adjacent to the ditch label grid zones were evaluated as true negatives and true positives, as they can be considered a part of correctly located ditch objects (\hyperref[fig:newlabels]{Figure} \ref{fig:newlabels}).

\begin{figure} [!htb]
    \centering
    \subfigure[]{
        \resizebox*{6.5cm}{!}{\includegraphics{./images/re_evaluation_1_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{6.5cm}{!}{\includegraphics{./images/re_evaluation_2_lo.jpg}}}
    \caption{Illustration of the modified evaluation labels where \textbf{a} shows the original result, and \textbf{b} shows the modified result. Green marks true positives, red marks false negatives, and blue marks false positives. False positives and false negatives that lay within one grid zone (3 $m^2$) of a ditch label were evaluated as true positives and true negatives.}
    \label{fig:newlabels}
\end{figure}

\section{Results and analysis}

\hyperref[recreatedpredictionperformance]{Table} \ref{recreatedpredictionperformance} shows the evaluation metrics from the experiments with the digital terrain indices separately. Of the four, the Impoundment Index and the HPMF index performed relatively close to each other and outperformed the other indices for most of the metrics.

\begin{table}[!htb]
    \tbl{Metrics from the total results of the four digital terrain indices experiments.}
    {\begin{tabular}{lcccc} \toprule
        Metric & Sky View Factor & Impoundment & HPMF & Slope\\ \midrule
        Accuracy \%     & 83.67 & 97.07 & 97.45 & 82.66 \\
        Recall \%       & 44.54 & 29.98 & 25.31 & 34.83 \\
        Precision \%    &{ 4.00}  & 18.51 & 20.10 & { 2.99} \\
        $\kappa$ rating & 0.048 & 0.215 & 0.211 & 0.029 \\
        AUPRC & 0.247 & 0.248 & 0.232 & 0.194 \\ \bottomrule
    \end{tabular}}
    \label{recreatedpredictionperformance}
\end{table}

\hyperref[predictionperformance]{Table} \ref{predictionperformance} displays all the evaluation metrics for the prediction of our method. The confidence intervals were calculated from the 11 different subsections that the experiment was performed on. Because the subsections have varying amounts of ditches in them, the confidence intervals will not be completely accurate, but will produce a close estimation.

From \hyperref[predictionperformance]{Table} \ref{predictionperformance}, it can be seen that averaging the results from the 11 subsections will yield a very similar value to the total results, indicating that the model generally performs equally well on subsections with a small amount of ditches as on subsections with a large amount of ditches.

\begin{table}[!htb]
    \tbl{Metrics for the prediction performance of our model.}{\begin{tabular}{lccc} 
        \toprule
        Metric & Total\textsuperscript{a} & Subsection\textsuperscript{b}& CI 95\%\textsuperscript{c} \\ 
        &&Average&\\ \midrule
        Accuracy     \% & 98.96 & 98.96 & [98.64 , 99.27] \\
        Recall       \% & 77.41 & 76.81 & [69.35 , 84.27] \\
        Precision    \% & 73.89 & 73.06 & [69.73 , 76.40] \\
        $\kappa$ rating & 0.751 & 0.739 & [0.693 , 0.786] \\
        AUPRC           & 0.759 & 0.752 & [0.706 , 0.798] \\ 
        \bottomrule
    \end{tabular}}
    \tabnote{
        \textsuperscript{a} The result of all 11 subsection experiments when combined. \newline
        \textsuperscript{b} An average score from the 11 different subsections that the experiment was performed on. \newline
        \textsuperscript{c} Confidence intervals at 95 \% confidence level.}
    \label{predictionperformance}
\end{table}

The $\kappa$ rating for our method can be seen in the \textit{Total} column in \hyperref[predictionperformance]{Table} \ref{predictionperformance} ($\kappa$ = 0.751). The $\kappa$ ratings from the four digital terrain indices experiments can be seen in  \hyperref[recreatedpredictionperformance]{Table} \ref{recreatedpredictionperformance} ($\kappa$ = 0.048, $\kappa$ = 0.215, $\kappa$ = 0.211, $\kappa$ = 0.029). Because the $\kappa$ rating from our method outperforms all digital terrain indices, our hypothesis has been confirmed by the experiments. Most false positives lie in stream channels (\hyperref[fig:resultsillustrations]{Figure} \ref{fig:resultsillustrations} \hyperref[fig:resultsillustrations]{a, b}) and most false negatives occurred either due to ditches being too shallow (\hyperref[fig:resultsillustrations]{Figure} \ref{fig:resultsillustrations} \hyperref[fig:resultsillustrations]{c, d}), or due to the attempt at removing streams using the input variables as explained in \ref{impoundmentstreamremoval} removing the deepest ditches  (\hyperref[fig:resultsillustrations]{Figure} \ref{fig:resultsillustrations} \hyperref[fig:resultsillustrations]{e, f}).

\begin{figure} [!htb]
    \centering
    \subfigure[]{
        \resizebox*{5.5cm}{!}{\includegraphics{./images/results_illustration_1_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{5.5cm}{!}{\includegraphics{./images/results_illustration_2_lo.jpg}}}
    \subfigure[]{
        \resizebox*{5.5cm}{!}{\includegraphics{./images/results_illustration_3_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{5.5cm}{!}{\includegraphics{./images/results_illustration_4_lo.jpg}}}
    \subfigure[]{
        \resizebox*{5.5cm}{!}{\includegraphics{./images/results_illustration_5_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{5.5cm}{!}{\includegraphics{./images/results_illustration_6_lo.jpg}}}
    \caption{The left panel shows the hillshade for a number of subsections, and the right panel shows the hillshade with the extracted ditches superimposed on top. Green marks correctly classified ditches (true positives), red marks missed ditches (false negatives), blue marks incorrectly classified ditches (false positives), and transparent areas mark correctly classified non-ditches (true negatives). \textbf{a} and \textbf{b} illustrate that the most common false positives were natural stream channels being classified as ditches. \textbf{c} and \textbf{d} illustrate that very shallow ditches that were barely visible in the DEM were not captured with our model. \textbf{e} and \textbf{f} illustrate that some of the deepest and more prominent ditches were also missed. This occurred as a result of attempting to get rid of more false positives in the form of streams by using the input variables, as explained in \ref{impoundmentstreamremoval}.}
    \label{fig:resultsillustrations}
\end{figure}
\clearpage

In \hyperref[featureimportancetable]{Table} \ref{featureimportancetable}, the top 20 input variables from the Random Forests model are presented with their importance percentages for the model making a successful prediction. From the table, it can be seen that input variables based on HPMF and Impoundment Index contributed the most to a successful prediction. The Gini importance was obtained using the Gini impurity for each variable \citep{gini}.

\begin{table} [!htb]
    \tbl{The top 20 input variables (out of 81) by importance when the model makes a prediction.}
        {\begin{tabular}{llr}
          \toprule
          Position & Variable\textsuperscript{a} & Importance (\%) \\ \midrule
          1.  & Impoundment mean 3                                  & 11.08\\
          2.  & Impoundment mean 4                                  & 7.44\\
          3.  & HPMF mean 4                                         & 7.03\\
          4.  & HPMF mean 3                                         & 3.97\\
          5.  & Impoundment median 4                                & 2.97\\
          6.  & Impoundment mean 2                                  & 2.53\\
          7.  & Sky View Factor Gabor - streams removed             & 2.03\\
          8.  & Impoundment ditch amplification                     & 1.91\\
          9.  & HPMF median 4                                       & 1.90\\
          10. & Impoundment ditch amplification - streams removed   & 1.76\\
          11. & Slope standard deviation 6                          & 1.62\\
          12. & Sky View Factor non-ditch amplification             & 1.59\\
          13. & HPMF Gabor - streams removed                        & 1.59\\
          14. & Sky View Factor Gabor                               & 1.48\\
          15. & HPMF Gabor                                          & 1.46\\
          16. & Sky View Factor max 6                               & 1.39\\
          17. & HPMF mean 6                                         & 1.39\\
          18. & Impoundment standard deviation 4                    & 1.39\\
          19. & Slope min 6                                         & 1.31\\
          20. & Slope non-ditch amplification                       & 1.25\\
          \bottomrule
        \end{tabular}}
        \tabnote{\textsuperscript{a} The number next to some of the variables indicates the circular radius used to select \newline what neighbouring pixels to use in the statistical aggregation method. The radii\newline \mbox{represents pixels with a 0.5 m resolution.\itshape\ignorespaces}}
    \label{featureimportancetable}
\end{table}

\section{Discussion}

In our study we have shown that it is possible to locate ditches automatically in high-resolution DEMs ($0.5  * 0.5 $ metres, in our case), and that more of the ditches can be detected if the information from several terrain indices (\hyperref[recreatedpredictionperformance]{Table} \ref{recreatedpredictionperformance}) is combined through machine learning (\hyperref[predictionperformance]{Table} \ref{predictionperformance}).
The difficulty of detecting the ditches in the study catchment vary greatly from ditch to ditch, this is illustrated in \hyperref[fig:ditchpictures]{Figure} \ref{fig:ditchpictures}.
By separating the road ditches from the other ditches, a comparison showed that the retrieval rate was slightly higher for the road ditches. This is likely due to the fact that they are found in open areas alongside roads, and also that road ditches have generally been better maintained as it has been more prioritised to keep these ditches functioning (\hyperref[fig:ditchpictures]{Figure} \ref{fig:ditchpictures} \hyperref[fig:ditchpictures]{a}). However, most of the ditches in Krycklan today are found below the canopy (e.g. \hyperref[fig:ditchpictures]{Figure} \ref{fig:ditchpictures} \hyperref[fig:ditchpictures]{c}), and the majority have not been maintained, causing difficulties in detecting them as some have grow back in. This, in combination with the surrounding peat having subsided \citep{heikurainen}, have rendered them almost impossible to detect in the DEM today (\hyperref[fig:ditchpictures]{Figure} \ref{fig:ditchpictures} \hyperref[fig:ditchpictures]{d}). Despite these issues we had a recall rate of 77.4\%, showing that the majority of the ditches could still be retrieved automatically.

\begin{figure} [!htb]
    \centering
    \subfigure[]{
        \resizebox*{6.85cm}{!}{\includegraphics{./images/road_ditch_lo.jpg}}}\hspace{7pt}
    \subfigure[]{
        \resizebox*{6.85cm}{!}{\includegraphics{./images/mire_ditch_lo.jpg}}}
    \subfigure[]{
        \resizebox*{6.85cm}{!}{\includegraphics{./images/forest_ditch_lo.jpg}}}\hspace{7pt}
    \subfigure[]{
        \resizebox*{6.85cm}{!}{\includegraphics{./images/overgrown_ditch_lo.jpg}}}
    \caption{All photos show ditches in the study catchment. They illustrate the difficulty of capturing the ditches based on their form (i.e. an elongated, narrow, hollow in the ground). \textbf{a: }A roadside ditch. These are generally easily detected as they are usually found in open areas along the roads, and have often been maintained. \textbf{b: }A ditch in a mire; this is still easily detected, as there is no canopy and there is a clear elevation difference between the ditch and surrounding area. \textbf{c: }A narrow, more or less  overgrown ditch in a dense forest stand. \textbf{d: }A ditch in a mire that has more or less grown back in with \textit{Sphagnum sp.} and \textit{Carex}, likely also in combination with subsiding surrounding soils. Such ditches can be found in the field as the vegetation differs, but as there is no longer an elevation difference between the ditch and the surrounding soils it will not be detected in the DEM. Photos a-c: E. Maher Hasselquist  Photo d: G. Norstedt}
    \label{fig:ditchpictures}
\end{figure}

Hypothetically, it should be quite easy to detect ditches using LiDAR data. In practice, however, there are many variables affecting the results. One is the point density and the interpolation method used to create the DEM. \citet{rapinel} found that the results were usually more sensitive to the point density (ranging 1-4 points $m^{2}$ in their study) than the interpolation method. They retrieved 54.8 and 63.8 \% of the ditches on the Acuey and Boucey marches in France, compared to 77.4 \% in our study. However, in our study we had a point density of 20 points $m^{2}$, so the quality of the DEM in our study ought to be robust, which may have contributed to our higher retrieval rate. \hyperref[fig:resultstreesbushes]{Figure} \ref{fig:resultstreesbushes} \hyperref[fig:resultstreesbushes]{(a, b)} illustrates the difference in the quality of the LiDAR data where the laser pulses have caught in trees or shrubs. When too few LiDAR points reach the ground, the DEM generation is forced to interpolate over a larger area, leading to inaccuracies in the ground DEM.

\begin{figure} [!htb]
    \centering
    \subfigure[]{
        \resizebox*{6.8cm}{!}{\includegraphics{./images/results_trees_bushes_1_lo.jpg}}}\hspace{7pt}
    \subfigure[]{
        \resizebox*{6.8cm}{!}{\includegraphics{./images/results_trees_bushes_2_lo.jpg}}}
    \subfigure[]{
        \resizebox*{6.8cm}{!}{\includegraphics{./images/results_trees_bushes_3_lo.jpg}}}\hspace{7pt}
    \subfigure[]{
        \resizebox*{6.8cm}{!}{\includegraphics{./images/results_trees_bushes_4_lo.jpg}}}
    \caption{\textbf{a} and \textbf{b} illustrate that despite a point density of 20 points $m^2$, not all points reach the ground in places with dense vegetation. In \textbf{a} there is a clear difference between the smooth surface of the DEM in the open area (black arrow), and in the areas covered by trees and shrubs (red arrows), where the DEM is much coarser due to interpolation with a lower number of LiDAR points here. The orthophoto (\textbf{b}), clearly shows that trees and shrubs are growing in the ditches in this area. This makes it more difficult to detect ditches. However, despite these coarser grid points in the DEM, \textbf{c} and \textbf{d} shows that the model still managed to capture a majority of the ditches (green), and only failed at certain points (red).}
    \label{fig:resultstreesbushes}
\end{figure}

The results from our method (\hyperref[predictionperformance]{Table} \ref{predictionperformance}) show that we managed to attain a Cohen's $\kappa$ rating in the substantial range, according to performance thresholds proposed by \citet{kappaanalysis}. This means that our model performed substantially better than a completely random model. The total recall value of 77.4 \% shows that we managed to find a majority of the ditch pixels that exist. The results of our method also showed a significant improvement for ditch detection over classifying with the digital terrain indices separately. Had we been able to use labels that were correctly labelled on a pixel basis, i.e. each ditch had its actual width recorded, the model could have learned with a better accuracy in the training phase, and the classification results would also most likely have benefitted from this.

Sometimes pixels in ditches were initially identified correctly as ditch pixels, but because not enough pixels in the surrounding area were detected, the post-processing algorithms identified them as noise and removed them. Incorrectly classified ditch pixels that caused the precision to decrease were generally the pixels that lay in streams (\hyperref[fig:resultsillustrations]{Figure} \ref{fig:resultsillustrations} \hyperref[fig:resultsillustrations]{a, b}). However, small sinks or hilly areas incorrectly classified as ditches by the model were generally removed in the post-processing. We managed to bridge gaps and exclude streams to some extent in our predictions. This was in part due to the post-processing of the prediction helping where the model was unsuccessful on its own, and in part due to the input variables developed. \hyperref[featureimportancetable]{Table} \ref{featureimportancetable} highlights the importance of the input variables that were developed by applying different statistical aggregations for neighbouring areas of pixels, similar to \citet{roelens}, as well as the custom variables using filters such as the Impoundment stream removal or Gabor filters.

\citet{roelens} showed that you can use machine learning to detect ditches directly from point-clouds instead of a DEM. One concern we had with our method was that valuable information may have been discarded when aggregating the elevations from the original point cloud to a DEM, making it more difficult to locate the ditches. However, we achieved a similar score to their method; $\kappa=0.75$ in our study compared to 0.77 for grassland and 0.73 for peri-urban area in Roelens study. This suggests that you can also use a DEM to automatically extract ditches, provided that you have enough input data to generate a high resolution DEM to capture the small scale features of ditches.

\citet{bailly} also showed that the vegetation cover significantly affects the ability to detect ditches from LiDAR data (\hyperref[fig:ditchpictures]{Figure} \ref{fig:ditchpictures} \& \ref{fig:resultstreesbushes}). Approximately 75\% of the ditches in their study were retrieved where no vegetation existed, whereas the detection levels sank to below 10\% under vegetation; especially when high vegetation, shrubs and trees covered the ditches \citep{bailly}. Hence, a $\kappa$ of 0.75 in the Krycklan catchment with  87\% forest cover \citep{krycklancatchment} is in that aspect surprisingly good, and could perhaps be a result of the higher point density of our LiDAR measurements (20 compared to 10 points $m^{2}$). Or, perhaps the generation of a DEM from the point cloud has made the detection more robust?

The hydrology of our study catchment also differs from the other mentioned studies as there are almost as many natural stream channels as there are ditches in the area \citep{hasselquist, mappingtemporal}. Other studies were conducted in smaller, predominately artificially drained areas \citep{roelens, bailly, rapinel}. Our method locates both artificial channels (ditches) and natural stream channels. This would not be a problem if the aim had been to locate all channels. However, our study focused on locating ditches specifically. Hence, we tried to remove the natural stream channels by manipulating some of the input variables, using the Impoundment Index to lower the values of pixels that had a high dam height, as explained in \ref{impoundmentstreamremoval}. This helped to remove the deepest streams and increased the overall accuracy, with the drawback of decreasing the recall slightly (\hyperref[fig:resultsillustrations]{Figure} \ref{fig:resultsillustrations} \hyperref[fig:resultsillustrations]{f}). This method did not remove the smaller streams (\hyperref[fig:resultsillustrations]{Figure} \ref{fig:resultsillustrations} \hyperref[fig:resultsillustrations]{b}). A future improvement to the model could be to use a shape index to remove small natural streams, as natural streams often form more complex channels whereas ditches are often straight. A potential issue with generalising our model for use in other geographical areas is that all the post-processing steps and input variables were developed based on occurrences in the Krycklan area. Different geographical compositions in other areas may require that the thresholds used to fill gaps in ditches and to remove noise would need to be adjusted.

\section{Conclusion}

This study investigates how to use digital elevation data together with manually labelled ditches to identify patterns and relationships that can be used for automatic ditch detection. The proposed method significantly outperforms classifying with the digital terrain indices separately. Although the method still has room for improvement, it performs well on the available data. In the future, a more extensive hyperparameter tuning could be performed, because, as \citet{lavesson} states; hyperparameter tuning is often more important than choice of algorithm. Some of the custom input variables could be reused in the future, with Random Forests or other machine learning algorithms. Several of the post-processing algorithms, such as the probability prediction noise removal or the cluster removal from the binary prediction could be generalised for use in any type of pixel classification of ditches.

The work in this study gives an insight into what input variables work well for ditch detection and what input variables do not, as well as an increased understanding of how machine learning can be applied to ditch detection. The proposed method may help in making ditch detection both faster and easier, reducing manual labour and cost. In conclusion, this study has shown that it is possible to use machine learning with digital elevation models to learn patterns that enable robust detection of ditches.

\section*{Data and codes availability statement}
The data and codes that support the findings of this study are available at the private links:\newline
\mbox{\href{https://figshare.com/s/c03de5594268843a8b50}{https://figshare.com/s/c03de5594268843a8b50} - Functions\itshape\ignorespaces}
\mbox{\href{https://figshare.com/s/741725088fe6292442c8}{https://figshare.com/s/741725088fe6292442c8} - Feature creation\itshape\ignorespaces}
\mbox{\href{https://figshare.com/s/f4e3bcf3bb2f8c18ecfa}{https://figshare.com/s/f4e3bcf3bb2f8c18ecfa} - Digital terrain indices experiment\itshape\ignorespaces}
\mbox{\href{https://figshare.com/s/c1326af09f600d18a9d7}{https://figshare.com/s/c1326af09f600d18a9d7} - Random Forests experiment\itshape\ignorespaces}
\mbox{\href{https://figshare.com/s/c0ae38ee5951498d21cd}{https://figshare.com/s/c0ae38ee5951498d21cd}} - Experiment data

\label{lidartodem}
The LiDAR dataset used in this study was produced by TerraTec Sweden AB in August 2015, on demand of the Department of Forest Resource Management at the Swedish University of Agricultural Sciences, and is freely available online at \href{https://gis.krycklan.se/}{https://gis.krycklan.se/}. The average density of the dataset was 20 pulses per $m^2$.  Ground point classification was performed with lasground (included in LAStools, a software suite produced by rapidlasso GmbH). The DEM was created with blast2dem (included in LAStools),  with cell size of $0.5*0.5$ m. Only ground points were imported  to generate a DEM of the ground. This was conducted for 488 las-files of the study area that were mosaicked to one large file using Mosaic to new raster in Arc GIS Pro \citep{EsriArcGisBook}.

\section*{Acknowledgements}
We wish to thank Sigurd Israelsson at the School of Engineering at J\"onk\"oping University for lending hardware resources for the experiments in this study. The work has been funded by VINNOVA, EU InterReg Baltic Sea project WAMBAF tools and Formas.

\label{references}

\bibliographystyle{tfv}
\bibliography{references.bib}

\end{document}

### 2020-06-23 Functions/post_processing.py
Adds failed experiment.

@@ -47,6 +47,7 @@ from numba import jit
 from numba import prange
 
 from Functions import general_functions
+from Functions import feature_creation
 
 @jit(nopython=True)
 def raster_to_zones(arr, zoneSize, threshold):
@@ -327,7 +328,7 @@ def conic_proba_post_processing(arr, maskRadius, threshold):
     masks = []
     maxArr = d_gf(da.from_array(arr,chunks = (800,800)), np.nanmax, footprint=general_functions.create_circular_mask(5))
     for i in range(0, 8):
-        masks.append(create_conic_mask(maskRadius, i))
+        masks.append(_create_conic_mask(maskRadius, i))
 
     return _conic_proba_post_processing(np.array(arr), np.array(maxArr), np.array(masks),threshold)
     

### 2020-06-24 Functions/post_processing.py
fix?

@@ -48,6 +48,7 @@ from numba import prange
 
 from Functions import general_functions
 from Functions import feature_creation
+from Functions.feature_creation import _create_conic_mask
 
 @jit(nopython=True)
 def raster_to_zones(arr, zoneSize, threshold):

### 2020-06-24 post_processing.py

@@ -47,6 +47,8 @@ from numba import jit
 from numba import prange
 
 from Functions import general_functions
+from Functions.feature_creation import _create_conic_mask
+
 
 @jit(nopython=True)
 def raster_to_zones(arr, zoneSize, threshold):
@@ -327,7 +329,7 @@ def conic_proba_post_processing(arr, maskRadius, threshold):
     masks = []
     maxArr = d_gf(da.from_array(arr,chunks = (800,800)), np.nanmax, footprint=general_functions.create_circular_mask(5))
     for i in range(0, 8):
-        masks.append(create_conic_mask(maskRadius, i))
+        masks.append(_create_conic_mask(maskRadius, i))
 
     return _conic_proba_post_processing(np.array(arr), np.array(maxArr), np.array(masks),threshold)
     

### 2020-06-27 Functions/feature_creation.py
Fixes (hopefully) the final experiment code.

@@ -47,7 +47,7 @@ from numba import jit
 from numba import prange
 
 
-import general_functions
+from . import general_functions
 
 
 def dem_ditch_detection(arr):

### 2020-06-27 general_functions.py

@@ -1,175 +1,0 @@
-# General arr lib
-import numpy as np
-
-# Compile py code
-from numba import jit
-from numba import prange
-
-# Multithreading
-import dask.array as da
-import dask as dk
-dk.config.set(scheduler='processes')
-
-#from dask_image.ndfilters import generic_filter as d_gf
-
-from collections import deque
-
-from skimage.filters import gabor
-from skimage.restoration import  denoise_bilateral
-
-import datetime
-
-import numpy as np
-from numpy import random, nanmax, argmax, unravel_index
-from scipy.spatial.distance import pdist, squareform
-from scipy.ndimage.filters import generic_filter as gf
-import scipy.stats.mstats as ms
-from scipy.stats import skew
-import scipy.ndimage.morphology as morph
-from scipy import ndimage
-from PIL import Image
-import scipy
-import matplotlib.pyplot as plt
-import re
-import os
-import pandas as pd
-import random
-import math
-from skimage.restoration import (denoise_tv_chambolle, denoise_bilateral, denoise_wavelet, estimate_sigma)
-from skimage import data, img_as_float
-from skimage.filters import gabor
-from skimage.util import random_noise
-from collections import deque
-from sklearn.ensemble import RandomForestClassifier
-from joblib import dump, load
-from sklearn.metrics import cohen_kappa_score, accuracy_score, recall_score, confusion_matrix, precision_score
-from numba import jit
-from numba import prange
-
-
-
-@jit
-def create_circular_mask(radius):
-    """
-    Creates a circular mask.
-    """
-    kernel = np.zeros((2*radius+1, 2*radius+1))
-    y, x = np.ogrid[-radius:radius+1, -radius:radius+1]
-    mask = x**2 + y**2 <= radius**2
-    mask[radius][radius] = 0
-    kernel[mask] = 1
-    return kernel
-
-def extract_numpy_files_in_folder(path, skip=[]):
-    """
-    Returns all the .npy files in a given directory.
-    Skips subdirectories.
-    """
-    root, _, files = next(os.walk(path))
-    holder = []
-        
-    for file in files:
-        if file[-3:] != ".npy":
-            continue
-        elif file in skip:
-            continue
-        holder.append(os.path.join(root,file))
-    return holder
-
-def generate_mask(small_radius, big_radius):
-    """
-    Generate a mask in a circular radius around a point.
-    """
-    height, width = big_radius*2,big_radius*2
[…]

### 2020-06-27 post_processing.py

@@ -1,368 +1,0 @@
-# General arr lib
-import numpy as np
-
-# Compile py code
-from numba import jit
-from numba import prange
-
-# Multithreading
-import dask.array as da
-import dask as dk
-dk.config.set(scheduler='processes')
-
-from dask_image.ndfilters import generic_filter as d_gf
-
-from collections import deque
-
-from skimage.filters import gabor
-from skimage.restoration import  denoise_bilateral
-
-import datetime
-
-import numpy as np
-from numpy import random, nanmax, argmax, unravel_index
-from scipy.spatial.distance import pdist, squareform
-from scipy.ndimage.filters import generic_filter as gf
-import scipy.stats.mstats as ms
-from scipy.stats import skew
-import scipy.ndimage.morphology as morph
-from scipy import ndimage
-from PIL import Image
-import scipy
-import matplotlib.pyplot as plt
-import re
-import os
-import pandas as pd
-import random
-import math
-from skimage.restoration import (denoise_tv_chambolle, denoise_bilateral, denoise_wavelet, estimate_sigma)
-from skimage import data, img_as_float
-from skimage.filters import gabor
-from skimage.util import random_noise
-from collections import deque
-from sklearn.ensemble import RandomForestClassifier
-from joblib import dump, load
-from sklearn.metrics import cohen_kappa_score, accuracy_score, recall_score, confusion_matrix, precision_score
-from numba import jit
-from numba import prange
-
-from Functions import general_functions
-from Functions.feature_creation import _create_conic_mask
-
-
-@jit(nopython=True)
-def raster_to_zones(arr, zoneSize, threshold):
-    """
-    Converts binary pixel labels to zones with a specified zone size and threshold.
-    """
-    new_arr = arr.copy()
-    for i in range(0, len(arr), zoneSize):
-        for j in range(0, len(arr[i]), zoneSize):
-            numberOfClassified = 0
-            if i < len(arr) - zoneSize and j < len(arr[i]) - zoneSize:
-                for k in range(zoneSize):
-                    for l in range(zoneSize):
-                        if arr[i + k][j + l] == 1:
-                            numberOfClassified += 1
-                if numberOfClassified > (zoneSize**2)/threshold:
-                    for k in range(zoneSize):
-                        for l in range(zoneSize):
-                            new_arr[i + k][j + l] = 1
-                else:
[…]

### 2020-06-27 LaTeX/Ditch detection in high-resolution digital elevation models.tex
Adds changes to article and rejoinder.

@@
 Values of all LiDAR attributes were represented for the neighbouring area for both 0.5, 1, 1.5, 2, 3 and 4 metre radii.
-We reproduce these neighbouring area variables when building the model in our study.
+We reproduce these neighbouring area variables when building the ditch locator in our study.
 In a pilot study, we compared several different algorithms (Random Forests, XGBoost, Naive Bayes, and Support Vector Machines), and it was found that Random Forests produced the most accurate results.
 Random Forests also computes Mean Decrease in Impurity, or Gini importance, highlighting what input variables are the most important for a given prediction \citep{gini}.
-Because of these factors, and the fact that it proved to be suitable for \citet{roelens}, a Random Forests model is used to locate the ditches.
+Because of these factors, and the fact that it proved to be suitable for \citet{roelens}, Random Forests models are used to locate the ditches.
 
@@
 To test this, we conduct a study using a high resolution DEM of the Krycklan Catchment in Sweden \citep{krycklancatchment}, where a high precision manual mapping has been performed.
-This mapping was used to both train the Random Forests model and to ground-truth the models.
-To evaluate the performance of the model, we capture the various aspects of detection performance through the following measures: accuracy, precision, recall, Cohen’s Kappa index, and Area under Precision-Recall curve score.
+This mapping was used to both train the Random Forests models and to produce the ground-truth.
+To evaluate the performance of the ditch locator, we capture the various aspects of detection performance through the following measures: accuracy, precision, recall, Cohen’s Kappa index, and Area under Precision-Recall curve score.
 
@@
 A high resolution DEM was generated from LiDAR data, see \hyperref[lidartodem]{Appendix} \ref{lidartodem} for specifications of the process.
-To train and evaluate the Random Forests model, a vector layer of all ditches in the Krycklan Catchment \citep{krycklancatchment} was digitised.
+To train and evaluate the Random Forests models, a vector layer of all ditches in the Krycklan Catchment \citep{krycklancatchment} was digitised.
 Road ditches, forest ditches, and agricultural ditches can be found in the catchment.
@@
     \caption{Map of Krycklan showing ditches and stream channels, as well as our division into 21 subsections.
-The 10 subsections with a white border were used for developing the model before the experiment.
+The 10 subsections with a white border were used for developing input variables and optimising parameters before the experiment.
 The 11 subsections with a black border were used in the experiment for evaluation with 11-fold cross validation.
@@
 \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{a} shows the ditches rasterised from vectors and \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{b} shows the ditches after widening.
-The data in \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{b} is the labelled data that was used to train the Random Forests model.
+The data in \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{b} is the labelled data that was used to train the Random Forests models.
 The ditches all have a generalised width based on a perceived average of ditch width in the area.
@@
 Each grid cell that contained at least 25\% ditch pixels was then labelled as a ditch, see \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{c}.
-This layer was used to ground-truth the Random Forests model.
+This layer was used to ground-truth the Random Forests models.
 
@@
 \textbf{b: }Ditches after a widening process, seven pixels (3.5 metres) wide.
-Used as input when training the model.
+Used as input when training the models.
 \textbf{c: }Ditches after a grid zone conversion.
-Used for evaluating the results from the model.} \label{sample-figure}
+Used for evaluating the results from the ditch locator.} \label{sample-figure}
     \label{fig:ditchpreprocess}
@@
 \subsubsection{Processing the digital terrain indices}
-Developing the Random Forests model involved examining how different kinds of input variables affected the prediction.
+To improve the performance of the Random Forests models, we examined how different kinds of input variables affected the prediction.
 Several possible data manipulation methods could theoretically produce a better prediction.
@@
 
-The digital terrain indices provided a satisfactory foundation for the model, but lacked in the generalisability of their predictions.
+The digital terrain indices provided a satisfactory foundation for the models, but lacked in the generalisability of their predictions.
 More diverse input variables were extracted using simple statistical aggregates such as mean, median, min, max, and standard deviation.
@@
 
-From the Slope and Sky View Factor indices, input variables for the model were also extracted to attempt to amplify the values of pixels that lay in streams or in hilly areas, but leave pixels in ditches or other geographical areas untouched, see \hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{i}.
+From the Slope and Sky View Factor indices, input variables for the models were also extracted to attempt to amplify the values of pixels that lay in streams or in hilly areas, but leave pixels in ditches or other geographical areas untouched, see \hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{i}.
 
@@
 The features marked with \textit{streams removed} in \hyperref[featuretable]{Table} \ref{featuretable} make use of this filter (\hyperref[fig:features]{Figure} \ref{fig:features} \hyperref[fig:features]{f} and \hyperref[fig:features]{l}).
+
+Overall, 81 input variables were extracted from the terrain indices.
+We then conducted a sub-experiment to find the best input variables, as well as the optimal number of variables to use.
+Random Forest's feature importance was used to select the most valuable input variables for different variable amounts.
+It was found that using 53 input variables produced the best results.
 
@@
         \resizebox*{4cm}{!}{\includegraphics{./images/feature_l_lo.jpg}}}\hspace{5pt}
-    \caption{Example of 11 of the 81 input variables, in addition to ditch labels used by the model for a small sample area.
+    \caption{Example of 11 of the 81 input variables, in addition to ditch labels used by the models for a small sample area.
 The radii represents pixels with a 0.5 m resolution.
@@
 \label{trainingvalidationdatasets}
-To develop and evaluate our model, the raster and ditch label data of Krycklan were manually divided into 21 subsections.
+To develop and evaluate our ditch locator, the raster and ditch label data of Krycklan were manually divided into 21 subsections.
 Each subsection represents an area of roughly 196 hectare.
 From this division, 11 of the subsections were put aside as hold-out data, only for use in the experiment to evaluate the performance of the predictions.
-The remaining 10 subsections were used solely before the experiment to develop the model to test different ways of preparing the data.
-This allowed the model to be evaluated on unseen data to strengthen the validity of the experiment.
+The remaining 10 subsections were used solely before the experiment to develop the ditch locator to test different ways of preparing the data.
+This allowed the ditch locator to be evaluated on unseen data to strengthen the validity of the experiment.
 \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan} shows which subsections were used for development and evaluation respectively.
@@
 You then use that subsection to evaluate the results \citep{crossvalidation}.
-Using this technique, shifting which subsection to leave out from the training, allowed us to train 11 different Random Forests classifying models with a large amount of data from the remaining 10 subsections in the hold-out data, producing 11 sub-experiments to evaluate the method on.
+A new model is trained using the training folds for each iteration (\hyperref[fig:crossvalidation]{Figure} \ref{fig:crossvalidation}).
+Using this technique, shifting which subsection to leave out from the training, allowed us to train 11 different Random Forests classifying models with a large amount of data from the remaining 10 subsections in the hold-out data, producing 11 independent sub-experiments to evaluate the method on.
 
-\subsection{Building the Random Forests model}
+\begin{figure}[!htb]
+    \centering
+    \includegraphics[width=1\linewidth]{images/cross_validation_lo.jpg}
+    \caption{11-fold cross validation. Each subsection in the hold-out data was used once for testing, with the remaining subsections being used to build a new model for each iteration.}
+    \label{fig:crossvalidation}
+\end{figure}
+
+\subsection{Developing the Random Forests model}
 
@@
 Evaluating a maximum of 25 input variables for each node, and using 200 trees produced the best results.
-Setting the class weight to \textit{balanced} also improved the performance of the classifier.
+Setting the class weight to \textit{balanced} also improved the performance of the classifiers.
 
@@
 
-The Random Forests algorithm from the Python library \textit{scikit-learn} was used in the experiment, and the classifier was trained on all the input variables seen in \hyperref[featuretable]{Table} \ref{featuretable}.
+The Random Forests algorithm from the Python library \textit{scikit-learn} was used in the experiment, and the classifiers were trained on all the input variables seen in \hyperref[featuretable]{Table} \ref{featuretable}.
 The probability predictions used in this Random Forests implementation are computed by counting the number of occurrences of a class output and dividing with the total amount of trees in the forest \citep{scikit-learn}.
-The testing phase showed that the classifier produced poor results when the ratio of ditch- versus non-ditch pixels in the training data was very high.
-A high ratio led to the model not being punished for mislabelling ditches as non-ditches, causing it to prioritise a high accuracy over a high recall.
+The testing phase showed that the classifiers produced poor results when the ratio of ditch- versus non-ditch pixels in the training data was very high.
+A high ratio led to the models not being punished for mislabelling ditches as non-ditches, causing it to prioritise a high accuracy over a high recall.
 According to \citet{balanced}, an imbalanced training dataset causes a minority class to be less accurately predicted.
-Because the ditch class is much less common than the non-ditch class, we attempted to train the model with a roughly equal amount of ditch pixels and non-ditch pixels to balance the model.
-The model was, however, still evaluated on all the pixels in the 11 validation subsections.
+Because the ditch class is much less common than the non-ditch class, we attempted to train the models with a roughly equal amount of ditch pixels and non-ditch pixels to balance the models.
+The models were, however, still evaluated on all the pixels in the 11 validation subsections.
 
 \begin{table} [!htb]
-    \tbl{The 81 input variables used when training the model.}
+    \tbl{The 81 input variables used when training the models.}
     {\begin{tabular}{ll} \toprule
@@
         \resizebox*{6.85cm}{!}{\includegraphics{./images/publ_balanced_masks_B_lo.jpg}}}
-    \caption{Black pixels here indicate the balanced masks used to determine what pixels were used when training the model. \textbf{a, b: }Examples of two subsection from the training dataset.}
+    \caption{Black pixels here indicate the balanced masks used to determine what pixels were used when training the models. \textbf{a, b: }Examples of two subsection from the training dataset.}
     \label{fig:balancedmasks}
@@
 \subsection{Post-processing}
-The model outputs a ditch class probability prediction for each pixel.
+The models output a ditch class probability prediction for each pixel.
 Values close to zero indicate a low probability of a pixel lying in a ditch, whereas values close to one equals a high ditch probability (\hyperref[fig:postprocessing1]{Figure} \ref{fig:postprocessing1} \hyperref[fig:postprocessing1]{a}).
@@
 
-The third step involved taking measures to try to fill gaps in ditches that the model failed to correctly predict.
+The third step involved taking measures to try to fill gaps in ditches that the models failed to correctly predict.
 This was done by aggregating pixels covered by cone masks expanding outwards in different directions from the examined pixel.
@@
 \subsubsection{Binarisation with grid zones}
-The model's ditch prediction is given in the original resolution raster format.
+The models' ditch prediction is given in the original resolution raster format.
 To allow for comparison to the evaluation data, the same grid conversion was performed on the prediction raster as on the evaluation data seen in \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{c}.
@@
 
-From \hyperref[predictionperformance]{Table} \ref{predictionperformance}, it can be seen that averaging the results from the 11 subsections will yield a very similar value to the total results, indicating that the model generally performs equally well on subsections with a small amount of ditches as on subsections with a large amount of ditches.
+From \hyperref[predictionperformance]{Table} \ref{predictionperformance}, it can be seen that averaging the results from the 11 subsections will yield a very similar value to the total results, indicating that the ditch locator generally performs equally well on subsections with a small amount of ditches as on subsections with a large amount of ditches.
 
 \begin{table}[!htb]
-    \tbl{Metrics for the prediction performance of our model.}{\begin{tabular}{lccc} 
+    \tbl{Metrics for the prediction performance of our ditch locator.}{\begin{tabular}{lccc} 
         \toprule
@@
 \textbf{a} and \textbf{b} illustrate that the most common false positives were natural stream channels being classified as ditches.
-\textbf{c} and \textbf{d} illustrate that very shallow ditches that were barely visible in the DEM were not captured with our model.
+\textbf{c} and \textbf{d} illustrate that very shallow ditches that were barely visible in the DEM were not captured with our ditch locator.
 \textbf{e} and \textbf{f} illustrate that some of the deepest and more prominent ditches were also missed.
@@
 
-In \hyperref[featureimportancetable]{Table} \ref{featureimportancetable}, the top 20 input variables from the Random Forests model are presented with their importance percentages for the model making a successful prediction.
+In \hyperref[featureimportancetable]{Table} \ref{featureimportancetable}, the top 20 input variables from the Random Forests models are presented with their importance percentages for the models making a successful prediction.
 From the table, it can be seen that input variables based on HPMF and Impoundment Index contributed the most to a successful prediction.
@@
 \begin{table} [!htb]
-    \tbl{The top 20 input variables (out of 81) by importance when the model makes a prediction.}
+    \tbl{The top 20 input variables (out of 81) by importance when the models make a prediction.}
         {\begin{tabular}{llr}
@@
 This makes it more difficult to detect ditches.
-However, despite these coarser grid points in the DEM, \textbf{c} and \textbf{d} shows that the model still managed to capture a majority of the ditches (green), and only failed at certain points (red).}
+However, despite these coarser grid points in the DEM, \textbf{c} and \textbf{d} shows that the ditch locator still managed to capture a majority of the ditches (green), and only failed at certain points (red).}
     \label{fig:resultstreesbushes}
@@
 The results from our method (\hyperref[predictionperformance]{Table} \ref{predictionperformance}) show that we managed to attain a Cohen's $\kappa$ rating in the substantial range, according to performance thresholds proposed by \citet{kappaanalysis}.
-This means that our model performed substantially better than a completely random model.
+This means that our ditch locator performed substantially better than a completely random ditch locator.
 The total recall value of 77.4 \% shows that we managed to find a majority of the ditch pixels that exist.
@@
 Had we been able to use labels that were correctly labelled on a pixel basis, i.e.
-each ditch had its actual width recorded, the model could have learned with a better accuracy in the training phase, and the classification results would also most likely have benefitted from this.
+each ditch had its actual width recorded, the models could have learned with a better accuracy in the training phase, and the classification results would also most likely have benefitted from this.
 
@@
 Incorrectly classified ditch pixels that caused the precision to decrease were generally the pixels that lay in streams (\hyperref[fig:resultsillustrations]{Figure} \ref{fig:resultsillustrations} \hyperref[fig:resultsillustrations]{a, b}).
-However, small sinks or hilly areas incorrectly classified as ditches by the model were generally removed in the post-processing.
+However, small sinks or hilly areas incorrectly classified as ditches by the Random Forests models were generally removed in the post-processing.
 We managed to bridge gaps and exclude streams to some extent in our predictions.
-This was in part due to the post-processing of the prediction helping where the model was unsuccessful on its own, and in part due to the input variables developed.
+This was in part due to the post-processing of the prediction helping where the models were unsuccessful on their own, and in part due to the input variables developed.
 \hyperref[featureimportancetable]{Table} \ref{featureimportancetable} highlights the importance of the input variables that were developed by applying different statistical aggregations for neighbouring areas of pixels, similar to \citet{roelens}, as well as the custom variables using filters such as the Impoundment stream removal or Gabor filters.
@@
 This method did not remove the smaller streams (\hyperref[fig:resultsillustrations]{Figure} \ref{fig:resultsillustrations} \hyperref[fig:resultsillustrations]{b}).
-A future improvement to the model could be to use a shape index to remove small natural streams, as natural streams often form more complex channels whereas ditches are often straight.
-A potential issue with generalising our model for use in other geographical areas is that all the post-processing steps and input variables were developed based on occurrences in the Krycklan area.
+A future improvement to the ditch locator could be to use a shape index to remove small natural streams, as natural streams often form more complex channels whereas ditches are often straight.
+A potential issue with generalising our ditch locator for use in other geographical areas is that all the post-processing steps and input variables were developed based on occurrences in the Krycklan area.
 Different geographical compositions in other areas may require that the thresholds used to fill gaps in ditches and to remove noise would need to be adjusted.

### 2020-07-04 Functions/general_functions.py
Adds new experiment for random forest optimisation with post processing.

@@ -126,7 +126,7 @@ def create_balanced_mask(ditchArr, height, width):
     """
     Creates a mask from a labeled zone to balance the ditch and non-ditch classes more.
     """
-    random.seed(89)
+    random.seed(42)
     new_arr = ditchArr.copy()
     for i in range(0, len(ditchArr), height):
         for j in range(0, len(ditchArr[i]), width):

### 2020-07-05 LaTeX/Ditch detection in high-resolution digital elevation models.tex
Adds more changes to article and additions to the rejoinder.

@@
 
-\title{Ditch detection in high-resolution digital elevation models}
+%\title{Ditch detection in high-resolution digital elevation models}
+
+\title{Detecting ditches using machine learning on high-resolution DEMs}
 
@@
 We reproduce these neighbouring area variables when building the ditch locator in our study.
+Several other input variables for the model are calculated from the indices to place emphasis on the clear linear directional characteristics of ditches.
+These characteristics are also used when processing the model's output, in order to remove noise and other inconsistencies in the predictions.
 In a pilot study, we compared several different algorithms (Random Forests, XGBoost, Naive Bayes, and Support Vector Machines), and it was found that Random Forests produced the most accurate results.
@@
 
-We hypothesise that by combining the information from all the digital terrain indices using a machine learner (Random Forests), we can improve the detection of the ditches.
+We hypothesise that by combining the information from all the digital terrain indices using a machine learner (Random Forests), we can improve the detection of the ditches, compared to if a single index would be used.
 To test this, we conduct a study using a high resolution DEM of the Krycklan Catchment in Sweden \citep{krycklancatchment}, where a high precision manual mapping has been performed.
@@
         \resizebox*{6.85cm}{!}{\includegraphics{./images/publ_post_process_step_2_lo.jpg}}}
-    \caption{Step one and two of the post-process.
-Darker pixels indicate a higher ditch probability.
-\textbf{a: }Raw probability prediction produced by the Random Forests model.
-\textbf{b: }Prediction after bilateral de-noising.}
+    \caption{Step one and two of the post-process. Darker pixels indicate a higher ditch probability. \textbf{a: }Random Forests probability prediction. \textbf{b: }Bilateral de-noising.}
     \label{fig:postprocessing1}
@@
 The second step for removing noise was to use a custom function to remove pixels that had a semi-high ditch probability, but that lay far away from any other high probability pixels.
-A threshold value was used to avoid removing pixels with a high enough ditch probability, helping to retain pixels that lay in or close to a ditch.
+In this function, a threshold value was used to avoid removing pixels with a high enough ditch probability, helping to retain pixels that lay in or close to a ditch.
 The max ditch probability value in a circular radius of 10 pixels was then calculated.
@@
         \resizebox*{6.85cm}{!}{\includegraphics{./images/publ_post_process_step_4_lo.jpg}}}
-    \caption{Step three and four of the post-process. Darker pixels indicate a higher ditch probability. \textbf{a: }Prediction after custom de-noising. \textbf{b: }Prediction after gap filling.}
+    \caption{Step three and four of the post-process. Darker pixels indicate a higher ditch probability. \textbf{a: }Custom de-noising. \textbf{b: }Gap filling.}
     \label{fig:postprocessing2}
@@
 To allow for comparison to the evaluation data, the same grid conversion was performed on the prediction raster as on the evaluation data seen in \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{c}.
-A mean probability rating was calculated for each $6*6$ grid zone, classifying the entire zone as part of a ditch if the mean probability exceeded 35 \%.
-This helped to fill in gaps where lone pixels in ditches had been incorrectly classified, and also helped in the next step of the post-processing (\hyperref[fig:postprocessing3]{Figure} \ref{fig:postprocessing3} \hyperref[fig:postprocessing3]{a}).
+A mean probability rating was calculated for each $6*6$ grid zone, classifying the entire zone as part of a ditch if the mean probability exceeded 40 \%.
+Different zone sizes and probability thresholds were evaluated in a pilot study, and these values were found to produce the best results.
+This zone partitioning helped to fill in gaps where lone pixels in ditches had been incorrectly classified, and also helped in the next step of the post-processing (\hyperref[fig:postprocessing3]{Figure} \ref{fig:postprocessing3} \hyperref[fig:postprocessing3]{a}).
 
@@
 A distance calculation was also performed in tandem with this method to find the largest distance of pixels inside each given cluster.
-This helped to remove sinks and hollows that were not removed by the initial small cluster removal, but that had a shape that indicated that they did not represent a ditch (\hyperref[fig:postprocessing3]{Figure} \ref{fig:postprocessing3} \hyperref[fig:postprocessing3]{b}).
+This helped to remove sinks and hollows that were not removed by the initial small cluster removal, but that did not have a linear directional characteristic, indicating that they did not represent a ditch (\hyperref[fig:postprocessing3]{Figure} \ref{fig:postprocessing3} \hyperref[fig:postprocessing3]{b}).
 
@@
         \resizebox*{6.85cm}{!}{\includegraphics{./images/publ_post_process_step_6_lo.jpg}}}
-    \caption{Step five and six of the post-process.
-Black pixels indicate a ditch prediction.
-\textbf{a: }Prediction after binarisation with grid zone probability.
-\textbf{b: }Final binary prediction after cluster removal.}
+    \caption{Step five and six of the post-process. Black pixels indicate a ditch prediction. \textbf{a: }Grid zone binarisation \textbf{b: }Cluster removal.}
     \label{fig:postprocessing3}
@@
         \resizebox*{6.5cm}{!}{\includegraphics{./images/re_evaluation_2_lo.jpg}}}
-    \caption{Illustration of the modified evaluation labels where \textbf{a} shows the original result, and \textbf{b} shows the modified result.
+    \caption{Illustration of the modified evaluation labels.
 Green marks true positives, red marks false negatives, and blue marks false positives.
-False positives and false negatives that lay within one grid zone (3 $m^2$) of a ditch label were evaluated as true positives and true negatives.}
+False positives and false negatives that lay within one grid zone (9 $m^2$) of a ditch label were evaluated as true positives and true negatives.
+\textbf{a:} Original results, \textbf{b:} Modified results.
+}
     \label{fig:newlabels}
@@
 
-In our study we have shown that it is possible to locate ditches automatically in high-resolution DEMs ($0.5  * 0.5 $ metres, in our case), and that more of the ditches can be detected if the information from several terrain indices (\hyperref[recreatedpredictionperformance]{Table} \ref{recreatedpredictionperformance}) is combined through machine learning (\hyperref[predictionperformance]{Table} \ref{predictionperformance}).
+In our study we have shown that it is possible to locate ditches automatically in high-resolution DEMs ($0.5  * 0.5 $ metres, in our case), and that more of the ditches can be detected if the information from several terrain indices (\hyperref[recreatedpredictionperformance]{Table} \ref{recreatedpredictionperformance}) is combined through machine learning than if indices are used separately (\hyperref[predictionperformance]{Table} \ref{predictionperformance}).
 The difficulty of detecting the ditches in the study catchment vary greatly from ditch to ditch, this is illustrated in \hyperref[fig:ditchpictures]{Figure} \ref{fig:ditchpictures}.
@@
 
-This study investigates how to use digital elevation data together with manually labelled ditches to identify patterns and relationships that can be used for automatic ditch detection.
+%This study investigates how to use digital elevation data together with manually labelled ditches to identify patterns and relationships that can be used for automatic ditch detection.
 The proposed method significantly outperforms classifying with the digital terrain indices separately.
@@
 
-The work in this study gives an insight into what input variables work well for ditch detection and what input variables do not, as well as an increased understanding of how machine learning can be applied to ditch detection.
+%The work in this study gives an insight into what input variables work well for ditch detection and what input variables do not, as well as an increased understanding of how machine learning can be applied to ditch detection.
+The proposed method may help in making ditch detection both faster and easier, reducing manual labour and cost.
+In conclusion, this study has shown that it is possible to use machine learning with digital elevation models to learn patterns that enable robust detection of ditches.
+
+This study investigates how to use digital elevation data together with manually labelled ditches to identify patterns and relationships that can be used for automatic ditch detection.
+Although the method still has room for improvement, it performs well on the available data, and significantly outperforms classifying with the digital terrain indices separately.
+Future work could introduce more input variables and post-processing steps, such as using pathfinding algorithms to fill gaps in the ditch model, or shape indices to remove stream channels from the prediction.
+Moving towards image segmentation with grids of pixels instead of pixel classification of tabular pixel values could be a suitable future approach.
+
+Many of the input variables, could be reused in the future, with Random Forests or other machine learning algorithms.
+Several of the post-processing algorithms, such as the probability prediction noise removal, and the cluster removal from the binary prediction could be generalised for use in any type of segmentation of ditches.
 The proposed method may help in making ditch detection both faster and easier, reducing manual labour and cost.

### 2020-07-09 LaTeX/Ditch detection in high-resolution digital elevation models.tex
Adds updates to article after read-through. Adds updates to rejoinder.

@@
 
-The confidence interval for the Cohen's Kappa index ranged \textbf{[0.693 , 0.786]} between the evaluation plots with a confidence level of \textbf{95\%}.
+The confidence interval for the Cohen's Kappa index ranged \textbf{[0.655 , 0.781]} between the evaluation plots with a confidence level of \textbf{95\%}.
 Input variables based on the Impoundment Index overall represented the strongest ditch predictors.
@@
 Values of all LiDAR attributes were represented for the neighbouring area for both 0.5, 1, 1.5, 2, 3 and 4 metre radii.
-We reproduce these neighbouring area variables when building the ditch locator in our study.
+We reproduce these neighbouring area variables when building the ditch detector in our study.
 Several other input variables for the model are calculated from the indices to place emphasis on the clear linear directional characteristics of ditches.
 These characteristics are also used when processing the model's output, in order to remove noise and other inconsistencies in the predictions.
-In a pilot study, we compared several different algorithms (Random Forests, XGBoost, Naive Bayes, and Support Vector Machines), and it was found that Random Forests produced the most accurate results.
+In a pilot study, we compared several different algorithms (Random Forests, Extreme Gradient Boosting, Naive Bayes, and Support Vector Machines), and it was found that Random Forests produced the most accurate results.
 Random Forests also computes Mean Decrease in Impurity, or Gini importance, highlighting what input variables are the most important for a given prediction \citep{gini}.
@@
 This mapping was used to both train the Random Forests models and to produce the ground-truth.
-To evaluate the performance of the ditch locator, we capture the various aspects of detection performance through the following measures: accuracy, precision, recall, Cohen’s Kappa index, and Area under Precision-Recall curve score.
+To evaluate the performance of the ditch detector, we capture the various aspects of detection performance through the following measures: accuracy, precision, recall, Cohen’s Kappa index, and Area under Precision-Recall curve score.
 
@@
 \subsection{Digitising the ground truth}
-A high resolution DEM was generated from LiDAR data, see \hyperref[lidartodem]{Appendix} \ref{lidartodem} for specifications of the process.
+A high resolution DEM was generated from LiDAR data, see \hyperref[lidartodem]{'Data and codes availability statement'} for specifications of the process.
 To train and evaluate the Random Forests models, a vector layer of all ditches in the Krycklan Catchment \citep{krycklancatchment} was digitised.
@@
 Though, the length of the stream network vary substantially between season, ranging from 140 km during baseflow to 630 km during high-flow conditions \citep{mappingtemporal}.
-The digitisation of the ditch networks showed that there were 107 km road ditches and 208 km forest/agricultural ditches (\hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan}).
+The digitisation of the ditch networks showed that there are 107 km road ditches and 208 km forest/agricultural ditches (\hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan}).
 The stream network in \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan}, which aims to show median flow conditions, was 197 km.
@@
 
-From the ditch mapping, the vector layer was rasterised so it could be compared to the automatically derived ditches from the $0.5*0.5$ m DEM.
+From the ditch mapping, the vector layer was rasterised so that it could be compared to the automatically derived ditches from the $0.5*0.5$ m DEM.
 Because the observed average width of ditches is larger than 0.5 metres, all pixels within a radius of three pixels (1.5 metres) of the vectors were labelled as ditch pixels.
@@
 \textbf{c: }Ditches after a grid zone conversion.
-Used for evaluating the results from the ditch locator.} \label{sample-figure}
+Used for evaluating the results from the ditch detector.} \label{sample-figure}
     \label{fig:ditchpreprocess}
@@
 Negative values indicate depressions in the DEM, such as ditches.
-A typical false positive with this index are small hollows that also show negative numbers.
+Typical false positives with this index are small hollows that also show negative numbers.
 HPMF was calculated using Whitebox Tools \citep{whiteboxtools} with a window size of $4.5 * 4.5$ m.
@@
 Using these statistical aggregations with the help of neighbouring areas around pixels aided in pruning pixels with outlier values, often smoothing out the data to represent ditches more accurately on a per-pixel basis.
-These variables were calculated by gathering all data points in different circular radii around the studied pixel, before computing one of the statistical aggregations, see \hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{b}, \hyperref[fig:features]{c}, \hyperref[fig:features]{h}, and \hyperref[fig:features]{j}.
+These variables were calculated by gathering all data points in different circular radii around the studied pixel, before computing one of the statistical aggregations (\hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{b}, \hyperref[fig:features]{c}, \hyperref[fig:features]{f}, and \hyperref[fig:features]{g}).
 
@@
 Several of these variables were calculated by looking at the separate digital terrain indices, attempting to amplify the ditches.
-This was done in several different ways for the indices by combining statistical aggregations, and by using thresholds to exclude, lower, or amplify the values of pixels that indicated ditches or non-ditches, see \hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{e} and \hyperref[fig:features]{k}.
+This was done in several different ways for the indices by combining statistical aggregations, and by using thresholds to exclude, lower, or amplify the values of pixels that indicated ditches or non-ditches (\hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{d} and \hyperref[fig:features]{h}).
 
-From the Slope and Sky View Factor indices, input variables for the models were also extracted to attempt to amplify the values of pixels that lay in streams or in hilly areas, but leave pixels in ditches or other geographical areas untouched, see \hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{i}.
-
-\newpage
+From the Slope and Sky View Factor indices, input variables for the models were also extracted to attempt to amplify the values of pixels that lay in streams or in hilly areas, but leave pixels in ditches or other geographical areas untouched.
 
 Both HPMF and Sky View Factor were used with the image processing \textit{gabor filter}, which can be used to detect lines of a certain orientation in an image \citep{gabor}.
-30 Gabor filters, which were rotated in different angles and with different frequencies, were combined to detect lines, amplifying ditches by utilising the fact that ditches have a linear elongated shape (\hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{d} and \hyperref[fig:features]{g}).
+30 Gabor filters, which were rotated in different angles and with different frequencies, were combined to detect lines, amplifying ditches by utilising the fact that ditches have a linear elongated shape (\hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{e}).
 
@@
 After widening the resulting area, this mask was used to attempt to remove streams from all the aforementioned custom input variables, generating one new input variable from each.
-The features marked with \textit{streams removed} in \hyperref[featuretable]{Table} \ref{featuretable} make use of this filter (\hyperref[fig:features]{Figure} \ref{fig:features} \hyperref[fig:features]{f} and \hyperref[fig:features]{l}).
+The features marked with \textit{streams removed} in \hyperref[featuretable]{Table} \ref{featuretable} make use of this filter (\hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{i}).
 
@@
 Random Forest's feature importance was used to select the most valuable input variables for different variable amounts.
-It was found that using 53 input variables produced the best results.
+It was found that using 40 input variables produced the best results.
 
@@
     \subfigure[]{
-        \resizebox*{4cm}{!}{\includegraphics{./images/feature_d_lo.jpg}}}\hspace{5pt}
+        \resizebox*{4cm}{!}{\includegraphics{./images/feature_d_lo.jpg}}}
     \subfigure[]{
-        \resizebox*{4cm}{!}{\includegraphics{./images/feature_e_lo.jpg}}}
+        \resizebox*{4cm}{!}{\includegraphics{./images/feature_e_lo.jpg}}}\hspace{5pt}
     \subfigure[]{
-        \resizebox*{4cm}{!}{\includegraphics{./images/feature_f_lo.jpg}}}\hspace{5pt}
+        \resizebox*{4cm}{!}{\includegraphics{./images/feature_f_lo.jpg}}}
     \subfigure[]{
@@
     \subfigure[]{
-        \resizebox*{4cm}{!}{\includegraphics{./images/feature_i_lo.jpg}}}\hspace{5pt}
-    \subfigure[]{
-        \resizebox*{4cm}{!}{\includegraphics{./images/feature_j_lo.jpg}}}\hspace{5pt}
-    \subfigure[]{
-        \resizebox*{4cm}{!}{\includegraphics{./images/feature_k_lo.jpg}}}
-    \subfigure[]{
-        \resizebox*{4cm}{!}{\includegraphics{./images/feature_l_lo.jpg}}}\hspace{5pt}
-    \caption{Example of 11 of the 81 input variables, in addition to ditch labels used by the models for a small sample area.
+        \resizebox*{4cm}{!}{\includegraphics{./images/feature_i_lo.jpg}}}
+    \caption{Example of 8 of the 40 input variables, in addition to ditch labels used by the models for a small sample area.
 The radii represents pixels with a 0.5 m resolution.
-\newline \textbf{a:} Labelled ditches, \textbf{b:} Slope standard deviation, radius 6, \textbf{c:} HPMF mean, radius 4, \textbf{d:} HPMF Gabor, \textbf{e:} HPMF ditch amplification, \textbf{f:} HPMF ditch amplification - streams removed \textbf{g:} Sky View Factor Gabor, \textbf{h:} Sky View Factor max, radius 6, \textbf{i:} Sky View Factor non-ditch amplification  \textbf{j:} Impoundment mean, radius 3, \textbf{k:} Impoundment ditch amplification, \textbf{l:} Impoundment ditch amplification - streams removed}
+\newline \textbf{a:} Labelled ditches, \textbf{b:} Slope standard deviation, radius 6, \textbf{c:} HPMF mean, radius 4, \textbf{d:} HPMF ditch amplification, \textbf{e:} Sky View Factor Gabor, \textbf{f:} Sky View Factor max, radius 6, \textbf{g:} Impoundment mean, radius 3, \textbf{h:} Impoundment ditch amplification, \textbf{i:} Impoundment ditch amplification - streams removed}
     \label{fig:features}
 \end{figure}
-\clearpage
 
@@
 \label{trainingvalidationdatasets}
-To develop and evaluate our ditch locator, the raster and ditch label data of Krycklan were manually divided into 21 subsections.
+To develop and evaluate our ditch detector, the raster and ditch label data of Krycklan were manually divided into 21 subsections.
 Each subsection represents an area of roughly 196 hectare.
 From this division, 11 of the subsections were put aside as hold-out data, only for use in the experiment to evaluate the performance of the predictions.
-The remaining 10 subsections were used solely before the experiment to develop the ditch locator to test different ways of preparing the data.
-This allowed the ditch locator to be evaluated on unseen data to strengthen the validity of the experiment.
+The remaining 10 subsections were used solely before the experiment to develop the ditch detector to test different ways of preparing the data.
+This allowed the ditch detector to be evaluated on unseen data to strengthen the validity of the experiment.
 \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan} shows which subsections were used for development and evaluation respectively.
@@
 For example, you can adjust the amount of decision trees to use, or the maximum amount of input variables to use for each subspace sampling \citep{scikit-learn}.
-We performed a simple hyperparameter tuning to determine what parameter values for the Random Forests algorithm would yield the best prediction.
-Evaluating a maximum of 25 input variables for each node, and using 200 trees produced the best results.
-Setting the class weight to \textit{balanced} also improved the performance of the classifiers.
-
-\newpage
+We performed a hyperparameter tuning to determine what parameter values for the Random Forests algorithm would yield the best prediction.
+Using 300 trees and \textit{gini} as the splitting criterion, with a minimum of 10 samples per node split, and no artificial class weight or max depth of trees produced the best results.
 
@@
 \begin{table} [!htb]
-    \tbl{The 81 input variables used when training the models.}
+    \tbl{The 40 input variables used when training the models.}
     {\begin{tabular}{ll} \toprule
@@
 
-      Sky View Factor raw & -\\
-      Sky View Factor mean & 2, 3, 4, 6 \\
-      Sky View Factor median &2, 4, 6 \\
-      Sky View Factor standard deviation & 2, 4, 6 \\
-      Sky View Factor min & 2, 4, 6 \\
+      Sky View Factor median &2, 6 \\
+      Sky View Factor standard deviation & 6 \\
+      Sky View Factor min & 6 \\
       Sky View Factor max & 2, 4, 6 \\
       Sky View Factor non ditch amplification & - \\ 
-      Sky View Factor conic filter - streams removed & - \\ 
       Sky View Factor Gabor & - \\
@@
 
-      Impoundment raw & - \\
       Impoundment mean & 2, 3, 4, 6 \\
       Impoundment median & 2, 4, 6 \\
-      Impoundment standard deviation & 2, 4, 6 \\
-      Impoundment min & 2, 4, 6 \\
-      Impoundment max & 2, 4, 6 \\
+      Impoundment standard deviation & 4, 6 \\
+      Impoundment max & 6 \\
       Impoundment ditch amplification & - \\
@@
 
-      HPMF raw & - \\
-      HPMF mean & 2, 3, 4, 6 \\
-      HPMF median & 2, 4, 6 \\
-      HPMF standard deviation & 2, 4, 6 \\
-      HPMF min & 2, 4, 6 \\
-      HPMF max & 2, 4, 6 \\
+      HPMF mean & 3, 4, 6 \\
+      HPMF median & 4 \\
+      HPMF standard deviation & 6 \\
+      HPMF min & 2, 4 \\
       HPMF ditch amplification & - \\
       HPMF ditch amplification - streams removed & - \\
-      HPMF Gabor & -\\
       HPMF Gabor - streams removed & -\\
 
-      Slope raw & - \\
-      Slope mean & 2, 3, 4, 6 \\
-      Slope median & 2, 4, 6 \\
-      Slope standard deviation & 2, 4, 6 \\
+      Slope mean & 6 \\
+      Slope median & 6 \\
+      Slope standard deviation & 4, 6 \\
       Slope min & 2, 4, 6 \\
-      Slope max & 2, 4, 6 \\
+      %Slope max & 2, 4, 6 \\
       Slope non-ditch amplification & - \\ \bottomrule
@@
 
-\newpage
-
 The random rating $P_c$ of a prediction of $n$ occurrences is calculated with:\footnote{ P = Positive (ditch pixel), N = Negative (non-ditch pixel), T = true, F = false}
@@
 
-From \hyperref[predictionperformance]{Table} \ref{predictionperformance}, it can be seen that averaging the results from the 11 subsections will yield a very similar value to the total results, indicating that the ditch locator generally performs equally well on subsections with a small amount of ditches as on subsections with a large amount of ditches.
+From \hyperref[predictionperformance]{Table} \ref{predictionperformance}, it can be seen that averaging the results from the 11 subsections will yield a very similar value to the total results, indicating that the ditch detector generally performs equally well on subsections with a small amount of ditches as on subsections with a large amount of ditches.
 
 \begin{table}[!htb]
-    \tbl{Metrics for the prediction performance of our ditch locator.}{\begin{tabular}{lccc} 
+    \tbl{Metrics for the prediction performance of our ditch detector.}{\begin{tabular}{lccc} 
         \toprule
@@
         &&Average&\\ \midrule
-        Accuracy     \% & 98.96 & 98.96 & [98.64 , 99.27] \\
-        Recall       \% & 77.41 & 76.81 & [69.35 , 84.27] \\
-        Precision    \% & 73.89 & 73.06 & [69.73 , 76.40] \\
-        $\kappa$ rating & 0.751 & 0.739 & [0.693 , 0.786] \\
-        AUPRC           & 0.759 & 0.752 & [0.706 , 0.798] \\ 
+        Accuracy     \% & 99.00 & 99.00 & [98.69 , 99.32] \\
+        Recall       \% & 70.28 & 70.19 & [61.28 , 79.09] \\
+        Precision    \% & 77.38 & 75.79 & [71.94 , 79.64] \\
+        $\kappa$ rating & 0.732 & 0.718 & [0.655 , 0.781] \\
+        AUPRC           & 0.741 & 0.733 & [0.674 , 0.791] \\ 
         \bottomrule
@@
 
-The $\kappa$ rating for our method can be seen in the \textit{Total} column in \hyperref[predictionperformance]{Table} \ref{predictionperformance} ($\kappa$ = 0.751).
+The $\kappa$ rating for our method can be seen in the \textit{Total} column in \hyperref[predictionperformance]{Table} \ref{predictionperformance} ($\kappa$ = 0.732).
 The $\kappa$ ratings from the four digital terrain indices experiments can be seen in  \hyperref[recreatedpredictionperformance]{Table} \ref{recreatedpredictionperformance} ($\kappa$ = 0.048, $\kappa$ = 0.215, $\kappa$ = 0.211, $\kappa$ = 0.029).
@@
 \textbf{a} and \textbf{b} illustrate that the most common false positives were natural stream channels being classified as ditches.
-\textbf{c} and \textbf{d} illustrate that very shallow ditches that were barely visible in the DEM were not captured with our ditch locator.
+\textbf{c} and \textbf{d} illustrate that very shallow ditches that were barely visible in the DEM were not captured with our ditch detector.
 \textbf{e} and \textbf{f} illustrate that some of the deepest and more prominent ditches were also missed.
@@
 \begin{table} [!htb]
-    \tbl{The top 20 input variables (out of 81) by importance when the models make a prediction.}
+    \tbl{The top 20 input variables (out of 40) by importance when the models make a prediction.}
         {\begin{tabular}{llr}
@@
           Position & Variable\textsuperscript{a} & Importance (\%) \\ \midrule
-          1.  & Impoundment mean 3                                  & 11.08\\
-          2.  & Impoundment mean 4                                  & 7.44\\
-          3.  & HPMF mean 4                                         & 7.03\\
-          4.  & HPMF mean 3                                         & 3.97\\
-          5.  & Impoundment median 4                                & 2.97\\
-          6.  & Impoundment mean 2                                  & 2.53\\
-          7.  & Sky View Factor Gabor - streams removed             & 2.03\\
-          8.  & Impoundment ditch amplification                     & 1.91\\
-          9.  & HPMF median 4                                       & 1.90\\
-          10. & Impoundment ditch amplification - streams removed   & 1.76\\
-          11. & Slope standard deviation 6                          & 1.62\\
-          12. & Sky View Factor non-ditch amplification             & 1.59\\
-          13. & HPMF Gabor - streams removed                        & 1.59\\
-          14. & Sky View Factor Gabor                               & 1.48\\
-          15. & HPMF Gabor                                          & 1.46\\
-          16. & Sky View Factor max 6                               & 1.39\\
-          17. & HPMF mean 6                                         & 1.39\\
-          18. & Impoundment standard deviation 4                    & 1.39\\
-          19. & Slope min 6                                         & 1.31\\
-          20. & Slope non-ditch amplification                       & 1.25\\
+          1.  & Impoundment mean 3                                  & 7.33\\
+          2.  & HPMF mean 4                                         & 6.23\\
+          3.  & Impoundment mean 4                                  & 5.57\\
+          4.  & Impoundment mean 2                                  & 5.41\\
+          5.  & HPMF mean 3                                         & 4.33\\
+          6.  & Impoundment median 4                                & 3.84\\
+          7.  & HPMF Gabor - streams removed                        & 3.54\\
+          8.  & HPMF median 4                                       & 3.40\\
+          9.  & Impoundment median 2                                & 3.01\\
+          10. & Sky View Factor Gabor - streams removed             & 2.84\\
+          11. & HPMF mean 6                                         & 2.73\\
+          12. & Impoundment median 6                                & 2.69\\
+          13. & Impoundment standard deviation 4                    & 2.52\\
+          14. & Sky View Factor Gabor                               & 2.44\\
+          15. & Impoundment ditch amplification                     & 2.43\\
+          16. & Impoundment mean 6                                  & 2.27\\
+          17. & Impoundment ditch amplification - streams removed   & 2.24\\
+          18. & HPMF min 2                                          & 2.23\\
+          19. & Slope standard deviation 6                          & 2.01\\
+          20. & Sky View Factor non-ditch amplification             & 1.88\\
           \bottomrule
@@
 This, in combination with the surrounding peat having subsided \citep{heikurainen}, have rendered them almost impossible to detect in the DEM today (\hyperref[fig:ditchpictures]{Figure} \ref{fig:ditchpictures} \hyperref[fig:ditchpictures]{d}).
-Despite these issues we had a recall rate of 77.4\%, showing that the majority of the ditches could still be retrieved automatically.
+Despite these issues we had a recall rate of 70.28\%, showing that the majority of the ditches could still be retrieved automatically.
 
@@
 \citet{rapinel} found that the results were usually more sensitive to the point density (ranging 1-4 points $m^{2}$ in their study) than the interpolation method.
-They retrieved 54.8 and 63.8 \% of the ditches on the Acuey and Boucey marches in France, compared to 77.4 \% in our study.
+They retrieved 54.8 and 63.8 \% of the ditches on the Acuey and Boucey marches in France, compared to 70.28 \% in our study.
 However, in our study we had a point density of 20 points $m^{2}$, so the quality of the DEM in our study ought to be robust, which may have contributed to our higher retrieval rate.
@@
 This makes it more difficult to detect ditches.
-However, despite these coarser grid points in the DEM, \textbf{c} and \textbf{d} shows that the ditch locator still managed to capture a majority of the ditches (green), and only failed at certain points (red).}
+However, despite these coarser grid points in the DEM, \textbf{c} and \textbf{d} shows that the ditch detector still managed to capture a majority of the ditches (green), and only failed at certain points (red).}
     \label{fig:resultstreesbushes}
@@
 The results from our method (\hyperref[predictionperformance]{Table} \ref{predictionperformance}) show that we managed to attain a Cohen's $\kappa$ rating in the substantial range, according to performance thresholds proposed by \citet{kappaanalysis}.
-This means that our ditch locator performed substantially better than a completely random ditch locator.
-The total recall value of 77.4 \% shows that we managed to find a majority of the ditch pixels that exist.
+This means that our ditch detector performed substantially better than a completely random ditch detector.
+The total recall value of 70.28 \% shows that we managed to find a majority of the ditch pixels that exist.
 The results of our method also showed a significant improvement for ditch detection over classifying with the digital terrain indices separately.
@@
 One concern we had with our method was that valuable information may have been discarded when aggregating the elevations from the original point cloud to a DEM, making it more difficult to locate the ditches.
-However, we achieved a similar score to their method; $\kappa=0.75$ in our study compared to 0.77 for grassland and 0.73 for peri-urban area in Roelens study.
+However, we achieved a similar score to their method; $\kappa=0.73$ in our study compared to 0.77 for grassland and 0.73 for peri-urban area in Roelens study.
 This suggests that you can also use a DEM to automatically extract ditches, provided that you have enough input data to generate a high resolution DEM to capture the small scale features of ditches.
@@
 Approximately 75\% of the ditches in their study were retrieved where no vegetation existed, whereas the detection levels sank to below 10\% under vegetation; especially when high vegetation, shrubs and trees covered the ditches \citep{bailly}.
-Hence, a $\kappa$ of 0.75 in the Krycklan catchment with  87\% forest cover \citep{krycklancatchment} is in that aspect surprisingly good, and could perhaps be a result of the higher point density of our LiDAR measurements (20 compared to 10 points $m^{2}$).
+Hence, a $\kappa$ of 0.73 in the Krycklan catchment with  87\% forest cover \citep{krycklancatchment} is in that aspect surprisingly good, and could perhaps be a result of the higher point density of our LiDAR measurements (20 compared to 10 points $m^{2}$).
 Or, perhaps the generation of a DEM from the point cloud has made the detection more robust?
@@
 This method did not remove the smaller streams (\hyperref[fig:resultsillustrations]{Figure} \ref{fig:resultsillustrations} \hyperref[fig:resultsillustrations]{b}).
-A future improvement to the ditch locator could be to use a shape index to remove small natural streams, as natural streams often form more complex channels whereas ditches are often straight.
-A potential issue with generalising our ditch locator for use in other geographical areas is that all the post-processing steps and input variables were developed based on occurrences in the Krycklan area.
+A future improvement to the ditch detector could be to use a shape index to remove small natural streams, as natural streams often form more complex channels whereas ditches are often straight.
+A potential issue with generalising our ditch detector for use in other geographical areas is that all the post-processing steps and input variables were developed based on occurrences in the Krycklan area.
 Different geographical compositions in other areas may require that the thresholds used to fill gaps in ditches and to remove noise would need to be adjusted.
@@
 
-Many of the input variables, could be reused in the future, with Random Forests or other machine learning algorithms.
+Many of the input variables could be reused in the future, with Random Forests or other machine learning algorithms.
 Several of the post-processing algorithms, such as the probability prediction noise removal, and the cluster removal from the binary prediction could be generalised for use in any type of segmentation of ditches.

### 2020-08-09 LaTeX/Ditch detection in high-resolution digital elevation models.tex
Performs GIS-related changes in the article.

@@
 Previous attempts \citep{uppsala} have shown that methods such as Sky View Factor \citep{zaksek}, Impoundment Index \citep{whiteboxtools} and High Pass Median Filter \citep{whiteboxtools} can be used to detect the ditches.
-However, all methods yield false negatives, and Sky View Factor tends to give a large amount of false positives in steep terrain, whereas Impoundment Index produces more false positives in flat terrain if volume is used as a measure, and will falsely detect natural stream channels if dam height is used as a measure.
-High Pass Median Filter tends to also detect small sinks as ditches.
+Different false positives and negatives can be observed when examining each terrain index manually.
+Sky View Factor tends to give a large amount of false positives in steep terrain, whereas Impoundment Index and High Pass Median Filter are more difficult to analyse.
+Due to the difference in false positives and negatives, we want to combine the indices to help prune the weaknesses and amplify the strengths of each index.
 This is in part due to the Light Detection and Ranging (LiDAR) measurements sometimes being interfered by trees, bushes, or grass - leaving gaps in the ditch model, or sometimes due to a poorly set threshold.
@@
 From the ditch mapping, the vector layer was rasterised so that it could be compared to the automatically derived ditches from the $0.5*0.5$ m DEM.
-Because the observed average width of ditches is larger than 0.5 metres, all pixels within a radius of three pixels (1.5 metres) of the vectors were labelled as ditch pixels.
+Although we have no data about the width of the ditches in the catchment, field observations have shown that the vast majority of ditches are between 0.5 and 3.5 metres wide.
+Because we wanted to ensure that the model received all ditch pixels correctly in the training phase, we widened the ditches so that all pixels within a radius of three pixels (1.5 metres) of the vectors were labelled as ditch pixels.
+This ensured that the edges of ditches were included, as well as that the raster converted from vector format actually covered the ditches.
 \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{a} shows the ditches rasterised from vectors and \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{b} shows the ditches after widening.
@@
 This method tends to give false positives in steep terrain, where the sky can be obscured by a hillside, or in narrow valleys or ravines that also show low numbers.
-The Sky View Factor was calculated in SAGA GIS, using a search radius of 10 m.
+The Sky View Factor was calculated in SAGA GIS, using a search radius of 10 metres to ensure that ditches were covered, and to not let steep hills close to ditches affect the results.
 
@@
 Additionally, the tool will output a 'dam height' raster that is particularly useful for mapping channels, such as ditches.
-Here, we used dam height calculated with a dam length of 3 m.
+Here, we used dam height calculated with a dam length of 3 metres, to cover the entire ditch which is typically not wider than 3 metres.
 Typical false positives with this method are natural stream channels with a clear incised channel.

### 2020-09-05 LaTeX/Ditch detection in high-resolution digital elevation models.tex
Adds new bar plot and some paragraphs describing watercourse mappings in Sweden.

@@
 \usepackage[parfill]{parskip}
+
+% Ploting
+\usepackage{pgfplots}
+\pgfplotsset{width=7cm,compat=1.8}
+\definecolor{custombeige}{HTML}{e0d4ba}
+\definecolor{custombeigeouter}{HTML}{968f7e}
+\definecolor{customgray}{HTML}{666666}
+\definecolor{customgrayouter}{HTML}{404040}
 
@@
 \label{method}
+
+\subsection{The need for better small scale channel mappings}
+To investigate the need for better maps of ditches and small stream channels we compared the data from a national field survey with available maps.
+The National Inventory of Landscapes in Sweden (NILS) has recorded small watercourses (width $<$ 6 m) by conducting a line inventory in 631 $5*5$ km squares randomly distributed in Sweden, to best capture a representative and varied sample of the Swedish landscape.
+The position of the small watercourses from this inventory were compared with the watercourses on Swedish Property Map (1: 12 500) (allowing for an uncertainty in the positioning of up to 15 m, by snapping the point observations to the lines on the map) (\hyperref[fig:watercoursebarplot]{Figure} \ref{fig:watercoursebarplot}).
 
@@
 
+A total of 1103 natural watercourses, 131 straightened watercourses/channels, and 2089 ditches were found in NILS field inventory.
+Analysing the data from NILS, two things are striking: Firstly, the ditches make up the majority of the small scale water channels in the Swedish landscape; almost twice as many as the natural streams.
+Secondly, most of the small scale channels are missing on the current maps; only 45 \% of the natural watercourses, 25 \% of the straightened watercourses/channels, and 9 \% of the ditches were mapped (\hyperref[fig:watercoursebarplot]{Figure} \ref{fig:watercoursebarplot}).
+
+\begin{figure}[htb]
+    \centering
+    \begin{tikzpicture}
+    \begin{axis}[
+        width={0.9\textwidth},
+        height={0.6\textwidth},
+        ylabel={fluorescence},
+        ybar stacked,
+        ymajorgrids=true,
+    	bar width=45pt,
+    	nodes near coords,
+    	every node near coord/.append style={font=\tiny},
+        enlarge y limits={0.15},
+        enlarge x limits=0.15,
+        legend style={at={(0.78,0.96)},
+          anchor=north,legend columns=-1},
+        ylabel={{}},
+        symbolic x coords={Ditches, Straightened  watercourses, Natural  watercourses},
+        xtick=data,
+        ytick={0,500,1000,1500,2000},
+        x tick label style={rotate=0,anchor=north},
+        ]
+    \addplot+[ybar,color=custombeige,fill=custombeige,draw=custombeigeouter,thick] plot coordinates {(Ditches,188) (Straightened  watercourses,33) 
+      (Natural  watercourses,496)};
+    \addplot+[ybar,color=customgrayouter,fill=customgray,draw=customgrayouter,thick] plot coordinates {(Ditches,1901) (Straightened  watercourses,98) 
+      (Natural  watercourses,607)};
+    \legend{\strut Mapped, \strut Unmapped}
+    \end{axis}
+    \end{tikzpicture}
+    \caption{The bars indicate the number of observed channels in the National Inventory of Landscapes in Sweden's (NILS) line inventory of small watercourses (width $<$ 6 m).
+The colours of the bars indicate if they are mapped or not on the Swedish Property map.}
+    \label{fig:watercoursebarplot}
+\end{figure}
+
 In \hyperref[featureimportancetable]{Table} \ref{featureimportancetable}, the top 20 input variables from the Random Forests models are presented with their importance percentages for the models making a successful prediction.
@@
 
-In our study we have shown that it is possible to locate ditches automatically in high-resolution DEMs ($0.5  * 0.5 $ metres, in our case), and that more of the ditches can be detected if the information from several terrain indices (\hyperref[recreatedpredictionperformance]{Table} \ref{recreatedpredictionperformance}) is combined through machine learning than if indices are used separately (\hyperref[predictionperformance]{Table} \ref{predictionperformance}).
+The comparison with the field observed channels and the available maps clearly indicate the need for better maps of ditches (\hyperref[fig:watercoursebarplot]{Figure} \ref{fig:watercoursebarplot}).
+Our study has shown that it is possible to locate ditches automatically in high-resolution DEMs ($0.5  * 0.5 $ metres, in our case), and that more of the ditches can be detected if the information from several terrain indices (\hyperref[recreatedpredictionperformance]{Table} \ref{recreatedpredictionperformance}) is combined through machine learning than if indices are used separately (\hyperref[predictionperformance]{Table} \ref{predictionperformance}).
 The difficulty of detecting the ditches in the study catchment vary greatly from ditch to ditch, this is illustrated in \hyperref[fig:ditchpictures]{Figure} \ref{fig:ditchpictures}.
@@
 
+With our method, natural streams were often classified as false positives (\hyperref[fig:resultsillustrations]{Figure} \ref{fig:resultsillustrations} \hyperref[fig:resultsillustrations]{b}) as our aim was focused on locating ditches.
+However, this indicates that our method could also be used to map small, previously unmapped natural stream channels.
+A major knowledge gap, both nationally  \citep{kuglerova} and internationally \citep{benstead} is how to best identify the location of small brooks and creeks in the landscape - a phenomenon called “Aqua Incognita, the unknown headwaters” \citep{bishop, kuglerova}.
+
 The results from our method (\hyperref[predictionperformance]{Table} \ref{predictionperformance}) show that we managed to attain a Cohen's $\kappa$ rating in the substantial range, according to performance thresholds proposed by \citet{kappaanalysis}.
@@
 
-%This study investigates how to use digital elevation data together with manually labelled ditches to identify patterns and relationships that can be used for automatic ditch detection.
-The proposed method significantly outperforms classifying with the digital terrain indices separately.
-Although the method still has room for improvement, it performs well on the available data.
-In the future, a more extensive hyperparameter tuning could be performed, because, as \citet{lavesson} states; hyperparameter tuning is often more important than choice of algorithm.
-Some of the custom input variables could be reused in the future, with Random Forests or other machine learning algorithms.
-Several of the post-processing algorithms, such as the probability prediction noise removal or the cluster removal from the binary prediction could be generalised for use in any type of pixel classification of ditches.
-
-%The work in this study gives an insight into what input variables work well for ditch detection and what input variables do not, as well as an increased understanding of how machine learning can be applied to ditch detection.
-The proposed method may help in making ditch detection both faster and easier, reducing manual labour and cost.
-In conclusion, this study has shown that it is possible to use machine learning with digital elevation models to learn patterns that enable robust detection of ditches.
-
 This study investigates how to use digital elevation data together with manually labelled ditches to identify patterns and relationships that can be used for automatic ditch detection.

### 2020-09-14 LaTeX/proofreading_rewrite.tex
Adds a lot of changes after Sidd's read-through.

--- 2020-09-05 LaTeX/Ditch detection in high-resolution digital elevation models.tex
@@
 Thresholding methods using digital terrain indices, such as Sky View Factor, Impoundment Index, or High Pass Median Filter can be used to detect ditches.
-However, a single threshold generally does not capture the variability in the landscape, generating many false positives and negatives.
-We hypothesised that, by combining the digital terrain indices using a machine learner, we could improve the ditch detection.
+However, a single threshold generally does not capture the variability in the landscape and generates many false positives and negatives.
+We hypothesised that, by combining the digital terrain indices using a machine learner, we could improve ditch detection at a landscape-scale.
 
@@
 
-The study shows that by combining the information in several digital terrain indices using machine learning, more ditch networks could be automatically retrieved.
+The study demonstrates that combining information from  a suite of digital terrain indices using machine learning provides an effective technique for automatic ditch detection at a landscape-scale.
 \end{abstract}
@@
 
-Over the last century many of the wetlands in North Europe and parts of North America have been ditched.
-Mapping and modelling the hydrology of a flat landscape is challenging because of the low topographic gradients.
-In many places both the landscape and the channel network have also been heavily modified by humans, especially in agricultural landscapes, for example in mid-western United States \citep{passalacqua}.
-In North Europe, ditches have also been dug to turn flat peatlands into productive forests, thereby profoundly altering the hydrology in North Europe \citep{peatlands}.
-Sweden is one of the North European countries with the most ditches; over five million hectares of peatland have been drained for forestry in the 20th century.
-This has resulted in almost one million km of ditches, which has more than doubled the length of watercourses in Sweden \citep{hasselquist}.
+Over the last century, the wetlands in North Europe and parts of North America have been substantially modified and ditched.
+In many places both the landscape and the channel network have also been heavily altered  by humans, especially in the agricultural landscapes, for example in mid-western United States \citep{passalacqua}.
+In North Europe, however, ditches have been dug to turn flat peatlands into productive forests, resulting in profound impacts on the hydrology of this region.
+\citep{peatlands}.
+In Sweden which is one of the North European countries with the most extensive ditch network, over five million hectares of peatland have been drained for forestry in the 20th century.
+As a result, almost one million km of ditches have been introduced and doubled the length of watercourses across Sweden  \citep{hasselquist}.
 
-A comparison between a field inventory of ditches (within the National Inventory of Landscapes in Sweden (NILS)) and current maps, shows that approximately 90 \% of the ditch networks are missing on current maps ({\AA}gren, unpublished).
-To better prioritise the restoration of ditched wetlands, it is necessary to know where the ditches are located, but field inventories are often too costly.
-The research objective of this study is therefore to automatically detect ditches from high-resolution digital elevation models (DEMs) using digital terrain indices.
-Previous attempts \citep{uppsala} have shown that methods such as Sky View Factor \citep{zaksek}, Impoundment Index \citep{whiteboxtools} and High Pass Median Filter \citep{whiteboxtools} can be used to detect the ditches.
-Different false positives and negatives can be observed when examining each terrain index manually.
-Sky View Factor tends to give a large amount of false positives in steep terrain, whereas Impoundment Index and High Pass Median Filter are more difficult to analyse.
-Due to the difference in false positives and negatives, we want to combine the indices to help prune the weaknesses and amplify the strengths of each index.
-This is in part due to the Light Detection and Ranging (LiDAR) measurements sometimes being interfered by trees, bushes, or grass - leaving gaps in the ditch model, or sometimes due to a poorly set threshold.
-Therefore, using solely the digital terrain indices for ditch mapping tends to yield an imprecise result.
+For effective restoration of ditched wetlands, it is highly important  to accurately locate the ditches on digital maps, whereas landscape-scale field inventories are often challenging and expensive.
+Automatic detection of ditches from high-resolution digital elevation models (DEMs) and terrain indices provides a feasible means for spatially explicit assessment of ditches across a landscape.
+Previous researches  \citep{uppsala} documented  that terrain indices such as Sky View Factor \citep{zaksek}, Impoundment Index \citep{whiteboxtools} and High Pass Median Filter \citep{whiteboxtools} can efficiently  detect the ditches.
+However, different false positives and negatives may be observed when examining each terrain index manually.
+For example, Sky View Factor tends to give a large amount of false positives in steep terrain whereas Impoundment Index and High Pass Median Filter are more difficult to analyse.
+Such inaccuracies and complexities are in part related to the Light Detection and Ranging (LiDAR) measurements, the mainstay of high resolution DEM and terrain mapping, sometimes being interfered by trees, bushes, or grass - leaving gaps in the ditch model.
+The inaccuracies can also arise from a poorly set threshold for the LiDAR derived terrain parameters.
+Thus, it would be useful to combine the terrain indices derived from high resolution LiDAR data to prune the weaknesses and amplify the strengths of each index and produce highly accurate detection of ditch networks.
 
-In this study, we introduce new methods for the automated mapping of ditches.
-From the high resolution DEM, we use machine learning to locate the ditch networks.
-We also examine which digital terrain indices are the most relevant for a correct ditch prediction.
-To accomplish this, we conduct experiments using a supervised learning algorithm with labelled ditch data to automatically extract ditches from externally supplied occurrences (data not before seen by the model) \citep{kotsiantis}.
-These occurrences are used as input variables to the algorithm, and these input variables consist of refined or raw digital terrain indices derived from a DEM.
-
-There are other studies where geographical data has been used with machine learning to map ditches.
-\citet{roelens} used the Random Forests algorithm to detect ditches in Belgium, using LiDAR point clouds.
-They used several different input variables where neighbouring points of a specific LiDAR point were examined.
-Values of all LiDAR attributes were represented for the neighbouring area for both 0.5, 1, 1.5, 2, 3 and 4 metre radii.
-We reproduce these neighbouring area variables when building the ditch detector in our study.
-Several other input variables for the model are calculated from the indices to place emphasis on the clear linear directional characteristics of ditches.
-These characteristics are also used when processing the model's output, in order to remove noise and other inconsistencies in the predictions.
-In a pilot study, we compared several different algorithms (Random Forests, Extreme Gradient Boosting, Naive Bayes, and Support Vector Machines), and it was found that Random Forests produced the most accurate results.
-Random Forests also computes Mean Decrease in Impurity, or Gini importance, highlighting what input variables are the most important for a given prediction \citep{gini}.
-Because of these factors, and the fact that it proved to be suitable for \citet{roelens}, Random Forests models are used to locate the ditches.
+In this study, we first investigate the need for better maps of ditch networks, then we introduce a new method for the automated mapping of ditches by integrating multiple high resolution terrain indices (derived from LiDAR DEM) using a machine learning algorithm.
+We also examine which digital terrain indices are the most relevant for an accurate prediction of ditches.
+We use a supervised learning algorithm with labelled ditch data to automatically extract ditches from instances not before seen by the model.
+These instances are used as input variables to the algorithm, and consist of refined or raw digital terrain indices derived from a DEM.
 
 We hypothesise that by combining the information from all the digital terrain indices using a machine learner (Random Forests), we can improve the detection of the ditches, compared to if a single index would be used.
-To test this, we conduct a study using a high resolution DEM of the Krycklan Catchment in Sweden \citep{krycklancatchment}, where a high precision manual mapping has been performed.
-This mapping was used to both train the Random Forests models and to produce the ground-truth.
-To evaluate the performance of the ditch detector, we capture the various aspects of detection performance through the following measures: accuracy, precision, recall, Cohen’s Kappa index, and Area under Precision-Recall curve score.
+To test this, we conducted a landscape-scale ditch detection analysis at the Krycklan Catchment in Sweden \citep{krycklancatchment}, where a high precision manual mapping has been performed previously.
+This manual map guided the model training and ground-truthing in our analysis and provided a test-bench for our methodology.
+Several other studies \citep{roelens, bailly, rapinel, kiss} that have located ditches from LiDAR data have worked on smaller areas (smaller than 2 $km^2$) where artificial drainage dominates.
+Here, we work on landscape scale in a mesoscale catchment.
+The  Krycklan Study Catchment in Sweden is 68 $km^2$, which means that we cover more landscape types.
 
@@
 \subsection{Study area}
-Several other studies \citep{roelens, bailly, rapinel, kiss} that have located ditches from LiDAR data have worked on smaller areas (smaller than 2 $km^2$) where artificial drainage dominates.
-Here, we work on landscape scale in a mesoscale catchment.
-The  Krycklan Study Catchment in Sweden is 68 $km^2$, which means that we cover more landscape types.
-The quaternary deposits are dominated by till (51\%) and sorted sediments (30\%).
+In the Krycklan Study Catchment the quaternary deposits are dominated by till (51\%) and sorted sediments (30\%).
 The catchment ranges in elevation from 114 to 405 m above sea level.
 Forest covers 87\% of the catchment, mires 9\%, thin soils 7\%, and rock outcrops 1\%.
-The land use is dominated by forestry; approximately 25\% of the Krycklan catchment has been protected since 1922, but most of the other area is second growth forest \citep{krycklancatchment}.
+The land use is dominated by forestry; approximately 25\% of the Krycklan catchment has been protected since 1922, but  the rest of the area is mostly second growth forest \citep{krycklancatchment}.
 
-A historical survey of the catchment (Gudrun Norstedt, unpublished) shows that the ditches in the catchment were dug between 1900 and 1993.
-The earliest ditches were used to drain mires to harvest hay.
-Later, many of the ditches were dug based on the notion that any mire could be turned into productive forest land by drainage \citep{paivanen}.
-Today we know that nutrient limitation also plays a key role, and some drained wetlands did not turn into productive forest \citep{sikstrom}.
-Once a canopy of trees and shrubs has been established in the forest, the increased evapotranspiration \citep{price, koivusalo}, increased interception \citep{price}, and reduced groundwater recharge will continue to naturally lower the water-table in these areas, making ditch maintenance unnecessary, at least until the clear-cut phase when the canopy is removed.
-Also,  lowering the water table, the oxygenated soils show increasing bulk density and decreasing hydraulic conductivity \citep{silins, price}, causing a subsidence of the peat layer near the ditch.
-This is likely the most important factor causing ditches to become shallow \citep{heikurainen}.
+A historical survey of the catchment (Gudrun Norstedt, unpublished) recorded  that the ditches in the catchment were dug between 1900 and 1993.
+Initially, the ditches were used to drain the mires for hay production while the drained mires were later attempted to turn into productive forest lands \citep{paivanen}.
+However, research later showed that many of the drained mires were not productive for forestry due to their nutrient deficient environment \citep{sikstrom}.
 
 \subsection{Digitising the ground truth}
-A high resolution DEM was generated from LiDAR data, see \hyperref[lidartodem]{'Data and codes availability statement'} for specifications of the process.
-To train and evaluate the Random Forests models, a vector layer of all ditches in the Krycklan Catchment \citep{krycklancatchment} was digitised.
-Road ditches, forest ditches, and agricultural ditches can be found in the catchment.
+A high resolution DEM was generated from LiDAR data, see \hyperref[lidartodem]{'Data and codes availability statement'}.
+To train and validate the machine learning models, a vector layer of all ditches in the Krycklan Catchment \citep{krycklancatchment} was digitised.
+The study catchment comprises several types of ditches including road ditches, forest ditches, and agricultural ditches.
 A multidirectional hillshade of the 0.5 m DEM was calculated in ArcGIS Pro to locate the ditches.
 From this layer, most of the ditches could be easily identified.
-To ensure that the digitised ditch was located correctly inside the ditch and not on the bank nearby, the digitisation was conducted by zooming in to a scale of 1:600.
-In addition to this, both current and historical records and orthophotos dating back to 1960’s were used to help locate all the ditches.
-In uncertain areas, we also found it useful to view the data in 3D.
-However, the digitising was conducted in 2D, and in uncertain areas the ditches were verified in the field.
-We are confident that we have captured all the ditches in the catchment according to the digitised  ditch layers in \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan}.
+We conducted the digitisation at a scale of 1:600 to ensure all the digitised ditches properly align the actual ditch on the ground and do not coincide with the edge of the banks.
+In addition, both current and historical records and orthophotos dating back to the 1960's were consulted to help locate all the ditches.
+In uncertain areas, the ditches were verified in the field.
 The stream network in \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan} shows a network with a 8 hectare flow initiation threshold.
-Though, the length of the stream network vary substantially between season, ranging from 140 km during baseflow to 630 km during high-flow conditions \citep{mappingtemporal}.
-The digitisation of the ditch networks showed that there are 107 km road ditches and 208 km forest/agricultural ditches (\hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan}).
-The stream network in \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan}, which aims to show median flow conditions, was 197 km.
+The digitisation of the ditch networks showed that there are 107 km of road ditches and 208 km of forest/agricultural ditches in the catchment (\hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan}).
+The stream network in \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan}, with median flow conditions was 197 km.
 
@@
 From the ditch mapping, the vector layer was rasterised so that it could be compared to the automatically derived ditches from the $0.5*0.5$ m DEM.
-Although we have no data about the width of the ditches in the catchment, field observations have shown that the vast majority of ditches are between 0.5 and 3.5 metres wide.
-Because we wanted to ensure that the model received all ditch pixels correctly in the training phase, we widened the ditches so that all pixels within a radius of three pixels (1.5 metres) of the vectors were labelled as ditch pixels.
-This ensured that the edges of ditches were included, as well as that the raster converted from vector format actually covered the ditches.
+Our field observations have shown that the  width of most ditches within the catchment ranges between 0.5 and 3.5 m.
+To ensure that the model received all ditch pixels correctly in the training phase, we widened the ditches so that all pixels within a radius of three pixels (1.5 metres) of the vectors were labelled as ditch pixels.
 \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{a} shows the ditches rasterised from vectors and \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{b} shows the ditches after widening.
-The data in \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{b} is the labelled data that was used to train the Random Forests models.
+The data in \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{b} is the labelled data that was used to train the machine learning models.
 The ditches all have a generalised width based on a perceived average of ditch width in the area.
@@
 Each grid cell that contained at least 25\% ditch pixels was then labelled as a ditch, see \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{c}.
-This layer was used to ground-truth the Random Forests models.
+This layer was used to ground-truth the machine learning models.
 
@@
 On an open field all of the sky would be visible, but at the bottom of a ditch much of the sky would be obscured, meaning low values would indicate ditches.
-This method tends to give false positives in steep terrain, where the sky can be obscured by a hillside, or in narrow valleys or ravines that also show low numbers.
 The Sky View Factor was calculated in SAGA GIS, using a search radius of 10 metres to ensure that ditches were covered, and to not let steep hills close to ditches affect the results.
@@
 \subsubsection{Processing the digital terrain indices}
-To improve the performance of the Random Forests models, we examined how different kinds of input variables affected the prediction.
+We examined how different  input variables affected the prediction of the machine learning models.
 Several possible data manipulation methods could theoretically produce a better prediction.
@@
 In addition to the general input variables, a number of custom input variables were also developed.
-Several of these variables were calculated by looking at the separate digital terrain indices, attempting to amplify the ditches.
-This was done in several different ways for the indices by combining statistical aggregations, and by using thresholds to exclude, lower, or amplify the values of pixels that indicated ditches or non-ditches (\hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{d} and \hyperref[fig:features]{h}).
-
-From the Slope and Sky View Factor indices, input variables for the models were also extracted to attempt to amplify the values of pixels that lay in streams or in hilly areas, but leave pixels in ditches or other geographical areas untouched.
+Some of these custom variables were produced by statistical aggregation of multiple indices whereas in some cases, they were generated by using thresholds to exclude, lower, or amplify the values of pixels that indicated ditches or non-ditches (\hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{d} and \hyperref[fig:features]{h}).
 
 Both HPMF and Sky View Factor were used with the image processing \textit{gabor filter}, which can be used to detect lines of a certain orientation in an image \citep{gabor}.
-30 Gabor filters, which were rotated in different angles and with different frequencies, were combined to detect lines, amplifying ditches by utilising the fact that ditches have a linear elongated shape (\hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{e}).
+30 Gabor filters which were rotated in different angles and with different frequencies were combined to detect lines, amplifying ditches by utilising the fact that ditches have a linear elongated shape (\hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{e}).
 
@@
 The raw Impoundment Index dam height was used to create a mask, attempting to mark out streams, but retain ditches.
-Only areas with a relatively high dam height, above 14 metres, which would indicate that these areas most likely contained streams, were marked out.
-After widening the resulting area, this mask was used to attempt to remove streams from all the aforementioned custom input variables, generating one new input variable from each.
+Only areas with a relatively high dam height, above 14 metres, which would indicate that these areas most likely contained streams, were masked out.
+After widening the resulting area, this mask was used to remove streams from all the aforementioned custom input variables, generating one new input variable from each.
 The features marked with \textit{streams removed} in \hyperref[featuretable]{Table} \ref{featuretable} make use of this filter (\hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{i}).
@@
 Overall, 81 input variables were extracted from the terrain indices.
+In a pilot study, we compared several different algorithms (Random Forests, Extreme Gradient Boosting, Naive Bayes, and Support Vector Machines), and it was found that Random Forests produced the most accurate results, here we therefore only present the results from the Random Forest model.
 We then conducted a sub-experiment to find the best input variables, as well as the optimal number of variables to use.
-Random Forest's feature importance was used to select the most valuable input variables for different variable amounts.
-It was found that using 40 input variables produced the best results.
+The variable importance score in Random Forests was used to select the most important input variables and it was found that using the top 40 input variables produced the best results.
 
@@
 The remaining 10 subsections were used solely before the experiment to develop the ditch detector to test different ways of preparing the data.
-This allowed the ditch detector to be evaluated on unseen data to strengthen the validity of the experiment.
+This allowed the ditch detector to be evaluated on  independent data and  strengthened the validity of the experiment.
 \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan} shows which subsections were used for development and evaluation respectively.
@@
 \subsubsection{Noise reduction and gap filling}
-The probability predictions contained a lot of noise in geographical locations far away from ditches, which needed to be excluded.
-The first step for removing noise was to use a bilateral de-noising filter on the entire prediction image.
+The probability predictions contained a lot of noise in  the areas largely distant from the ditches.
+The first step for removing this noise was to use a bilateral de-noising filter on the entire prediction image.
 This left linear properties and pixels with a very high value intact, while lowering the value of pixels that did not contribute to an accurate prediction (\hyperref[fig:postprocessing1]{Figure} \ref{fig:postprocessing1} \hyperref[fig:postprocessing1]{b}).
@@
 In this function, a threshold value was used to avoid removing pixels with a high enough ditch probability, helping to retain pixels that lay in or close to a ditch.
-The max ditch probability value in a circular radius of 10 pixels was then calculated.
-If this max value was too low, the probability value of the examined pixel was lowered (\hyperref[fig:postprocessing2]{Figure} \ref{fig:postprocessing2} \hyperref[fig:postprocessing2]{a}).
+The maximum ditch probability value in a circular radius of 10 pixels was then calculated.
+If this maximum value was too low, the probability value of the examined pixel was lowered (\hyperref[fig:postprocessing2]{Figure} \ref{fig:postprocessing2} \hyperref[fig:postprocessing2]{a}).
 
@@
 The models' ditch prediction is given in the original resolution raster format.
-To allow for comparison to the evaluation data, the same grid conversion was performed on the prediction raster as on the evaluation data seen in \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{c}.
-A mean probability rating was calculated for each $6*6$ grid zone, classifying the entire zone as part of a ditch if the mean probability exceeded 40 \%.
-Different zone sizes and probability thresholds were evaluated in a pilot study, and these values were found to produce the best results.
-This zone partitioning helped to fill in gaps where lone pixels in ditches had been incorrectly classified, and also helped in the next step of the post-processing (\hyperref[fig:postprocessing3]{Figure} \ref{fig:postprocessing3} \hyperref[fig:postprocessing3]{a}).
+To allow for comparison to the evaluation data, the same grid conversion was performed on the prediction raster as on the evaluation data displayed in \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{c}.
+After testing different combinations of zone sizes and probability values, a mean probability rating was calculated for each $6*6$ grid zone, classifying the entire zone as part of a ditch if the mean probability exceeded 40 \%.
+This zone partitioning helped to fill in gaps where lone pixels in ditches had been incorrectly classified.
+(\hyperref[fig:postprocessing3]{Figure} \ref{fig:postprocessing3} \hyperref[fig:postprocessing3]{a}).
 
@@
 \section{Results and analysis}
+In the national survey (NILS), a total of 1103 natural watercourses, 131 straightened watercourses/channels, and 2089 ditches were found in the field.
+This simple investigation highlights two things: Firstly, the ditches make up the majority of the small scale water channels in the Swedish landscape; almost twice as many as the natural streams.
+Secondly, most of the small scale channels are missing on the current maps; only 45 \% of the natural watercourses, 25 \% of the straightened watercourses/channels, and 9 \% of the ditches were mapped (\hyperref[fig:watercoursebarplot]{Figure} \ref{fig:watercoursebarplot}).
+
+\begin{figure}[htb]
+    \centering
+    \begin{tikzpicture}
+    \begin{axis}[
+        width={0.9\textwidth},
+        height={0.6\textwidth},
+        ylabel={fluorescence},
+        ybar stacked,
+        ymajorgrids=true,
+    	bar width=45pt,
+    	nodes near coords,
+    	every node near coord/.append style={font=\tiny},
+        enlarge y limits={0.15},
+        enlarge x limits=0.15,
+        legend style={at={(0.78,0.96)},
+          anchor=north,legend columns=-1},
+        ylabel={{}},
+        symbolic x coords={Ditches, Straightened  watercourses, Natural  watercourses},
+        xtick=data,
+        ytick={0,500,1000,1500,2000},
+        x tick label style={rotate=0,anchor=north},
+        ]
+    \addplot+[ybar,color=custombeige,fill=custombeige,draw=custombeigeouter,thick] plot coordinates {(Ditches,188) (Straightened  watercourses,33) 
+      (Natural  watercourses,496)};
+    \addplot+[ybar,color=customgrayouter,fill=customgray,draw=customgrayouter,thick] plot coordinates {(Ditches,1901) (Straightened  watercourses,98) 
+      (Natural  watercourses,607)};
+    \legend{\strut Mapped, \strut Unmapped}
+    \end{axis}
+    \end{tikzpicture}
+    \caption{The bars indicate the number of observed channels in the National Inventory of Landscapes in Sweden's (NILS) line inventory of small watercourses (width $<$ 6 m).
+The colours of the bars indicate if they are mapped or not on the Swedish Property map.}
+    \label{fig:watercoursebarplot}
+\end{figure}
 
@@
 
-From \hyperref[predictionperformance]{Table} \ref{predictionperformance}, it can be seen that averaging the results from the 11 subsections will yield a very similar value to the total results, indicating that the ditch detector generally performs equally well on subsections with a small amount of ditches as on subsections with a large amount of ditches.
+From \hyperref[predictionperformance]{Table} \ref{predictionperformance}, it is observed that averaging the results from the 11 subsections would yield a very similar value to the total results, indicating that the ditch detector generally performs equally well on subsections with a small amount of ditches as on subsections with a large amount of ditches.
 
@@
 
-A total of 1103 natural watercourses, 131 straightened watercourses/channels, and 2089 ditches were found in NILS field inventory.
-Analysing the data from NILS, two things are striking: Firstly, the ditches make up the majority of the small scale water channels in the Swedish landscape; almost twice as many as the natural streams.
-Secondly, most of the small scale channels are missing on the current maps; only 45 \% of the natural watercourses, 25 \% of the straightened watercourses/channels, and 9 \% of the ditches were mapped (\hyperref[fig:watercoursebarplot]{Figure} \ref{fig:watercoursebarplot}).
-
-\begin{figure}[htb]
-    \centering
-    \begin{tikzpicture}
-    \begin{axis}[
-        width={0.9\textwidth},
-        height={0.6\textwidth},
-        ylabel={fluorescence},
-        ybar stacked,
-        ymajorgrids=true,
-    	bar width=45pt,
-    	nodes near coords,
-    	every node near coord/.append style={font=\tiny},
-        enlarge y limits={0.15},
-        enlarge x limits=0.15,
-        legend style={at={(0.78,0.96)},
-          anchor=north,legend columns=-1},
-        ylabel={{}},
-        symbolic x coords={Ditches, Straightened  watercourses, Natural  watercourses},
-        xtick=data,
-        ytick={0,500,1000,1500,2000},
-        x tick label style={rotate=0,anchor=north},
-        ]
-    \addplot+[ybar,color=custombeige,fill=custombeige,draw=custombeigeouter,thick] plot coordinates {(Ditches,188) (Straightened  watercourses,33) 
-      (Natural  watercourses,496)};
-    \addplot+[ybar,color=customgrayouter,fill=customgray,draw=customgrayouter,thick] plot coordinates {(Ditches,1901) (Straightened  watercourses,98) 
-      (Natural  watercourses,607)};
-    \legend{\strut Mapped, \strut Unmapped}
-    \end{axis}
-    \end{tikzpicture}
-    \caption{The bars indicate the number of observed channels in the National Inventory of Landscapes in Sweden's (NILS) line inventory of small watercourses (width $<$ 6 m).
-The colours of the bars indicate if they are mapped or not on the Swedish Property map.}
-    \label{fig:watercoursebarplot}
-\end{figure}
-
-In \hyperref[featureimportancetable]{Table} \ref{featureimportancetable}, the top 20 input variables from the Random Forests models are presented with their importance percentages for the models making a successful prediction.
-From the table, it can be seen that input variables based on HPMF and Impoundment Index contributed the most to a successful prediction.
+In \hyperref[featureimportancetable]{Table} \ref{featureimportancetable}, the top 20 (out of 40) input variables from the Random Forests models are presented with their importance percentages for the models making a successful prediction.
+Input variables based on HPMF and Impoundment Index contributed the most to a successful prediction.
 The Gini importance was obtained using the Gini impurity for each variable \citep{gini}.
@@
 
-The comparison with the field observed channels and the available maps clearly indicate the need for better maps of ditches (\hyperref[fig:watercoursebarplot]{Figure} \ref{fig:watercoursebarplot}).
+The comparison with the field observed channels and the available maps clearly indicate the need for better maps of ditches, as only 9\% of ditches are mapped.
+(\hyperref[fig:watercoursebarplot]{Figure} \ref{fig:watercoursebarplot}).
 Our study has shown that it is possible to locate ditches automatically in high-resolution DEMs ($0.5  * 0.5 $ metres, in our case), and that more of the ditches can be detected if the information from several terrain indices (\hyperref[recreatedpredictionperformance]{Table} \ref{recreatedpredictionperformance}) is combined through machine learning than if indices are used separately (\hyperref[predictionperformance]{Table} \ref{predictionperformance}).
-The difficulty of detecting the ditches in the study catchment vary greatly from ditch to ditch, this is illustrated in \hyperref[fig:ditchpictures]{Figure} \ref{fig:ditchpictures}.
-By separating the road ditches from the other ditches, a comparison showed that the retrieval rate was slightly higher for the road ditches.
-This is likely due to the fact that they are found in open areas alongside roads, and also that road ditches have generally been better maintained as it has been more prioritised to keep these ditches functioning (\hyperref[fig:ditchpictures]{Figure} \ref{fig:ditchpictures} \hyperref[fig:ditchpictures]{a}).
+The ditch detection performance varied depending on the type of ditches, this is illustrated in \hyperref[fig:ditchpictures]{Figure} \ref{fig:ditchpictures}.
+For example, the retrieval rate was slightly higher for the road ditches compared to forest or agricultural ditches.
+This is likely due to the fact that  road ditches are found in open areas alongside roads, and also,   they are generally well maintained for optimum functioning (\hyperref[fig:ditchpictures]{Figure} \ref{fig:ditchpictures} \hyperref[fig:ditchpictures]{a}).
 However, most of the ditches in Krycklan today are found below the canopy (e.g.
-\hyperref[fig:ditchpictures]{Figure} \ref{fig:ditchpictures} \hyperref[fig:ditchpictures]{c}), and the majority have not been maintained, causing difficulties in detecting them as some have grow back in.
+\hyperref[fig:ditchpictures]{Figure} \ref{fig:ditchpictures} \hyperref[fig:ditchpictures]{c}), and the majority have not been maintained, causing difficulties in detecting them as some have grown back in with peat, grasses, or shrubs (fig 12b, 13b).
 This, in combination with the surrounding peat having subsided \citep{heikurainen}, have rendered them almost impossible to detect in the DEM today (\hyperref[fig:ditchpictures]{Figure} \ref{fig:ditchpictures} \hyperref[fig:ditchpictures]{d}).
@@
 Hypothetically, it should be quite easy to detect ditches using LiDAR data.
-In practice, however, there are many variables affecting the results.
-One is the point density and the interpolation method used to create the DEM.
+In practice, however, there are many variables affecting the results, such as the point density and the interpolation method used to create the DEM.
 \citet{rapinel} found that the results were usually more sensitive to the point density (ranging 1-4 points $m^{2}$ in their study) than the interpolation method.
@@
 However, this indicates that our method could also be used to map small, previously unmapped natural stream channels.
-A major knowledge gap, both nationally  \citep{kuglerova} and internationally \citep{benstead} is how to best identify the location of small brooks and creeks in the landscape - a phenomenon called “Aqua Incognita, the unknown headwaters” \citep{bishop, kuglerova}.
+The national survey showed that 55\% of the natural stream channels and 75\% of the straightened water courses were missing on the maps.
+This is in line with other national \citep{kuglerova} and international \citep{benstead} studies highlighting this phenomenon  called \“Aqua Incognita - the unknown headwaters\” \citep{bishop, kuglerova}.
 
 The results from our method (\hyperref[predictionperformance]{Table} \ref{predictionperformance}) show that we managed to attain a Cohen's $\kappa$ rating in the substantial range, according to performance thresholds proposed by \citet{kappaanalysis}.
-This means that our ditch detector performed substantially better than a completely random ditch detector.
 The total recall value of 70.28 \% shows that we managed to find a majority of the ditch pixels that exist.
-The results of our method also showed a significant improvement for ditch detection over classifying with the digital terrain indices separately.
+The results of our method also showed a significant improvement for ditch detection over classifying with the digital terrain indices separately (\hyperref[recreatedpredictionperformance]{Table} \ref{recreatedpredictionperformance}).
 Had we been able to use labels that were correctly labelled on a pixel basis, i.e.
@@
 
-\citet{roelens} showed that you can use machine learning to detect ditches directly from point-clouds instead of a DEM.
-One concern we had with our method was that valuable information may have been discarded when aggregating the elevations from the original point cloud to a DEM, making it more difficult to locate the ditches.
-However, we achieved a similar score to their method; $\kappa=0.73$ in our study compared to 0.77 for grassland and 0.73 for peri-urban area in Roelens study.
-This suggests that you can also use a DEM to automatically extract ditches, provided that you have enough input data to generate a high resolution DEM to capture the small scale features of ditches.
+One concern with our method was that valuable information may have been discarded when aggregating the elevations from the original point cloud to a DEM.
+\citet{roelens} used  machine learning to detect ditches directly from point-clouds instead of a DEM.
+However, we achieved a similar score to their method; $\kappa=0.73$ in our study compared to 0.77 for grassland and 0.73 for peri-urban area in \citet{roelens} study.
+This suggests that DEM can also be used to automatically extract ditches, provided that enough input data is available to generate a high resolution DEM for capturing the small scale features of ditches.
 
 \citet{bailly} also showed that the vegetation cover significantly affects the ability to detect ditches from LiDAR data (\hyperref[fig:ditchpictures]{Figure} \ref{fig:ditchpictures} \& \ref{fig:resultstreesbushes}).
-Approximately 75\% of the ditches in their study were retrieved where no vegetation existed, whereas the detection levels sank to below 10\% under vegetation; especially when high vegetation, shrubs and trees covered the ditches \citep{bailly}.
-Hence, a $\kappa$ of 0.73 in the Krycklan catchment with  87\% forest cover \citep{krycklancatchment} is in that aspect surprisingly good, and could perhaps be a result of the higher point density of our LiDAR measurements (20 compared to 10 points $m^{2}$).
-Or, perhaps the generation of a DEM from the point cloud has made the detection more robust?
+Approximately 75\% of the ditches in their study were retrieved where no vegetation existed, whereas the detection accuracy went below 10\% under vegetation; especially when the ditches were covered by high vegetation, shrubs, and trees \citep{bailly}.
+Hence, a $\kappa$ of 0.73 in our study  with  87\% forest cover \citep{krycklancatchment} is substantially better than that of previous research, and this could perhaps be a result of the higher point density of our LiDAR measurements (20 compared to 10 points $m^{2}$).
 
 The hydrology of our study catchment also differs from the other mentioned studies as there are almost as many natural stream channels as there are ditches in the area \citep{hasselquist, mappingtemporal}.
-Other studies were conducted in smaller, predominately artificially drained areas \citep{roelens, bailly, rapinel}.
+Other studies were conducted in smaller, predominantly artificially drained areas \citep{roelens, bailly, rapinel}.
 Our method locates both artificial channels (ditches) and natural stream channels.
@@
 A potential issue with generalising our ditch detector for use in other geographical areas is that all the post-processing steps and input variables were developed based on occurrences in the Krycklan area.
-Different geographical compositions in other areas may require that the thresholds used to fill gaps in ditches and to remove noise would need to be adjusted.
+Different geographical compositions in other areas may require the adjustment of the thresholds used to fill gaps in ditches and to remove noise.
 

### 2020-09-15 LaTeX/proofreading_rewrite.tex
Adds more changes from Sidd's suggestions.

@@
 However, different false positives and negatives may be observed when examining each terrain index manually.
-For example, Sky View Factor tends to give a large amount of false positives in steep terrain whereas Impoundment Index and High Pass Median Filter are more difficult to analyse.
 Such inaccuracies and complexities are in part related to the Light Detection and Ranging (LiDAR) measurements, the mainstay of high resolution DEM and terrain mapping, sometimes being interfered by trees, bushes, or grass - leaving gaps in the ditch model.
@@
 This manual map guided the model training and ground-truthing in our analysis and provided a test-bench for our methodology.
-Several other studies \citep{roelens, bailly, rapinel, kiss} that have located ditches from LiDAR data have worked on smaller areas (smaller than 2 $km^2$) where artificial drainage dominates.
-Here, we work on landscape scale in a mesoscale catchment.
-The  Krycklan Study Catchment in Sweden is 68 $km^2$, which means that we cover more landscape types.
+Previous studies \citep{roelens, bailly, rapinel, kiss} that have located ditches from LiDAR data have worked on smaller areas (smaller than 2 $km^2$), predominantly characterised by artificial drainage.
+However, we herein conducted a landscape-scale analysis in a mesoscale catchment.
+The study  catchment  is 68 $km^2$, representing diverse landscape features and geomorphological characteristics.
 
@@
 The ditches all have a generalised width based on a perceived average of ditch width in the area.
-Due to all ditches varying in width, it was impossible to produce a perfect representation of each ditch.
-However, this made for a good compromise for the average ditch.
+Due to all ditches varying in width, it was impossible to produce a perfect representation of each ditch, but this widening helped ensure that almost all ditch pixels were covered by the labels.
 
-Because our priority was to detect ditches, rather than accurately detecting each pixel labelled as a ditch, some adjustments were made when producing the prediction results.
-The dataset was divided into a lower resolution grid of $6*6$ pixels (9 $m^2$) for each grid.
+%--OLD--
+%Because our priority was to detect ditches, rather than accurately detecting each pixel labelled as a ditch, some adjustments were made when producing the prediction results.
+
+To remove outliers in our predictions, the dataset was downscaled by dividing it into a lower resolution grid of $6*6$ pixels (9 $m^2$) for each grid.
 Each grid cell that contained at least 25\% ditch pixels was then labelled as a ditch, see \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{c}.
@@
   \item \textbf{Sky View Factor} \newline
-  The Sky View Factor is a digital terrain index with a value ranging from 0 to 1, representing how much of the sky that is visible from a certain point on the ground \citep{zaksek}.
+%--OLD--
+  %The Sky View Factor is a digital terrain index with a value ranging from 0 to 1, representing how much of the sky that is visible from a certain point on the ground \citep{zaksek}.
 On an open field all of the sky would be visible, but at the bottom of a ditch much of the sky would be obscured, meaning low values would indicate ditches.
@@
 
+%--NEW--
+The Sky View Factor represents how much of the sky that is visible from a certain point on the ground \citep{zaksek}.
+A point in a ditch would yield a low value, as more of the sky is obscured by the ditch bank.
+However, it tends to produce false positives in steep terrain.
+The Sky View Factor was calculated in SAGA GIS, using a search radius of 10 metres to ensure that ditches were covered, and to not let steep hills close to ditches affect the results.
+
   \item \textbf{Impoundment Index}\newline
-  The tool Impoundment Index in Whitebox Tools \citep{whiteboxtools} is a digital terrain index designed to serve as a measure of topographic incision and flow constriction.
+%--OLD--
+  %The tool Impoundment Index in Whitebox Tools \citep{whiteboxtools} is a digital terrain index designed to serve as a measure of topographic incision and flow constriction.
 The index effectively maps the area and average depth of the reservoir that would be created by placing a dam of a user-specified length at each grid cell in the DEM.
@@
 
+%--NEW--
+The Impoundment Index in Whitebox Tools \citep{whiteboxtools} can be used as a measure of flow constriction, which is useful for mapping channels, such as ditches.
+When tuned to output the flooded volume, it effectively maps the area and average depth of the reservoir that would be created by placing a dam of a user-specified length at each grid cell in the DEM.
+Here, we used dam height calculated with a dam length of 3 metres to cover the entire ditches.
+Typical false positives with this method are natural stream channels with a clear incised channel.
+
   \item \textbf{High Pass Median Filter} \newline
-  High Pass Median Filters (HPMF) highlight ditches by indicating local depressions in the DEM.
+%--OLD--  
+%High Pass Median Filters (HPMF) highlight ditches by indicating local depressions in the DEM.
 The algorithm operates essentially by subtracting the value at the grid cell at the centre of the window from the median value in the window.
@@
 Typical false positives with this index are small hollows that also show negative numbers.
+HPMF was calculated using Whitebox Tools \citep{whiteboxtools} with a window size of $4.5 * 4.5$ m.
+
+%--NEW--
+High Pass Median Filters (HPMF) highlight ditches by indicating local depressions (such as ditches) in the DEM by subtracting the value at the grid cell at the centre of the window from the median value in the window.
+Typical false positives with this index are small hollows.
 HPMF was calculated using Whitebox Tools \citep{whiteboxtools} with a window size of $4.5 * 4.5$ m.
@@
 \label{impoundmentstreamremoval}
-The raw Impoundment Index dam height was used to create a mask, attempting to mark out streams, but retain ditches.
+The raw Impoundment Index dam height was used to create a mask, attempting to mask out streams, but retain ditches.
 Only areas with a relatively high dam height, above 14 metres, which would indicate that these areas most likely contained streams, were masked out.
@@
 
-\textit{Model ensembles} is a category of machine learning where a set of models can be combined to increase diversity and robustness of predictions \citep{flach}.
+%--NEW--
+Random Forests is an ensemble machine learning method which builds multiple diverse decision trees to increase the robustness of its predictions \citep{breiman,flach}.
+The individual trees are trained by minimising the entropy of the data from each node’s output \citep{kotsiantis}.
+A byproduct of the training is the Gini importance, which denotes the most important input variables \citep{gini}.
+For our study, a hyperparameter tuning showed that using 300 trees and \textit{gini} as the splitting criterion, with a minimum of 10 samples per node split, and no artificial class weight or max depth of trees yielded the best predictions.
+The Random Forests algorithm from the Python library \textit{scikit-learn} \citep{scikit-learn} was used in the experiment, and the classifiers were trained on all the input variables seen in \hyperref[featuretable]{Table} \ref{featuretable}.
+
+%--OLD BELOW--
+%\textit{Model ensembles} is a category of machine learning where a set of models can be combined to increase diversity and robustness of predictions \citep{flach}.
 Random Forests is a supervised learning algorithm, first developed by \citet{breiman}, that makes use of the model ensemble technique by using a set of decision trees to build a \textit{forest} of trees.
@@
 
-The decision tree data structure is a \textit{tree} of connected nodes where the end node of each branch (leaf) produces an output value.
+%The decision tree data structure is a \textit{tree} of connected nodes where the end node of each branch (leaf) produces an output value.
 Each step in the tree aims to minimise the entropy of the outputs \citep{kotsiantis}.
@@
 
-Random Forests uses two techniques called \textit{subspace sampling} and \textit{bootstrap aggregation} (bagging) in tandem \citep{breiman}.
+%Random Forests uses two techniques called \textit{subspace sampling} and \textit{bootstrap aggregation} (bagging) in tandem \citep{breiman}.
 Subspace sampling is the process of drawing a random subset of variables from the input variable set \citet{ho}.
@@
 
-Random Forests computes Gini importance, which shows how often a variable is selected for a split for a certain classification \citep{gini}.
+%Random Forests computes Gini importance, which shows how often a variable is selected for a split for a certain classification \citep{gini}.
 Each input variable is given a Gini impurity rating, which measures the frequency of incorrect labelling occurring from its use.
@@
 
-The Random Forests algorithm from the Python library \textit{scikit-learn} was used in the experiment, and the classifiers were trained on all the input variables seen in \hyperref[featuretable]{Table} \ref{featuretable}.
+%The Random Forests algorithm from the Python library \textit{scikit-learn} was used in the experiment, and the classifiers were trained on all the input variables seen in \hyperref[featuretable]{Table} \ref{featuretable}.
 The probability predictions used in this Random Forests implementation are computed by counting the number of occurrences of a class output and dividing with the total amount of trees in the forest \citep{scikit-learn}.
@@
 
-The first step to create a more balanced training dataset was to extract all pixels labelled as ditches, as well as pixels within close proximity of ditches.
+%--OLD--
+%The first step to create a more balanced training dataset was to extract all pixels labelled as ditches, as well as pixels within close proximity of ditches.
 Secondly, pixels were sampled randomly from the entire subsection.
 This allowed the training dataset to be fairly balanced while still containing most of the geographical features of each subsection (\hyperref[fig:balancedmasks]{Figure} \ref{fig:balancedmasks}).
+
+%--NEW--
+The testing phase showed that the classifiers produced poor results when the ratio of ditch- versus non-ditch pixels in the training data was very high, as the models were not being punished for mislabelling ditches as non-ditches.
+To increase the recall, we evened out the amount of ditch and non-ditch instances in the training data.
+First, we extracted all pixels labelled as ditches as well as pixels within close proximity of ditches.
+Secondly, pixels were sampled randomly from the entire subsection to still contain most of the geographical features of each area (\hyperref[fig:balancedmasks]{Figure} \ref{fig:balancedmasks}).
 
@@
 
-The comparison with the field observed channels and the available maps clearly indicate the need for better maps of ditches, as only 9\% of ditches are mapped.
+The comparison with the field observed channels and the available maps clearly indicate the need for better maps of ditches, as only 9% of ditches are mapped.
 (\hyperref[fig:watercoursebarplot]{Figure} \ref{fig:watercoursebarplot}).
@@
 However, this indicates that our method could also be used to map small, previously unmapped natural stream channels.
-The national survey showed that 55\% of the natural stream channels and 75\% of the straightened water courses were missing on the maps.
+The national survey showed that 55% of the natural stream channels and 75% of the straightened water courses were missing on the maps.
 This is in line with other national \citep{kuglerova} and international \citep{benstead} studies highlighting this phenomenon  called \“Aqua Incognita - the unknown headwaters\” \citep{bishop, kuglerova}.
@@
 \section*{Acknowledgements}
-We wish to thank Sigurd Israelsson at the School of Engineering at J\"onk\"oping University for lending hardware resources for the experiments in this study.
+We wish to thank the two anonymous reviewers and Siddhartho Paul whose suggestions helped improve and clarify this manuscript.
+We also thank Sigurd Israelsson at the School of Engineering at J\"onk\"oping University for lending hardware resources for the experiments in this study.
 The work has been funded by VINNOVA, EU InterReg Baltic Sea project WAMBAF tools and Formas.

### 2020-09-19 LaTeX/proofreading_rewrite.tex
Adds rewritten and fully proofread and grammatically version after Sidd and Anneli's comments.

@@
 
-% Ploting
+% Plotting
 \usepackage{pgfplots}
@@
 
-%\title{Ditch detection in high-resolution digital elevation models}
-
 \title{Detecting ditches using machine learning on high-resolution DEMs}
@@
 However, a single threshold generally does not capture the variability in the landscape and generates many false positives and negatives.
-We hypothesised that, by combining the digital terrain indices using a machine learner, we could improve ditch detection at a landscape-scale.
+We hypothesised that, by combining the digital terrain indices using machine learning, we could improve ditch detection at a landscape-scale.
 
@@
 \citep{peatlands}.
-In Sweden which is one of the North European countries with the most extensive ditch network, over five million hectares of peatland have been drained for forestry in the 20th century.
-As a result, almost one million km of ditches have been introduced and doubled the length of watercourses across Sweden  \citep{hasselquist}.
+In Sweden, which is one of the North European countries with the most extensive ditch network, over five million hectares of peatland have been drained for forestry in the 20th century.
+As a result, almost one million km of ditches have been introduced, which has doubled the length of watercourses across Sweden  \citep{hasselquist}.
 
-For effective restoration of ditched wetlands, it is highly important  to accurately locate the ditches on digital maps, whereas landscape-scale field inventories are often challenging and expensive.
+For effective restoration of ditched wetlands, it is highly important  to accurately locate the ditches on digital maps.
+However, landscape-scale field inventories are often challenging and expensive.
 Automatic detection of ditches from high-resolution digital elevation models (DEMs) and terrain indices provides a feasible means for spatially explicit assessment of ditches across a landscape.
@@
 We hypothesise that by combining the information from all the digital terrain indices using a machine learner (Random Forests), we can improve the detection of the ditches, compared to if a single index would be used.
-To test this, we conducted a landscape-scale ditch detection analysis at the Krycklan Catchment in Sweden \citep{krycklancatchment}, where a high precision manual mapping has been performed previously.
+To test this, we conduct a landscape-scale ditch detection analysis at the Krycklan Catchment in Sweden \citep{krycklancatchment}, where a high precision manual mapping has been performed previously.
 This manual map guided the model training and ground-truthing in our analysis and provided a test-bench for our methodology.
 Previous studies \citep{roelens, bailly, rapinel, kiss} that have located ditches from LiDAR data have worked on smaller areas (smaller than 2 $km^2$), predominantly characterised by artificial drainage.
-However, we herein conducted a landscape-scale analysis in a mesoscale catchment.
+However, we herein conduct a landscape-scale analysis in a mesoscale catchment.
 The study  catchment  is 68 $km^2$, representing diverse landscape features and geomorphological characteristics.
@@
 The National Inventory of Landscapes in Sweden (NILS) has recorded small watercourses (width $<$ 6 m) by conducting a line inventory in 631 $5*5$ km squares randomly distributed in Sweden, to best capture a representative and varied sample of the Swedish landscape.
-The position of the small watercourses from this inventory were compared with the watercourses on Swedish Property Map (1: 12 500) (allowing for an uncertainty in the positioning of up to 15 m, by snapping the point observations to the lines on the map) (\hyperref[fig:watercoursebarplot]{Figure} \ref{fig:watercoursebarplot}).
+The positions of the small watercourses from this inventory were compared with the watercourses on Swedish Property Map (1: 12 500) (allowing for an uncertainty in the positioning of up to 15 m, by snapping the point observations to the lines on the map) (\hyperref[fig:watercoursebarplot]{Figure} \ref{fig:watercoursebarplot}).
 
@@
 A historical survey of the catchment (Gudrun Norstedt, unpublished) recorded  that the ditches in the catchment were dug between 1900 and 1993.
-Initially, the ditches were used to drain the mires for hay production while the drained mires were later attempted to turn into productive forest lands \citep{paivanen}.
+Initially, the ditches were used to drain the mires for hay production.
+These drained mires were later attempted to turn into productive forest lands \citep{paivanen}.
 However, research later showed that many of the drained mires were not productive for forestry due to their nutrient deficient environment \citep{sikstrom}.
@@
 
-%--OLD--
-%Because our priority was to detect ditches, rather than accurately detecting each pixel labelled as a ditch, some adjustments were made when producing the prediction results.
-
-To remove outliers in our predictions, the dataset was downscaled by dividing it into a lower resolution grid of $6*6$ pixels (9 $m^2$) for each grid.
-Each grid cell that contained at least 25\% ditch pixels was then labelled as a ditch, see \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{c}.
+To match the labels with out predictions (where we downscale the predictions into grids to remove outliers), the ground-truth was also downscaled into grids of $6*6$ pixels (9 $m^2$) each.
+Each grid cell that contained at least 25\% ditch pixels was labelled as a ditch, see \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{c}.
 This layer was used to ground-truth the machine learning models.
@@
   \item \textbf{Sky View Factor} \newline
-%--OLD--
-  %The Sky View Factor is a digital terrain index with a value ranging from 0 to 1, representing how much of the sky that is visible from a certain point on the ground \citep{zaksek}.
-On an open field all of the sky would be visible, but at the bottom of a ditch much of the sky would be obscured, meaning low values would indicate ditches.
-The Sky View Factor was calculated in SAGA GIS, using a search radius of 10 metres to ensure that ditches were covered, and to not let steep hills close to ditches affect the results.
-
-%--NEW--
     The Sky View Factor represents how much of the sky that is visible from a certain point on the ground \citep{zaksek}.
@@
   \item \textbf{Impoundment Index}\newline
-%--OLD--
-  %The tool Impoundment Index in Whitebox Tools \citep{whiteboxtools} is a digital terrain index designed to serve as a measure of topographic incision and flow constriction.
-The index effectively maps the area and average depth of the reservoir that would be created by placing a dam of a user-specified length at each grid cell in the DEM.
-The tool can be tuned to output the flooded area or volume (in which case it highlights areas of flow constriction), or the average flooded depth (which seems to measure topographic incision).
-Additionally, the tool will output a 'dam height' raster that is particularly useful for mapping channels, such as ditches.
-Here, we used dam height calculated with a dam length of 3 metres, to cover the entire ditch which is typically not wider than 3 metres.
-Typical false positives with this method are natural stream channels with a clear incised channel.
-
-%--NEW--
     The Impoundment Index in Whitebox Tools \citep{whiteboxtools} can be used as a measure of flow constriction, which is useful for mapping channels, such as ditches.
@@
   \item \textbf{High Pass Median Filter} \newline
-%--OLD--  
-%High Pass Median Filters (HPMF) highlight ditches by indicating local depressions in the DEM.
-The algorithm operates essentially by subtracting the value at the grid cell at the centre of the window from the median value in the window.
-Negative values indicate depressions in the DEM, such as ditches.
-Typical false positives with this index are small hollows that also show negative numbers.
-HPMF was calculated using Whitebox Tools \citep{whiteboxtools} with a window size of $4.5 * 4.5$ m.
-
-%--NEW--
     High Pass Median Filters (HPMF) highlight ditches by indicating local depressions (such as ditches) in the DEM by subtracting the value at the grid cell at the centre of the window from the median value in the window.
@@
 In addition to the general input variables, a number of custom input variables were also developed.
-Some of these custom variables were produced by statistical aggregation of multiple indices whereas in some cases, they were generated by using thresholds to exclude, lower, or amplify the values of pixels that indicated ditches or non-ditches (\hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{d} and \hyperref[fig:features]{h}).
+Some of these custom variables were produced by statistical aggregation of multiple indices, whereas in some cases, they were generated by using thresholds to exclude, lower, or amplify the values of pixels that indicated ditches or non-ditches (\hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{d} and \hyperref[fig:features]{h}).
 
@@
 
-%--NEW--
 Random Forests is an ensemble machine learning method which builds multiple diverse decision trees to increase the robustness of its predictions \citep{breiman,flach}.
@@
 The Random Forests algorithm from the Python library \textit{scikit-learn} \citep{scikit-learn} was used in the experiment, and the classifiers were trained on all the input variables seen in \hyperref[featuretable]{Table} \ref{featuretable}.
-
-%--OLD BELOW--
-%\textit{Model ensembles} is a category of machine learning where a set of models can be combined to increase diversity and robustness of predictions \citep{flach}.
-Random Forests is a supervised learning algorithm, first developed by \citet{breiman}, that makes use of the model ensemble technique by using a set of decision trees to build a \textit{forest} of trees.
-The outputs from these trees are then examined, and a class probability estimation ranging from 0 to 1 can be produced for each sample by dividing the amount of tree outputs of a class with the total amount of trees \citep{breiman}.
-
-%The decision tree data structure is a \textit{tree} of connected nodes where the end node of each branch (leaf) produces an output value.
-Each step in the tree aims to minimise the entropy of the outputs \citep{kotsiantis}.
-Each node in a decision tree represents a variable or an abstraction of variables from the inputs and each split is made by using this variable to separate the inputs.
-
-%Random Forests uses two techniques called \textit{subspace sampling} and \textit{bootstrap aggregation} (bagging) in tandem \citep{breiman}.
-Subspace sampling is the process of drawing a random subset of variables from the input variable set \citet{ho}.
-The bagging technique takes, with replacement, a set of different random samples for constructing each tree \citep{flach}.
-There are several hyperparameters that can be tuned to achieve better results from the model.
-For example, you can adjust the amount of decision trees to use, or the maximum amount of input variables to use for each subspace sampling \citep{scikit-learn}.
-We performed a hyperparameter tuning to determine what parameter values for the Random Forests algorithm would yield the best prediction.
-Using 300 trees and \textit{gini} as the splitting criterion, with a minimum of 10 samples per node split, and no artificial class weight or max depth of trees produced the best results.
-
-%Random Forests computes Gini importance, which shows how often a variable is selected for a split for a certain classification \citep{gini}.
-Each input variable is given a Gini impurity rating, which measures the frequency of incorrect labelling occurring from its use.
-As the Gini impurity rating for a variable increases, its Gini importance decreases \citep{gini}.
-When the model is trained, it examines the Gini importance from all trees and produces the total Gini importance as a value between 0 and 1 \citep{gini}.
-The Gini importance can help when developing a model by indirectly giving advice on new input variables to introduce.
-
-%The Random Forests algorithm from the Python library \textit{scikit-learn} was used in the experiment, and the classifiers were trained on all the input variables seen in \hyperref[featuretable]{Table} \ref{featuretable}.
-The probability predictions used in this Random Forests implementation are computed by counting the number of occurrences of a class output and dividing with the total amount of trees in the forest \citep{scikit-learn}.
-The testing phase showed that the classifiers produced poor results when the ratio of ditch- versus non-ditch pixels in the training data was very high.
-A high ratio led to the models not being punished for mislabelling ditches as non-ditches, causing it to prioritise a high accuracy over a high recall.
-According to \citet{balanced}, an imbalanced training dataset causes a minority class to be less accurately predicted.
-Because the ditch class is much less common than the non-ditch class, we attempted to train the models with a roughly equal amount of ditch pixels and non-ditch pixels to balance the models.
-The models were, however, still evaluated on all the pixels in the 11 validation subsections.
 
@@
       Slope min & 2, 4, 6 \\
-      %Slope max & 2, 4, 6 \\
       Slope non-ditch amplification & - \\ \bottomrule
@@
 
-%--OLD--
-%The first step to create a more balanced training dataset was to extract all pixels labelled as ditches, as well as pixels within close proximity of ditches.
-Secondly, pixels were sampled randomly from the entire subsection.
-This allowed the training dataset to be fairly balanced while still containing most of the geographical features of each subsection (\hyperref[fig:balancedmasks]{Figure} \ref{fig:balancedmasks}).
-
-%--NEW--
-The testing phase showed that the classifiers produced poor results when the ratio of ditch- versus non-ditch pixels in the training data was very high, as the models were not being punished for mislabelling ditches as non-ditches.
+The testing phase showed that the classifiers produced poor results when the ratio of ditch versus non-ditch pixels in the training data was very high, as the models were not being punished for mislabelling ditches as non-ditches.
 To increase the recall, we evened out the amount of ditch and non-ditch instances in the training data.
@@
 \subsubsection{Noise reduction and gap filling}
+
 The probability predictions contained a lot of noise in the areas largely distant from the ditches.
-The first step for removing this noise was to use a bilateral de-noising filter on the entire prediction image.
-This left linear properties and pixels with a very high value intact, while lowering the value of pixels that did not contribute to an accurate prediction (\hyperref[fig:postprocessing1]{Figure} \ref{fig:postprocessing1} \hyperref[fig:postprocessing1]{b}).
+This noise was removed by first using a bilateral de-noising filter, and then removing outlier pixels that were far away from any other high probability ditch pixels.
+This left linear properties and pixels with a very high value intact, while lowering the value of pixels that did not contribute to an accurate prediction (\hyperref[fig:postprocessing1]{Figure} \ref{fig:postprocessing1} \hyperref[fig:postprocessing1]{b}) (\hyperref[fig:postprocessing2]{Figure} \ref{fig:postprocessing2} \hyperref[fig:postprocessing2]{a}).
+
+To fill gaps in ditches that the models failed to correctly predict, we aggregated pixel values by covering them with cone masks expanding outwards in different directions from the examined pixel.
+This amplified some of the noise that was left, but filling the gaps in the ditches was judged to be more important (\hyperref[fig:postprocessing2]{Figure} \ref{fig:postprocessing2} \hyperref[fig:postprocessing2]{b}).
 
@@
 \end{figure}
-
-The second step for removing noise was to use a custom function to remove pixels that had a semi-high ditch probability, but that lay far away from any other high probability pixels.
-In this function, a threshold value was used to avoid removing pixels with a high enough ditch probability, helping to retain pixels that lay in or close to a ditch.
-The maximum ditch probability value in a circular radius of 10 pixels was then calculated.
-If this maximum value was too low, the probability value of the examined pixel was lowered (\hyperref[fig:postprocessing2]{Figure} \ref{fig:postprocessing2} \hyperref[fig:postprocessing2]{a}).
-
-The third step involved taking measures to try to fill gaps in ditches that the models failed to correctly predict.
-This was done by aggregating pixels covered by cone masks expanding outwards in different directions from the examined pixel.
-This step also amplified some of the noise that was left, but filling the gaps in the ditches was judged to be more important to help make the next step more effective (\hyperref[fig:postprocessing2]{Figure} \ref{fig:postprocessing2} \hyperref[fig:postprocessing2]{b}).
 
@@
 To circumvent some of the performance evaluation issues arising from the use of pixel classification for ditch objects as well as not having completely accurate labels on a pixel basis (due to uncertainties in the width of the ditches), we modified the evaluation labels to allow for some error in close proximity to ditches in the prediction.
-False negatives and false positive grid zones that lay adjacent to the ditch label grid zones were evaluated as true negatives and true positives, as they can be considered a part of correctly located ditch objects (\hyperref[fig:newlabels]{Figure} \ref{fig:newlabels}).
+False negative and false positive grid zones that lay adjacent to the ditch label grid zones were evaluated as true negatives and true positives, as they can be considered a part of correctly located ditch objects (\hyperref[fig:newlabels]{Figure} \ref{fig:newlabels}).
 
@@
 
-\begin{figure}[htb]
+\begin{figure}[!htb]
     \centering
@@
 
-The comparison with the field observed channels and the available maps clearly indicate the need for better maps of ditches, as only 9% of ditches are mapped.
+The comparison with the field observed channels and the available maps clearly indicate the need for better maps of ditches, as only 9\% of ditches are mapped.
 (\hyperref[fig:watercoursebarplot]{Figure} \ref{fig:watercoursebarplot}).
@@
 However, this indicates that our method could also be used to map small, previously unmapped natural stream channels.
-The national survey showed that 55% of the natural stream channels and 75% of the straightened water courses were missing on the maps.
-This is in line with other national \citep{kuglerova} and international \citep{benstead} studies highlighting this phenomenon  called \“Aqua Incognita - the unknown headwaters\” \citep{bishop, kuglerova}.
+The national survey showed that 55\% of the natural stream channels and 75\% of the straightened water courses were missing on the maps.
+This is in line with other national \citep{kuglerova} and international \citep{benstead} studies highlighting this phenomenon called "Aqua Incognita - the unknown headwaters" \citep{bishop,kuglerova}.
 
@@
 Hence, a $\kappa$ of 0.73 in our study  with  87\% forest cover \citep{krycklancatchment} is substantially better than that of previous research, and this could perhaps be a result of the higher point density of our LiDAR measurements (20 compared to 10 points $m^{2}$).
+The hydrology of our study catchment also differs from that of other studies, as we have almost as many natural stream channels as ditches, whereas their studies were conducted in smaller, predominantly artificially drained areas \citep{bailly, roelens, rapinel}.
+Because our method locates both ditches and natural stream channels, we have to take additional measures to remove the stream channels (affecting recall negatively), as explained in \ref{impoundmentstreamremoval}.
 
-The hydrology of our study catchment also differs from the other mentioned studies as there are almost as many natural stream channels as there are ditches in the area \citep{hasselquist, mappingtemporal}.
-Other studies were conducted in smaller, predominantly artificially drained areas \citep{roelens, bailly, rapinel}.
-Our method locates both artificial channels (ditches) and natural stream channels.
-This would not be a problem if the aim had been to locate all channels.
-However, our study focused on locating ditches specifically.
-Hence, we tried to remove the natural stream channels by manipulating some of the input variables, using the Impoundment Index to lower the values of pixels that had a high dam height, as explained in \ref{impoundmentstreamremoval}.
-This helped to remove the deepest streams and increased the overall accuracy, with the drawback of decreasing the recall slightly (\hyperref[fig:resultsillustrations]{Figure} \ref{fig:resultsillustrations} \hyperref[fig:resultsillustrations]{f}).
-This method did not remove the smaller streams (\hyperref[fig:resultsillustrations]{Figure} \ref{fig:resultsillustrations} \hyperref[fig:resultsillustrations]{b}).
 A future improvement to the ditch detector could be to use a shape index to remove small natural streams, as natural streams often form more complex channels whereas ditches are often straight.
@@
 This study investigates how to use digital elevation data together with manually labelled ditches to identify patterns and relationships that can be used for automatic ditch detection.
-Although the method still has room for improvement, it performs well on the available data, and significantly outperforms classifying with the digital terrain indices separately.
+We have shown that it is possible to use machine learning with digital elevation models to learn patterns that enable robust detection of ditches.
+Designing a robust ditch detection in a study area with substantial natural stream coverage was challenging, both in regards to model input variables and post-processing measures.
 Future work could introduce more input variables and post-processing steps, such as using pathfinding algorithms to fill gaps in the ditch model, or shape indices to remove stream channels from the prediction.
-Moving towards image segmentation with grids of pixels instead of pixel classification of tabular pixel values could be a suitable future approach.
-
-Many of the input variables could be reused in the future, with Random Forests or other machine learning algorithms.
-Several of the post-processing algorithms, such as the probability prediction noise removal, and the cluster removal from the binary prediction could be generalised for use in any type of segmentation of ditches.
-The proposed method may help in making ditch detection both faster and easier, reducing manual labour and cost.
-In conclusion, this study has shown that it is possible to use machine learning with digital elevation models to learn patterns that enable robust detection of ditches.
+Moving towards image segmentation with grids of pixels instead of pixel classification of tabular pixel values could also be a suitable future approach.
+With minor adjustments, our proposed method can be employed in a general setting, reducing manual labour and cost, and enabling faster and easier mapping of ditches.
 
@@
 \section*{Acknowledgements}
-We wish to thank the two anonymous reviewers and Siddhartho Paul whose suggestions helped improve and clarify this manuscript.
+We wish to thank the two anonymous reviewers and Dr.
+Siddhartho Paul at the Swedish University of Agricultural Sciences, whose suggestions helped improve and clarify this manuscript.
 We also thank Sigurd Israelsson at the School of Engineering at J\"onk\"oping University for lending hardware resources for the experiments in this study.

### 2020-09-19 LaTeX/Ditch detection in high-resolution digital elevation models.tex
Adds new results illustration from new experiment.
Adds final (?) changes to article and rejoinder.

@@
 Thresholding methods using digital terrain indices, such as Sky View Factor, Impoundment Index, or High Pass Median Filter can be used to detect ditches.
-However, a single threshold generally does not capture the variability in the landscape, generating many false positives and negatives.
-We hypothesised that, by combining the digital terrain indices using a machine learner, we could improve the ditch detection.
+However, a single threshold generally does not capture the variability in the landscape and generates many false positives and negatives.
+We hypothesised that, by combining the digital terrain indices using machine learning, we could improve ditch detection at a landscape-scale.
 
@@
 
-Over the last century many of the wetlands in North Europe and parts of North America have been ditched.
-Mapping and modelling the hydrology of a flat landscape is challenging because of the low topographic gradients.
-In many places both the landscape and the channel network have also been heavily modified by humans, especially in agricultural landscapes, for example in mid-western United States \citep{passalacqua}.
-In North Europe, ditches have also been dug to turn flat peatlands into productive forests, thereby profoundly altering the hydrology in North Europe \citep{peatlands}.
-Sweden is one of the North European countries with the most ditches; over five million hectares of peatland have been drained for forestry in the 20th century.
-This has resulted in almost one million km of ditches, which has more than doubled the length of watercourses in Sweden \citep{hasselquist}.
+Over the last century, the wetlands in North Europe and parts of North America have been substantially modified and ditched.
+In many places both the landscape and the channel network have also been heavily altered  by humans, especially in the agricultural landscapes, for example in mid-western United States \citep{passalacqua}.
+In North Europe, however, ditches have been dug to turn flat peatlands into productive forests, resulting in profound impacts on the hydrology of this region.
+\citep{peatlands}.
+In Sweden, which is one of the North European countries with the most extensive ditch network, over five million hectares of peatland have been drained for forestry in the 20th century.
+As a result, almost one million km of ditches have been introduced, which has doubled the length of watercourses across Sweden  \citep{hasselquist}.
 
-A comparison between a field inventory of ditches (within the National Inventory of Landscapes in Sweden (NILS)) and current maps, shows that approximately 90 \% of the ditch networks are missing on current maps ({\AA}gren, unpublished).
-To better prioritise the restoration of ditched wetlands, it is necessary to know where the ditches are located, but field inventories are often too costly.
-The research objective of this study is therefore to automatically detect ditches from high-resolution digital elevation models (DEMs) using digital terrain indices.
-Previous attempts \citep{uppsala} have shown that methods such as Sky View Factor \citep{zaksek}, Impoundment Index \citep{whiteboxtools} and High Pass Median Filter \citep{whiteboxtools} can be used to detect the ditches.
-Different false positives and negatives can be observed when examining each terrain index manually.
-Sky View Factor tends to give a large amount of false positives in steep terrain, whereas Impoundment Index and High Pass Median Filter are more difficult to analyse.
-Due to the difference in false positives and negatives, we want to combine the indices to help prune the weaknesses and amplify the strengths of each index.
-This is in part due to the Light Detection and Ranging (LiDAR) measurements sometimes being interfered by trees, bushes, or grass - leaving gaps in the ditch model, or sometimes due to a poorly set threshold.
-Therefore, using solely the digital terrain indices for ditch mapping tends to yield an imprecise result.
+For effective restoration of ditched wetlands, it is highly important  to accurately locate the ditches on digital maps.
+However, landscape-scale field inventories are often challenging and expensive.
+Automatic detection of ditches from high-resolution digital elevation models (DEMs) and terrain indices provides a feasible means for spatially explicit assessment of ditches across a landscape.
+Previous researches  \citep{uppsala} documented  that terrain indices such as Sky View Factor \citep{zaksek}, Impoundment Index \citep{whiteboxtools} and High Pass Median Filter \citep{whiteboxtools} can efficiently  detect the ditches.
+However, different false positives and negatives may be observed when examining each terrain index manually.
+Such inaccuracies and complexities are in part related to the Light Detection and Ranging (LiDAR) measurements, the mainstay of high resolution DEM and terrain mapping, sometimes being interfered by trees, bushes, or grass - leaving gaps in the ditch model.
+The inaccuracies can also arise from a poorly set threshold for the LiDAR derived terrain parameters.
+Thus, it would be useful to combine the terrain indices derived from high resolution LiDAR data to prune the weaknesses and amplify the strengths of each index and produce highly accurate detection of ditch networks.
 
-In this study, we introduce new methods for the automated mapping of ditches.
-From the high resolution DEM, we use machine learning to locate the ditch networks.
-We also examine which digital terrain indices are the most relevant for a correct ditch prediction.
-To accomplish this, we conduct experiments using a supervised learning algorithm with labelled ditch data to automatically extract ditches from externally supplied occurrences (data not before seen by the model) \citep{kotsiantis}.
-These occurrences are used as input variables to the algorithm, and these input variables consist of refined or raw digital terrain indices derived from a DEM.
-
-There are other studies where geographical data has been used with machine learning to map ditches.
-\citet{roelens} used the Random Forests algorithm to detect ditches in Belgium, using LiDAR point clouds.
-They used several different input variables where neighbouring points of a specific LiDAR point were examined.
-Values of all LiDAR attributes were represented for the neighbouring area for both 0.5, 1, 1.5, 2, 3 and 4 metre radii.
-We reproduce these neighbouring area variables when building the ditch detector in our study.
-Several other input variables for the model are calculated from the indices to place emphasis on the clear linear directional characteristics of ditches.
+In this study, we first investigate the need for better maps of ditch networks, then we introduce a new method for the automated mapping of ditches by integrating multiple high resolution terrain indices (derived from LiDAR DEM) using a machine learning algorithm.
+We also examine which digital terrain indices are the most relevant for an accurate prediction of ditches.
+Several input variables for the model are calculated from the indices to place emphasis on the clear linear directional characteristics of ditches.
 These characteristics are also used when processing the model's output, in order to remove noise and other inconsistencies in the predictions.
-In a pilot study, we compared several different algorithms (Random Forests, Extreme Gradient Boosting, Naive Bayes, and Support Vector Machines), and it was found that Random Forests produced the most accurate results.
-Random Forests also computes Mean Decrease in Impurity, or Gini importance, highlighting what input variables are the most important for a given prediction \citep{gini}.
-Because of these factors, and the fact that it proved to be suitable for \citet{roelens}, Random Forests models are used to locate the ditches.
+We use a supervised learning algorithm with labelled ditch data to automatically extract ditches from instances not before seen by the model.
+These instances are used as input variables to the algorithm, and consist of refined or raw digital terrain indices derived from a DEM.
 
 We hypothesise that by combining the information from all the digital terrain indices using a machine learner (Random Forests), we can improve the detection of the ditches, compared to if a single index would be used.
-To test this, we conduct a study using a high resolution DEM of the Krycklan Catchment in Sweden \citep{krycklancatchment}, where a high precision manual mapping has been performed.
-This mapping was used to both train the Random Forests models and to produce the ground-truth.
-To evaluate the performance of the ditch detector, we capture the various aspects of detection performance through the following measures: accuracy, precision, recall, Cohen’s Kappa index, and Area under Precision-Recall curve score.
+To test this, we conduct a landscape-scale ditch detection analysis at the Krycklan Catchment in Sweden \citep{krycklancatchment}, where a high precision manual mapping has been performed previously.
+This manual map guided the model training and ground-truthing in our analysis and provided a test-bench for our methodology.
+Previous studies \citep{roelens, bailly, rapinel, kiss} that have located ditches from LiDAR data have worked on smaller areas (smaller than 2 $km^2$), predominantly characterised by artificial drainage.
+However, we herein conduct a landscape-scale analysis in a mesoscale catchment.
+The study  catchment  is 68 $km^2$, representing diverse landscape features and geomorphological characteristics.
 
@@
 The National Inventory of Landscapes in Sweden (NILS) has recorded small watercourses (width $<$ 6 m) by conducting a line inventory in 631 $5*5$ km squares randomly distributed in Sweden, to best capture a representative and varied sample of the Swedish landscape.
-The position of the small watercourses from this inventory were compared with the watercourses on Swedish Property Map (1: 12 500) (allowing for an uncertainty in the positioning of up to 15 m, by snapping the point observations to the lines on the map) (\hyperref[fig:watercoursebarplot]{Figure} \ref{fig:watercoursebarplot}).
+The positions of the small watercourses from this inventory were compared with the watercourses on Swedish Property Map (1: 12 500) (allowing for an uncertainty in the positioning of up to 15 m, by snapping the point observations to the lines on the map) (\hyperref[fig:watercoursebarplot]{Figure} \ref{fig:watercoursebarplot}).
 
 \subsection{Study area}
-Several other studies \citep{roelens, bailly, rapinel, kiss} that have located ditches from LiDAR data have worked on smaller areas (smaller than 2 $km^2$) where artificial drainage dominates.
-Here, we work on landscape scale in a mesoscale catchment.
-The  Krycklan Study Catchment in Sweden is 68 $km^2$, which means that we cover more landscape types.
-The quaternary deposits are dominated by till (51\%) and sorted sediments (30\%).
+In the Krycklan Study Catchment the quaternary deposits are dominated by till (51\%) and sorted sediments (30\%).
 The catchment ranges in elevation from 114 to 405 m above sea level.
 Forest covers 87\% of the catchment, mires 9\%, thin soils 7\%, and rock outcrops 1\%.
-The land use is dominated by forestry; approximately 25\% of the Krycklan catchment has been protected since 1922, but most of the other area is second growth forest \citep{krycklancatchment}.
+The land use is dominated by forestry; approximately 25\% of the Krycklan catchment has been protected since 1922, but  the rest of the area is mostly second growth forest \citep{krycklancatchment}.
 
-A historical survey of the catchment (Gudrun Norstedt, unpublished) shows that the ditches in the catchment were dug between 1900 and 1993.
-The earliest ditches were used to drain mires to harvest hay.
-Later, many of the ditches were dug based on the notion that any mire could be turned into productive forest land by drainage \citep{paivanen}.
-Today we know that nutrient limitation also plays a key role, and some drained wetlands did not turn into productive forest \citep{sikstrom}.
-Once a canopy of trees and shrubs has been established in the forest, the increased evapotranspiration \citep{price, koivusalo}, increased interception \citep{price}, and reduced groundwater recharge will continue to naturally lower the water-table in these areas, making ditch maintenance unnecessary, at least until the clear-cut phase when the canopy is removed.
-Also,  lowering the water table, the oxygenated soils show increasing bulk density and decreasing hydraulic conductivity \citep{silins, price}, causing a subsidence of the peat layer near the ditch.
-This is likely the most important factor causing ditches to become shallow \citep{heikurainen}.
+A historical survey of the catchment (Gudrun Norstedt, unpublished) recorded  that the ditches in the catchment were dug between 1900 and 1993.
+Initially, the ditches were used to drain the mires for hay production.
+These drained mires were later attempted to turn into productive forest lands \citep{paivanen}.
+However, research later showed that many of the drained mires were not productive for forestry due to their nutrient deficient environment \citep{sikstrom}.
 
 \subsection{Digitising the ground truth}
-A high resolution DEM was generated from LiDAR data, see \hyperref[lidartodem]{'Data and codes availability statement'} for specifications of the process.
-To train and evaluate the Random Forests models, a vector layer of all ditches in the Krycklan Catchment \citep{krycklancatchment} was digitised.
-Road ditches, forest ditches, and agricultural ditches can be found in the catchment.
+A high resolution DEM was generated from LiDAR data, see \hyperref[lidartodem]{'Data and codes availability statement'}.
+To train and validate the machine learning models, a vector layer of all ditches in the Krycklan Catchment \citep{krycklancatchment} was digitised.
+The study catchment comprises several types of ditches including road ditches, forest ditches, and agricultural ditches.
 A multidirectional hillshade of the 0.5 m DEM was calculated in ArcGIS Pro to locate the ditches.
 From this layer, most of the ditches could be easily identified.
-To ensure that the digitised ditch was located correctly inside the ditch and not on the bank nearby, the digitisation was conducted by zooming in to a scale of 1:600.
-In addition to this, both current and historical records and orthophotos dating back to 1960’s were used to help locate all the ditches.
-In uncertain areas, we also found it useful to view the data in 3D.
-However, the digitising was conducted in 2D, and in uncertain areas the ditches were verified in the field.
-We are confident that we have captured all the ditches in the catchment according to the digitised  ditch layers in \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan}.
+We conducted the digitisation at a scale of 1:600 to ensure all the digitised ditches properly align the actual ditch on the ground and do not coincide with the edge of the banks.
+In addition, both current and historical records and orthophotos dating back to the 1960's were consulted to help locate all the ditches.
+In uncertain areas, the ditches were verified in the field.
 The stream network in \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan} shows a network with a 8 hectare flow initiation threshold.
-Though, the length of the stream network vary substantially between season, ranging from 140 km during baseflow to 630 km during high-flow conditions \citep{mappingtemporal}.
-The digitisation of the ditch networks showed that there are 107 km road ditches and 208 km forest/agricultural ditches (\hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan}).
-The stream network in \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan}, which aims to show median flow conditions, was 197 km.
+The digitisation of the ditch networks showed that there are 107 km of road ditches and 208 km of forest/agricultural ditches in the catchment (\hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan}).
+The stream network in \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan}, with median flow conditions was 197 km.
 
@@
 From the ditch mapping, the vector layer was rasterised so that it could be compared to the automatically derived ditches from the $0.5*0.5$ m DEM.
-Although we have no data about the width of the ditches in the catchment, field observations have shown that the vast majority of ditches are between 0.5 and 3.5 metres wide.
-Because we wanted to ensure that the model received all ditch pixels correctly in the training phase, we widened the ditches so that all pixels within a radius of three pixels (1.5 metres) of the vectors were labelled as ditch pixels.
-This ensured that the edges of ditches were included, as well as that the raster converted from vector format actually covered the ditches.
+Our field observations have shown that the  width of most ditches within the catchment ranges between 0.5 and 3.5 m.
+To ensure that the model received all ditch pixels correctly in the training phase, we widened the ditches so that all pixels within a radius of three pixels (1.5 metres) of the vectors were labelled as ditch pixels.
 \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{a} shows the ditches rasterised from vectors and \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{b} shows the ditches after widening.
-The data in \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{b} is the labelled data that was used to train the Random Forests models.
+The data in \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{b} is the labelled data that was used to train the machine learning models.
 The ditches all have a generalised width based on a perceived average of ditch width in the area.
-Due to all ditches varying in width, it was impossible to produce a perfect representation of each ditch.
-However, this made for a good compromise for the average ditch.
+Due to all ditches varying in width, it was impossible to produce a perfect representation of each ditch, but this widening helped ensure that almost all ditch pixels were covered by the labels.
 
-Because our priority was to detect ditches, rather than accurately detecting each pixel labelled as a ditch, some adjustments were made when producing the prediction results.
-The dataset was divided into a lower resolution grid of $6*6$ pixels (9 $m^2$) for each grid.
-Each grid cell that contained at least 25\% ditch pixels was then labelled as a ditch, see \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{c}.
-This layer was used to ground-truth the Random Forests models.
+To match the labels with out predictions (where we downscale the predictions into grids to remove outliers), the ground-truth was also downscaled into grids of $6*6$ pixels (9 $m^2$) each.
+Each grid cell that contained at least 25\% ditch pixels was labelled as a ditch, see \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{c}.
+This layer was used to ground-truth the machine learning models.
 
@@
   \item \textbf{Sky View Factor} \newline
-  The Sky View Factor is a digital terrain index with a value ranging from 0 to 1, representing how much of the sky that is visible from a certain point on the ground \citep{zaksek}.
-On an open field all of the sky would be visible, but at the bottom of a ditch much of the sky would be obscured, meaning low values would indicate ditches.
-This method tends to give false positives in steep terrain, where the sky can be obscured by a hillside, or in narrow valleys or ravines that also show low numbers.
+    The Sky View Factor represents how much of the sky that is visible from a certain point on the ground \citep{zaksek}.
+A point in a ditch would yield a low value, as more of the sky is obscured by the ditch bank.
+However, it tends to produce false positives in steep terrain.
 The Sky View Factor was calculated in SAGA GIS, using a search radius of 10 metres to ensure that ditches were covered, and to not let steep hills close to ditches affect the results.
@@
   \item \textbf{Impoundment Index}\newline
-  The tool Impoundment Index in Whitebox Tools \citep{whiteboxtools} is a digital terrain index designed to serve as a measure of topographic incision and flow constriction.
-The index effectively maps the area and average depth of the reservoir that would be created by placing a dam of a user-specified length at each grid cell in the DEM.
-The tool can be tuned to output the flooded area or volume (in which case it highlights areas of flow constriction), or the average flooded depth (which seems to measure topographic incision).
-Additionally, the tool will output a 'dam height' raster that is particularly useful for mapping channels, such as ditches.
-Here, we used dam height calculated with a dam length of 3 metres, to cover the entire ditch which is typically not wider than 3 metres.
+    The Impoundment Index in Whitebox Tools \citep{whiteboxtools} can be used as a measure of flow constriction, which is useful for mapping channels, such as ditches.
+When tuned to output the flooded volume, it effectively maps the area and average depth of the reservoir that would be created by placing a dam of a user-specified length at each grid cell in the DEM.
+Here, we used dam height calculated with a dam length of 3 metres to cover the entire ditches.
 Typical false positives with this method are natural stream channels with a clear incised channel.
@@
   \item \textbf{High Pass Median Filter} \newline
-  High Pass Median Filters (HPMF) highlight ditches by indicating local depressions in the DEM.
-The algorithm operates essentially by subtracting the value at the grid cell at the centre of the window from the median value in the window.
-Negative values indicate depressions in the DEM, such as ditches.
-Typical false positives with this index are small hollows that also show negative numbers.
+    High Pass Median Filters (HPMF) highlight ditches by indicating local depressions (such as ditches) in the DEM by subtracting the value at the grid cell at the centre of the window from the median value in the window.
+Typical false positives with this index are small hollows.
 HPMF was calculated using Whitebox Tools \citep{whiteboxtools} with a window size of $4.5 * 4.5$ m.
@@
 In addition to the general input variables, a number of custom input variables were also developed.
-Several of these variables were calculated by looking at the separate digital terrain indices, attempting to amplify the ditches.
-This was done in several different ways for the indices by combining statistical aggregations, and by using thresholds to exclude, lower, or amplify the values of pixels that indicated ditches or non-ditches (\hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{d} and \hyperref[fig:features]{h}).
-
-From the Slope and Sky View Factor indices, input variables for the models were also extracted to attempt to amplify the values of pixels that lay in streams or in hilly areas, but leave pixels in ditches or other geographical areas untouched.
+Some of these custom variables were produced by statistical aggregation of multiple indices, whereas in some cases, they were generated by using thresholds to exclude, lower, or amplify the values of pixels that indicated ditches or non-ditches (\hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{d} and \hyperref[fig:features]{h}).
 
 Both HPMF and Sky View Factor were used with the image processing \textit{gabor filter}, which can be used to detect lines of a certain orientation in an image \citep{gabor}.
-30 Gabor filters, which were rotated in different angles and with different frequencies, were combined to detect lines, amplifying ditches by utilising the fact that ditches have a linear elongated shape (\hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{e}).
+30 Gabor filters which were rotated in different angles and with different frequencies were combined to detect lines, amplifying ditches by utilising the fact that ditches have a linear elongated shape (\hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{e}).
 
 \label{impoundmentstreamremoval}
-The raw Impoundment Index dam height was used to create a mask, attempting to mark out streams, but retain ditches.
-Only areas with a relatively high dam height, above 14 metres, which would indicate that these areas most likely contained streams, were marked out.
-After widening the resulting area, this mask was used to attempt to remove streams from all the aforementioned custom input variables, generating one new input variable from each.
+The raw Impoundment Index dam height was used to create a mask, attempting to mask out streams, but retain ditches.
+Only areas with a relatively high dam height, above 14 metres, which would indicate that these areas most likely contained streams, were masked out.
+After widening the resulting area, this mask was used to remove streams from all the aforementioned custom input variables, generating one new input variable from each.
 The features marked with \textit{streams removed} in \hyperref[featuretable]{Table} \ref{featuretable} make use of this filter (\hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{i}).
@@
 
-\textit{Model ensembles} is a category of machine learning where a set of models can be combined to increase diversity and robustness of predictions \citep{flach}.
-Random Forests is a supervised learning algorithm, first developed by \citet{breiman}, that makes use of the model ensemble technique by using a set of decision trees to build a \textit{forest} of trees.
-The outputs from these trees are then examined, and a class probability estimation ranging from 0 to 1 can be produced for each sample by dividing the amount of tree outputs of a class with the total amount of trees \citep{breiman}.
-
-The decision tree data structure is a \textit{tree} of connected nodes where the end node of each branch (leaf) produces an output value.
-Each step in the tree aims to minimise the entropy of the outputs \citep{kotsiantis}.
-Each node in a decision tree represents a variable or an abstraction of variables from the inputs and each split is made by using this variable to separate the inputs.
-
-Random Forests uses two techniques called \textit{subspace sampling} and \textit{bootstrap aggregation} (bagging) in tandem \citep{breiman}.
-Subspace sampling is the process of drawing a random subset of variables from the input variable set \citet{ho}.
-The bagging technique takes, with replacement, a set of different random samples for constructing each tree \citep{flach}.
-There are several hyperparameters that can be tuned to achieve better results from the model.
-For example, you can adjust the amount of decision trees to use, or the maximum amount of input variables to use for each subspace sampling \citep{scikit-learn}.
-We performed a hyperparameter tuning to determine what parameter values for the Random Forests algorithm would yield the best prediction.
-Using 300 trees and \textit{gini} as the splitting criterion, with a minimum of 10 samples per node split, and no artificial class weight or max depth of trees produced the best results.
-
-Random Forests computes Gini importance, which shows how often a variable is selected for a split for a certain classification \citep{gini}.
-Each input variable is given a Gini impurity rating, which measures the frequency of incorrect labelling occurring from its use.
-As the Gini impurity rating for a variable increases, its Gini importance decreases \citep{gini}.
-When the model is trained, it examines the Gini importance from all trees and produces the total Gini importance as a value between 0 and 1 \citep{gini}.
-The Gini importance can help when developing a model by indirectly giving advice on new input variables to introduce.
-
-The Random Forests algorithm from the Python library \textit{scikit-learn} was used in the experiment, and the classifiers were trained on all the input variables seen in \hyperref[featuretable]{Table} \ref{featuretable}.
-The probability predictions used in this Random Forests implementation are computed by counting the number of occurrences of a class output and dividing with the total amount of trees in the forest \citep{scikit-learn}.
-The testing phase showed that the classifiers produced poor results when the ratio of ditch- versus non-ditch pixels in the training data was very high.
-A high ratio led to the models not being punished for mislabelling ditches as non-ditches, causing it to prioritise a high accuracy over a high recall.
-According to \citet{balanced}, an imbalanced training dataset causes a minority class to be less accurately predicted.
-Because the ditch class is much less common than the non-ditch class, we attempted to train the models with a roughly equal amount of ditch pixels and non-ditch pixels to balance the models.
-The models were, however, still evaluated on all the pixels in the 11 validation subsections.
+Random Forests is an ensemble machine learning method which builds multiple diverse decision trees to increase the robustness of its predictions \citep{breiman,flach}.
+The individual trees are trained by minimising the entropy of the data from each node’s output \citep{kotsiantis}.
+A byproduct of the training is the Gini importance, which denotes the most important input variables \citep{gini}.
+For our study, a hyperparameter tuning showed that using 300 trees and \textit{gini} as the splitting criterion, with a minimum of 10 samples per node split, and no artificial class weight or max depth of trees yielded the best predictions.
+The Random Forests algorithm from the Python library \textit{scikit-learn} \citep{scikit-learn} was used in the experiment, and the classifiers were trained on all the input variables seen in \hyperref[featuretable]{Table} \ref{featuretable}.
 
@@
 
-The first step to create a more balanced training dataset was to extract all pixels labelled as ditches, as well as pixels within close proximity of ditches.
-Secondly, pixels were sampled randomly from the entire subsection.
-This allowed the training dataset to be fairly balanced while still containing most of the geographical features of each subsection (\hyperref[fig:balancedmasks]{Figure} \ref{fig:balancedmasks}).
+The testing phase showed that the classifiers produced poor results when the ratio of ditch versus non-ditch pixels in the training data was very high, as the models were not being punished for mislabelling ditches as non-ditches.
+To increase the recall, we evened out the amount of ditch and non-ditch instances in the training data.
+First, we extracted all pixels labelled as ditches as well as pixels within close proximity of ditches.
+Secondly, pixels were sampled randomly from the entire subsection to still contain most of the geographical features of each area (\hyperref[fig:balancedmasks]{Figure} \ref{fig:balancedmasks}).
 
@@
 \subsubsection{Noise reduction and gap filling}
-The probability predictions contained a lot of noise in geographical locations far away from ditches, which needed to be excluded.
-The first step for removing noise was to use a bilateral de-noising filter on the entire prediction image.
-This left linear properties and pixels with a very high value intact, while lowering the value of pixels that did not contribute to an accurate prediction (\hyperref[fig:postprocessing1]{Figure} \ref{fig:postprocessing1} \hyperref[fig:postprocessing1]{b}).
+
+The probability predictions contained a lot of noise in the areas largely distant from the ditches.
+This noise was removed by first using a bilateral de-noising filter, and then removing outlier pixels that were far away from any other high probability ditch pixels.
+This left linear properties and pixels with a very high value intact, while lowering the value of pixels that did not contribute to an accurate prediction (\hyperref[fig:postprocessing1]{Figure} \ref{fig:postprocessing1} \hyperref[fig:postprocessing1]{b}) (\hyperref[fig:postprocessing2]{Figure} \ref{fig:postprocessing2} \hyperref[fig:postprocessing2]{a}).
+
+To fill gaps in ditches that the models failed to correctly predict, we aggregated pixel values by covering them with cone masks expanding outwards in different directions from the examined pixel.
+This amplified some of the noise that was left, but filling the gaps in the ditches was judged to be more important (\hyperref[fig:postprocessing2]{Figure} \ref{fig:postprocessing2} \hyperref[fig:postprocessing2]{b}).
 
@@
 \end{figure}
-
-The second step for removing noise was to use a custom function to remove pixels that had a semi-high ditch probability, but that lay far away from any other high probability pixels.
-In this function, a threshold value was used to avoid removing pixels with a high enough ditch probability, helping to retain pixels that lay in or close to a ditch.
-The max ditch probability value in a circular radius of 10 pixels was then calculated.
-If this max value was too low, the probability value of the examined pixel was lowered (\hyperref[fig:postprocessing2]{Figure} \ref{fig:postprocessing2} \hyperref[fig:postprocessing2]{a}).
-
-The third step involved taking measures to try to fill gaps in ditches that the models failed to correctly predict.
-This was done by aggregating pixels covered by cone masks expanding outwards in different directions from the examined pixel.
-This step also amplified some of the noise that was left, but filling the gaps in the ditches was judged to be more important to help make the next step more effective (\hyperref[fig:postprocessing2]{Figure} \ref{fig:postprocessing2} \hyperref[fig:postprocessing2]{b}).
 
@@
 \section{Results and analysis}
+In the national survey (NILS), a total of 1103 natural watercourses, 131 straightened watercourses/channels, and 2089 ditches were found in the field.
+This simple investigation highlights two things: Firstly, the ditches make up the majority of the small scale water channels in the Swedish landscape; almost twice as many as the natural streams.
+Secondly, most of the small scale channels are missing on the current maps; only 45 \% of the natural watercourses, 25 \% of the straightened watercourses/channels, and 9 \% of the ditches were mapped (\hyperref[fig:watercoursebarplot]{Figure} \ref{fig:watercoursebarplot}).
+
+\begin{figure}[!htb]
+    \centering
+    \begin{tikzpicture}
+    \begin{axis}[
+        width={0.9\textwidth},
+        height={0.6\textwidth},
+        ylabel={fluorescence},
+        ybar stacked,
+        ymajorgrids=true,
+    	bar width=45pt,
+    	nodes near coords,
+    	every node near coord/.append style={font=\tiny},
+        enlarge y limits={0.15},
+        enlarge x limits=0.15,
+        legend style={at={(0.78,0.96)},
+          anchor=north,legend columns=-1},
+        ylabel={{}},
+        symbolic x coords={Ditches, Straightened  watercourses, Natural  watercourses},
+        xtick=data,
+        ytick={0,500,1000,1500,2000},
+        x tick label style={rotate=0,anchor=north},
+        ]
+    \addplot+[ybar,color=custombeige,fill=custombeige,draw=custombeigeouter,thick] plot coordinates {(Ditches,188) (Straightened  watercourses,33) 
+      (Natural  watercourses,496)};
+    \addplot+[ybar,color=customgrayouter,fill=customgray,draw=customgrayouter,thick] plot coordinates {(Ditches,1901) (Straightened  watercourses,98) 
+      (Natural  watercourses,607)};
+    \legend{\strut Mapped, \strut Unmapped}
+    \end{axis}
+    \end{tikzpicture}
+    \caption{The bars indicate the number of observed channels in the National Inventory of Landscapes in Sweden's (NILS) line inventory of small watercourses (width $<$ 6 m).
+The colours of the bars indicate if they are mapped or not on the Swedish Property map.}
+    \label{fig:watercoursebarplot}
+\end{figure}
 
@@
 However, this indicates that our method could also be used to map small, previously unmapped natural stream channels.
-A major knowledge gap, both nationally  \citep{kuglerova} and internationally \citep{benstead} is how to best identify the location of small brooks and creeks in the landscape - a phenomenon called “Aqua Incognita, the unknown headwaters” \citep{bishop, kuglerova}.
+The national survey showed that 55\% of the natural stream channels and 75\% of the straightened water courses were missing on the maps.
+This is in line with other national \citep{kuglerova} and international \citep{benstead} studies highlighting this phenomenon called "Aqua Incognita - the unknown headwaters" \citep{bishop,kuglerova}.
 
 The results from our method (\hyperref[predictionperformance]{Table} \ref{predictionperformance}) show that we managed to attain a Cohen's $\kappa$ rating in the substantial range, according to performance thresholds proposed by \citet{kappaanalysis}.
-This means that our ditch detector performed substantially better than a completely random ditch detector.
 The total recall value of 70.28 \% shows that we managed to find a majority of the ditch pixels that exist.
-The results of our method also showed a significant improvement for ditch detection over classifying with the digital terrain indices separately.
+The results of our method also showed a significant improvement for ditch detection over classifying with the digital terrain indices separately (\hyperref[recreatedpredictionperformance]{Table} \ref{recreatedpredictionperformance}).
 Had we been able to use labels that were correctly labelled on a pixel basis, i.e.
@@
 
-\citet{roelens} showed that you can use machine learning to detect ditches directly from point-clouds instead of a DEM.
-One concern we had with our method was that valuable information may have been discarded when aggregating the elevations from the original point cloud to a DEM, making it more difficult to locate the ditches.
-However, we achieved a similar score to their method; $\kappa=0.73$ in our study compared to 0.77 for grassland and 0.73 for peri-urban area in Roelens study.
-This suggests that you can also use a DEM to automatically extract ditches, provided that you have enough input data to generate a high resolution DEM to capture the small scale features of ditches.
+One concern with our method was that valuable information may have been discarded when aggregating the elevations from the original point cloud to a DEM.
+\citet{roelens} used  machine learning to detect ditches directly from point-clouds instead of a DEM.
+However, we achieved a similar score to their method; $\kappa=0.73$ in our study compared to 0.77 for grassland and 0.73 for peri-urban area in \citet{roelens} study.
+This suggests that DEM can also be used to automatically extract ditches, provided that enough input data is available to generate a high resolution DEM for capturing the small scale features of ditches.
 
 \citet{bailly} also showed that the vegetation cover significantly affects the ability to detect ditches from LiDAR data (\hyperref[fig:ditchpictures]{Figure} \ref{fig:ditchpictures} \& \ref{fig:resultstreesbushes}).
-Approximately 75\% of the ditches in their study were retrieved where no vegetation existed, whereas the detection levels sank to below 10\% under vegetation; especially when high vegetation, shrubs and trees covered the ditches \citep{bailly}.
-Hence, a $\kappa$ of 0.73 in the Krycklan catchment with  87\% forest cover \citep{krycklancatchment} is in that aspect surprisingly good, and could perhaps be a result of the higher point density of our LiDAR measurements (20 compared to 10 points $m^{2}$).
-Or, perhaps the generation of a DEM from the point cloud has made the detection more robust?
+Approximately 75\% of the ditches in their study were retrieved where no vegetation existed, whereas the detection accuracy went below 10\% under vegetation; especially when the ditches were covered by high vegetation, shrubs, and trees \citep{bailly}.
+Hence, a $\kappa$ of 0.73 in our study  with  87\% forest cover \citep{krycklancatchment} is substantially better than that of previous research, and this could perhaps be a result of the higher point density of our LiDAR measurements (20 compared to 10 points $m^{2}$).
+The hydrology of our study catchment also differs from that of other studies, as we have almost as many natural stream channels as ditches, whereas their studies were conducted in smaller, predominantly artificially drained areas \citep{bailly, roelens, rapinel}.
+Because our method locates both ditches and natural stream channels, we have to take additional measures to remove the stream channels (affecting recall negatively), as explained in \ref{impoundmentstreamremoval}.
 
-The hydrology of our study catchment also differs from the other mentioned studies as there are almost as many natural stream channels as there are ditches in the area \citep{hasselquist, mappingtemporal}.
-Other studies were conducted in smaller, predominately artificially drained areas \citep{roelens, bailly, rapinel}.
-Our method locates both artificial channels (ditches) and natural stream channels.
-This would not be a problem if the aim had been to locate all channels.
-However, our study focused on locating ditches specifically.
-Hence, we tried to remove the natural stream channels by manipulating some of the input variables, using the Impoundment Index to lower the values of pixels that had a high dam height, as explained in \ref{impoundmentstreamremoval}.
-This helped to remove the deepest streams and increased the overall accuracy, with the drawback of decreasing the recall slightly (\hyperref[fig:resultsillustrations]{Figure} \ref{fig:resultsillustrations} \hyperref[fig:resultsillustrations]{f}).
-This method did not remove the smaller streams (\hyperref[fig:resultsillustrations]{Figure} \ref{fig:resultsillustrations} \hyperref[fig:resultsillustrations]{b}).
 A future improvement to the ditch detector could be to use a shape index to remove small natural streams, as natural streams often form more complex channels whereas ditches are often straight.
 A potential issue with generalising our ditch detector for use in other geographical areas is that all the post-processing steps and input variables were developed based on occurrences in the Krycklan area.
-Different geographical compositions in other areas may require that the thresholds used to fill gaps in ditches and to remove noise would need to be adjusted.
+Different geographical compositions in other areas may require the adjustment of the thresholds used to fill gaps in ditches and to remove noise.
 
@@
 The data and codes that support the findings of this study are available at the private links:\newline
-\mbox{\href{https://figshare.com/s/c03de5594268843a8b50}{https://figshare.com/s/c03de5594268843a8b50} - Functions\itshape\ignorespaces}
+\mbox{\href{https://figshare.com/s/23742e1f20fae3e404de}{https://figshare.com/s/23742e1f20fae3e404de} - Functions\itshape\ignorespaces}
 \mbox{\href{https://figshare.com/s/741725088fe6292442c8}{https://figshare.com/s/741725088fe6292442c8} - Feature creation\itshape\ignorespaces}
 \mbox{\href{https://figshare.com/s/f4e3bcf3bb2f8c18ecfa}{https://figshare.com/s/f4e3bcf3bb2f8c18ecfa} - Digital terrain indices experiment\itshape\ignorespaces}
-\mbox{\href{https://figshare.com/s/c1326af09f600d18a9d7}{https://figshare.com/s/c1326af09f600d18a9d7} - Random Forests experiment\itshape\ignorespaces}
+\mbox{\href{https://figshare.com/s/88cd93c3b4aea6776c8b}{https://figshare.com/s/88cd93c3b4aea6776c8b} - Pilot experiment\itshape\ignorespaces}
+\mbox{\href{https://figshare.com/s/60631095d26699f65f53}{https://figshare.com/s/60631095d26699f65f53} - Random Forests experiment\itshape\ignorespaces}
 \mbox{\href{https://figshare.com/s/c0ae38ee5951498d21cd}{https://figshare.com/s/c0ae38ee5951498d21cd}} - Experiment data
@@
 \section*{Acknowledgements}
-We wish to thank Sigurd Israelsson at the School of Engineering at J\"onk\"oping University for lending hardware resources for the experiments in this study.
+We wish to thank the two anonymous reviewers and Dr.
+Siddhartho Paul at the Swedish University of Agricultural Sciences, whose suggestions helped improve and clarify this manuscript.
+We also thank Sigurd Israelsson at the School of Engineering at J\"onk\"oping University for lending hardware resources for the experiments in this study.
 The work has been funded by VINNOVA, EU InterReg Baltic Sea project WAMBAF tools and Formas.

### 2020-09-22 LaTeX/Ditch detection in high-resolution digital elevation models.tex
Adds minor tweaks in article, and changes a bit to clarify the rejoinder.

@@
 \name{Filip Andersson\textsuperscript{a,1}, Jonatan Flyckt\textsuperscript{a,1}, Niklas Lavesson\textsuperscript{a}\thanks{CONTACT Niklas Lavesson.
-Email: niklas.lavesson@ju.se}, Liselott Nilsson\textsuperscript{b}, and Anneli \AA gren\textsuperscript{c}} \affil{\textsuperscript{a}J\"onk\"oping Artificial Intelligence Laboratory, J\"onk\"oping University, Gjuterigatan 5, 551 11, J\"onk\"oping, Sweden; \textsuperscript{b}Forest Department, Swedish Forest Agency, Skeppargatan 17, 931 32 Skellefte\aa, Sweden; \textsuperscript{c}Department of Forest Ecology and Management, Swedish University of Agricultural Sciences, SLU, Skogsmarksgr\"and 17, 901 83, Ume\aa, Sweden}
+Email: niklas.lavesson@ju.se}, Liselott Nilsson\textsuperscript{b}, and Anneli M.
+\AA gren\textsuperscript{c}} \affil{\textsuperscript{a}J\"onk\"oping Artificial Intelligence Laboratory, J\"onk\"oping University, Gjuterigatan 5, 551 11, J\"onk\"oping, Sweden; \textsuperscript{b}Forest Department, Swedish Forest Agency, Skeppargatan 17, 931 32 Skellefte\aa, Sweden; \textsuperscript{c}Department of Forest Ecology and Management, Swedish University of Agricultural Sciences, SLU, Skogsmarksgr\"and 17, 901 83, Ume\aa, Sweden}
 \footnote{\textsuperscript{1} Filip Andersson and Jonatan Flyckt contributed equally to this work as first authors.}}
@@
 In many places both the landscape and the channel network have also been heavily altered  by humans, especially in the agricultural landscapes, for example in mid-western United States \citep{passalacqua}.
-In North Europe, however, ditches have been dug to turn flat peatlands into productive forests, resulting in profound impacts on the hydrology of this region.
+In North Europe, however, ditches have mainly been dug to turn flat peatlands into productive forests, resulting in profound impacts on the hydrology of this region.
 \citep{peatlands}.
@@
 These drained mires were later attempted to turn into productive forest lands \citep{paivanen}.
-However, research later showed that many of the drained mires were not productive for forestry due to their nutrient deficient environment \citep{sikstrom}.
+However, research later showed that many of the drained mires were not productive for forestry due to lack of nutrients \citep{sikstrom}.
 
@@
 The digitisation of the ditch networks showed that there are 107 km of road ditches and 208 km of forest/agricultural ditches in the catchment (\hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan}).
-The stream network in \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan}, with median flow conditions was 197 km.
+The stream network in \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan}, which represents median flow conditions, was 197 km.
 
@@
 \subsection{Extracting ditches with digital terrain indices}
-From the DEM, several digital terrain indices were calculated:
+From the DEM, several digital terrain indices were calculated. The strengths and weaknesses of these indices have been obtained from observing their performance in different topological environments:
 \begin{itemize} \label{data_attributes}
@@
 However, it tends to produce false positives in steep terrain.
-The Sky View Factor was calculated in SAGA GIS, using a search radius of 10 metres to ensure that ditches were covered, and to not let steep hills close to ditches affect the results.
+The Sky View Factor was calculated in SAGA GIS, using a search radius of 10 metres to ensure that it was large enough to capture the ditches, but small enough to not include large scale topographical features such as hills.
 
@@
 However, most of the ditches in Krycklan today are found below the canopy (e.g.
-\hyperref[fig:ditchpictures]{Figure} \ref{fig:ditchpictures} \hyperref[fig:ditchpictures]{c}), and the majority have not been maintained, causing difficulties in detecting them as some have grown back in with peat, grasses, or shrubs (fig 12b, 13b).
+\hyperref[fig:ditchpictures]{Figure} \ref{fig:ditchpictures} \hyperref[fig:ditchpictures]{c}), and the majority have not been maintained, causing difficulties in detecting them as some have grown back in with peat, grasses, or shrubs (\hyperref[fig:resultstreesbushes]{Figure} \ref{fig:resultstreesbushes} \hyperref[fig:resultstreesbushes]{b}).
 This, in combination with the surrounding peat having subsided \citep{heikurainen}, have rendered them almost impossible to detect in the DEM today (\hyperref[fig:ditchpictures]{Figure} \ref{fig:ditchpictures} \hyperref[fig:ditchpictures]{d}).
@@
 
-This study investigates how to use digital elevation data together with manually labelled ditches to identify patterns and relationships that can be used for automatic ditch detection.
-We have shown that it is possible to use machine learning with digital elevation models to learn patterns that enable robust detection of ditches.
-Designing a robust ditch detection in a study area with substantial natural stream coverage was challenging, both in regards to model input variables and post-processing measures.
+Detection of artificial ditches in a study area with substantial natural stream coverage is
+challenging but highly desired for effective forest management.
+This study developed and applied a methodology using multiple terrain indices and machine learning for automatic ditch detection at the Krycklan catchment in Sweden.
+The confidence interval for the Cohen's Kappa index ranged [0.655 , 0.781] between the evaluation plots with a confidence level of 95\%.
 Future work could introduce more input variables and post-processing steps, such as using pathfinding algorithms to fill gaps in the ditch model, or shape indices to remove stream channels from the prediction.

### 2020-10-05 LaTeX/anonymised.tex
Adds new conclusion and updated abstract.

--- 2020-09-22 LaTeX/Ditch detection in high-resolution digital elevation models.tex
@@
 \author{
-\name{Filip Andersson\textsuperscript{a,1}, Jonatan Flyckt\textsuperscript{a,1}, Niklas Lavesson\textsuperscript{a}\thanks{CONTACT Niklas Lavesson.
-Email: niklas.lavesson@ju.se}, Liselott Nilsson\textsuperscript{b}, and Anneli M.
-\AA gren\textsuperscript{c}} \affil{\textsuperscript{a}J\"onk\"oping Artificial Intelligence Laboratory, J\"onk\"oping University, Gjuterigatan 5, 551 11, J\"onk\"oping, Sweden; \textsuperscript{b}Forest Department, Swedish Forest Agency, Skeppargatan 17, 931 32 Skellefte\aa, Sweden; \textsuperscript{c}Department of Forest Ecology and Management, Swedish University of Agricultural Sciences, SLU, Skogsmarksgr\"and 17, 901 83, Ume\aa, Sweden}
-\footnote{\textsuperscript{1} Filip Andersson and Jonatan Flyckt contributed equally to this work as first authors.}}
+\name{Anonymised.}}
 
@@
 A historical survey of the catchment (Gudrun Norstedt, unpublished) recorded  that the ditches in the catchment were dug between 1900 and 1993.
-Initially, the ditches were used to drain the mires for hay production.
-These drained mires were later attempted to turn into productive forest lands \citep{paivanen}.
+Initially, the ditches were dug to drain the mires for hay production.
+Later on, wet areas were drained to attempt to turn them into productive forest lands \citep{paivanen}.
 However, research later showed that many of the drained mires were not productive for forestry due to lack of nutrients \citep{sikstrom}.
@@
 The raw Impoundment Index dam height was used to create a mask, attempting to mask out streams, but retain ditches.
-Only areas with a relatively high dam height, above 14 metres, which would indicate that these areas most likely contained streams, were masked out.
+Only areas with a relatively high value, which would indicate that these areas most likely contained streams, were masked out.
 After widening the resulting area, this mask was used to remove streams from all the aforementioned custom input variables, generating one new input variable from each.
@@
 The $\kappa$ ratings from the four digital terrain indices experiments can be seen in  \hyperref[recreatedpredictionperformance]{Table} \ref{recreatedpredictionperformance} ($\kappa$ = 0.048, $\kappa$ = 0.215, $\kappa$ = 0.211, $\kappa$ = 0.029).
-Because the $\kappa$ rating from our method outperforms all digital terrain indices, our hypothesis has been confirmed by the experiments.
+Because the $\kappa$ rating from our method outperforms all digital terrain indices, our hypothesis that combining digital terrain indices with a machine learning produces better ditch detection than using single terrain indices has been confirmed by the experiments.
 Most false positives lie in stream channels (\hyperref[fig:resultsillustrations]{Figure} \ref{fig:resultsillustrations} \hyperref[fig:resultsillustrations]{a, b}) and most false negatives occurred either due to ditches being too shallow (\hyperref[fig:resultsillustrations]{Figure} \ref{fig:resultsillustrations} \hyperref[fig:resultsillustrations]{c, d}), or due to the attempt at removing streams using the input variables as explained in \ref{impoundmentstreamremoval} removing the deepest ditches  (\hyperref[fig:resultsillustrations]{Figure} \ref{fig:resultsillustrations} \hyperref[fig:resultsillustrations]{e, f}).
@@
 \label{lidartodem}
-The LiDAR dataset used in this study was produced by TerraTec Sweden AB in August 2015, on demand of the Department of Forest Resource Management at the Swedish University of Agricultural Sciences, and is freely available online at \href{https://gis.krycklan.se/}{https://gis.krycklan.se/}.
+The LiDAR dataset used in this study (anonymised), and is freely available online at \href{https://gis.krycklan.se/}{https://gis.krycklan.se/}.
 The average density of the dataset was 20 pulses per $m^2$.
@@
 \section*{Acknowledgements}
-We wish to thank the two anonymous reviewers and Dr.
-Siddhartho Paul at the Swedish University of Agricultural Sciences, whose suggestions helped improve and clarify this manuscript.
-We also thank Sigurd Israelsson at the School of Engineering at J\"onk\"oping University for lending hardware resources for the experiments in this study.
-The work has been funded by VINNOVA, EU InterReg Baltic Sea project WAMBAF tools and Formas.
+Anonymised.
 

### 2020-10-10 LaTeX/anonymised.tex
Completes major revision.

@@
 
-\author{
-\name{Anonymised.}}
+\author{Anonymised.}
 
@@
 \begin{abstract}
-In this study, a method was developed for detecting ditches in high resolution digital elevation models derived from LiDAR scans.
-
+In this study, we develop a method for detecting ditches in high resolution digital elevation models derived from LiDAR scans.
 Thresholding methods using digital terrain indices, such as Sky View Factor, Impoundment Index, or High Pass Median Filter can be used to detect ditches.
 However, a single threshold generally does not capture the variability in the landscape and generates many false positives and negatives.
-We hypothesised that, by combining the digital terrain indices using machine learning, we could improve ditch detection at a landscape-scale.
-
-In addition to the raw terrain indices, additional input variables were generated by transforming the data by including neighbouring cells to improve the predictability of the ditches.
-Random Forests was used to locate the ditches, and the probability output from this classifier was processed to remove noise, and binarised to produce the final ditch prediction.
-
-The confidence interval for the Cohen's Kappa index ranged \textbf{[0.655 , 0.781]} between the evaluation plots with a confidence level of \textbf{95\%}.
-Input variables based on the Impoundment Index overall represented the strongest ditch predictors.
-
+We hypothesise that, by combining the digital terrain indices using machine learning, we can improve ditch detection at a landscape-scale.
+In addition to the raw terrain indices, additional input variables are generated by transforming the data by including neighbouring cells to improve the predictability of the ditches.
+Random Forests is used to locate the ditches, and the probability output from this classifier is processed to remove noise, and binarised to produce the final ditch prediction.
+The confidence interval for the Cohen's Kappa index ranges \textbf{[0.655 , 0.781]} between the evaluation plots with a confidence level of \textbf{95\%}.
+Input variables based on the Impoundment Index overall represents the strongest ditch predictors.
 The study demonstrates that combining information from  a suite of digital terrain indices using machine learning provides an effective technique for automatic ditch detection at a landscape-scale.
@@
 
-Detection of artificial ditches in a study area with substantial natural stream coverage is
-challenging but highly desired for effective forest management.
+%Detection of artificial ditches in a study area with substantial natural stream coverage is challenging but highly desired for effective forest management.
 This study developed and applied a methodology using multiple terrain indices and machine learning for automatic ditch detection at the Krycklan catchment in Sweden.
-The confidence interval for the Cohen's Kappa index ranged [0.655 , 0.781] between the evaluation plots with a confidence level of 95\%.
+Despite a substantial vegetation (87\% forest cover), our method achieved good results; the confidence interval for the Cohen's Kappa index ranged [0.655 , 0.781] between the evaluation plots with a confidence level of 95\%.
+With the developed input variables and extensive post-processing, we managed to remove many stream channels and fill gaps in the ditches where the LiDAR scans were weak.
+However, these issues are still present to some extent.
 Future work could introduce more input variables and post-processing steps, such as using pathfinding algorithms to fill gaps in the ditch model, or shape indices to remove stream channels from the prediction.
@@
 With minor adjustments, our proposed method can be employed in a general setting, reducing manual labour and cost, and enabling faster and easier mapping of ditches.
+
+Detection of artificial ditches over large areas with substantial natural stream coverage is challenging, especially under tree canopy, but highly desired for effective forest management.
+Ditches represent elongated depressions in a DEM; however, the exact shape and size differs between ditches.
+We combined recently developed digital terrain indices (the most important being Impoundment Index and High Pass Medium Filter) but, unlike previous studies, did not apply generic thresholding on them.
+Instead, we included local variability in the indices by using different radii to capture ditches with varied shapes.
+This integration of a large set of terrain indices using machine learning produced good results, as evidenced by the confidence interval for the Cohen's $\kappa$ index ranging from 0.655 to 0.781 with a confidence level of 95\%.
+With the developed input variables and extensive post-processing, we also managed to remove many stream channels and fill gaps in the ditches where the LiDAR scans were weak.
+
+Although there is room for further improvement, the methodology is a significant step forward as it performed on a par with previous studies conducted in more homogeneous agricultural landscapes.
+Our approach enables accurate ditch detection in complex forested landscapes with tree canopy and many natural streams.
+Future work could introduce more input variables and post-processing steps, such as using pathfinding algorithms to fill gaps in the ditch model or shape indices to remove stream channels from the prediction.
+Moving towards image segmentation instead of pixel classification of tabular pixel values could also be a suitable future approach.
+Our developed methodology can be adjusted to apply to any setting for artificial ditch detection, reducing time and labour investment compared to the conventional manual mapping techniques.
+The accurate ditch maps can significantly contribute to practical forestry and operational land management at a landscape-scale.
 
@@
 \label{lidartodem}
-The LiDAR dataset used in this study (anonymised), and is freely available online at \href{https://gis.krycklan.se/}{https://gis.krycklan.se/}.
+The LiDAR dataset used in this study was produced by (anonymised), and is freely available online at \href{https://gis.krycklan.se/}{https://gis.krycklan.se/}.
 The average density of the dataset was 20 pulses per $m^2$.

### 2021-01-04 LaTeX ESWA/article.tex
Starts work on porting article to ESWA format.

\documentclass[review]{elsarticle}
\graphicspath{ {./images/} }
\usepackage{hyperref}
\usepackage{float}
\usepackage{verbatim} %comments
\usepackage{apalike}
\restylefloat{figure}
\restylefloat{table}

\journal{Expert Systems with Applications}

\bibliographystyle{model5-names}\biboptions{authoryear}

\begin{document}
\begin{frontmatter}

\begin{titlepage}
\begin{center}
\vspace*{1cm}

\textbf{ \large Detecting ditches using supervised learning on high-resolution digital elevation models}

\vspace{1.5cm}

Jonatan Flyckt$^{a}$ (fljo1589@student.ju.se), Filip Andersson$^{a}$ (anfi1632@student.ju.se), Niklas Lavesson$^a$ (niklas.lavesson@ju.se), Liselott Nilsson$^b$ (liselott.nilsson@skogsstyrelsen.se), Anneli M. \AA gren$^c$ (anneli.agren@slu.se) \\

\hspace{10pt}

\begin{flushleft}
\small  
$^a$ J\"onk\"oping Artificial Intelligence Laboratory, J\"onk\"oping University, Gjuterigatan 5, 551 11, J\"onk\"oping, Sweden \\
$^b$ Forest Department, Swedish Forest Agency, Skeppargatan 17, 931 32 Skellefte\aa, Sweden \\
$^c$ Department of Forest Ecology and Management, Swedish University of Agricultural Sciences, SLU, Skogsmarksgr\"and 17, 901 83, Ume\aa, Sweden

\begin{comment}
Clearly indicate who will handle correspondence at all stages of refereeing and publication, also post-publication. Ensure that phone numbers (with country and area code) are provided in addition to the e-mail address and the complete postal address. Contact details must be kept up to date by the corresponding author.
\end{comment}

\vspace{1cm}
\textbf{Corresponding Author:} \\
Niklas Lavesson \\
J\"onk\"oping Artificial Intelligence Laboratory, J\"onk\"oping University, Gjuterigatan 5, 551 11, J\"onk\"oping, Sweden \\
Tel: (+46) 70-5383338 \textbf{Ska detta nummer stå här? (corresponding author måste ha telefonnummer)} \\
Email: niklas.lavesson@ju.se

\end{flushleft}        
\end{center}
\end{titlepage}

\title{Detecting ditches using supervised learning on high-resolution digital elevation models}

\author[ju]{Jonatan Flyckt\corref{equal_contribution}}
\ead{fljo1589@student.ju.se}

\author[ju]{Filip Andersson\corref{equal_contribution}}
\ead{anfi1632@student.ju.se}

\author[ju]{Niklas Lavesson \corref{cor1}}
\ead{niklas.lavesson@ju.se}

\author[fa]{Liselott Nilsson}
\ead{liselott.nilsson@skogsstyrelsen.se}

\author[slu]{Anneli M. \AA gren}
\ead{anneli.agren@slu.se}

\cortext[cor1]{Corresponding author.}
\address[ju]{J\"onk\"oping Artificial Intelligence Laboratory, J\"onk\"oping University, Gjuterigatan 5, 551 11, J\"onk\"oping, Sweden}
\address[fa]{Forest Department, Swedish Forest Agency, Skeppargatan 17, 931 32 Skellefte\aa, Sweden}
\address[slu]{Department of Forest Ecology and Management, Swedish University of Agricultural Sciences, SLU, Skogsmarksgr\"and 17, 901 83, Ume\aa, Sweden}
\cortext[equal_contribution]{Jonatan Flyckt and Filip Andersson contributed equally to this work as first authors.}

\begin{abstract}
In this study, we develop a method for detecting ditches in high resolution digital elevation models derived from LiDAR scans. Thresholding methods using digital terrain indices, such as Sky View Factor, Impoundment Index, or High Pass Median Filter can be used to detect ditches. However, a single threshold generally does not capture the variability in the landscape and generates many false positives and negatives. We hypothesise that, by combining the digital terrain indices using machine learning, we can improve ditch detection at a landscape-scale. In addition to the raw terrain indices, additional input variables are generated by transforming the data by including neighbouring cells to improve the predictability of the ditches. Random Forests is used to locate the ditches, and the probability output from this classifier is processed to remove noise, and binarised to produce the final ditch prediction. The confidence interval for the Cohen's Kappa index ranges \textbf{[0.655 , 0.781]} between the evaluation plots with a confidence level of \textbf{95\%}. Input variables based on the Impoundment Index overall represents the strongest ditch predictors. The study demonstrates that combining information from  a suite of digital terrain indices using machine learning provides an effective technique for automatic ditch detection at a landscape-scale.
\end{abstract}

\begin{keyword}
Machine learning \sep Geographic information systems \sep Classification and regression trees \sep Supervised learning by classification
\end{keyword}

\end{frontmatter}

\section{Introduction}
\label{introduction}

\section*{Acknowledgements}

\bibliography{references}

\end{document}

### 2021-01-06 LaTeX ESWA/article.tex
Adds WIP on converting to ESWA template.

@@
 \usepackage{apalike}
+
+%custom:
+\usepackage{mathastext}
+\usepackage{pgfplots}
+\pgfplotsset{width=7cm,compat=1.8}
+\definecolor{custombeige}{HTML}{e0d4ba}
+\definecolor{custombeigeouter}{HTML}{968f7e}
+\definecolor{customgray}{HTML}{666666}
+\definecolor{customgrayouter}{HTML}{404040}
+
 \restylefloat{figure}
@@
 
-\section{Introduction}
-\label{introduction}
+\section{Introduction}\label{introduction}
+
+Over the last century, the wetlands in North Europe and parts of North America have been substantially modified and ditched.
+In many places both the landscape and the channel network have also been heavily altered  by humans, especially in the agricultural landscapes, for example in mid-western United States \citep{passalacqua}.
+In North Europe, however, ditches have mainly been dug to turn flat peatlands into productive forests, resulting in profound impacts on the hydrology of this region.
+\citep{peatlands}.
+In Sweden, which is one of the North European countries with the most extensive ditch network, over five million hectares of peatland have been drained for forestry in the 20th century.
+As a result, almost one million km of ditches have been introduced, which has doubled the length of watercourses across Sweden  \citep{hasselquist}.
+
+For effective restoration of ditched wetlands, it is highly important  to accurately locate the ditches on digital maps.
+However, landscape-scale field inventories are often challenging and expensive.
+Automatic detection of ditches from high-resolution digital elevation models (DEMs) and terrain indices provides a feasible means for spatially explicit assessment of ditches across a landscape.
+Previous researches  \citep{uppsala} documented  that terrain indices such as Sky View Factor \citep{zaksek}, Impoundment Index \citep{whiteboxtools} and High Pass Median Filter \citep{whiteboxtools} can efficiently  detect the ditches.
+However, different false positives and negatives may be observed when examining each terrain index manually.
+Such inaccuracies and complexities are in part related to the Light Detection and Ranging (LiDAR) measurements, the mainstay of high resolution DEM and terrain mapping, sometimes being interfered by trees, bushes, or grass - leaving gaps in the ditch model.
+The inaccuracies can also arise from a poorly set threshold for the LiDAR derived terrain parameters.
+Thus, it would be useful to combine the terrain indices derived from high resolution LiDAR data to prune the weaknesses and amplify the strengths of each index and produce highly accurate detection of ditch networks.
+
+In this study, we first investigate the need for better maps of ditch networks, then we introduce a new method for the automated mapping of ditches by integrating multiple high resolution terrain indices (derived from LiDAR DEM) using a machine learning algorithm.
+We also examine which digital terrain indices are the most relevant for an accurate prediction of ditches.
+Several input variables for the model are calculated from the indices to place emphasis on the clear linear directional characteristics of ditches.
+These characteristics are also used when processing the model's output, in order to remove noise and other inconsistencies in the predictions.
+We use a supervised learning algorithm with labelled ditch data to automatically extract ditches from instances not before seen by the model.
+These instances are used as input variables to the algorithm, and consist of refined or raw digital terrain indices derived from a DEM.
+
+We hypothesise that by combining the information from all the digital terrain indices using a machine learner (Random Forests), we can improve the detection of the ditches, compared to if a single index would be used.
+To test this, we conduct a landscape-scale ditch detection analysis at the Krycklan Catchment in Sweden \citep{krycklancatchment}, where a high precision manual mapping has been performed previously.
+This manual map guided the model training and ground-truthing in our analysis and provided a test-bench for our methodology.
+Previous studies \citep{roelens, bailly, rapinel, kiss} that have located ditches from LiDAR data have worked on smaller areas (smaller than 2 $km^2$), predominantly characterised by artificial drainage.
+However, we herein conduct a landscape-scale analysis in a mesoscale catchment.
+The study  catchment  is 68 $km^2$, representing diverse landscape features and geomorphological characteristics.
+
+\section{Materials and method}
+\label{method}
+
+\subsection{The need for better small scale channel mappings}
+To investigate the need for better maps of ditches and small stream channels we compared the data from a national field survey with available maps.
+The National Inventory of Landscapes in Sweden (NILS) has recorded small watercourses (width $<$ 6 m) by conducting a line inventory in 631 $5*5$ km squares randomly distributed in Sweden, to best capture a representative and varied sample of the Swedish landscape.
+The positions of the small watercourses from this inventory were compared with the watercourses on Swedish Property Map (1: 12 500) (allowing for an uncertainty in the positioning of up to 15 m, by snapping the point observations to the lines on the map) (\hyperref[fig:watercoursebarplot]{Figure} \ref{fig:watercoursebarplot}).
+
+\subsection{Study area}
+In the Krycklan Study Catchment the quaternary deposits are dominated by till (51\%) and sorted sediments (30\%).
+The catchment ranges in elevation from 114 to 405 m above sea level.
+Forest covers 87\% of the catchment, mires 9\%, thin soils 7\%, and rock outcrops 1\%.
+The land use is dominated by forestry; approximately 25\% of the Krycklan catchment has been protected since 1922, but  the rest of the area is mostly second growth forest \citep{krycklancatchment}.
+
+A historical survey of the catchment (Gudrun Norstedt, unpublished) recorded  that the ditches in the catchment were dug between 1900 and 1993.
+Initially, the ditches were dug to drain the mires for hay production.
+Later on, wet areas were drained to attempt to turn them into productive forest lands \citep{paivanen}.
+However, research later showed that many of the drained mires were not productive for forestry due to lack of nutrients \citep{sikstrom}.
+
+\subsection{Digitising the ground truth}
+A high resolution DEM was generated from LiDAR data, see \hyperref[lidartodem]{'Data and codes availability statement'}.
+To train and validate the machine learning models, a vector layer of all ditches in the Krycklan Catchment \citep{krycklancatchment} was digitised.
+The study catchment comprises several types of ditches including road ditches, forest ditches, and agricultural ditches.
+A multidirectional hillshade of the 0.5 m DEM was calculated in ArcGIS Pro to locate the ditches.
+From this layer, most of the ditches could be easily identified.
+We conducted the digitisation at a scale of 1:600 to ensure all the digitised ditches properly align the actual ditch on the ground and do not coincide with the edge of the banks.
+In addition, both current and historical records and orthophotos dating back to the 1960's were consulted to help locate all the ditches.
+In uncertain areas, the ditches were verified in the field.
+The stream network in \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan} shows a network with a 8 hectare flow initiation threshold.
+The digitisation of the ditch networks showed that there are 107 km of road ditches and 208 km of forest/agricultural ditches in the catchment (\hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan}).
+The stream network in \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan}, which represents median flow conditions, was 197 km.
+
+\begin{figure}[!htb]
+    \centering
+    \includegraphics[width=1\linewidth]{./images/Krycklan_lo.jpg}
+    \caption{Map of Krycklan showing ditches and stream channels, as well as our division into 21 subsections.
+The 10 subsections with a white border were used for developing input variables and optimising parameters before the experiment.
+The 11 subsections with a black border were used in the experiment for evaluation with 11-fold cross validation.
+Each rectangular subsection represents $2997 \cdot 2620$ pixels (roughly 196 hectare).}
+    \label{fig:swedenkrycklan}
+\end{figure}
+
+From the ditch mapping, the vector layer was rasterised so that it could be compared to the automatically derived ditches from the $0.5*0.5$ m DEM.
+Our field observations have shown that the  width of most ditches within the catchment ranges between 0.5 and 3.5 m.
+To ensure that the model received all ditch pixels correctly in the training phase, we widened the ditches so that all pixels within a radius of three pixels (1.5 metres) of the vectors were labelled as ditch pixels.
+\hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{a} shows the ditches rasterised from vectors and \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{b} shows the ditches after widening.
+The data in \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{b} is the labelled data that was used to train the machine learning models.
+The ditches all have a generalised width based on a perceived average of ditch width in the area.
+Due to all ditches varying in width, it was impossible to produce a perfect representation of each ditch, but this widening helped ensure that almost all ditch pixels were covered by the labels.
+
+To match the labels with out predictions (where we downscale the predictions into grids to remove outliers), the ground-truth was also downscaled into grids of $6*6$ pixels (9 $m^2$) each.
+Each grid cell that contained at least 25\% ditch pixels was labelled as a ditch, see \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{c}.
+This layer was used to ground-truth the machine learning models.
+
+\begin{figure} [htb!]
+    \centering
+    a{
+        \resizebox*{3.5cm}{!}{\includegraphics{./images/publ_ditch_preprocess_A_lo.jpg}}}\hspace{5pt}
+    b{
+        \resizebox*{3.5cm}{!}{\includegraphics{./images/publ_ditch_preprocess_B_lo.jpg}}}
+    c{
+        \resizebox*{3.5cm}{!}{\includegraphics{./images/publ_ditch_preprocess_C_lo.jpg}}}
+    \caption{Processing of ditch labels.
+\textbf{a: }Rasterised ditches with a width of one pixel.
+\textbf{b: }Ditches after a widening process, seven pixels (3.5 metres) wide.
+Used as input when training the models.
+\textbf{c: }Ditches after a grid zone conversion.
+Used for evaluating the results from the ditch detector.} \label{sample-figure}
+    \label{fig:ditchpreprocess}
+\end{figure}
+
+\subsection{Extracting ditches with digital terrain indices}
+From the DEM, several digital terrain indices were calculated. The strengths and weaknesses of these indices have been obtained from observing their performance in different topological environments:
+\begin{itemize} \label{data_attributes}
+
+  \item \textbf{Sky View Factor} \newline
+    The Sky View Factor represents how much of the sky that is visible from a certain point on the ground \citep{zaksek}.
+A point in a ditch would yield a low value, as more of the sky is obscured by the ditch bank.
+However, it tends to produce false positives in steep terrain.
+The Sky View Factor was calculated in SAGA GIS, using a search radius of 10 metres to ensure that it was large enough to capture the ditches, but small enough to not include large scale topographical features such as hills.
+
+  \item \textbf{Impoundment Index}\newline
+    The Impoundment Index in Whitebox Tools \citep{whiteboxtools} can be used as a measure of flow constriction, which is useful for mapping channels, such as ditches.
+When tuned to output the flooded volume, it effectively maps the area and average depth of the reservoir that would be created by placing a dam of a user-specified length at each grid cell in the DEM.
+Here, we used dam height calculated with a dam length of 3 metres to cover the entire ditches.
+Typical false positives with this method are natural stream channels with a clear incised channel.
+
+  \item \textbf{High Pass Median Filter} \newline
+    High Pass Median Filters (HPMF) highlight ditches by indicating local depressions (such as ditches) in the DEM by subtracting the value at the grid cell at the centre of the window from the median value in the window.
+Typical false positives with this index are small hollows.
+HPMF was calculated using Whitebox Tools \citep{whiteboxtools} with a window size of $4.5 * 4.5$ m.
+
+    \item \textbf{Slope} \newline
+    The Slope index represents the degree of slope at a certain point, with a value ranging from $0^{\circ}$ to $90^{\circ}$.
+It was calculated on the 0.5 m grid, using ArcGIS \citep{EsriArcGisBook}.
+This index contains no information about the direction of the slope.
+\end{itemize}
+
+\subsubsection{Thresholding the digital terrain indices}
+To benchmark our method against a non-machine learning method, we binarised the original four digital terrain indices using thresholds.
+Different threshold values were tested and the value that yielded the highest Cohen's Kappa score for each respective index was chosen.
+This test was conducted for the 11 subsections described in \ref{trainingvalidationdatasets}.
+The Sky View Factor index was classified to only include values below 0.955, the Impoundment Index Dam Height using values above 0.27 $m$, the HPMF index using values below -0.18 m, and the Slope index using values above $14 ^{\circ}$ (Table 2).
+
+\subsection{Data preparation}
+
+\subsubsection{Processing the digital terrain indices}
+We examined how different  input variables affected the prediction of the machine learning models.
+Several possible data manipulation methods could theoretically produce a better prediction.
+To improve the prediction, steps were taken where neighbouring pixels were included to give a representation of the area surrounding a specific pixel, similar to the approach of \citet{roelens}.
+
+The digital terrain indices provided a satisfactory foundation for the models, but lacked in the generalisability of their predictions.
+More diverse input variables were extracted using simple statistical aggregates such as mean, median, min, max, and standard deviation.
+Using these statistical aggregations with the help of neighbouring areas around pixels aided in pruning pixels with outlier values, often smoothing out the data to represent ditches more accurately on a per-pixel basis.
+These variables were calculated by gathering all data points in different circular radii around the studied pixel, before computing one of the statistical aggregations (\hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{b}, \hyperref[fig:features]{c}, \hyperref[fig:features]{f}, and \hyperref[fig:features]{g}).
+
+In addition to the general input variables, a number of custom input variables were also developed.
+Some of these custom variables were produced by statistical aggregation of multiple indices, whereas in some cases, they were generated by using thresholds to exclude, lower, or amplify the values of pixels that indicated ditches or non-ditches (\hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{d} and \hyperref[fig:features]{h}).
+
+Both HPMF and Sky View Factor were used with the image processing \textit{gabor filter}, which can be used to detect lines of a certain orientation in an image \citep{gabor}.
+30 Gabor filters which were rotated in different angles and with different frequencies were combined to detect lines, amplifying ditches by utilising the fact that ditches have a linear elongated shape (\hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{e}).
+
+\label{impoundmentstreamremoval}
+The raw Impoundment Index dam height was used to create a mask, attempting to mask out streams, but retain ditches.
+Only areas with a relatively high value, which would indicate that these areas most likely contained streams, were masked out.
+After widening the resulting area, this mask was used to remove streams from all the aforementioned custom input variables, generating one new input variable from each.
+The features marked with \textit{streams removed} in \hyperref[featuretable]{Table} \ref{featuretable} make use of this filter (\hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{i}).
+
+Overall, 81 input variables were extracted from the terrain indices.
+In a pilot study, we compared several different algorithms (Random Forests, Extreme Gradient Boosting, Naive Bayes, and Support Vector Machines), and it was found that Random Forests produced the most accurate results, here we therefore only present the results from the Random Forest model.
+We then conducted a sub-experiment to find the best input variables, as well as the optimal number of variables to use.
+The variable importance score in Random Forests was used to select the most important input variables and it was found that using the top 40 input variables produced the best results.
+
+\begin{figure} [!htb]
+    \centering
+    a{
+        \resizebox*{3.5cm}{!}{\includegraphics{./images/feature_a_lo.jpg}}}
+    b{
+        \resizebox*{3.5cm}{!}{\includegraphics{./images/feature_b_lo.jpg}}}
+    c{
+        \resizebox*{3.5cm}{!}{\includegraphics{./images/feature_c_lo.jpg}}}
+    \newline d{
+        \resizebox*{3.5cm}{!}{\includegraphics{./images/feature_d_lo.jpg}}}
+    e{
+        \resizebox*{3.5cm}{!}{\includegraphics{./images/feature_e_lo.jpg}}}
+    f{
+        \resizebox*{3.5cm}{!}{\includegraphics{./images/feature_f_lo.jpg}}}
+    \newline g{
+        \resizebox*{3.5cm}{!}{\includegraphics{./images/feature_g_lo.jpg}}}
+    h{
+        \resizebox*{3.5cm}{!}{\includegraphics{./images/feature_h_lo.jpg}}}
+    i{
+        \resizebox*{3.5cm}{!}{\includegraphics{./images/feature_i_lo.jpg}}}
+    \caption{Example of 8 of the 40 input variables, in addition to ditch labels used by the models for a small sample area.
+The radii represents pixels with a 0.5 m resolution.
+\newline \textbf{a:} Labelled ditches, \textbf{b:} Slope standard deviation, radius 6, \textbf{c:} HPMF mean, radius 4, \textbf{d:} HPMF ditch amplification, \textbf{e:} Sky View Factor Gabor, \textbf{f:} Sky View Factor max, radius 6, \textbf{g:} Impoundment mean, radius 3, \textbf{h:} Impoundment ditch amplification, \textbf{i:} Impoundment ditch amplification - streams removed}
+    \label{fig:features}
+\end{figure}
+
+\subsubsection{Training and validation datasets}
+\label{trainingvalidationdatasets}
+To develop and evaluate our ditch detector, the raster and ditch label data of Krycklan were manually divided into 21 subsections.
+Each subsection represents an area of roughly 196 hectare.
+From this division, 11 of the subsections were put aside as hold-out data, only for use in the experiment to evaluate the performance of the predictions.
+The remaining 10 subsections were used solely before the experiment to develop the ditch detector to test different ways of preparing the data.
+This allowed the ditch detector to be evaluated on  independent data and  strengthened the validity of the experiment.
+\hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan} shows which subsections were used for development and evaluation respectively.
+
+Using the 11 subsections in the hold-out data for the final experiment, a process called \textit{k-fold cross validation} was employed (11-fold cross validation in our case).
+K-fold cross validation is a method where you divide your dataset into folds of similar size and train a model on all but one of your folds (subsections).
+You then use that subsection to evaluate the results \citep{crossvalidation}.
+A new model is trained using the training folds for each iteration (\hyperref[fig:crossvalidation]{Figure} \ref{fig:crossvalidation}).
+Using this technique, shifting which subsection to leave out from the training, allowed us to train 11 different Random Forests classifying models with a large amount of data from the remaining 10 subsections in the hold-out data, producing 11 independent sub-experiments to evaluate the method on.
+
+\begin{figure}[!htb]
+    \centering
+    \includegraphics[width=1\linewidth]{images/cross_validation_lo.jpg}
+    \caption{11-fold cross validation. Each subsection in the hold-out data was used once for testing, with the remaining subsections being used to build a new model for each iteration.}
+    \label{fig:crossvalidation}
+\end{figure}
+
+\subsection{Developing the Random Forests model}
+
+Random Forests is an ensemble machine learning method which builds multiple diverse decision trees to increase the robustness of its predictions \citep{breiman,flach}.
+The individual trees are trained by minimising the entropy of the data from each node’s output \citep{kotsiantis}.
+A byproduct of the training is the Gini importance, which denotes the most important input variables \citep{gini}.
+For our study, a hyperparameter tuning showed that using 300 trees and \textit{gini} as the splitting criterion, with a minimum of 10 samples per node split, and no artificial class weight or max depth of trees yielded the best predictions.
+The Random Forests algorithm from the Python library \textit{scikit-learn} \citep{scikit-learn} was used in the experiment, and the classifiers were trained on all the input variables seen in \hyperref[featuretable]{Table} \ref{featuretable}.
+
+\begin{table} [!htb]
+\centering
+    %\tbl{The 40 input variables used when training the models.}
+    {\begin{tabular}{l|l} %\toprule
+    %\hline
+      \textbf{Variable/Algorithm}\textsuperscript{a} & \textbf{Circular radii}\textsuperscript{b} \\ %\midrule
+      \hline
+      Sky View Factor median &2, 6 \\
+      Sky View Factor standard deviation & 6 \\
+      Sky View Factor min & 6 \\
+      Sky View Factor max & 2, 4, 6 \\
+      Sky View Factor non ditch amplification & - \\ 
+      Sky View Factor Gabor & - \\
+      Sky View Factor Gabor - streams removed & -\\
+
+      Impoundment mean & 2, 3, 4, 6 \\
+      Impoundment median & 2, 4, 6 \\
+      Impoundment standard deviation & 4, 6 \\
+      Impoundment max & 6 \\
+      Impoundment ditch amplification & - \\
+      Impoundment ditch amplification - streams removed & - \\
+
+      HPMF mean & 3, 4, 6 \\
+      HPMF median & 4 \\
+      HPMF standard deviation & 6 \\
+      HPMF min & 2, 4 \\
+      HPMF ditch amplification & - \\
+      HPMF ditch amplification - streams removed & - \\
+      HPMF Gabor - streams removed & -\\
+
+      Slope mean & 6 \\
+      Slope median & 6 \\
+      Slope standard deviation & 4, 6 \\
+      Slope min & 2, 4, 6 \\
+      Slope non-ditch amplification & - \\ %\bottomrule 
+      \hline
+    \end{tabular}}
+    \caption{ The 40 input variables used when training the models.
+    \newline \textsuperscript{a} Displays algorithm used to produce the variable. \newline
+        \textsuperscript{b} Represents the radius of the circular mask (if one was used) to determine \newline what neighbouring pixels to use in an algorithm.
+The radii represents pixels \newline \mbox{with a 0.5 m resolution.\itshape\ignorespaces}}
+    \label{featuretable}
+\end{table}
+
+The testing phase showed that the classifiers produced poor results when the ratio of ditch versus non-ditch pixels in the training data was very high, as the models were not being punished for mislabelling ditches as non-ditches.
+To increase the recall, we evened out the amount of ditch and non-ditch instances in the training data.
+First, we extracted all pixels labelled as ditches as well as pixels within close proximity of ditches.
+Secondly, pixels were sampled randomly from the entire subsection to still contain most of the geographical features of each area (\hyperref[fig:balancedmasks]{Figure} \ref{fig:balancedmasks}).
+
+\begin{figure} [!htb]
+    \centering
+    a{
+        \resizebox*{5.6cm}{!}{\includegraphics{./images/publ_balanced_masks_A_lo.jpg}}}\hspace{5pt}
+    b{
+        \resizebox*{5.6cm}{!}{\includegraphics{./images/publ_balanced_masks_B_lo.jpg}}}
+    \caption{Black pixels here indicate the balanced masks used to determine what pixels were used when training the models. \textbf{a, b: }Examples of two subsection from the training dataset.}
+    \label{fig:balancedmasks}
+\end{figure}
+
+\subsection{Post-processing}
+The models output a ditch class probability prediction for each pixel.
+Values close to zero indicate a low probability of a pixel lying in a ditch, whereas values close to one equals a high ditch probability (\hyperref[fig:postprocessing1]{Figure} \ref{fig:postprocessing1} \hyperref[fig:postprocessing1]{a}).
+
+\subsubsection{Noise reduction and gap filling}
+
+The probability predictions contained a lot of noise in the areas largely distant from the ditches.
+This noise was removed by first using a bilateral de-noising filter, and then removing outlier pixels that were far away from any other high probability ditch pixels.
+This left linear properties and pixels with a very high value intact, while lowering the value of pixels that did not contribute to an accurate prediction (\hyperref[fig:postprocessing1]{Figure} \ref{fig:postprocessing1} \hyperref[fig:postprocessing1]{b}) (\hyperref[fig:postprocessing2]{Figure} \ref{fig:postprocessing2} \hyperref[fig:postprocessing2]{a}).
+
+To fill gaps in ditches that the models failed to correctly predict, we aggregated pixel values by covering them with cone masks expanding outwards in different directions from the examined pixel.
+This amplified some of the noise that was left, but filling the gaps in the ditches was judged to be more important (\hyperref[fig:postprocessing2]{Figure} \ref{fig:postprocessing2} \hyperref[fig:postprocessing2]{b}).
+
+\begin{figure} [!htb]
+    \centering
+    a{
+        \resizebox*{5.6cm}{!}{\includegraphics{./images/publ_post_process_step_1_lo.jpg}}}\hspace{5pt}
+    b{
+        \resizebox*{5.6cm}{!}{\includegraphics{./images/publ_post_process_step_2_lo.jpg}}}
+    \caption{Step one and two of the post-process. Darker pixels indicate a higher ditch probability. \textbf{a: }Random Forests probability prediction. \textbf{b: }Bilateral de-noising.}
+    \label{fig:postprocessing1}
+\end{figure}
+
+\begin{figure} [!htb]
+    \centering
+    a{
+        \resizebox*{5.6cm}{!}{\includegraphics{./images/publ_post_process_step_3_lo.jpg}}}\hspace{5pt}
+    b{
+        \resizebox*{5.6cm}{!}{\includegraphics{./images/publ_post_process_step_4_lo.jpg}}}
+    \caption{Step three and four of the post-process. Darker pixels indicate a higher ditch probability. \textbf{a: }Custom de-noising. \textbf{b: }Gap filling.}
+    \label{fig:postprocessing2}
+\end{figure}
+
+\subsubsection{Binarisation with grid zones}
+The models' ditch prediction is given in the original resolution raster format.
+To allow for comparison to the evaluation data, the same grid conversion was performed on the prediction raster as on the evaluation data displayed in \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{c}.
+After testing different combinations of zone sizes and probability values, a mean probability rating was calculated for each $6*6$ grid zone, classifying the entire zone as part of a ditch if the mean probability exceeded 40 \%.
+This zone partitioning helped to fill in gaps where lone pixels in ditches had been incorrectly classified.
+(\hyperref[fig:postprocessing3]{Figure} \ref{fig:postprocessing3} \hyperref[fig:postprocessing3]{a}).
+
+\subsubsection{Cluster removal}
+To remove noise from the binary prediction, a custom cluster detection algorithm was used.
+By finding the number of connected pixels with a true value and removing those whose cluster size were below a given threshold, minor noise in the prediction could be removed while still retaining most of the ditch pixels.
+Actual ditches that have a low probability and therefore create small clusters may still be excluded by this method, but the noise removal advantages were judged to outweigh the loss in recall.
+A distance calculation was also performed in tandem with this method to find the largest distance of pixels inside each given cluster.
+This helped to remove sinks and hollows that were not removed by the initial small cluster removal, but that did not have a linear directional characteristic, indicating that they did not represent a ditch (\hyperref[fig:postprocessing3]{Figure} \ref{fig:postprocessing3} \hyperref[fig:postprocessing3]{b}).
+
+\begin{figure} [!htb]
+    \centering
+    a{
+        \resizebox*{5.6cm}{!}{\includegraphics{./images/publ_post_process_step_5_lo.jpg}}}\hspace{5pt}
+    b{
+        \resizebox*{5.6cm}{!}{\includegraphics{./images/publ_post_process_step_6_lo.jpg}}}
+    \caption{Step five and six of the post-process. Black pixels indicate a ditch prediction. \textbf{a: }Grid zone binarisation \textbf{b: }Cluster removal.}
+    \label{fig:postprocessing3}
+\end{figure}
+
+\subsection{Evaluation} \label{evaluation}
+
+Using only accuracy as an evaluation metric when dealing with an imbalanced dataset (roughly 98 \% of all pixels are non-ditch) would produce a poor performance assessment \citep{balanced}; by simply classifying all pixels as non-ditches, we would by default attain 98 \% accuracy.
+For this reason, the results were mainly evaluated using Cohen's Kappa (Cohen's $\kappa$) index, and the Area under Precision-Recall curve.
+Cohen's $\kappa$ index measures how much better a prediction is compared to a completely random prediction, where random would yield a value of zero \citep{kappa123}.
+With our data, a $\kappa$ value close to zero would be attained by predicting 2 \% of the occurrences as ditch pixels completely at random.
+Values above zero are better than random, and values below zero are worse than random.
+
+The random rating $P_c$ of a prediction of $n$ occurrences is calculated with:\footnote{ P = Positive (ditch pixel), N = Negative (non-ditch pixel), T = true, F = false}
+
+$$
+P_c = \frac{\left(\frac{(TP + FN) \cdot (TP + FP)}{n}\right) + \left(\frac{(FN + TN) \cdot (FP + TN)}{n}\right)}{n}
+$$
+
+Cohen's $\kappa$ is then calculated as a value between -1 and 1 with:
+
+$$\kappa = \frac{Accuracy - P_c}{1 - P_c}$$
+
+The Precision-Recall curve and the Area under Precision-Recall curve (AUPRC) are additional metrics that can be used when evaluating datasets with a largely imbalanced class distribution \citep{precision_recall_curve}.
+The Precision-Recall curve has the recall value on the x-axis and the precision value on the y-axis, and the area under the curve that is defined by this point gives the AUPRC value.
+The area under this curve is given as a value between zero and one, where a value closer to one is better.
+The weighting causes the Precision-Recall curve to not place an equal value on true negatives and true positives \citep{precision_recall_curve}.
+For our ditch detection problem, this means that the AUPRC evaluation metric favours accurately classifying ditch pixels over accurately classifying non-ditch pixels.
+
+To circumvent some of the performance evaluation issues arising from the use of pixel classification for ditch objects as well as not having completely accurate labels on a pixel basis (due to uncertainties in the width of the ditches), we modified the evaluation labels to allow for some error in close proximity to ditches in the prediction.
+False negative and false positive grid zones that lay adjacent to the ditch label grid zones were evaluated as true negatives and true positives, as they can be considered a part of correctly located ditch objects (\hyperref[fig:newlabels]{Figure} \ref{fig:newlabels}).
+
+\begin{figure} [!htb]
+    \centering
+    a{
+        \resizebox*{5.6cm}{!}{\includegraphics{./images/re_evaluation_1_lo.jpg}}}\hspace{5pt}
+    b{
+        \resizebox*{5.6cm}{!}{\includegraphics{./images/re_evaluation_2_lo.jpg}}}
+    \caption{Illustration of the modified evaluation labels.
+Green marks true positives, red marks false negatives, and blue marks false positives.
+False positives and false negatives that lay within one grid zone (9 $m^2$) of a ditch label were evaluated as true positives and true negatives.
+\textbf{a:} Original results, \textbf{b:} Modified results.
+}
+    \label{fig:newlabels}
+\end{figure}
+
+\section{Results and analysis}
+In the national survey (NILS), a total of 1103 natural watercourses, 131 straightened watercourses/channels, and 2089 ditches were found in the field.
+This simple investigation highlights two things: Firstly, the ditches make up the majority of the small scale water channels in the Swedish landscape; almost twice as many as the natural streams.
+Secondly, most of the small scale channels are missing on the current maps; only 45 \% of the natural watercourses, 25 \% of the straightened watercourses/channels, and 9 \% of the ditches were mapped (\hyperref[fig:watercoursebarplot]{Figure} \ref{fig:watercoursebarplot}).
+
+\begin{figure}[!htb]
+    \centering
+    \begin{tikzpicture}
+    \begin{axis}[
+        width={0.9\textwidth},
+        height={0.6\textwidth},
+        ylabel={fluorescence},
+        ybar stacked,
+        ymajorgrids=true,
+    	bar width=45pt,
+    	nodes near coords,
+    	every node near coord/.append style={font=\tiny},
+        enlarge y limits={0.15},
+        enlarge x limits=0.15,
+        legend style={at={(0.78,0.96)},
+          anchor=north,legend columns=-1},
+        ylabel={{}},
+        symbolic x coords={Ditches, Straightened  watercourses, Natural  watercourses},
+        xtick=data,
+        ytick={0,500,1000,1500,2000},
+        x tick label style={rotate=0,anchor=north},
+        ]
+    \addplot+[ybar,color=custombeige,fill=custombeige,draw=custombeigeouter,thick] plot coordinates {(Ditches,188) (Straightened  watercourses,33) 
+      (Natural  watercourses,496)};
+    \addplot+[ybar,color=customgrayouter,fill=customgray,draw=customgrayouter,thick] plot coordinates {(Ditches,1901) (Straightened  watercourses,98) 
+      (Natural  watercourses,607)};
+    \legend{\strut Mapped, \strut Unmapped}
+    \end{axis}
+    \end{tikzpicture}
+    \caption{The bars indicate the number of observed channels in the National Inventory of Landscapes in Sweden's (NILS) line inventory of small watercourses (width $<$ 6 m).
+The colours of the bars indicate if they are mapped or not on the Swedish Property map.}
+    \label{fig:watercoursebarplot}
+\end{figure}
+
+\hyperref[recreatedpredictionperformance]{Table} \ref{recreatedpredictionperformance} shows the evaluation metrics from the experiments with the digital terrain indices separately.
+Of the four, the Impoundment Index and the HPMF index performed relatively close to each other and outperformed the other indices for most of the metrics.
+
+\begin{table}[!htb]
+\centering
+    {\begin{tabular}{l|cccc} %\toprule
+        \textbf{Metric} & \textbf{Sky View Factor} & \textbf{Impoundment} & \textbf{HPMF} & \textbf{Slope}\\ %\midrule
+        \hline
+        Accuracy \%     & 83.67 & 97.07 & 97.45 & 82.66 \\
+        Recall \%       & 44.54 & 29.98 & 25.31 & 34.83 \\
+        Precision \%    &{ 4.00}  & 18.51 & 20.10 & { 2.99} \\
+        $\kappa$ rating & 0.048 & 0.215 & 0.211 & 0.029 \\
+        AUPRC & 0.247 & 0.248 & 0.232 & 0.194 \\ %\bottomrule
+        \hline
+    \end{tabular}}
+    \caption{Metrics from the total results of the four digital terrain indices experiments.}
+    \label{recreatedpredictionperformance}
+\end{table}
+
+\hyperref[predictionperformance]{Table} \ref{predictionperformance} displays all the evaluation metrics for the prediction of our method.
+The confidence intervals were calculated from the 11 different subsections that the experiment was performed on.
+Because the subsections have varying amounts of ditches in them, the confidence intervals will not be completely accurate, but will produce a close estimation.
+
+From \hyperref[predictionperformance]{Table} \ref{predictionperformance}, it is observed that averaging the results from the 11 subsections would yield a very similar value to the total results, indicating that the ditch detector generally performs equally well on subsections with a small amount of ditches as on subsections with a large amount of ditches.
+
+\begin{table}[!htb]
+\centering
+    {\begin{tabular}{l|ccc}
+        \textbf{Metric} & \textbf{Total}\textsuperscript{a} & \textbf{Subsection Average}\textsuperscript{b}& \textbf{CI 95\%}\textsuperscript{c} \\ 
+        %&&\textbf{Average}&\\
+        \hline
+        Accuracy     \% & 99.00 & 99.00 & [98.69 , 99.32] \\
+        Recall       \% & 70.28 & 70.19 & [61.28 , 79.09] \\
+        Precision    \% & 77.38 & 75.79 & [71.94 , 79.64] \\
+        $\kappa$ rating & 0.732 & 0.718 & [0.655 , 0.781] \\
+        AUPRC           & 0.741 & 0.733 & [0.674 , 0.791] \\
+        \hline
+    \end{tabular}}
+    \caption{Metrics for the prediction performance of our ditch detector. \newline
+    \textsuperscript{a} The result of all 11 subsection experiments when combined. \newline
+    \textsuperscript{b} An average score from the 11 different subsections that the experiment was performed on. \newline
+    \textsuperscript{c} Confidence intervals at 95 \% confidence level.}
+    \label{predictionperformance}
+\end{table}
+
+The $\kappa$ rating for our method can be seen in the \textit{Total} column in \hyperref[predictionperformance]{Table} \ref{predictionperformance} ($\kappa$ = 0.732).
+The $\kappa$ ratings from the four digital terrain indices experiments can be seen in  \hyperref[recreatedpredictionperformance]{Table} \ref{recreatedpredictionperformance} ($\kappa$ = 0.048, $\kappa$ = 0.215, $\kappa$ = 0.211, $\kappa$ = 0.029).
+Because the $\kappa$ rating from our method outperforms all digital terrain indices, our hypothesis that combining digital terrain indices with a machine learning produces better ditch detection than using single terrain indices has been confirmed by the experiments.
+Most false positives lie in stream channels (\hyperref[fig:resultsillustrations]{Figure} \ref{fig:resultsillustrations} \hyperref[fig:resultsillustrations]{a, b}) and most false negatives occurred either due to ditches being too shallow (\hyperref[fig:resultsillustrations]{Figure} \ref{fig:resultsillustrations} \hyperref[fig:resultsillustrations]{c, d}), or due to the attempt at removing streams using the input variables as explained in \ref{impoundmentstreamremoval} removing the deepest ditches  (\hyperref[fig:resultsillustrations]{Figure} \ref{fig:resultsillustrations} \hyperref[fig:resultsillustrations]{e, f}).
+
+\begin{figure} [!htb]
+    a{
+        \resizebox*{4cm}{!}{\includegraphics{./images/results_illustration_1_lo.jpg}}}\hspace{5pt}
+    b{
+        \resizebox*{4cm}{!}{\includegraphics{./images/results_illustration_2_lo.jpg}}}
+    \newline c{
+        \resizebox*{4cm}{!}{\includegraphics{./images/results_illustration_3_lo.jpg}}}\hspace{5pt}
+    d{
+        \resizebox*{4cm}{!}{\includegraphics{./images/results_illustration_4_lo.jpg}}}
+    \newline e{
+        \resizebox*{4cm}{!}{\includegraphics{./images/results_illustration_5_lo.jpg}}}\hspace{5pt}
+    f{
+        \resizebox*{4cm}{!}{\includegraphics{./images/results_illustration_6_lo.jpg}}}
+    \caption{The left panel shows the hillshade for a number of subsections, and the right panel shows the hillshade with the extracted ditches superimposed on top.
+Green marks correctly classified ditches (true positives), red marks missed ditches (false negatives), blue marks incorrectly classified ditches (false positives), and transparent areas mark correctly classified non-ditches (true negatives).
+\textbf{a} and \textbf{b} illustrate that the most common false positives were natural stream channels being classified as ditches.
+\textbf{c} and \textbf{d} illustrate that very shallow ditches that were barely visible in the DEM were not captured with our ditch detector.
+\textbf{e} and \textbf{f} illustrate that some of the deepest and more prominent ditches were also missed.
+This occurred as a result of attempting to get rid of more false positives in the form of streams by using the input variables, as explained in \ref{impoundmentstreamremoval}.}
+    \label{fig:resultsillustrations}
+\end{figure}
+
+In \hyperref[featureimportancetable]{Table} \ref{featureimportancetable}, the top 20 (out of 40) input variables from the Random Forests models are presented with their importance percentages for the models making a successful prediction.
+Input variables based on HPMF and Impoundment Index contributed the most to a successful prediction.
+The Gini importance was obtained using the Gini impurity for each variable \citep{gini}.
+
+\begin{table} [!htb]
+        {\begin{tabular}{r|ll}
+          Pos & Variable\textsuperscript{a} & Importance (\%) \\
+          \hline
+          1.  & Impoundment mean 3                                  & 7.33\\
+          2.  & HPMF mean 4                                         & 6.23\\
+          3.  & Impoundment mean 4                                  & 5.57\\
+          4.  & Impoundment mean 2                                  & 5.41\\
+          5.  & HPMF mean 3                                         & 4.33\\
+          6.  & Impoundment median 4                                & 3.84\\
+          7.  & HPMF Gabor - streams removed                        & 3.54\\
+          8.  & HPMF median 4                                       & 3.40\\
+          9.  & Impoundment median 2                                & 3.01\\
+          10. & Sky View Factor Gabor - streams removed             & 2.84\\
+          11. & HPMF mean 6                                         & 2.73\\
+          12. & Impoundment median 6                                & 2.69\\
+          13. & Impoundment standard deviation 4                    & 2.52\\
+          14. & Sky View Factor Gabor                               & 2.44\\
+          15. & Impoundment ditch amplification                     & 2.43\\
+          16. & Impoundment mean 6                                  & 2.27\\
+          17. & Impoundment ditch amplification - streams removed   & 2.24\\
+          18. & HPMF min 2                                          & 2.23\\
+          19. & Slope standard deviation 6                          & 2.01\\
+          20. & Sky View Factor non-ditch amplification             & 1.88\\
+          \hline
+        \end{tabular}}
+        \caption{The top 20 input variables (out of 40) by importance when the models make a prediction.
+\newline \textsuperscript{a} The number next to some of the variables indicates the circular radius used to select what neighbouring pixels to use in the statistical aggregation method.
+The radii represents pixels with a 0.5 m resolution.}
+    \label{featureimportancetable}
+\end{table}
+
+\section{Discussion}
+
+The comparison with the field observed channels and the available maps clearly indicate the need for better maps of ditches, as only 9\% of ditches are mapped.
+(\hyperref[fig:watercoursebarplot]{Figure} \ref{fig:watercoursebarplot}).
+Our study has shown that it is possible to locate ditches automatically in high-resolution DEMs ($0.5  * 0.5 $ metres, in our case), and that more of the ditches can be detected if the information from several terrain indices (\hyperref[recreatedpredictionperformance]{Table} \ref{recreatedpredictionperformance}) is combined through machine learning than if indices are used separately (\hyperref[predictionperformance]{Table} \ref{predictionperformance}).
+The ditch detection performance varied depending on the type of ditches, this is illustrated in \hyperref[fig:ditchpictures]{Figure} \ref{fig:ditchpictures}.
+For example, the retrieval rate was slightly higher for the road ditches compared to forest or agricultural ditches.
+This is likely due to the fact that  road ditches are found in open areas alongside roads, and also,   they are generally well maintained for optimum functioning (\hyperref[fig:ditchpictures]{Figure} \ref{fig:ditchpictures} \hyperref[fig:ditchpictures]{a}).
+However, most of the ditches in Krycklan today are found below the canopy (e.g.
+\hyperref[fig:ditchpictures]{Figure} \ref{fig:ditchpictures} \hyperref[fig:ditchpictures]{c}), and the majority have not been maintained, causing difficulties in detecting them as some have grown back in with peat, grasses, or shrubs (\hyperref[fig:resultstreesbushes]{Figure} \ref{fig:resultstreesbushes} \hyperref[fig:resultstreesbushes]{b}).
+This, in combination with the surrounding peat having subsided \citep{heikurainen}, have rendered them almost impossible to detect in the DEM today (\hyperref[fig:ditchpictures]{Figure} \ref{fig:ditchpictures} \hyperref[fig:ditchpictures]{d}).
+Despite these issues we had a recall rate of 70.28\%, showing that the majority of the ditches could still be retrieved automatically.
+
+\begin{figure} [!htb]
+    a{
+        \resizebox*{5.5cm}{!}{\includegraphics{./images/road_ditch_lo.jpg}}}\hspace{7pt}
+    b{
+        \resizebox*{5.5cm}{!}{\includegraphics{./images/mire_ditch_lo.jpg}}}
+    \newline c{
+        \resizebox*{5.5cm}{!}{\includegraphics{./images/forest_ditch_lo.jpg}}}\hspace{7pt}
+    d{
+        \resizebox*{5.5cm}{!}{\includegraphics{./images/overgrown_ditch_lo.jpg}}}
+    \caption{All photos show ditches in the study catchment.
+They illustrate the difficulty of capturing the ditches based on their form (i.e.
+an elongated, narrow, hollow in the ground).
+\textbf{a: }A roadside ditch.
+These are generally easily detected as they are usually found in open areas along the roads, and have often been maintained.
+\textbf{b: }A ditch in a mire; this is still easily detected, as there is no canopy and there is a clear elevation difference between the ditch and surrounding area.
+\textbf{c: }A narrow, more or less  overgrown ditch in a dense forest stand.
+\textbf{d: }A ditch in a mire that has more or less grown back in with \textit{Sphagnum sp.} and \textit{Carex}, likely also in combination with subsiding surrounding soils.
+Such ditches can be found in the field as the vegetation differs, but as there is no longer an elevation difference between the ditch and the surrounding soils it will not be detected in the DEM.
+Photos a-c: E.
+Maher Hasselquist  Photo d: G.
+Norstedt}
+    \label{fig:ditchpictures}
+\end{figure}
+
+Hypothetically, it should be quite easy to detect ditches using LiDAR data.
+In practice, however, there are many variables affecting the results, such as the point density and the interpolation method used to create the DEM.
+\citet{rapinel} found that the results were usually more sensitive to the point density (ranging 1-4 points $m^{2}$ in their study) than the interpolation method.
+They retrieved 54.8 and 63.8 \% of the ditches on the Acuey and Boucey marches in France, compared to 70.28 \% in our study.
+However, in our study we had a point density of 20 points $m^{2}$, so the quality of the DEM in our study ought to be robust, which may have contributed to our higher retrieval rate.
+\hyperref[fig:resultstreesbushes]{Figure} \ref{fig:resultstreesbushes} \hyperref[fig:resultstreesbushes]{(a, b)} illustrates the difference in the quality of the LiDAR data where the laser pulses have caught in trees or shrubs.
+When too few LiDAR points reach the ground, the DEM generation is forced to interpolate over a larger area, leading to inaccuracies in the ground DEM.
+
+\begin{figure} [!htb]
+    a{
+        \resizebox*{5.5cm}{!}{\includegraphics{./images/results_trees_bushes_1_lo.jpg}}}\hspace{7pt}
+    b{
+        \resizebox*{5.5cm}{!}{\includegraphics{./images/results_trees_bushes_2_lo.jpg}}}
+    \newline c{
+        \resizebox*{5.5cm}{!}{\includegraphics{./images/results_trees_bushes_3_lo.jpg}}}\hspace{7pt}
+    d{
+        \resizebox*{5.5cm}{!}{\includegraphics{./images/results_trees_bushes_4_lo.jpg}}}
+    \caption{\textbf{a} and \textbf{b} illustrate that despite a point density of 20 points $m^2$, not all points reach the ground in places with dense vegetation.
+In \textbf{a} there is a clear difference between the smooth surface of the DEM in the open area (black arrow), and in the areas covered by trees and shrubs (red arrows), where the DEM is much coarser due to interpolation with a lower number of LiDAR points here.
+The orthophoto (\textbf{b}), clearly shows that trees and shrubs are growing in the ditches in this area.
+This makes it more difficult to detect ditches.
+However, despite these coarser grid points in the DEM, \textbf{c} and \textbf{d} shows that the ditch detector still managed to capture a majority of the ditches (green), and only failed at certain points (red).}
+    \label{fig:resultstreesbushes}
+\end{figure}
+
+With our method, natural streams were often classified as false positives (\hyperref[fig:resultsillustrations]{Figure} \ref{fig:resultsillustrations} \hyperref[fig:resultsillustrations]{b}) as our aim was focused on locating ditches.
+However, this indicates that our method could also be used to map small, previously unmapped natural stream channels.
+The national survey showed that 55\% of the natural stream channels and 75\% of the straightened water courses were missing on the maps.
+This is in line with other national \citep{kuglerova} and international \citep{benstead} studies highlighting this phenomenon called "Aqua Incognita - the unknown headwaters" \citep{bishop,kuglerova}.
+
+The results from our method (\hyperref[predictionperformance]{Table} \ref{predictionperformance}) show that we managed to attain a Cohen's $\kappa$ rating in the substantial range, according to performance thresholds proposed by \citet{kappaanalysis}.
+The total recall value of 70.28 \% shows that we managed to find a majority of the ditch pixels that exist.
+The results of our method also showed a significant improvement for ditch detection over classifying with the digital terrain indices separately (\hyperref[recreatedpredictionperformance]{Table} \ref{recreatedpredictionperformance}).
+Had we been able to use labels that were correctly labelled on a pixel basis, i.e.
+each ditch had its actual width recorded, the models could have learned with a better accuracy in the training phase, and the classification results would also most likely have benefitted from this.
+
+Sometimes pixels in ditches were initially identified correctly as ditch pixels, but because not enough pixels in the surrounding area were detected, the post-processing algorithms identified them as noise and removed them.
+Incorrectly classified ditch pixels that caused the precision to decrease were generally the pixels that lay in streams (\hyperref[fig:resultsillustrations]{Figure} \ref{fig:resultsillustrations} \hyperref[fig:resultsillustrations]{a, b}).
+However, small sinks or hilly areas incorrectly classified as ditches by the Random Forests models were generally removed in the post-processing.
+We managed to bridge gaps and exclude streams to some extent in our predictions.
+This was in part due to the post-processing of the prediction helping where the models were unsuccessful on their own, and in part due to the input variables developed.
+\hyperref[featureimportancetable]{Table} \ref{featureimportancetable} highlights the importance of the input variables that were developed by applying different statistical aggregations for neighbouring areas of pixels, similar to \citet{roelens}, as well as the custom variables using filters such as the Impoundment stream removal or Gabor filters.
+
+One concern with our method was that valuable information may have been discarded when aggregating the elevations from the original point cloud to a DEM.
+\citet{roelens} used  machine learning to detect ditches directly from point-clouds instead of a DEM.
+However, we achieved a similar score to their method; $\kappa=0.73$ in our study compared to 0.77 for grassland and 0.73 for peri-urban area in \citet{roelens} study.
+This suggests that DEM can also be used to automatically extract ditches, provided that enough input data is available to generate a high resolution DEM for capturing the small scale features of ditches.
+
+\citet{bailly} also showed that the vegetation cover significantly affects the ability to detect ditches from LiDAR data (\hyperref[fig:ditchpictures]{Figure} \ref{fig:ditchpictures} \& \ref{fig:resultstreesbushes}).
+Approximately 75\% of the ditches in their study were retrieved where no vegetation existed, whereas the detection accuracy went below 10\% under vegetation; especially when the ditches were covered by high vegetation, shrubs, and trees \citep{bailly}.
+Hence, a $\kappa$ of 0.73 in our study  with  87\% forest cover \citep{krycklancatchment} is substantially better than that of previous research, and this could perhaps be a result of the higher point density of our LiDAR measurements (20 compared to 10 points $m^{2}$).
+The hydrology of our study catchment also differs from that of other studies, as we have almost as many natural stream channels as ditches, whereas their studies were conducted in smaller, predominantly artificially drained areas \citep{bailly, roelens, rapinel}.
+Because our method locates both ditches and natural stream channels, we have to take additional measures to remove the stream channels (affecting recall negatively), as explained in \ref{impoundmentstreamremoval}.
+
+A future improvement to the ditch detector could be to use a shape index to remove small natural streams, as natural streams often form more complex channels whereas ditches are often straight.
+A potential issue with generalising our ditch detector for use in other geographical areas is that all the post-processing steps and input variables were developed based on occurrences in the Krycklan area.
+Different geographical compositions in other areas may require the adjustment of the thresholds used to fill gaps in ditches and to remove noise.
+
+\section{Conclusion}
+
+Detection of artificial ditches over large areas with substantial natural stream coverage is challenging, especially under tree canopy, but highly desired for effective forest management.
+Ditches represent elongated depressions in a DEM; however, the exact shape and size differs between ditches.
+We combined recently developed digital terrain indices (the most important being Impoundment Index and High Pass Medium Filter) but, unlike previous studies, did not apply generic thresholding on them.
+Instead, we included local variability in the indices by using different radii to capture ditches with varied shapes.
+This integration of a large set of terrain indices using machine learning produced good results, as evidenced by the confidence interval for the Cohen's $\kappa$ index ranging from 0.655 to 0.781 with a confidence level of 95\%.
+With the developed input variables and extensive post-processing, we also managed to remove many stream channels and fill gaps in the ditches where the LiDAR scans were weak.
+
+Although there is room for further improvement, the methodology is a significant step forward as it performed on a par with previous studies conducted in more homogeneous agricultural landscapes.
+Our approach enables accurate ditch detection in complex forested landscapes with tree canopy and many natural streams.
+Future work could introduce more input variables and post-processing steps, such as using pathfinding algorithms to fill gaps in the ditch model or shape indices to remove stream channels from the prediction.
+Moving towards image segmentation instead of pixel classification of tabular pixel values could also be a suitable future approach.
+Our developed methodology can be adjusted to apply to any setting for artificial ditch detection, reducing time and labour investment compared to the conventional manual mapping techniques.
+The accurate ditch maps can significantly contribute to practical forestry and operational land management at a landscape-scale.
 
 \section*{Acknowledgements}
+We wish to thank Dr.
+Siddhartho Paul at the Swedish University of Agricultural Sciences, whose suggestions helped improve and clarify this manuscript.
+We also thank Sigurd Israelsson at the School of Engineering at J\"onk\"oping University for lending hardware resources for the experiments in this study.
+The work has been funded by VINNOVA, EU InterReg Baltic Sea project WAMBAF tools and Formas.
 

### 2021-01-11 LaTeX ESWA/article.tex
Adds more template stuff.

@@
 
-\subsection{Study area}
+\subsection{Study area and data}
 In the Krycklan Study Catchment the quaternary deposits are dominated by till (51\%) and sorted sediments (30\%).
@@
 
+The LiDAR dataset used in this study was produced by TerraTec Sweden AB in August 2015, on demand of the Department of Forest Resource Management at the Swedish University of Agricultural Sciences, and is freely available online \citep{dataset}.
+The average density of the dataset was 20 pulses per $m^2$.
+Ground point classification was performed with lasground (included in LAStools, a software suite produced by rapidlasso GmbH).
+The DEM was created with blast2dem (included in LAStools), with a cell size of $0.5*0.5$ m.
+Only ground points were imported to generate a DEM of the ground.
+
 \subsection{Digitising the ground truth}
-A high resolution DEM was generated from LiDAR data, see \hyperref[lidartodem]{'Data and codes availability statement'}.
 To train and validate the machine learning models, a vector layer of all ditches in the Krycklan Catchment \citep{krycklancatchment} was digitised.
@@
 
-\begin{table} [!htb]
+\begin{table} [H]
 \centering
@@
 
-\begin{table}[!htb]
+\begin{table}[H]
 \centering
@@
 
-\begin{figure} [!htb]
+\begin{figure} [H]
     a{
@@
 \end{table}
+\clearpage
 
@@
 
-\begin{figure} [!htb]
+\begin{figure}[H]
     a{
@@
 
+\section*{CRediT authorship contribution statement}
+Jag tolkar det som att denna sektion är frivillig. De flesta artiklar jag har hittat listar här vad alla enskilda författare har bidragit med.
+
+\section*{Declaration of Competing Interest}
+The authors declare that they have no known competing financial interests or personal relationships that could have appeared to influence the work reported in this paper.
+
 \section*{Acknowledgements}
@@
 We also thank Sigurd Israelsson at the School of Engineering at J\"onk\"oping University for lending hardware resources for the experiments in this study.
-The work has been funded by VINNOVA, EU InterReg Baltic Sea project WAMBAF tools and Formas.
+This work was supported by VINNOVA, EU InterReg Baltic Sea project WAMBAF tools and Formas.
+\textbf{Här ska varje grant definieras med "grant number[xxxx] efter varje aktör"}
 

### 2021-01-13 LaTeX ESWA/article.tex
Shortens some overexplained concepts to better fit new journal.

@@
 \definecolor{customgrayouter}{HTML}{404040}
+
+\usepackage{xcolor}
+\usepackage{soul}
+\newcommand{\hlc}[2][yellow]{{%
+    \colorlet{foo}{#1}%
+    \sethlcolor{foo}\hl{#2}}%
+}
 
@@
 We hypothesise that, by combining the digital terrain indices using machine learning, we can improve ditch detection at a landscape-scale.
-In addition to the raw terrain indices, additional input variables are generated by transforming the data by including neighbouring cells to improve the predictability of the ditches.
+In addition to the raw terrain indices, additional features are generated by transforming the data by including neighbouring cells to improve the predictability of the ditches.
 Random Forests is used to locate the ditches, and the probability output from this classifier is processed to remove noise, and binarised to produce the final ditch prediction.
 The confidence interval for the Cohen's Kappa index ranges \textbf{[0.655 , 0.781]} between the evaluation plots with a confidence level of \textbf{95\%}.
-Input variables based on the Impoundment Index overall represents the strongest ditch predictors.
+features based on the Impoundment Index overall represents the strongest ditch predictors.
 The study demonstrates that combining information from  a suite of digital terrain indices using machine learning provides an effective technique for automatic ditch detection at a landscape-scale.
@@
 We also examine which digital terrain indices are the most relevant for an accurate prediction of ditches.
-Several input variables for the model are calculated from the indices to place emphasis on the clear linear directional characteristics of ditches.
+Several features for the model are calculated from the indices to place emphasis on the clear linear directional characteristics of ditches.
 These characteristics are also used when processing the model's output, in order to remove noise and other inconsistencies in the predictions.
 We use a supervised learning algorithm with labelled ditch data to automatically extract ditches from instances not before seen by the model.
-These instances are used as input variables to the algorithm, and consist of refined or raw digital terrain indices derived from a DEM.
+These instances are used as features to the algorithm, and consist of refined or raw digital terrain indices derived from a DEM.
 
@@
 
-\begin{figure}[!htb]
+\begin{figure}[H]
     \centering
@@
     \caption{Map of Krycklan showing ditches and stream channels, as well as our division into 21 subsections.
-The 10 subsections with a white border were used for developing input variables and optimising parameters before the experiment.
+The 10 subsections with a white border were used for developing features and optimising parameters before the experiment.
 The 11 subsections with a black border were used in the experiment for evaluation with 11-fold cross validation.
@@
 \subsubsection{Processing the digital terrain indices}
-We examined how different  input variables affected the prediction of the machine learning models.
+We examined how different  features affected the prediction of the machine learning models.
 Several possible data manipulation methods could theoretically produce a better prediction.
@@
 The digital terrain indices provided a satisfactory foundation for the models, but lacked in the generalisability of their predictions.
-More diverse input variables were extracted using simple statistical aggregates such as mean, median, min, max, and standard deviation.
+More diverse features were extracted using simple statistical aggregates such as mean, median, min, max, and standard deviation.
 Using these statistical aggregations with the help of neighbouring areas around pixels aided in pruning pixels with outlier values, often smoothing out the data to represent ditches more accurately on a per-pixel basis.
@@
 
-In addition to the general input variables, a number of custom input variables were also developed.
+In addition to the general features, a number of custom features were also developed.
 Some of these custom variables were produced by statistical aggregation of multiple indices, whereas in some cases, they were generated by using thresholds to exclude, lower, or amplify the values of pixels that indicated ditches or non-ditches (\hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{d} and \hyperref[fig:features]{h}).
@@
 Only areas with a relatively high value, which would indicate that these areas most likely contained streams, were masked out.
-After widening the resulting area, this mask was used to remove streams from all the aforementioned custom input variables, generating one new input variable from each.
+After widening the resulting area, this mask was used to remove streams from all the aforementioned custom features, generating one new feature from each.
 The features marked with \textit{streams removed} in \hyperref[featuretable]{Table} \ref{featuretable} make use of this filter (\hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{i}).
 
-Overall, 81 input variables were extracted from the terrain indices.
+Overall, 81 features were extracted from the terrain indices.
 In a pilot study, we compared several different algorithms (Random Forests, Extreme Gradient Boosting, Naive Bayes, and Support Vector Machines), and it was found that Random Forests produced the most accurate results, here we therefore only present the results from the Random Forest model.
-We then conducted a sub-experiment to find the best input variables, as well as the optimal number of variables to use.
-The variable importance score in Random Forests was used to select the most important input variables and it was found that using the top 40 input variables produced the best results.
+We then conducted a sub-experiment to find the best features, as well as the optimal number of variables to use.
+The variable importance score in Random Forests was used to select the most important features and it was found that using the top 40 features produced the best results.
 
-\begin{figure} [!htb]
+\begin{figure} [H]
     \centering
@@
         \resizebox*{3.5cm}{!}{\includegraphics{./images/feature_i_lo.jpg}}}
-    \caption{Example of 8 of the 40 input variables, in addition to ditch labels used by the models for a small sample area.
+    \caption{Example of 8 of the 40 features, in addition to ditch labels used by the models for a small sample area.
 The radii represents pixels with a 0.5 m resolution.
@@
 To develop and evaluate our ditch detector, the raster and ditch label data of Krycklan were manually divided into 21 subsections.
-Each subsection represents an area of roughly 196 hectare.
+\hlc{This division into zones was necessary because the method relies on surrounding pixels, preventing the possibility of random sole pixels being used for training and testing.} Each subsection represents an area of roughly 196 hectare.
 From this division, 11 of the subsections were put aside as hold-out data, only for use in the experiment to evaluate the performance of the predictions.
@@
 \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan} shows which subsections were used for development and evaluation respectively.
-
-Using the 11 subsections in the hold-out data for the final experiment, a process called \textit{k-fold cross validation} was employed (11-fold cross validation in our case).
-K-fold cross validation is a method where you divide your dataset into folds of similar size and train a model on all but one of your folds (subsections).
-You then use that subsection to evaluate the results \citep{crossvalidation}.
-A new model is trained using the training folds for each iteration (\hyperref[fig:crossvalidation]{Figure} \ref{fig:crossvalidation}).
-Using this technique, shifting which subsection to leave out from the training, allowed us to train 11 different Random Forests classifying models with a large amount of data from the remaining 10 subsections in the hold-out data, producing 11 independent sub-experiments to evaluate the method on.
-
-\begin{figure}[!htb]
-    \centering
-    \includegraphics[width=1\linewidth]{images/cross_validation_lo.jpg}
-    \caption{11-fold cross validation. Each subsection in the hold-out data was used once for testing, with the remaining subsections being used to build a new model for each iteration.}
-    \label{fig:crossvalidation}
-\end{figure}
+\hlc{The 11 hold-out subsections were used as folds in an 11-fold cross validation in the final experiment.}
 
@@
 
-Random Forests is an ensemble machine learning method which builds multiple diverse decision trees to increase the robustness of its predictions \citep{breiman,flach}.
-The individual trees are trained by minimising the entropy of the data from each node’s output \citep{kotsiantis}.
-A byproduct of the training is the Gini importance, which denotes the most important input variables \citep{gini}.
+The Random Forests \citep{breiman} algorithm from the Python library \textit{scikit-learn} \citep{scikit-learn} was used in the experiment, and the classifiers were trained on the 40 features seen in \hyperref[featuretable]{Table} \ref{featuretable}.
+A byproduct of the training is the Gini importance, which denotes the most important features \citep{gini}.
 For our study, a hyperparameter tuning showed that using 300 trees and \textit{gini} as the splitting criterion, with a minimum of 10 samples per node split, and no artificial class weight or max depth of trees yielded the best predictions.
-The Random Forests algorithm from the Python library \textit{scikit-learn} \citep{scikit-learn} was used in the experiment, and the classifiers were trained on all the input variables seen in \hyperref[featuretable]{Table} \ref{featuretable}.
 
@@
 \centering
-    %\tbl{The 40 input variables used when training the models.}
-    {\begin{tabular}{l|l} %\toprule
-    %\hline
-      \textbf{Variable/Algorithm}\textsuperscript{a} & \textbf{Circular radii}\textsuperscript{b} \\ %\midrule
+    {\begin{tabular}{l|l}
+      \textbf{Terrain Index/Algorithm}\textsuperscript{a} & \textbf{Circular radii}\textsuperscript{b} \\ %\midrule
       \hline
@@
       Slope min & 2, 4, 6 \\
-      Slope non-ditch amplification & - \\ %\bottomrule 
+      Slope non-ditch amplification & - \\
       \hline
     \end{tabular}}
-    \caption{ The 40 input variables used when training the models.
-    \newline \textsuperscript{a} Displays algorithm used to produce the variable. \newline
+    \caption{ The 40 features used when training the models.
+    \newline \textsuperscript{a} Displays algorithm used to produce the feature. \newline
         \textsuperscript{b} Represents the radius of the circular mask (if one was used) to determine \newline what neighbouring pixels to use in an algorithm.
@@
 
-\begin{figure} [!htb]
+\begin{figure} [H]
     \centering
@@
 Because the $\kappa$ rating from our method outperforms all digital terrain indices, our hypothesis that combining digital terrain indices with a machine learning produces better ditch detection than using single terrain indices has been confirmed by the experiments.
-Most false positives lie in stream channels (\hyperref[fig:resultsillustrations]{Figure} \ref{fig:resultsillustrations} \hyperref[fig:resultsillustrations]{a, b}) and most false negatives occurred either due to ditches being too shallow (\hyperref[fig:resultsillustrations]{Figure} \ref{fig:resultsillustrations} \hyperref[fig:resultsillustrations]{c, d}), or due to the attempt at removing streams using the input variables as explained in \ref{impoundmentstreamremoval} removing the deepest ditches  (\hyperref[fig:resultsillustrations]{Figure} \ref{fig:resultsillustrations} \hyperref[fig:resultsillustrations]{e, f}).
+Most false positives lie in stream channels (\hyperref[fig:resultsillustrations]{Figure} \ref{fig:resultsillustrations} \hyperref[fig:resultsillustrations]{a, b}) and most false negatives occurred either due to ditches being too shallow (\hyperref[fig:resultsillustrations]{Figure} \ref{fig:resultsillustrations} \hyperref[fig:resultsillustrations]{c, d}), or due to the attempt at removing streams using the features as explained in \ref{impoundmentstreamremoval} removing the deepest ditches  (\hyperref[fig:resultsillustrations]{Figure} \ref{fig:resultsillustrations} \hyperref[fig:resultsillustrations]{e, f}).
 
@@
 \textbf{e} and \textbf{f} illustrate that some of the deepest and more prominent ditches were also missed.
-This occurred as a result of attempting to get rid of more false positives in the form of streams by using the input variables, as explained in \ref{impoundmentstreamremoval}.}
+This occurred as a result of attempting to get rid of more false positives in the form of streams by using the features, as explained in \ref{impoundmentstreamremoval}.}
     \label{fig:resultsillustrations}
@@
 
-In \hyperref[featureimportancetable]{Table} \ref{featureimportancetable}, the top 20 (out of 40) input variables from the Random Forests models are presented with their importance percentages for the models making a successful prediction.
-Input variables based on HPMF and Impoundment Index contributed the most to a successful prediction.
+In \hyperref[featureimportancetable]{Table} \ref{featureimportancetable}, the top 20 (out of 40) features from the Random Forests models are presented with their importance percentages for the models making a successful prediction.
+features based on HPMF and Impoundment Index contributed the most to a successful prediction.
 The Gini importance was obtained using the Gini impurity for each variable \citep{gini}.
 
-\begin{table} [!htb]
+\begin{table} [H]
         {\begin{tabular}{r|ll}
@@
         \end{tabular}}
-        \caption{The top 20 input variables (out of 40) by importance when the models make a prediction.
+        \caption{The top 20 features (out of 40) by importance when the models make a prediction.
 \newline \textsuperscript{a} The number next to some of the variables indicates the circular radius used to select what neighbouring pixels to use in the statistical aggregation method.
@@
 We managed to bridge gaps and exclude streams to some extent in our predictions.
-This was in part due to the post-processing of the prediction helping where the models were unsuccessful on their own, and in part due to the input variables developed.
-\hyperref[featureimportancetable]{Table} \ref{featureimportancetable} highlights the importance of the input variables that were developed by applying different statistical aggregations for neighbouring areas of pixels, similar to \citet{roelens}, as well as the custom variables using filters such as the Impoundment stream removal or Gabor filters.
+This was in part due to the post-processing of the prediction helping where the models were unsuccessful on their own, and in part due to the features developed.
+\hyperref[featureimportancetable]{Table} \ref{featureimportancetable} highlights the importance of the features that were developed by applying different statistical aggregations for neighbouring areas of pixels, similar to \citet{roelens}, as well as the custom variables using filters such as the Impoundment stream removal or Gabor filters.
 
@@
 A future improvement to the ditch detector could be to use a shape index to remove small natural streams, as natural streams often form more complex channels whereas ditches are often straight.
-A potential issue with generalising our ditch detector for use in other geographical areas is that all the post-processing steps and input variables were developed based on occurrences in the Krycklan area.
+A potential issue with generalising our ditch detector for use in other geographical areas is that all the post-processing steps and features were developed based on occurrences in the Krycklan area.
 Different geographical compositions in other areas may require the adjustment of the thresholds used to fill gaps in ditches and to remove noise.
@@
 This integration of a large set of terrain indices using machine learning produced good results, as evidenced by the confidence interval for the Cohen's $\kappa$ index ranging from 0.655 to 0.781 with a confidence level of 95\%.
-With the developed input variables and extensive post-processing, we also managed to remove many stream channels and fill gaps in the ditches where the LiDAR scans were weak.
+With the developed features and extensive post-processing, we also managed to remove many stream channels and fill gaps in the ditches where the LiDAR scans were weak.
 
@@
 Our approach enables accurate ditch detection in complex forested landscapes with tree canopy and many natural streams.
-Future work could introduce more input variables and post-processing steps, such as using pathfinding algorithms to fill gaps in the ditch model or shape indices to remove stream channels from the prediction.
+Future work could introduce more features and post-processing steps, such as using pathfinding algorithms to fill gaps in the ditch model or shape indices to remove stream channels from the prediction.
 Moving towards image segmentation instead of pixel classification of tabular pixel values could also be a suitable future approach.

### 2021-01-14 LaTeX ESWA/article.tex
Fixes some formatting.

@@
-\documentclass[review]{elsarticle}
+\documentclass[11pt, review]{elsarticle} % first argument is font-size for document
 \graphicspath{ {./images/} }
@@
 
+\usepackage % Sets margin distance left and right
+[
+        left=3.5cm,
+        right=3.5cm,
+]
+{geometry}
+
+\usepackage{caption}
+\captionsetup{font=footnotesize}
+
 \usepackage{xcolor}
@@
 \bibliographystyle{model5-names}\biboptions{authoryear}
+
+% Fixes row-to-row distance
+\renewcommand{\baselinestretch}{1.0}
 
@@
 \begin{abstract}
-In this study, we develop a method for detecting ditches in high resolution digital elevation models derived from LiDAR scans.
+{\footnotesize In this study, we develop a method for detecting ditches in high resolution digital elevation models derived from LiDAR scans.
 Thresholding methods using digital terrain indices, such as Sky View Factor, Impoundment Index, or High Pass Median Filter can be used to detect ditches.
@@
 features based on the Impoundment Index overall represents the strongest ditch predictors.
-The study demonstrates that combining information from  a suite of digital terrain indices using machine learning provides an effective technique for automatic ditch detection at a landscape-scale.
+The study demonstrates that combining information from  a suite of digital terrain indices using machine learning provides an effective technique for automatic ditch detection at a landscape-scale.}
 \end{abstract}
@@
 \begin{keyword}
-Machine learning \sep Geographic information systems \sep Classification and regression trees \sep Supervised learning by classification
+{\footnotesize
+Machine learning \sep Geographic information systems \sep Classification and regression trees \sep Supervised learning by classification}
 \end{keyword}
@@
 
+\newpage
 \section{Introduction}\label{introduction}

### 2021-01-14 LaTeX ESWA/article.tex
Adds WIP on restructuring the introduction.

@@
 \usepackage{soul}
-\newcommand{\hlc}[2][yellow]{{%
-    \colorlet{foo}{#1}%
-    \sethlcolor{foo}\hl{#2}}%
-}
+
+\usepackage[bordercolor=white,backgroundcolor=gray!30,linecolor=black,colorinlistoftodos]{todonotes}
+\newcommand{\hlc}[1]{\todo[color=yellow,inline]{#1}}
 
@@
 
-\author[ju]{Niklas Lavesson \corref{cor1}}
+\author[ju]{Niklas Lavesson\corref{cor1}}
 \ead{niklas.lavesson@ju.se}
@@
 \begin{abstract}
-{\footnotesize In this study, we develop a method for detecting ditches in high resolution digital elevation models derived from LiDAR scans.
+{\footnotesize
+In this study, we develop a method for detecting ditches in high resolution digital elevation models derived from LiDAR scans.
 Thresholding methods using digital terrain indices, such as Sky View Factor, Impoundment Index, or High Pass Median Filter can be used to detect ditches.
@@
 The confidence interval for the Cohen's Kappa index ranges \textbf{[0.655 , 0.781]} between the evaluation plots with a confidence level of \textbf{95\%}.
-features based on the Impoundment Index overall represents the strongest ditch predictors.
+Features based on the Impoundment Index overall represents the strongest ditch predictors.
 The study demonstrates that combining information from  a suite of digital terrain indices using machine learning provides an effective technique for automatic ditch detection at a landscape-scale.}
@@
 
-For effective restoration of ditched wetlands, it is highly important  to accurately locate the ditches on digital maps.
+The forest industry and agencies have a demand for better maps of the ditch networks in the forest landscape.
+Drained afforested peatlands contribute strongly to Swedish wood production, and consequently to the Swedish economy.
+At the same time, Swedish authorities have identified drained wetlands as a major target for greenhouse gas mitigation \textbf{REF}.
+However, research on the effects of afforested drained wetlands on greenhouse gas emission is inconclusive, and they can constitute both a source and a sink for greenhouse gases \textbf{REF?}.
+Management effects on the greenhouse gas balance are controlled by complex interactions of various factors.
+There is a need for science-based tools to optimise forest production, as well as decision support systems for mitigation of greenhouse gas emissions from drained forested wetlands.
+A key step to accomplish this is to map the ditch networks in the forest landscape.
 However, landscape-scale field inventories are often challenging and expensive.
-Automatic detection of ditches from high-resolution digital elevation models (DEMs) and terrain indices provides a feasible means for spatially explicit assessment of ditches across a landscape.
-Previous researches  \citep{uppsala} documented  that terrain indices such as Sky View Factor \citep{zaksek}, Impoundment Index \citep{whiteboxtools} and High Pass Median Filter \citep{whiteboxtools} can efficiently  detect the ditches.
-However, different false positives and negatives may be observed when examining each terrain index manually.
-Such inaccuracies and complexities are in part related to the Light Detection and Ranging (LiDAR) measurements, the mainstay of high resolution DEM and terrain mapping, sometimes being interfered by trees, bushes, or grass - leaving gaps in the ditch model.
-The inaccuracies can also arise from a poorly set threshold for the LiDAR derived terrain parameters.
+
+The National Inventory of Landscapes in Sweden (NILS) has recorded small watercourses (width $<$ 6 m) by conducting a line inventory in 631 $5*5$ km squares randomly distributed in Sweden, to best capture a representative and varied sample of the Swedish landscape.
+The positions of the small watercourses from this inventory were compared with the watercourses on Swedish Property Map (1: 12 500) (\hyperref[fig:watercoursebarplot]{Figure} \ref{fig:watercoursebarplot}).
+A total of 1103 natural watercourses, 131 straightened watercourses/channels, and 2089 ditches were found in the field.
+This simple investigation highlights two things: First, ditches make up the majority of the small scale water channels in the Swedish landscape; almost twice as many as the natural streams.
+Second, most of the small scale channels are missing on the current maps; only 45 \% of the natural watercourses, 25 \% of the straightened watercourses/channels, and 9 \% of the ditches were mapped (\hyperref[fig:watercoursebarplot]{Figure} \ref{fig:watercoursebarplot}).
+
+\begin{figure}[!htb]
+    \centering
+    \begin{tikzpicture}
+    \begin{axis}[
+        width={0.9\textwidth},
+        height={0.6\textwidth},
+        ylabel={fluorescence},
+        ybar stacked,
+        ymajorgrids=true,
+    	bar width=45pt,
+    	nodes near coords,
+    	every node near coord/.append style={font=\tiny},
+        enlarge y limits={0.15},
+        enlarge x limits=0.15,
+        legend style={at={(0.78,0.96)},
+          anchor=north,legend columns=-1},
+        ylabel={{}},
+        symbolic x coords={Ditches, Straightened  watercourses, Natural  watercourses},
+        xtick=data,
+        ytick={0,500,1000,1500,2000},
+        x tick label style={rotate=0,anchor=north},
+        ]
+    \addplot+[ybar,color=custombeige,fill=custombeige,draw=custombeigeouter,thick] plot coordinates {(Ditches,188) (Straightened  watercourses,33) 
+      (Natural  watercourses,496)};
+    \addplot+[ybar,color=customgrayouter,fill=customgray,draw=customgrayouter,thick] plot coordinates {(Ditches,1901) (Straightened  watercourses,98) 
+      (Natural  watercourses,607)};
+    \legend{\strut Mapped, \strut Unmapped}
+    \end{axis}
+    \end{tikzpicture}
+    \caption{The bars indicate the number of observed channels in the National Inventory of Landscapes in Sweden's (NILS) line inventory of small watercourses (width $<$ 6 m).
+The colours of the bars indicate whether they are mapped or not on the Swedish Property map.}
+    \label{fig:watercoursebarplot}
+\end{figure}
+
+Several previous studies have located ditches from Light Detection and Ranging (LiDAR) data in smaller agricultural areas \citep{roelens, bailly} or smaller forest areas \citep{rapinel, kiss} (smaller than 2 $km^2$), predominantly characterised by artificial drainage.
+However, we herein conduct a landscape-scale analysis in a mesoscale catchment.
+The study  catchment  is 68 $km^2$, representing diverse landscape features and geomorphological characteristics.
+Additionally, the forest landscape introduces many challenges that are not present when detecting ditches along roads or in agricultural areas.
+The canopy cover in forest areas prevents the use of satellite imagery or aerial photos, as well as lowers the quality of the digital elevation models (DEMs), as fewer LiDAR points reach the ground, instead being caught in trees, bushes, or grass.
+Forest ditches are also generally not maintained to the same extent as agricultural or road ditches, often being overgrown or covered by windthrows or windsnaps.
+Previous researches documented  that terrain indices such as Sky View Factor \citep{zaksek}, Impoundment Index \citep{whiteboxtools} and High Pass Median Filter \citep{whiteboxtools} can detect ditches in some cases \citep{uppsala}.
+However, different false positives and negatives can be observed when examining each terrain index manually.
+These inaccuracies can stem either from the gaps in the LiDAR scans, or from a poorly set threshold for the LiDAR derived terrain parameters.
 Thus, it would be useful to combine the terrain indices derived from high resolution LiDAR data to prune the weaknesses and amplify the strengths of each index and produce highly accurate detection of ditch networks.
 
-In this study, we first investigate the need for better maps of ditch networks, then we introduce a new method for the automated mapping of ditches by integrating multiple high resolution terrain indices (derived from LiDAR DEM) using a machine learning algorithm.
+In this study, we introduce a new method for the automated mapping of ditches by integrating multiple high resolution terrain indices (derived from LiDAR DEM) using a supervised learning approach.
 We also examine which digital terrain indices are the most relevant for an accurate prediction of ditches.
@@
 These characteristics are also used when processing the model's output, in order to remove noise and other inconsistencies in the predictions.
-We use a supervised learning algorithm with labelled ditch data to automatically extract ditches from instances not before seen by the model.
-These instances are used as features to the algorithm, and consist of refined or raw digital terrain indices derived from a DEM.
-
 We hypothesise that by combining the information from all the digital terrain indices using a machine learner (Random Forests), we can improve the detection of the ditches, compared to if a single index would be used.
@@
 This manual map guided the model training and ground-truthing in our analysis and provided a test-bench for our methodology.
-Previous studies \citep{roelens, bailly, rapinel, kiss} that have located ditches from LiDAR data have worked on smaller areas (smaller than 2 $km^2$), predominantly characterised by artificial drainage.
-However, we herein conduct a landscape-scale analysis in a mesoscale catchment.
-The study  catchment  is 68 $km^2$, representing diverse landscape features and geomorphological characteristics.
 
@@
 
-\subsection{The need for better small scale channel mappings}
-To investigate the need for better maps of ditches and small stream channels we compared the data from a national field survey with available maps.
-The National Inventory of Landscapes in Sweden (NILS) has recorded small watercourses (width $<$ 6 m) by conducting a line inventory in 631 $5*5$ km squares randomly distributed in Sweden, to best capture a representative and varied sample of the Swedish landscape.
-The positions of the small watercourses from this inventory were compared with the watercourses on Swedish Property Map (1: 12 500) (allowing for an uncertainty in the positioning of up to 15 m, by snapping the point observations to the lines on the map) (\hyperref[fig:watercoursebarplot]{Figure} \ref{fig:watercoursebarplot}).
-
-\subsection{Study area and data}
+\subsection{Study area and dataset}
 In the Krycklan Study Catchment the quaternary deposits are dominated by till (51\%) and sorted sediments (30\%).
@@
 
-\begin{figure}[H]
+\begin{figure}[!htb]
     \centering
@@
 
-\begin{figure} [htb!]
+\begin{figure} [!htb]
     \centering
@@
 
-\subsubsection{Thresholding the digital terrain indices}
 To benchmark our method against a non-machine learning method, we binarised the original four digital terrain indices using thresholds.
@@
 
-\begin{figure} [H]
+\begin{figure} [!htb]
     \centering
@@
 
-\begin{table} [H]
+\begin{table} [!htb]
 \centering
@@
 \section{Results and analysis}
-In the national survey (NILS), a total of 1103 natural watercourses, 131 straightened watercourses/channels, and 2089 ditches were found in the field.
-This simple investigation highlights two things: Firstly, the ditches make up the majority of the small scale water channels in the Swedish landscape; almost twice as many as the natural streams.
-Secondly, most of the small scale channels are missing on the current maps; only 45 \% of the natural watercourses, 25 \% of the straightened watercourses/channels, and 9 \% of the ditches were mapped (\hyperref[fig:watercoursebarplot]{Figure} \ref{fig:watercoursebarplot}).
-
-\begin{figure}[H]
-    \centering
-    \begin{tikzpicture}
-    \begin{axis}[
-        width={0.9\textwidth},
-        height={0.6\textwidth},
-        ylabel={fluorescence},
-        ybar stacked,
-        ymajorgrids=true,
-    	bar width=45pt,
-    	nodes near coords,
-    	every node near coord/.append style={font=\tiny},
-        enlarge y limits={0.15},
-        enlarge x limits=0.15,
-        legend style={at={(0.78,0.96)},
-          anchor=north,legend columns=-1},
-        ylabel={{}},
-        symbolic x coords={Ditches, Straightened  watercourses, Natural  watercourses},
-        xtick=data,
-        ytick={0,500,1000,1500,2000},
-        x tick label style={rotate=0,anchor=north},
-        ]
-    \addplot+[ybar,color=custombeige,fill=custombeige,draw=custombeigeouter,thick] plot coordinates {(Ditches,188) (Straightened  watercourses,33) 
-      (Natural  watercourses,496)};
-    \addplot+[ybar,color=customgrayouter,fill=customgray,draw=customgrayouter,thick] plot coordinates {(Ditches,1901) (Straightened  watercourses,98) 
-      (Natural  watercourses,607)};
-    \legend{\strut Mapped, \strut Unmapped}
-    \end{axis}
-    \end{tikzpicture}
-    \caption{The bars indicate the number of observed channels in the National Inventory of Landscapes in Sweden's (NILS) line inventory of small watercourses (width $<$ 6 m).
-The colours of the bars indicate if they are mapped or not on the Swedish Property map.}
-    \label{fig:watercoursebarplot}
-\end{figure}
 
@@
 
-\begin{table}[H]
+\begin{table}[!htb]
 \centering
@@
 
-\begin{figure} [H]
+\begin{figure} [!htb]
     a{
@@
 
-\begin{table} [H]
+\begin{table} [!htb]
         {\begin{tabular}{r|ll}
@@
 
-\begin{figure}[H]
+\begin{figure}[!htb]
     a{
@@
 \section*{CRediT authorship contribution statement}
-Jag tolkar det som att denna sektion är frivillig. De flesta artiklar jag har hittat listar här vad alla enskilda författare har bidragit med.
+\hlc{Jag tolkar det som att denna sektion är frivillig. De flesta artiklar jag har hittat listar här vad alla enskilda författare har bidragit med.}
 
@@
 This work was supported by VINNOVA, EU InterReg Baltic Sea project WAMBAF tools and Formas.
-\textbf{Här ska varje grant definieras med "grant number[xxxx] efter varje aktör"}
+\hlc{ \textbf{ Här ska varje grant definieras med "grant number[xxxx] efter varje aktör"} }
 

### 2021-01-14 LaTeX ESWA/article.tex
Adds more WIP on the introduction, as well as adds a broader conclusion at the end of the article.

@@
 Drained afforested peatlands contribute strongly to Swedish wood production, and consequently to the Swedish economy.
-At the same time, Swedish authorities have identified drained wetlands as a major target for greenhouse gas mitigation \textbf{REF}.
-However, research on the effects of afforested drained wetlands on greenhouse gas emission is inconclusive, and they can constitute both a source and a sink for greenhouse gases \textbf{REF?}.
-Management effects on the greenhouse gas balance are controlled by complex interactions of various factors.
-There is a need for science-based tools to optimise forest production, as well as decision support systems for mitigation of greenhouse gas emissions from drained forested wetlands.
+At the same time, Swedish authorities have identified drained wetlands as a major target for greenhouse gas mitigation \citep{greenhouse_skogsstyrelsen, greenhouse_jordbruksverket}.
+However, research on the effects of afforested drained wetlands on greenhouse gas emission is inconclusive, and they can constitute both a source and a sink for greenhouse gases \citep{greenhouse_yes, greenhouse_no}.
+There is a need for science-based tools to optimise both forest production and decision support systems for mitigation of greenhouse gas emissions from drained forested wetlands.
 A key step to accomplish this is to map the ditch networks in the forest landscape.
-However, landscape-scale field inventories are often challenging and expensive.
+However, it is unfeasible to conduct landscape-scale field inventories, as they are often both challenging and expensive.
 
@@
 
-Several previous studies have located ditches from Light Detection and Ranging (LiDAR) data in smaller agricultural areas \citep{roelens, bailly} or smaller forest areas \citep{rapinel, kiss} (smaller than 2 $km^2$), predominantly characterised by artificial drainage.
-However, we herein conduct a landscape-scale analysis in a mesoscale catchment.
+Several previous studies have located ditches from Light Detection and Ranging (LiDAR) data in smaller agricultural areas \citep{roelens, bailly} or smaller forest areas (smaller than 2 $km^2$) \citep{rapinel, kiss}, predominantly characterised by artificial drainage.
+In this study, however, we conduct a landscape-scale analysis in a mesoscale catchment.
 The study  catchment  is 68 $km^2$, representing diverse landscape features and geomorphological characteristics.
 Additionally, the forest landscape introduces many challenges that are not present when detecting ditches along roads or in agricultural areas.
-The canopy cover in forest areas prevents the use of satellite imagery or aerial photos, as well as lowers the quality of the digital elevation models (DEMs), as fewer LiDAR points reach the ground, instead being caught in trees, bushes, or grass.
+The canopy cover in forest areas prevents the use of satellite imagery or aerial photos, as well as lowers the quality of the digital elevation models (DEMs), as fewer LiDAR points reach the ground, instead frequently catching in trees, bushes, or grass.
 Forest ditches are also generally not maintained to the same extent as agricultural or road ditches, often being overgrown or covered by windthrows or windsnaps.
-Previous researches documented  that terrain indices such as Sky View Factor \citep{zaksek}, Impoundment Index \citep{whiteboxtools} and High Pass Median Filter \citep{whiteboxtools} can detect ditches in some cases \citep{uppsala}.
+Previous studies have found that terrain indices such as Sky View Factor \citep{zaksek}, Impoundment Index \citep{whiteboxtools} and High Pass Median Filter \citep{whiteboxtools} can detect ditches in some cases \citep{uppsala}.
 However, different false positives and negatives can be observed when examining each terrain index manually.
-These inaccuracies can stem either from the gaps in the LiDAR scans, or from a poorly set threshold for the LiDAR derived terrain parameters.
-Thus, it would be useful to combine the terrain indices derived from high resolution LiDAR data to prune the weaknesses and amplify the strengths of each index and produce highly accurate detection of ditch networks.
+These inaccuracies can stem either from the gaps in the LiDAR scans, or from a poorly set threshold for the terrain index parameters.
+Thus, it would be useful to combine these LiDAR derived terrain indices to prune the weaknesses, and amplify the strengths of each index and produce highly accurate detection of ditch networks.
 
-In this study, we introduce a new method for the automated mapping of ditches by integrating multiple high resolution terrain indices (derived from LiDAR DEM) using a supervised learning approach.
-We also examine which digital terrain indices are the most relevant for an accurate prediction of ditches.
-Several features for the model are calculated from the indices to place emphasis on the clear linear directional characteristics of ditches.
-These characteristics are also used when processing the model's output, in order to remove noise and other inconsistencies in the predictions.
-We hypothesise that by combining the information from all the digital terrain indices using a machine learner (Random Forests), we can improve the detection of the ditches, compared to if a single index would be used.
+In this study, we introduce a new method for the automated mapping of ditches by integrating multiple high resolution terrain indices using a supervised learning approach.
+We hypothesise that by combining the information from all the digital terrain indices using a machine learner (Random Forests), we can improve the detection of the ditches, compared to if a single terrain index is used.
 To test this, we conduct a landscape-scale ditch detection analysis at the Krycklan Catchment in Sweden \citep{krycklancatchment}, where a high precision manual mapping has been performed previously.
-This manual map guided the model training and ground-truthing in our analysis and provided a test-bench for our methodology.
+This manual mapping guided the model training and ground-truthing in our analysis and provided a test-bench for our methodology.
 
@@
 \subsection{Study area and dataset}
-In the Krycklan Study Catchment the quaternary deposits are dominated by till (51\%) and sorted sediments (30\%).
+The quaternary deposits in the Krycklan Study Catchment are dominated by till (51\%) and sorted sediments (30\%).
 The catchment ranges in elevation from 114 to 405 m above sea level.
@@
 Instead, we included local variability in the indices by using different radii to capture ditches with varied shapes.
-This integration of a large set of terrain indices using machine learning produced good results, as evidenced by the confidence interval for the Cohen's $\kappa$ index ranging from 0.655 to 0.781 with a confidence level of 95\%.
+This integration of a large set of terrain indices using machine learning produced good results, as evidenced by the confidence interval for the Cohen's $\kappa$ index, which ranged from 0.655 to 0.781 with a confidence level of 95\%.
 With the developed features and extensive post-processing, we also managed to remove many stream channels and fill gaps in the ditches where the LiDAR scans were weak.
@@
 Our developed methodology can be adjusted to apply to any setting for artificial ditch detection, reducing time and labour investment compared to the conventional manual mapping techniques.
-The accurate ditch maps can significantly contribute to practical forestry and operational land management at a landscape-scale.
+\hl{The accurate ditch maps can both significantly contribute to practical forestry and operational land management, as well as allow stakeholders, including government agencies, forest companies, and regional and local authorities to predict the climate consequences of potential management actions, which will be critical in reaching global sustainability goals.}
 

### 2021-01-14 LaTeX ESWA/article.tex
Adds more WIP on introduction. (Still needs more foruc on previous work).

@@
 In this study, we introduce a new method for the automated mapping of ditches by integrating multiple high resolution terrain indices using a supervised learning approach.
+Several features and post-processing steps, specifically tailored for ditch detection are also developed and employed.
 We hypothesise that by combining the information from all the digital terrain indices using a machine learner (Random Forests), we can improve the detection of the ditches, compared to if a single terrain index is used.
@@
     a{
-        \resizebox*{5.6cm}{!}{\includegraphics{./images/re_evaluation_1_lo.jpg}}}\hspace{5pt}
+        \resizebox*{6.5cm}{!}{\includegraphics{./images/re_evaluation_1_lo.jpg}}}\hspace{5pt}
     b{
-        \resizebox*{5.6cm}{!}{\includegraphics{./images/re_evaluation_2_lo.jpg}}}
+        \resizebox*{6.5cm}{!}{\includegraphics{./images/re_evaluation_2_lo.jpg}}}
     \caption{Illustration of the modified evaluation labels.
@@
 \centering
-    {\begin{tabular}{l|cccc} %\toprule
-        \textbf{Metric} & \textbf{Sky View Factor} & \textbf{Impoundment} & \textbf{HPMF} & \textbf{Slope}\\ %\midrule
+    {\begin{tabular}{l|cccc}
+        \textbf{Metric} & \textbf{Sky View Factor} & \textbf{Impoundment} & \textbf{HPMF} & \textbf{Slope}\\
         \hline
@@
         $\kappa$ rating & 0.048 & 0.215 & 0.211 & 0.029 \\
-        AUPRC & 0.247 & 0.248 & 0.232 & 0.194 \\ %\bottomrule
+        AUPRC & 0.247 & 0.248 & 0.232 & 0.194 \\
         \hline
     \end{tabular}}
-    \caption{Metrics from the total results of the four digital terrain indices experiments.}
+    \caption{Metrics from the total results of the four digital terrain indices experiments, where only a sole terrain index is used with a threshold to detect ditches.}
     \label{recreatedpredictionperformance}
@@
 Because the subsections have varying amounts of ditches in them, the confidence intervals will not be completely accurate, but will produce a close estimation.
-
-From \hyperref[predictionperformance]{Table} \ref{predictionperformance}, it is observed that averaging the results from the 11 subsections would yield a very similar value to the total results, indicating that the ditch detector generally performs equally well on subsections with a small amount of ditches as on subsections with a large amount of ditches.
+Averaging the results from the 11 subsections would yield a very similar value to the total results, indicating that the ditch detector generally performs equally well on subsections with a small amount of ditches as on subsections with a large amount of ditches.
 
@@
         \textbf{Metric} & \textbf{Total}\textsuperscript{a} & \textbf{Subsection Average}\textsuperscript{b}& \textbf{CI 95\%}\textsuperscript{c} \\ 
-        %&&\textbf{Average}&\\
         \hline
@@
     a{
-        \resizebox*{4cm}{!}{\includegraphics{./images/results_illustration_1_lo.jpg}}}\hspace{5pt}
+        \resizebox*{5.3cm}{!}{\includegraphics{./images/results_illustration_1_lo.jpg}}}\hspace{5pt}
     b{
-        \resizebox*{4cm}{!}{\includegraphics{./images/results_illustration_2_lo.jpg}}}
+        \resizebox*{5.3cm}{!}{\includegraphics{./images/results_illustration_2_lo.jpg}}}
     \newline c{
-        \resizebox*{4cm}{!}{\includegraphics{./images/results_illustration_3_lo.jpg}}}\hspace{5pt}
+        \resizebox*{5.3cm}{!}{\includegraphics{./images/results_illustration_3_lo.jpg}}}\hspace{5pt}
     d{
-        \resizebox*{4cm}{!}{\includegraphics{./images/results_illustration_4_lo.jpg}}}
+        \resizebox*{5.3cm}{!}{\includegraphics{./images/results_illustration_4_lo.jpg}}}
     \newline e{
-        \resizebox*{4cm}{!}{\includegraphics{./images/results_illustration_5_lo.jpg}}}\hspace{5pt}
+        \resizebox*{5.3cm}{!}{\includegraphics{./images/results_illustration_5_lo.jpg}}}\hspace{5pt}
     f{
-        \resizebox*{4cm}{!}{\includegraphics{./images/results_illustration_6_lo.jpg}}}
+        \resizebox*{5.3cm}{!}{\includegraphics{./images/results_illustration_6_lo.jpg}}}
     \caption{The left panel shows the hillshade for a number of subsections, and the right panel shows the hillshade with the extracted ditches superimposed on top.

### 2021-01-16 LaTeX ESWA/article.tex
Adds more references to the introduction, and more tweaking to the overall article.

@@
 
-\title{Detecting ditches using supervised learning on high-resolution digital elevation models}
+\title{Detecting ditches using \hl{supervised} learning on high-resolution \hl{digital elevation models}}
 
@@
 {\footnotesize
-In this study, we develop a method for detecting ditches in high resolution digital elevation models derived from LiDAR scans.
-Thresholding methods using digital terrain indices, such as Sky View Factor, Impoundment Index, or High Pass Median Filter can be used to detect ditches.
-However, a single threshold generally does not capture the variability in the landscape and generates many false positives and negatives.
-We hypothesise that, by combining the digital terrain indices using machine learning, we can improve ditch detection at a landscape-scale.
-In addition to the raw terrain indices, additional features are generated by transforming the data by including neighbouring cells to improve the predictability of the ditches.
-Random Forests is used to locate the ditches, and the probability output from this classifier is processed to remove noise, and binarised to produce the final ditch prediction.
+\hl{Drained wetlands can constitute a large source of greenhouse gas emissions, but the drainage networks in these wetlands are largely unmapped, and better maps are needed to predict the climate consequences and aid in forest management decisions.} We develop a method for detecting ditches in high resolution digital elevation models derived from LiDAR scans.
+\hl{Thresholding methods using digital terrain indices can be used to detect ditches,} however, a single threshold generally does not capture the variability in the landscape and generates many false positives and negatives.
+We hypothesise that, by combining the digital terrain indices using \hl{supervised} learning, we can improve ditch detection at a landscape-scale.
+In addition to \hl{digital} terrain indices, additional features are generated by transforming the data to include neighbouring cells \hl{for better ditch predictions}.
+\hl{A Random Forests classifier} is used to locate the ditches, and \hl{its probability output} is processed to remove noise, and binarised to produce the final ditch prediction.
 The confidence interval for the Cohen's Kappa index ranges \textbf{[0.655 , 0.781]} between the evaluation plots with a confidence level of \textbf{95\%}.
-Features based on the Impoundment Index overall represents the strongest ditch predictors.
-The study demonstrates that combining information from  a suite of digital terrain indices using machine learning provides an effective technique for automatic ditch detection at a landscape-scale.}
+The study demonstrates that combining information from  a suite of digital terrain indices using machine learning provides an effective technique for automatic ditch detection at a landscape-scale\hl{, aiding in both practical forest management and in combatting climate change}.}
 \end{abstract}
@@
 Forest ditches are also generally not maintained to the same extent as agricultural or road ditches, often being overgrown or covered by windthrows or windsnaps.
-Previous studies have found that terrain indices such as Sky View Factor \citep{zaksek}, Impoundment Index \citep{whiteboxtools} and High Pass Median Filter \citep{whiteboxtools} can detect ditches in some cases \citep{uppsala}.
+Using raw LiDAR point clouds to detect ditches, as \citet{roelens} and \citet{bailly} did in their studies can thus be problematic in a forest landscape, where a DEM constructed to map the points on the ground should be more suitable.
+\citet{cazorzi} used a DEM with image processing techniques aimed at detecting the elongated ditch shapes and excluding large landscape features.
+This approach was suitable for their open agricultural areas, but is difficult to generalise to a diverse forest landscape.
+Previous studies have found that terrain indices such as Sky View Factor \citep{zaksek}, Impoundment Index \citep{whiteboxtools} and High Pass Median Filter \citep{whiteboxtools} can detect ditches in forest landscapes in some cases \citep{uppsala}.
 However, different false positives and negatives can be observed when examining each terrain index manually.
@@
 Several features and post-processing steps, specifically tailored for ditch detection are also developed and employed.
+These features aim to both highlight elongated shapes similarly to \citet{cazorzi}, as well as include a map of the neighbouring area as \citet{roelens} did in their study.
 We hypothesise that by combining the information from all the digital terrain indices using a machine learner (Random Forests), we can improve the detection of the ditches, compared to if a single terrain index is used.
@@
 To develop and evaluate our ditch detector, the raster and ditch label data of Krycklan were manually divided into 21 subsections.
-\hlc{This division into zones was necessary because the method relies on surrounding pixels, preventing the possibility of random sole pixels being used for training and testing.} Each subsection represents an area of roughly 196 hectare.
+\hl{This was necessary because our method relies on the pixels surrounding a pixel, meaning it is not possible to select random pixels for training and testing.} Each subsection represents an area of roughly 196 hectare.
 From this division, 11 of the subsections were put aside as hold-out data, only for use in the experiment to evaluate the performance of the predictions.
@@
 \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan} shows which subsections were used for development and evaluation respectively.
-\hlc{The 11 hold-out subsections were used as folds in an 11-fold cross validation in the final experiment.}
+\hl{The 11 hold-out subsections were used as folds in an 11-fold cross validation in the final experiment.}
 
@@
     \end{tabular}}
-    \caption{Metrics from the total results of the four digital terrain indices experiments, where only a sole terrain index is used with a threshold to detect ditches.}
+    \caption{Metrics from the total results of the four digital terrain indices experiments\hl{, where only a sole terrain index is used with a threshold to detect ditches}.}
     \label{recreatedpredictionperformance}
@@
 
-The comparison with the field observed channels and the available maps clearly indicate the need for better maps of ditches, as only 9\% of ditches are mapped.
-(\hyperref[fig:watercoursebarplot]{Figure} \ref{fig:watercoursebarplot}).
-Our study has shown that it is possible to locate ditches automatically in high-resolution DEMs ($0.5  * 0.5 $ metres, in our case), and that more of the ditches can be detected if the information from several terrain indices (\hyperref[recreatedpredictionperformance]{Table} \ref{recreatedpredictionperformance}) is combined through machine learning than if indices are used separately (\hyperref[predictionperformance]{Table} \ref{predictionperformance}).
+Our study has shown that it is possible to locate ditches automatically in high-resolution DEMs ($0.5  * 0.5 $ metres, in our case), and that more of the ditches can be detected if several terrain indices (\hyperref[recreatedpredictionperformance]{Table} \ref{recreatedpredictionperformance}) are combined through machine learning than if indices are used separately (\hyperref[predictionperformance]{Table} \ref{predictionperformance}).
 The ditch detection performance varied depending on the type of ditches, this is illustrated in \hyperref[fig:ditchpictures]{Figure} \ref{fig:ditchpictures}.
@@
 However, this indicates that our method could also be used to map small, previously unmapped natural stream channels.
-The national survey showed that 55\% of the natural stream channels and 75\% of the straightened water courses were missing on the maps.
+The national survey showed that 55\% of the natural stream channels and 75\% of the straightened water courses were missing on the maps (\autoref{fig:watercoursebarplot}).
 This is in line with other national \citep{kuglerova} and international \citep{benstead} studies highlighting this phenomenon called "Aqua Incognita - the unknown headwaters" \citep{bishop,kuglerova}.
@@
 This was in part due to the post-processing of the prediction helping where the models were unsuccessful on their own, and in part due to the features developed.
-\hyperref[featureimportancetable]{Table} \ref{featureimportancetable} highlights the importance of the features that were developed by applying different statistical aggregations for neighbouring areas of pixels, similar to \citet{roelens}, as well as the custom variables using filters such as the Impoundment stream removal or Gabor filters.
+\hyperref[featureimportancetable]{Table} \ref{featureimportancetable} highlights the importance of the features that were developed by applying different statistical aggregations for neighbouring areas of pixels (similar to \citet{roelens}) as well as the custom features using filters such as the Impoundment stream removal or Gabor filters.
 
@@
 The hydrology of our study catchment also differs from that of other studies, as we have almost as many natural stream channels as ditches, whereas their studies were conducted in smaller, predominantly artificially drained areas \citep{bailly, roelens, rapinel}.
-Because our method locates both ditches and natural stream channels, we have to take additional measures to remove the stream channels (affecting recall negatively), as explained in \ref{impoundmentstreamremoval}.
+Because our method locates both ditches and natural stream channels, we have to take additional measures to remove the stream channels (affecting ditch detection negatively), as explained in \ref{impoundmentstreamremoval}.
 
@@
 Our developed methodology can be adjusted to apply to any setting for artificial ditch detection, reducing time and labour investment compared to the conventional manual mapping techniques.
-\hl{The accurate ditch maps can both significantly contribute to practical forestry and operational land management, as well as allow stakeholders, including government agencies, forest companies, and regional and local authorities to predict the climate consequences of potential management actions, which will be critical in reaching global sustainability goals.}
+\hl{The accurate ditch maps can both significantly contribute to practical forestry and operational land management, as well as allow stakeholders including government agencies, forest companies, and regional and local authorities to predict the climate consequences of potential management actions, which will be critical in reaching global sustainability goals.}
 
 \section*{CRediT authorship contribution statement}
-\hlc{Jag tolkar det som att denna sektion är frivillig. De flesta artiklar jag har hittat listar här vad alla enskilda författare har bidragit med.}
+\hlc{Jag tolkar det som att denna sektion är frivillig.
+De flesta artiklar jag har hittat listar här vad alla enskilda författare har bidragit med.
+Kategorier som andra har med här är: Conceptualisation, Data curation, Formal analysis, Investigation, Methodology, Software, Validation, Visualisation, Writing - original draft, Writing - review & editing, Supervision, Project administration}
 

### 2021-01-16 LaTeX ESWA/article.tex
Basically completes the introduction.

@@
 {\footnotesize
-\hl{Drained wetlands can constitute a large source of greenhouse gas emissions, but the drainage networks in these wetlands are largely unmapped, and better maps are needed to predict the climate consequences and aid in forest management decisions.} We develop a method for detecting ditches in high resolution digital elevation models derived from LiDAR scans.
-\hl{Thresholding methods using digital terrain indices can be used to detect ditches,} however, a single threshold generally does not capture the variability in the landscape and generates many false positives and negatives.
+\hl{Drained wetlands can constitute a large source of greenhouse gas emissions, but the drainage networks in these wetlands are largely unmapped, and better maps are needed to aid in forest management decisions, and to predict the climate consequences.} We develop a method for detecting ditches in high resolution digital elevation models derived from LiDAR scans.
+\hl{Thresholding methods using digital terrain indices can be used to detect ditches,} however, a single threshold generally does not capture the variability in the landscape, and generates many false positives and negatives.
 We hypothesise that, by combining the digital terrain indices using \hl{supervised} learning, we can improve ditch detection at a landscape-scale.
-In addition to \hl{digital} terrain indices, additional features are generated by transforming the data to include neighbouring cells \hl{for better ditch predictions}.
+In addition to digital terrain indices, additional features are generated by transforming the data to include neighbouring cells \hl{for better ditch predictions}.
 \hl{A Random Forests classifier} is used to locate the ditches, and \hl{its probability output} is processed to remove noise, and binarised to produce the final ditch prediction.
@@
 
-Several previous studies have located ditches from Light Detection and Ranging (LiDAR) data in smaller agricultural areas \citep{roelens, bailly} or smaller forest areas (smaller than 2 $km^2$) \citep{rapinel, kiss}, predominantly characterised by artificial drainage.
+Several previous studies have located ditches from remote sensing Light Detection and Ranging (LiDAR) data in smaller agricultural areas \citep{roelens, bailly} or smaller forest areas (smaller than 2 $km^2$) \citep{rapinel, kiss}, predominantly characterised by artificial drainage.
 In this study, however, we conduct a landscape-scale analysis in a mesoscale catchment.
@@
 \citet{cazorzi} used a DEM with image processing techniques aimed at detecting the elongated ditch shapes and excluding large landscape features.
-This approach was suitable for their open agricultural areas, but is difficult to generalise to a diverse forest landscape.
+This approach worked well for their open agricultural areas, but is difficult to generalise to a diverse forest landscape.
+Instead, we will focus on the topographical incision of the ditches in the DEM, which will be highlighted with a set of digital terrain indices.
 Previous studies have found that terrain indices such as Sky View Factor \citep{zaksek}, Impoundment Index \citep{whiteboxtools} and High Pass Median Filter \citep{whiteboxtools} can detect ditches in forest landscapes in some cases \citep{uppsala}.
@@
 We hypothesise that by combining the information from all the digital terrain indices using a machine learner (Random Forests), we can improve the detection of the ditches, compared to if a single terrain index is used.
-To test this, we conduct a landscape-scale ditch detection analysis at the Krycklan Catchment in Sweden \citep{krycklancatchment}, where a high precision manual mapping has been performed previously.
+To test this, we conduct a landscape-scale ditch detection analysis at the Krycklan Catchment in Sweden \citep{krycklancatchment}, where a high precision manual ditch mapping has been performed.
 This manual mapping guided the model training and ground-truthing in our analysis and provided a test-bench for our methodology.
@@
 To develop and evaluate our ditch detector, the raster and ditch label data of Krycklan were manually divided into 21 subsections.
-\hl{This was necessary because our method relies on the pixels surrounding a pixel, meaning it is not possible to select random pixels for training and testing.} Each subsection represents an area of roughly 196 hectare.
+\hl{This was necessary because our method relies on the area surrounding a pixel being examined, meaning it is not possible to select random pixels for training and testing.} Each subsection represents an area of roughly 196 hectare.
 From this division, 11 of the subsections were put aside as hold-out data, only for use in the experiment to evaluate the performance of the predictions.
@@
 \begin{table} [!htb]
+\centering
     {\begin{tabular}{r|ll}

### 2021-01-16 LaTeX ESWA/article.tex
Rewrites up to section 2.4.2 to be more suited for ESWA and more concise.

@@
 
-\title{Detecting ditches using \hl{supervised} learning on high-resolution \hl{digital elevation models}}
+\title{Detecting ditches using supervised learning on high-resolution digital elevation models}
 
@@
 {\footnotesize
-\hl{Drained wetlands can constitute a large source of greenhouse gas emissions, but the drainage networks in these wetlands are largely unmapped, and better maps are needed to aid in forest management decisions, and to predict the climate consequences.} We develop a method for detecting ditches in high resolution digital elevation models derived from LiDAR scans.
-\hl{Thresholding methods using digital terrain indices can be used to detect ditches,} however, a single threshold generally does not capture the variability in the landscape, and generates many false positives and negatives.
-We hypothesise that, by combining the digital terrain indices using \hl{supervised} learning, we can improve ditch detection at a landscape-scale.
-In addition to digital terrain indices, additional features are generated by transforming the data to include neighbouring cells \hl{for better ditch predictions}.
-\hl{A Random Forests classifier} is used to locate the ditches, and \hl{its probability output} is processed to remove noise, and binarised to produce the final ditch prediction.
+Drained wetlands can constitute a large source of greenhouse gas emissions, but the drainage networks in these wetlands are largely unmapped, and better maps are needed to aid in forest management decisions, and to predict the climate consequences.
+We develop a method for detecting ditches in high resolution digital elevation models derived from LiDAR scans.
+Thresholding methods using digital terrain indices can be used to detect ditches, however, a single threshold generally does not capture the variability in the landscape, and generates many false positives and negatives.
+We hypothesise that, by combining the digital terrain indices using supervised learning, we can improve ditch detection at a landscape-scale.
+In addition to digital terrain indices, additional features are generated by transforming the data to include neighbouring cells for better ditch predictions.
+A Random Forests classifier is used to locate the ditches, and its probability output is processed to remove noise, and binarised to produce the final ditch prediction.
 The confidence interval for the Cohen's Kappa index ranges \textbf{[0.655 , 0.781]} between the evaluation plots with a confidence level of \textbf{95\%}.
-The study demonstrates that combining information from  a suite of digital terrain indices using machine learning provides an effective technique for automatic ditch detection at a landscape-scale\hl{, aiding in both practical forest management and in combatting climate change}.}
+The study demonstrates that combining information from  a suite of digital terrain indices using machine learning provides an effective technique for automatic ditch detection at a landscape-scale, aiding in both practical forest management and in combatting climate change.}
 \end{abstract}
@@
 The land use is dominated by forestry; approximately 25\% of the Krycklan catchment has been protected since 1922, but  the rest of the area is mostly second growth forest \citep{krycklancatchment}.
-
 A historical survey of the catchment (Gudrun Norstedt, unpublished) recorded  that the ditches in the catchment were dug between 1900 and 1993.
@@
 The LiDAR dataset used in this study was produced by TerraTec Sweden AB in August 2015, on demand of the Department of Forest Resource Management at the Swedish University of Agricultural Sciences, and is freely available online \citep{dataset}.
-The average density of the dataset was 20 pulses per $m^2$.
+The average density of the dataset is 20 pulses per $m^2$.
 Ground point classification was performed with lasground (included in LAStools, a software suite produced by rapidlasso GmbH).
@@
 \subsection{Digitising the ground truth}
-To train and validate the machine learning models, a vector layer of all ditches in the Krycklan Catchment \citep{krycklancatchment} was digitised.
+For training and validation, a vector layer of all ditches in the Krycklan Catchment \citep{krycklancatchment} was digitised.
 The study catchment comprises several types of ditches including road ditches, forest ditches, and agricultural ditches.
-A multidirectional hillshade of the 0.5 m DEM was calculated in ArcGIS Pro to locate the ditches.
-From this layer, most of the ditches could be easily identified.
-We conducted the digitisation at a scale of 1:600 to ensure all the digitised ditches properly align the actual ditch on the ground and do not coincide with the edge of the banks.
+A multidirectional hillshade of the 0.5 m DEM was calculated in ArcGIS Pro, and from this layer, most of the ditches could be easily identified.
+We conducted the digitisation at a scale of 1:600 to ensure that all the digitised ditches are properly aligned with the actual ditch on the ground, and do not coincide with the edge of the banks.
 In addition, both current and historical records and orthophotos dating back to the 1960's were consulted to help locate all the ditches.
 In uncertain areas, the ditches were verified in the field.
-The stream network in \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan} shows a network with a 8 hectare flow initiation threshold.
 The digitisation of the ditch networks showed that there are 107 km of road ditches and 208 km of forest/agricultural ditches in the catchment (\hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan}).
-The stream network in \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan}, which represents median flow conditions, was 197 km.
+The stream network in \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan} shows a 197 km long network with a 8 hectare flow initiation threshold, representing median flow conditions.
 
@@
     \includegraphics[width=1\linewidth]{./images/Krycklan_lo.jpg}
-    \caption{Map of Krycklan showing ditches and stream channels, as well as our division into 21 subsections.
+    \caption{Map of Krycklan showing ditches and stream channels, as well as our division into 21 subsections of $2997 \cdot 2620$ pixels (roughly 196 hectare).
 The 10 subsections with a white border were used for developing features and optimising parameters before the experiment.
-The 11 subsections with a black border were used in the experiment for evaluation with 11-fold cross validation.
-Each rectangular subsection represents $2997 \cdot 2620$ pixels (roughly 196 hectare).}
+The 11 subsections with a black border were used in the experiment for evaluation with 11-fold cross validation.}
     \label{fig:swedenkrycklan}
@@
 
-From the ditch mapping, the vector layer was rasterised so that it could be compared to the automatically derived ditches from the $0.5*0.5$ m DEM.
-Our field observations have shown that the  width of most ditches within the catchment ranges between 0.5 and 3.5 m.
-To ensure that the model received all ditch pixels correctly in the training phase, we widened the ditches so that all pixels within a radius of three pixels (1.5 metres) of the vectors were labelled as ditch pixels.
-\hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{a} shows the ditches rasterised from vectors and \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{b} shows the ditches after widening.
-The data in \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{b} is the labelled data that was used to train the machine learning models.
-The ditches all have a generalised width based on a perceived average of ditch width in the area.
+From the ditch mapping, the vector layer was rasterised with a resolution of $0.5*0.5$ metres for use as a ground-truth for our ditch detector.
+Our field observations have shown that most of the ditches in the catchment range between 0.5 and 3.5 metres in width.
+To ensure that all ditch pixels were labelled correctly for the model in the training phase, we labelled all pixels within three pixels (1.5 metres) of the vectors as ditch pixels, making the labels 3.5 metres wide (\hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess}).
 Due to all ditches varying in width, it was impossible to produce a perfect representation of each ditch, but this widening helped ensure that almost all ditch pixels were covered by the labels.
-
-To match the labels with out predictions (where we downscale the predictions into grids to remove outliers), the ground-truth was also downscaled into grids of $6*6$ pixels (9 $m^2$) each.
-Each grid cell that contained at least 25\% ditch pixels was labelled as a ditch, see \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{c}.
-This layer was used to ground-truth the machine learning models.
+To match the labels with our predictions (where we downscale the predictions into grids to remove outliers), the ground-truth was also downscaled into grids of $6*6$ pixels (9 $m^2$) each by labelling each grid cell containing at least 25\% ditch pixels as a ditch (\hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{c}).
 
@@
 \textbf{b: }Ditches after a widening process, seven pixels (3.5 metres) wide.
-Used as input when training the models.
+Used as labels when training the models.
 \textbf{c: }Ditches after a grid zone conversion.
@@
 \subsection{Extracting ditches with digital terrain indices}
-From the DEM, several digital terrain indices were calculated. The strengths and weaknesses of these indices have been obtained from observing their performance in different topological environments:
+From the DEM, several digital terrain indices were calculated:
 \begin{itemize} \label{data_attributes}
@@
     The Sky View Factor represents how much of the sky that is visible from a certain point on the ground \citep{zaksek}.
-A point in a ditch would yield a low value, as more of the sky is obscured by the ditch bank.
-However, it tends to produce false positives in steep terrain.
-The Sky View Factor was calculated in SAGA GIS, using a search radius of 10 metres to ensure that it was large enough to capture the ditches, but small enough to not include large scale topographical features such as hills.
+A point inside a ditch would produce a low value, as more of the sky is obscured by the ditch bank.
+Sky View Factor tends to produce false positives in steep terrain.
+The index was calculated in SAGA GIS, using a search radius of 10 metres to ensure that it was large enough to capture the ditches, but small enough to not include large scale topographical features such as hills.
 
   \item \textbf{Impoundment Index}\newline
-    The Impoundment Index in Whitebox Tools \citep{whiteboxtools} can be used as a measure of flow constriction, which is useful for mapping channels, such as ditches.
+    The Impoundment Index in Whitebox Tools \citep{whiteboxtools} can be used as a measure of flow constriction, which is useful for mapping channels such as ditches.
 When tuned to output the flooded volume, it effectively maps the area and average depth of the reservoir that would be created by placing a dam of a user-specified length at each grid cell in the DEM.
-Here, we used dam height calculated with a dam length of 3 metres to cover the entire ditches.
-Typical false positives with this method are natural stream channels with a clear incised channel.
+Here, we used dam height calculated with a dam length of 3 metres to attempt to cover the entire ditches.
+Typical false positives with this index are natural stream channels with a clear incised channel.
 
@@
     \item \textbf{Slope} \newline
-    The Slope index represents the degree of slope at a certain point, with a value ranging from $0^{\circ}$ to $90^{\circ}$.
+    The Slope index represents the degree of slope at a certain point, with a value ranging from $0^{\circ}$ to $90^{\circ}$ with no information about the direction of the slope.
 It was calculated on the 0.5 m grid, using ArcGIS \citep{EsriArcGisBook}.
-This index contains no information about the direction of the slope.
 \end{itemize}
@@
 This test was conducted for the 11 subsections described in \ref{trainingvalidationdatasets}.
-The Sky View Factor index was classified to only include values below 0.955, the Impoundment Index Dam Height using values above 0.27 $m$, the HPMF index using values below -0.18 m, and the Slope index using values above $14 ^{\circ}$ (Table 2).
+The Sky View Factor index was classified to only include values below 0.955, the Impoundment Index Dam Height using values above 0.27 $m$, the HPMF index using values below -0.18 m, and the Slope index using values above $14 ^{\circ}$ (\autoref{recreatedpredictionperformance}).
 
@@
 
-\subsubsection{Processing the digital terrain indices}
-We examined how different  features affected the prediction of the machine learning models.
-Several possible data manipulation methods could theoretically produce a better prediction.
-To improve the prediction, steps were taken where neighbouring pixels were included to give a representation of the area surrounding a specific pixel, similar to the approach of \citet{roelens}.
+\subsubsection{Feature engineering}
 
-The digital terrain indices provided a satisfactory foundation for the models, but lacked in the generalisability of their predictions.
-More diverse features were extracted using simple statistical aggregates such as mean, median, min, max, and standard deviation.
-Using these statistical aggregations with the help of neighbouring areas around pixels aided in pruning pixels with outlier values, often smoothing out the data to represent ditches more accurately on a per-pixel basis.
-These variables were calculated by gathering all data points in different circular radii around the studied pixel, before computing one of the statistical aggregations (\hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{b}, \hyperref[fig:features]{c}, \hyperref[fig:features]{f}, and \hyperref[fig:features]{g}).
+The digital terrain indices provided a satisfactory foundation for the models, but could not always generalise well to outlier situations.
+To give a representation of the area surrounding a specific pixel, we extracted more diverse features using simple statistical aggregates such as mean, median, min, max, and standard deviation in different circular radii around the pixel (similar to the approach of \citet{roelens}).
+This aided in pruning pixels with outlier values, often smoothing out the data to represent ditches more accurately on a per-pixel basis (\hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{b}, \hyperref[fig:features]{c}, \hyperref[fig:features]{f}, and \hyperref[fig:features]{g}).
+Additionally, several features were generated by using thresholds on one or multiple indices to exclude, lower, or amplify the values of pixels that indicated ditches or non-ditches (\hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{d} and \hyperref[fig:features]{h}).
 
-In addition to the general features, a number of custom features were also developed.
-Some of these custom variables were produced by statistical aggregation of multiple indices, whereas in some cases, they were generated by using thresholds to exclude, lower, or amplify the values of pixels that indicated ditches or non-ditches (\hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{d} and \hyperref[fig:features]{h}).
-
-Both HPMF and Sky View Factor were used with the image processing \textit{gabor filter}, which can be used to detect lines of a certain orientation in an image \citep{gabor}.
+Features were generated for both HPMF and Sky View Factor using the image processing \textit{gabor filter}, which can be used to detect lines of a certain orientation in an image \citep{gabor}.
 30 Gabor filters which were rotated in different angles and with different frequencies were combined to detect lines, amplifying ditches by utilising the fact that ditches have a linear elongated shape (\hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{e}).
@@
 \label{impoundmentstreamremoval}
-The raw Impoundment Index dam height was used to create a mask, attempting to mask out streams, but retain ditches.
-Only areas with a relatively high value, which would indicate that these areas most likely contained streams, were masked out.
-After widening the resulting area, this mask was used to remove streams from all the aforementioned custom features, generating one new feature from each.
+In an attempt to exclude streams from the predictions, the Impoundment Index was used to mask out areas with a very large dam height (which would indicate that the area most likely contained a stream).
+This mask was applied on several features to smooth out the values under the mask, generating one new feature from each of them.
 The features marked with \textit{streams removed} in \hyperref[featuretable]{Table} \ref{featuretable} make use of this filter (\hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{i}).
@@
 Overall, 81 features were extracted from the terrain indices.
-In a pilot study, we compared several different algorithms (Random Forests, Extreme Gradient Boosting, Naive Bayes, and Support Vector Machines), and it was found that Random Forests produced the most accurate results, here we therefore only present the results from the Random Forest model.
-We then conducted a sub-experiment to find the best features, as well as the optimal number of variables to use.
-The variable importance score in Random Forests was used to select the most important features and it was found that using the top 40 features produced the best results.
+In a pilot study, we compared several different algorithms (Random Forests, Extreme Gradient Boosting, Naive Bayes, and Support Vector Machines), and it was found that Random Forests produced the most accurate results.
+We then conducted a sub-experiment using Random Forests' feature importance to find the best features, as well as the optimal number of features to use, resulting in a final feature set of 40 features (\autoref{featuretable}).
 
@@
 To develop and evaluate our ditch detector, the raster and ditch label data of Krycklan were manually divided into 21 subsections.
-\hl{This was necessary because our method relies on the area surrounding a pixel being examined, meaning it is not possible to select random pixels for training and testing.} Each subsection represents an area of roughly 196 hectare.
+This was necessary because our method relies on the area surrounding a pixel being examined, meaning it is not possible to select random pixels for training and testing.
+Each subsection represents an area of roughly 196 hectare.
 From this division, 11 of the subsections were put aside as hold-out data, only for use in the experiment to evaluate the performance of the predictions.
@@
 \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan} shows which subsections were used for development and evaluation respectively.
-\hl{The 11 hold-out subsections were used as folds in an 11-fold cross validation in the final experiment.}
+The 11 hold-out subsections were used as folds in an 11-fold cross validation in the final experiment.
 
@@
     \end{tabular}}
-    \caption{ The 40 features used when training the models.
-    \newline \textsuperscript{a} Displays algorithm used to produce the feature. \newline
-        \textsuperscript{b} Represents the radius of the circular mask (if one was used) to determine \newline what neighbouring pixels to use in an algorithm.
-The radii represents pixels \newline \mbox{with a 0.5 m resolution.\itshape\ignorespaces}}
+    \caption{ The 40 features used to train the models.
+    \newline \textsuperscript{a} The algorithm used to produce the feature. \newline
+        \textsuperscript{b} Represents the radius of the circular mask (if one was used) to determine which neighbouring pixels to use in the aggregation.
+The radii represents pixels with a 0.5 m resolution.}
     \label{featuretable}
@@
     \end{tabular}}
-    \caption{Metrics from the total results of the four digital terrain indices experiments\hl{, where only a sole terrain index is used with a threshold to detect ditches}.}
+    \caption{Metrics from the total results of the four digital terrain indices experiments, where only a sole terrain index is used with a threshold to detect ditches.}
     \label{recreatedpredictionperformance}
@@
 Our developed methodology can be adjusted to apply to any setting for artificial ditch detection, reducing time and labour investment compared to the conventional manual mapping techniques.
-\hl{The accurate ditch maps can both significantly contribute to practical forestry and operational land management, as well as allow stakeholders including government agencies, forest companies, and regional and local authorities to predict the climate consequences of potential management actions, which will be critical in reaching global sustainability goals.}
+The accurate ditch maps can both significantly contribute to practical forestry and operational land management, as well as allow stakeholders including government agencies, forest companies, and regional and local authorities to predict the climate consequences of potential management actions, which will be critical in reaching global sustainability goals.
 
 \section*{CRediT authorship contribution statement}
-\hlc{Jag tolkar det som att denna sektion är frivillig.
+\textbf{Jag tolkar det som att denna sektion är frivillig.
 De flesta artiklar jag har hittat listar här vad alla enskilda författare har bidragit med.
@@
 This work was supported by VINNOVA, EU InterReg Baltic Sea project WAMBAF tools and Formas.
-\hlc{ \textbf{ Här ska varje grant definieras med "grant number[xxxx] efter varje aktör"} }
+\textbf{ Här ska varje grant definieras med "grant number[xxxx] efter varje aktör"}
 

### 2021-01-16 LaTeX/test.tex
Rewrites up to section 2.4.2 to be more suited for ESWA and more concise.

--- 2020-10-10 LaTeX/anonymised.tex
@@
 
-\author{Anonymised.}
+\author{
+\name{Filip Andersson\textsuperscript{a,1}, Jonatan Flyckt\textsuperscript{a,1}, Niklas Lavesson\textsuperscript{a}\thanks{CONTACT Niklas Lavesson.
+Email: niklas.lavesson@ju.se}, Liselott Nilsson\textsuperscript{b}, and Anneli M.
+\AA gren\textsuperscript{c}} \affil{\textsuperscript{a}J\"onk\"oping Artificial Intelligence Laboratory, J\"onk\"oping University, Gjuterigatan 5, 551 11, J\"onk\"oping, Sweden; \textsuperscript{b}Forest Department, Swedish Forest Agency, Skeppargatan 17, 931 32 Skellefte\aa, Sweden; \textsuperscript{c}Department of Forest Ecology and Management, Swedish University of Agricultural Sciences, SLU, Skogsmarksgr\"and 17, 901 83, Ume\aa, Sweden}
+\footnote{\textsuperscript{1} Filip Andersson and Jonatan Flyckt contributed equally to this work as first authors.}}
 
@@
 
-%Detection of artificial ditches in a study area with substantial natural stream coverage is challenging but highly desired for effective forest management.
-This study developed and applied a methodology using multiple terrain indices and machine learning for automatic ditch detection at the Krycklan catchment in Sweden.
-Despite a substantial vegetation (87\% forest cover), our method achieved good results; the confidence interval for the Cohen's Kappa index ranged [0.655 , 0.781] between the evaluation plots with a confidence level of 95\%.
-With the developed input variables and extensive post-processing, we managed to remove many stream channels and fill gaps in the ditches where the LiDAR scans were weak.
-However, these issues are still present to some extent.
-Future work could introduce more input variables and post-processing steps, such as using pathfinding algorithms to fill gaps in the ditch model, or shape indices to remove stream channels from the prediction.
-Moving towards image segmentation with grids of pixels instead of pixel classification of tabular pixel values could also be a suitable future approach.
-With minor adjustments, our proposed method can be employed in a general setting, reducing manual labour and cost, and enabling faster and easier mapping of ditches.
-
 Detection of artificial ditches over large areas with substantial natural stream coverage is challenging, especially under tree canopy, but highly desired for effective forest management.
@@
 \label{lidartodem}
-The LiDAR dataset used in this study was produced by (anonymised), and is freely available online at \href{https://gis.krycklan.se/}{https://gis.krycklan.se/}.
+The LiDAR dataset used in this study was produced by TerraTec Sweden AB in August 2015, on demand of the Department of Forest Resource Management at the Swedish University of Agricultural Sciences, and is freely available online at \href{https://gis.krycklan.se/}{https://gis.krycklan.se/}.
 The average density of the dataset was 20 pulses per $m^2$.
@@
 \section*{Acknowledgements}
-Anonymised.
+We wish to thank the two anonymous reviewers and Dr.
+Siddhartho Paul at the Swedish University of Agricultural Sciences, whose suggestions helped improve and clarify this manuscript.
+We also thank Sigurd Israelsson at the School of Engineering at J\"onk\"oping University for lending hardware resources for the experiments in this study.
+The work has been funded by VINNOVA, EU InterReg Baltic Sea project WAMBAF tools and Formas.
 

### 2021-01-17 LaTeX ESWA/article.tex
Completes read-through and adds many attempts at more concise language.

@@
 
-  \item \textbf{Sky View Factor} \newline
+  \item \textbf{Sky View Factor} \label{skyviewfactor} \newline
     The Sky View Factor represents how much of the sky that is visible from a certain point on the ground \citep{zaksek}.
@@
 
-  \item \textbf{Impoundment Index}\newline
+  \item \textbf{Impoundment Index} \label{impoundment} \newline
     The Impoundment Index in Whitebox Tools \citep{whiteboxtools} can be used as a measure of flow constriction, which is useful for mapping channels such as ditches.
@@
 
-  \item \textbf{High Pass Median Filter} \newline
+  \item \textbf{High Pass Median Filter} \label{hpmf} \newline
     High Pass Median Filters (HPMF) highlight ditches by indicating local depressions (such as ditches) in the DEM by subtracting the value at the grid cell at the centre of the window from the median value in the window.
@@
 
-    \item \textbf{Slope} \newline
+    \item \textbf{Slope} \label{slope} \newline
     The Slope index represents the degree of slope at a certain point, with a value ranging from $0^{\circ}$ to $90^{\circ}$ with no information about the direction of the slope.
@@
         \resizebox*{3.5cm}{!}{\includegraphics{./images/feature_i_lo.jpg}}}
-    \caption{Example of 8 of the 40 features, in addition to ditch labels used by the models for a small sample area.
+    \caption{\textbf{Feature examples.} Shows the ditch labels as well as 8 of the 40 features for a small sample area.
 The radii represents pixels with a 0.5 m resolution.
@@
 \label{trainingvalidationdatasets}
-To develop and evaluate our ditch detector, the raster and ditch label data of Krycklan were manually divided into 21 subsections.
+To develop and evaluate our ditch detector, the raster and ditch label data of Krycklan were manually divided into 21 subsections, each representing an area of roughly 196 hectare.
 This was necessary because our method relies on the area surrounding a pixel being examined, meaning it is not possible to select random pixels for training and testing.
-Each subsection represents an area of roughly 196 hectare.
-From this division, 11 of the subsections were put aside as hold-out data, only for use in the experiment to evaluate the performance of the predictions.
-The remaining 10 subsections were used solely before the experiment to develop the ditch detector to test different ways of preparing the data.
-This allowed the ditch detector to be evaluated on  independent data and  strengthened the validity of the experiment.
+11 of the subsections were put aside as hold-out data, only for use in the experiment to evaluate the performance of the ditch detector.
+The remaining 10 subsections were used solely before the experiment to develop the ditch detector, and to test different ways of preparing the data.
+This allowed the ditch detector to be evaluated on independent data , strengthening the validity of the experiment.
 \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan} shows which subsections were used for development and evaluation respectively.
@@
 For our study, a hyperparameter tuning showed that using 300 trees and \textit{gini} as the splitting criterion, with a minimum of 10 samples per node split, and no artificial class weight or max depth of trees yielded the best predictions.
+The testing phase showed that, due to the class imbalance (only 2\% of pixels being ditch pixels), the models were not being punished for mislabelling ditches as non-ditches.
+To combat this, we evened out the amount of ditch and non-ditch instances in the training data by first extracting all pixels labelled as ditches as well as pixels within close proximity of ditches.
+Second, pixels were sampled randomly from the entire subsection to still contain most of the geographical features of each area (\hyperref[fig:balancedmasks]{Figure} \ref{fig:balancedmasks}).
 
@@
     {\begin{tabular}{l|l}
-      \textbf{Terrain Index/Algorithm}\textsuperscript{a} & \textbf{Circular radii}\textsuperscript{b} \\ %\midrule
+      \textbf{Terrain Index/Algorithm}\textsuperscript{a} & \textbf{Circular radii}\textsuperscript{b} \\ 
       \hline
@@
 
-The testing phase showed that the classifiers produced poor results when the ratio of ditch versus non-ditch pixels in the training data was very high, as the models were not being punished for mislabelling ditches as non-ditches.
-To increase the recall, we evened out the amount of ditch and non-ditch instances in the training data.
-First, we extracted all pixels labelled as ditches as well as pixels within close proximity of ditches.
-Secondly, pixels were sampled randomly from the entire subsection to still contain most of the geographical features of each area (\hyperref[fig:balancedmasks]{Figure} \ref{fig:balancedmasks}).
-
 \begin{figure} [!htb]
@@
 \subsection{Post-processing}
-The models output a ditch class probability prediction for each pixel.
-Values close to zero indicate a low probability of a pixel lying in a ditch, whereas values close to one equals a high ditch probability (\hyperref[fig:postprocessing1]{Figure} \ref{fig:postprocessing1} \hyperref[fig:postprocessing1]{a}).
+
+The output from the model is a ditch probability ranging from zero to one for each pixel (\hyperref[fig:postprocessing1]{Figure} \ref{fig:postprocessing1} \hyperref[fig:postprocessing1]{a}).
 
@@
 
-To fill gaps in ditches that the models failed to correctly predict, we aggregated pixel values by covering them with cone masks expanding outwards in different directions from the examined pixel.
+To fill gaps in ditches that the model failed to correctly predict due to gaps in the LiDAR scans, we aggregated pixel values by covering them with cone masks expanding outwards in different directions from the examined pixel.
 This amplified some of the noise that was left, but filling the gaps in the ditches was judged to be more important (\hyperref[fig:postprocessing2]{Figure} \ref{fig:postprocessing2} \hyperref[fig:postprocessing2]{b}).
@@
 
-\subsubsection{Binarisation with grid zones}
-The models' ditch prediction is given in the original resolution raster format.
-To allow for comparison to the evaluation data, the same grid conversion was performed on the prediction raster as on the evaluation data displayed in \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{c}.
-After testing different combinations of zone sizes and probability values, a mean probability rating was calculated for each $6*6$ grid zone, classifying the entire zone as part of a ditch if the mean probability exceeded 40 \%.
-This zone partitioning helped to fill in gaps where lone pixels in ditches had been incorrectly classified.
-(\hyperref[fig:postprocessing3]{Figure} \ref{fig:postprocessing3} \hyperref[fig:postprocessing3]{a}).
+\subsubsection{Binarisation and cluster removal}
 
-\subsubsection{Cluster removal}
-To remove noise from the binary prediction, a custom cluster detection algorithm was used.
+To correct errors where lone pixels inside ditches had been incorrectly classified, the resolution was lowered by binary classifying each $6*6$ pixel grid as a ditch or a non-ditch if the mean probability of the pixels inside the grid exceeded 40 \% (\hyperref[fig:postprocessing3]{Figure} \ref{fig:postprocessing3} \hyperref[fig:postprocessing3]{a}).
+The same zone partitioning was performed on the labels (\hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{c}).
+
+A cluster detection algorithm was developed to remove noise from the final ditch prediction.
 By finding the number of connected pixels with a true value and removing those whose cluster size were below a given threshold, minor noise in the prediction could be removed while still retaining most of the ditch pixels.
-Actual ditches that have a low probability and therefore create small clusters may still be excluded by this method, but the noise removal advantages were judged to outweigh the loss in recall.
 A distance calculation was also performed in tandem with this method to find the largest distance of pixels inside each given cluster.
@@
 For this reason, the results were mainly evaluated using Cohen's Kappa (Cohen's $\kappa$) index, and the Area under Precision-Recall curve.
-Cohen's $\kappa$ index measures how much better a prediction is compared to a completely random prediction, where random would yield a value of zero \citep{kappa123}.
-With our data, a $\kappa$ value close to zero would be attained by predicting 2 \% of the occurrences as ditch pixels completely at random.
+Cohen's $\kappa$ index measures how much better a prediction is compared to a completely random prediction, where predicting 2 \% of the occurrences as ditch pixels completely at random would yield a value of zero \citep{kappa123}.
 Values above zero are better than random, and values below zero are worse than random.
@@
 The Precision-Recall curve and the Area under Precision-Recall curve (AUPRC) are additional metrics that can be used when evaluating datasets with a largely imbalanced class distribution \citep{precision_recall_curve}.
-The Precision-Recall curve has the recall value on the x-axis and the precision value on the y-axis, and the area under the curve that is defined by this point gives the AUPRC value.
-The area under this curve is given as a value between zero and one, where a value closer to one is better.
 The weighting causes the Precision-Recall curve to not place an equal value on true negatives and true positives \citep{precision_recall_curve}.
@@
 
-To circumvent some of the performance evaluation issues arising from the use of pixel classification for ditch objects as well as not having completely accurate labels on a pixel basis (due to uncertainties in the width of the ditches), we modified the evaluation labels to allow for some error in close proximity to ditches in the prediction.
+To circumvent some of the evaluation issues arising from the use of pixel classification for ditch objects, as well as not having completely accurate labels on a pixel basis (due to uncertainties in the width of the ditches), we modified the evaluation labels to allow for some error in close proximity to ditches in the prediction.
 False negative and false positive grid zones that lay adjacent to the ditch label grid zones were evaluated as true negatives and true positives, as they can be considered a part of correctly located ditch objects (\hyperref[fig:newlabels]{Figure} \ref{fig:newlabels}).
@@
 
-\hyperref[recreatedpredictionperformance]{Table} \ref{recreatedpredictionperformance} shows the evaluation metrics from the experiments with the digital terrain indices separately.
-Of the four, the Impoundment Index and the HPMF index performed relatively close to each other and outperformed the other indices for most of the metrics.
+\hyperref[recreatedpredictionperformance]{Table} \ref{recreatedpredictionperformance} shows the evaluation metrics from the experiments where the digital terrain indices were thresholded separately.
+Of the four, the Impoundment Index and the HPMF index performed relatively close to each other, and outperformed the other indices for most of the metrics.
 
@@
 \hyperref[predictionperformance]{Table} \ref{predictionperformance} displays all the evaluation metrics for the prediction of our method.
-The confidence intervals were calculated from the 11 different subsections that the experiment was performed on.
-Because the subsections have varying amounts of ditches in them, the confidence intervals will not be completely accurate, but will produce a close estimation.
-Averaging the results from the 11 subsections would yield a very similar value to the total results, indicating that the ditch detector generally performs equally well on subsections with a small amount of ditches as on subsections with a large amount of ditches.
+The confidence intervals were calculated from the results from 11 cross-validation subsections.
+Because the subsections have varying amounts of ditches in them, the confidence intervals will not be completely conclusive, but will produce a close estimation.
+Averaging the results from the 11 subsections yields a very similar value to the total results, indicating that the ditch detector generally performs equally well on subsections with a small amount of ditches as on subsections with a large amount of ditches.
 
@@
 
-The $\kappa$ rating for our method can be seen in the \textit{Total} column in \hyperref[predictionperformance]{Table} \ref{predictionperformance} ($\kappa$ = 0.732).
-The $\kappa$ ratings from the four digital terrain indices experiments can be seen in  \hyperref[recreatedpredictionperformance]{Table} \ref{recreatedpredictionperformance} ($\kappa$ = 0.048, $\kappa$ = 0.215, $\kappa$ = 0.211, $\kappa$ = 0.029).
-Because the $\kappa$ rating from our method outperforms all digital terrain indices, our hypothesis that combining digital terrain indices with a machine learning produces better ditch detection than using single terrain indices has been confirmed by the experiments.
-Most false positives lie in stream channels (\hyperref[fig:resultsillustrations]{Figure} \ref{fig:resultsillustrations} \hyperref[fig:resultsillustrations]{a, b}) and most false negatives occurred either due to ditches being too shallow (\hyperref[fig:resultsillustrations]{Figure} \ref{fig:resultsillustrations} \hyperref[fig:resultsillustrations]{c, d}), or due to the attempt at removing streams using the features as explained in \ref{impoundmentstreamremoval} removing the deepest ditches  (\hyperref[fig:resultsillustrations]{Figure} \ref{fig:resultsillustrations} \hyperref[fig:resultsillustrations]{e, f}).
+The $\kappa$ rating from our method ($\kappa$ = 0.732) (\hyperref[predictionperformance]{Table} \ref{predictionperformance}), which is in the substantial range, according to $\kappa$ performance thresholds proposed by \citet{kappaanalysis}, outperforms all digital terrain indices ($\kappa$ = 0.048, $\kappa$ = 0.215, $\kappa$ = 0.211, $\kappa$ = 0.029) (\hyperref[recreatedpredictionperformance]{Table} \ref{recreatedpredictionperformance}).
+This confirms our hypothesis that combining digital terrain indices with machine learning produces better ditch detection than using a single thresholded terrain index.
+Most false positives lie in stream channels (\hyperref[fig:resultsillustrations]{Figure} \ref{fig:resultsillustrations} \hyperref[fig:resultsillustrations]{a, b}), and most false negatives occur either due to ditches being too shallow (\hyperref[fig:resultsillustrations]{Figure} \ref{fig:resultsillustrations} \hyperref[fig:resultsillustrations]{c, d}), or due to the deepest ditches being removed as a result of the attempt at removing streams as explained in \hyperref[impoundmentstreamremoval]{section} \ref{impoundmentstreamremoval} (\hyperref[fig:resultsillustrations]{Figure} \ref{fig:resultsillustrations} \hyperref[fig:resultsillustrations]{e, f}).
 
@@
 \end{figure}
+\clearpage
 
-In \hyperref[featureimportancetable]{Table} \ref{featureimportancetable}, the top 20 (out of 40) features from the Random Forests models are presented with their importance percentages for the models making a successful prediction.
-features based on HPMF and Impoundment Index contributed the most to a successful prediction.
-The Gini importance was obtained using the Gini impurity for each variable \citep{gini}.
+In \hyperref[featureimportancetable]{Table} \ref{featureimportancetable}, the top 20 (out of 40) features from the Random Forests models are presented with their Gini importances \citep{gini}.
+Features derived from the \hyperref[hpmf]{HPMF} and \hyperref[impoundment]{Impoundment} terrain indices contributed the most to a successful prediction.
+The features that used statistical aggregation methods on the neighbouring area around pixels, as well as the stream removal and Gabor filter features also performed well.
 
@@
     \end{tabular}}
-    \caption{The top 20 features (out of 40) by importance when the models make a prediction.
+    \caption{The top 20 features (out of 40) by Gini importance.
 \newline \textsuperscript{a} The number next to some of the variables indicates the circular radius used to select what neighbouring pixels to use in the statistical aggregation method.
@@
 \end{table}
-\clearpage
 
@@
     a{
-        \resizebox*{5.5cm}{!}{\includegraphics{./images/road_ditch_lo.jpg}}}\hspace{7pt}
+        \resizebox*{6.65cm}{!}{\includegraphics{./images/road_ditch_lo.jpg}}}\hspace{7pt}
     b{
-        \resizebox*{5.5cm}{!}{\includegraphics{./images/mire_ditch_lo.jpg}}}
+        \resizebox*{6.65cm}{!}{\includegraphics{./images/mire_ditch_lo.jpg}}}
     \newline c{
-        \resizebox*{5.5cm}{!}{\includegraphics{./images/forest_ditch_lo.jpg}}}\hspace{7pt}
+        \resizebox*{6.65cm}{!}{\includegraphics{./images/forest_ditch_lo.jpg}}}\hspace{7pt}
     d{
-        \resizebox*{5.5cm}{!}{\includegraphics{./images/overgrown_ditch_lo.jpg}}}
-    \caption{All photos show ditches in the study catchment.
-They illustrate the difficulty of capturing the ditches based on their form (i.e.
+        \resizebox*{6.65cm}{!}{\includegraphics{./images/overgrown_ditch_lo.jpg}}}
+    \caption{\textbf{Ditches in the study catchment.} These photos illustrate the difficulty of capturing the ditches based on their form (i.e.
 an elongated, narrow, hollow in the ground).
@@
 Such ditches can be found in the field as the vegetation differs, but as there is no longer an elevation difference between the ditch and the surrounding soils it will not be detected in the DEM.
-Photos a-c: E.
-Maher Hasselquist  Photo d: G.
-Norstedt}
+\newline Photos a-c: E.
+Maher Hasselquist, Photo d: G.
+Norstedt.}
     \label{fig:ditchpictures}
@@
 \citet{rapinel} found that the results were usually more sensitive to the point density (ranging 1-4 points $m^{2}$ in their study) than the interpolation method.
-They retrieved 54.8 and 63.8 \% of the ditches on the Acuey and Boucey marches in France, compared to 70.28 \% in our study.
-However, in our study we had a point density of 20 points $m^{2}$, so the quality of the DEM in our study ought to be robust, which may have contributed to our higher retrieval rate.
-\hyperref[fig:resultstreesbushes]{Figure} \ref{fig:resultstreesbushes} \hyperref[fig:resultstreesbushes]{(a, b)} illustrates the difference in the quality of the LiDAR data where the laser pulses have caught in trees or shrubs.
-When too few LiDAR points reach the ground, the DEM generation is forced to interpolate over a larger area, leading to inaccuracies in the ground DEM.
+They retrieved 54.8 and 63.8 \% of the ditches on the Aucey and Boucey marshes in France, compared to 70.28 \% in our study.
+However, in our study we had a LiDAR point density of 20 points $m^{2}$, so the quality of the DEM in our study ought to be robust, which may have contributed to our higher retrieval rate.
+\hyperref[fig:resultstreesbushes]{Figure} \ref{fig:resultstreesbushes} \hyperref[fig:resultstreesbushes]{(a, b)} illustrates the difference in the quality of the ground DEM when it is generated from LiDAR where the laser pulses have caught in trees or shrubs, and fewer points have reached the ground.
 
@@
         \resizebox*{5.5cm}{!}{\includegraphics{./images/results_trees_bushes_4_lo.jpg}}}
-    \caption{\textbf{a} and \textbf{b} illustrate that despite a point density of 20 points $m^2$, not all points reach the ground in places with dense vegetation.
+    \caption{\textbf{a} and \textbf{b} illustrate that despite a LiDAR point density of 20 points $m^2$, not all points reach the ground in places with dense vegetation.
 In \textbf{a} there is a clear difference between the smooth surface of the DEM in the open area (black arrow), and in the areas covered by trees and shrubs (red arrows), where the DEM is much coarser due to interpolation with a lower number of LiDAR points here.
-The orthophoto (\textbf{b}), clearly shows that trees and shrubs are growing in the ditches in this area.
-This makes it more difficult to detect ditches.
+The orthophoto (\textbf{b}), clearly shows that trees and shrubs are growing in the ditches in this area, making it more difficult to detect ditches.
 However, despite these coarser grid points in the DEM, \textbf{c} and \textbf{d} shows that the ditch detector still managed to capture a majority of the ditches (green), and only failed at certain points (red).}
@@
 
-With our method, natural streams were often classified as false positives (\hyperref[fig:resultsillustrations]{Figure} \ref{fig:resultsillustrations} \hyperref[fig:resultsillustrations]{b}) as our aim was focused on locating ditches.
-However, this indicates that our method could also be used to map small, previously unmapped natural stream channels.
+A common false positive with our ditch detector is natural streams (\hyperref[fig:resultsillustrations]{Figure} \ref{fig:resultsillustrations} \hyperref[fig:resultsillustrations]{b}).
+However, this indicates that with some adjustments, our method could also be used to map small, previously unmapped natural stream channels.
 The national survey showed that 55\% of the natural stream channels and 75\% of the straightened water courses were missing on the maps (\autoref{fig:watercoursebarplot}).
@@
 
-The results from our method (\hyperref[predictionperformance]{Table} \ref{predictionperformance}) show that we managed to attain a Cohen's $\kappa$ rating in the substantial range, according to performance thresholds proposed by \citet{kappaanalysis}.
-The total recall value of 70.28 \% shows that we managed to find a majority of the ditch pixels that exist.
-The results of our method also showed a significant improvement for ditch detection over classifying with the digital terrain indices separately (\hyperref[recreatedpredictionperformance]{Table} \ref{recreatedpredictionperformance}).
+Sometimes pixels in ditches were initially identified correctly as ditch pixels, but because not enough pixels in the surrounding area were detected, the post-processing algorithms identified them as noise and removed them.
+However, these steps were necessary as they helped to remove small sinks or hilly areas that had been incorrectly classified as ditches.
 Had we been able to use labels that were correctly labelled on a pixel basis, i.e.
 each ditch had its actual width recorded, the models could have learned with a better accuracy in the training phase, and the classification results would also most likely have benefitted from this.
+We managed to bridge gaps and exclude streams to some extent in our predictions.
+This was in part due to the post-processing of the predictions helping where the model was unsuccessful on its own, and in part due to the features developed.
 
-Sometimes pixels in ditches were initially identified correctly as ditch pixels, but because not enough pixels in the surrounding area were detected, the post-processing algorithms identified them as noise and removed them.
-Incorrectly classified ditch pixels that caused the precision to decrease were generally the pixels that lay in streams (\hyperref[fig:resultsillustrations]{Figure} \ref{fig:resultsillustrations} \hyperref[fig:resultsillustrations]{a, b}).
-However, small sinks or hilly areas incorrectly classified as ditches by the Random Forests models were generally removed in the post-processing.
-We managed to bridge gaps and exclude streams to some extent in our predictions.
-This was in part due to the post-processing of the prediction helping where the models were unsuccessful on their own, and in part due to the features developed.
-\hyperref[featureimportancetable]{Table} \ref{featureimportancetable} highlights the importance of the features that were developed by applying different statistical aggregations for neighbouring areas of pixels (similar to \citet{roelens}) as well as the custom features using filters such as the Impoundment stream removal or Gabor filters.
-
-One concern with our method was that valuable information may have been discarded when aggregating the elevations from the original point cloud to a DEM.
+One concern with our method is that valuable information may have been discarded when aggregating the elevations from the original point cloud to a DEM.
 \citet{roelens} used  machine learning to detect ditches directly from point-clouds instead of a DEM.
@@
 \citet{bailly} also showed that the vegetation cover significantly affects the ability to detect ditches from LiDAR data (\hyperref[fig:ditchpictures]{Figure} \ref{fig:ditchpictures} \& \ref{fig:resultstreesbushes}).
-Approximately 75\% of the ditches in their study were retrieved where no vegetation existed, whereas the detection accuracy went below 10\% under vegetation; especially when the ditches were covered by high vegetation, shrubs, and trees \citep{bailly}.
-Hence, a $\kappa$ of 0.73 in our study  with  87\% forest cover \citep{krycklancatchment} is substantially better than that of previous research, and this could perhaps be a result of the higher point density of our LiDAR measurements (20 compared to 10 points $m^{2}$).
+Approximately 75\% of the ditches in their study were retrieved where no vegetation existed, whereas the detection accuracy went below 10\% under the cover of high vegetation, shrubs, and trees \citep{bailly}.
+Thus, a $\kappa$ of 0.73 in our study catchment with 87\% forest cover \citep{krycklancatchment} is substantially better than that of previous research.
+This could possibly be a result of the higher point density of our LiDAR measurements (20 compared to 10 points per $m^{2}$).
 The hydrology of our study catchment also differs from that of other studies, as we have almost as many natural stream channels as ditches, whereas their studies were conducted in smaller, predominantly artificially drained areas \citep{bailly, roelens, rapinel}.
-Because our method locates both ditches and natural stream channels, we have to take additional measures to remove the stream channels (affecting ditch detection negatively), as explained in \ref{impoundmentstreamremoval}.
+Because our model locates both ditches and natural stream channels, we have to take additional measures to remove the stream channels (affecting ditch detection negatively), as explained in \hyperref[impoundmentstreamremoval]{section} \ref{impoundmentstreamremoval}.
 
-A future improvement to the ditch detector could be to use a shape index to remove small natural streams, as natural streams often form more complex channels whereas ditches are often straight.
+A future improvement to the ditch detector could be to use a shape index to remove small natural streams, as natural streams often form complex channels, whereas ditches are often straight.
 A potential issue with generalising our ditch detector for use in other geographical areas is that all the post-processing steps and features were developed based on occurrences in the Krycklan area.
@@
 Our developed methodology can be adjusted to apply to any setting for artificial ditch detection, reducing time and labour investment compared to the conventional manual mapping techniques.
-The accurate ditch maps can both significantly contribute to practical forestry and operational land management, as well as allow stakeholders including government agencies, forest companies, and regional and local authorities to predict the climate consequences of potential management actions, which will be critical in reaching global sustainability goals.
+The accurate ditch maps can both significantly contribute to practical forestry and operational land management, as well as allow government agencies, forest companies, and regional and local authorities to predict the climate consequences of potential management actions, which will be critical in reaching global sustainability goals.
 
 \section*{CRediT authorship contribution statement}
+
 \textbf{Jag tolkar det som att denna sektion är frivillig.
@@
 \section*{Declaration of Competing Interest}
+
 The authors declare that they have no known competing financial interests or personal relationships that could have appeared to influence the work reported in this paper.

### 2021-01-17 LaTeX ESWA/article.tex
Completes draft of restructuring article for ESWA.

@@
 \captionsetup{font=footnotesize}
+
+\usepackage{subfigure}
 
@@
 J\"onk\"oping Artificial Intelligence Laboratory, J\"onk\"oping University, Gjuterigatan 5, 551 11, J\"onk\"oping, Sweden \\
-Tel: (+46) 70-5383338 \textbf{Ska detta nummer stå här? (corresponding author måste ha telefonnummer)} \\
+Tel: (+46) 70-5383338 \hl{Ska detta nummer st\aa  h\"ar? (corresponding author m\aa ste ha telefonnummer)} \\
 Email: niklas.lavesson@ju.se
@@
 {\footnotesize
-Drained wetlands can constitute a large source of greenhouse gas emissions, but the drainage networks in these wetlands are largely unmapped, and better maps are needed to aid in forest management decisions, and to predict the climate consequences.
+Drained wetlands can constitute a large source of greenhouse gas emissions, but the drainage networks in these wetlands are largely unmapped, and better maps are needed to aid in forest production and to better understand the climate consequences.
 We develop a method for detecting ditches in high resolution digital elevation models derived from LiDAR scans.
@@
 This simple investigation highlights two things: First, ditches make up the majority of the small scale water channels in the Swedish landscape; almost twice as many as the natural streams.
-Second, most of the small scale channels are missing on the current maps; only 45 \% of the natural watercourses, 25 \% of the straightened watercourses/channels, and 9 \% of the ditches were mapped (\hyperref[fig:watercoursebarplot]{Figure} \ref{fig:watercoursebarplot}).
+Second, most of the small scale channels are missing on the current maps; only 45 \% of the natural watercourses, 25 \% of the straightened watercourses/channels, and 9 \% of the ditches are mapped (\hyperref[fig:watercoursebarplot]{Figure} \ref{fig:watercoursebarplot}).
 
@@
     \end{tikzpicture}
-    \caption{The bars indicate the number of observed channels in the National Inventory of Landscapes in Sweden's (NILS) line inventory of small watercourses (width $<$ 6 m).
+    \caption{\textbf{Channel mapping status.} The bars indicate the number of observed channels in the National Inventory of Landscapes in Sweden's (NILS) line inventory of small watercourses (width $<$ 6 m).
 The colours of the bars indicate whether they are mapped or not on the Swedish Property map.}
@@
 Instead, we will focus on the topographical incision of the ditches in the DEM, which will be highlighted with a set of digital terrain indices.
-Previous studies have found that terrain indices such as Sky View Factor \citep{zaksek}, Impoundment Index \citep{whiteboxtools} and High Pass Median Filter \citep{whiteboxtools} can detect ditches in forest landscapes in some cases \citep{uppsala}.
+Previous studies have found that terrain indices such as \hyperref[skyviewfactor]{Sky View Factor} \citep{zaksek}, \hyperref[impoundment]{Impoundment Index} \citep{whiteboxtools} and High Pass Median Filter \citep{whiteboxtools} can detect ditches in forest landscapes in some cases \citep{uppsala}.
 However, different false positives and negatives can be observed when examining each terrain index manually.
 These inaccuracies can stem either from the gaps in the LiDAR scans, or from a poorly set threshold for the terrain index parameters.
-Thus, it would be useful to combine these LiDAR derived terrain indices to prune the weaknesses, and amplify the strengths of each index and produce highly accurate detection of ditch networks.
+Thus, it would be useful to combine these terrain indices to prune the weaknesses and amplify the strengths of each index to produce a highly accurate detection of ditch networks.
 
@@
 The study catchment comprises several types of ditches including road ditches, forest ditches, and agricultural ditches.
-A multidirectional hillshade of the 0.5 m DEM was calculated in ArcGIS Pro, and from this layer, most of the ditches could be easily identified.
+A multidirectional hillshade of the 0.5 m DEM was calculated in ArcGIS Pro, and from this layer, most of the ditches could be easily identified by a human.
 We conducted the digitisation at a scale of 1:600 to ensure that all the digitised ditches are properly aligned with the actual ditch on the ground, and do not coincide with the edge of the banks.
@@
     \includegraphics[width=1\linewidth]{./images/Krycklan_lo.jpg}
-    \caption{Map of Krycklan showing ditches and stream channels, as well as our division into 21 subsections of $2997 \cdot 2620$ pixels (roughly 196 hectare).
+    \caption{\textbf{Krycklan channel map.} Shows ditches and stream channels in the Krycklan catchment, as well as our division into 21 subsections of $2997 \cdot 2620$ pixels (roughly 196 hectare each).
 The 10 subsections with a white border were used for developing features and optimising parameters before the experiment.
@@
     \centering
-    a{
+    \subfigure[]{
         \resizebox*{3.5cm}{!}{\includegraphics{./images/publ_ditch_preprocess_A_lo.jpg}}}\hspace{5pt}
-    b{
+    \subfigure[]{
         \resizebox*{3.5cm}{!}{\includegraphics{./images/publ_ditch_preprocess_B_lo.jpg}}}
-    c{
+    \subfigure[]{
         \resizebox*{3.5cm}{!}{\includegraphics{./images/publ_ditch_preprocess_C_lo.jpg}}}
-    \caption{Processing of ditch labels.
-\textbf{a: }Rasterised ditches with a width of one pixel.
+    \caption{\textbf{Processing of ditch labels.} \textbf{a: }Rasterised ditches with a width of one pixel.
 \textbf{b: }Ditches after a widening process, seven pixels (3.5 metres) wide.
@@
 Different threshold values were tested and the value that yielded the highest Cohen's Kappa score for each respective index was chosen.
-This test was conducted for the 11 subsections described in \ref{trainingvalidationdatasets}.
-The Sky View Factor index was classified to only include values below 0.955, the Impoundment Index Dam Height using values above 0.27 $m$, the HPMF index using values below -0.18 m, and the Slope index using values above $14 ^{\circ}$ (\autoref{recreatedpredictionperformance}).
+This test was conducted for the 11 subsections described in \hyperref[trainingvalidationdatasets]{section} \ref{trainingvalidationdatasets}.
+\hyperref[skyviewfactor]{Sky View Factor} was classified to only include values below 0.955, the \hyperref[impoundment]{Impoundment Index} Dam Height using values above 0.27 $m$, \hyperref[hpmf]{HPMF} using values below -0.18 m, and the \hyperref[slope]{Slope} index using values above $14 ^{\circ}$ (\autoref{recreatedpredictionperformance}).
 
@@
 
-Features were generated for both HPMF and Sky View Factor using the image processing \textit{gabor filter}, which can be used to detect lines of a certain orientation in an image \citep{gabor}.
+Features were generated for both \hyperref[hpmf]{HPMF} and \hyperref[skyviewfactor]{Sky View Factor} using the image processing \textit{gabor filter}, which can be used to detect lines of a certain orientation in an image \citep{gabor}.
 30 Gabor filters which were rotated in different angles and with different frequencies were combined to detect lines, amplifying ditches by utilising the fact that ditches have a linear elongated shape (\hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{e}).
@@
 \label{impoundmentstreamremoval}
-In an attempt to exclude streams from the predictions, the Impoundment Index was used to mask out areas with a very large dam height (which would indicate that the area most likely contained a stream).
+In an attempt to exclude streams from the predictions, the \hyperref[impoundment]{Impoundment Index} was used to mask out areas with a very large dam height (which would indicate that the area most likely contained a stream).
 This mask was applied on several features to smooth out the values under the mask, generating one new feature from each of them.
@@
     \centering
-    a{
+    \subfigure[]{
         \resizebox*{3.5cm}{!}{\includegraphics{./images/feature_a_lo.jpg}}}
-    b{
+    \subfigure[]{
         \resizebox*{3.5cm}{!}{\includegraphics{./images/feature_b_lo.jpg}}}
-    c{
+    \subfigure[]{
         \resizebox*{3.5cm}{!}{\includegraphics{./images/feature_c_lo.jpg}}}
-    \newline d{
+    \subfigure[]{
         \resizebox*{3.5cm}{!}{\includegraphics{./images/feature_d_lo.jpg}}}
-    e{
+    \subfigure[]{
         \resizebox*{3.5cm}{!}{\includegraphics{./images/feature_e_lo.jpg}}}
-    f{
+    \subfigure[]{
         \resizebox*{3.5cm}{!}{\includegraphics{./images/feature_f_lo.jpg}}}
-    \newline g{
+    \subfigure[]{
         \resizebox*{3.5cm}{!}{\includegraphics{./images/feature_g_lo.jpg}}}
-    h{
+    \subfigure[]{
         \resizebox*{3.5cm}{!}{\includegraphics{./images/feature_h_lo.jpg}}}
-    i{
+    \subfigure[]{
         \resizebox*{3.5cm}{!}{\includegraphics{./images/feature_i_lo.jpg}}}
@@
 \label{trainingvalidationdatasets}
-To develop and evaluate our ditch detector, the raster and ditch label data of Krycklan were manually divided into 21 subsections, each representing an area of roughly 196 hectare.
-This was necessary because our method relies on the area surrounding a pixel being examined, meaning it is not possible to select random pixels for training and testing.
+To develop and evaluate our ditch detector, the raster and ditch label data of Krycklan were manually divided into 21 subsections, each representing an area of roughly 196 hectare (\hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan}).
+This was necessary because our method relies on the area surrounding a pixel for feature extraction, meaning it is not possible to select random pixels for training and testing.
 11 of the subsections were put aside as hold-out data, only for use in the experiment to evaluate the performance of the ditch detector.
 The remaining 10 subsections were used solely before the experiment to develop the ditch detector, and to test different ways of preparing the data.
-This allowed the ditch detector to be evaluated on independent data , strengthening the validity of the experiment.
+This allowed the ditch detector to be evaluated on independent data, strengthening the validity of the experiment.
 \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan} shows which subsections were used for development and evaluation respectively.
@@
 For our study, a hyperparameter tuning showed that using 300 trees and \textit{gini} as the splitting criterion, with a minimum of 10 samples per node split, and no artificial class weight or max depth of trees yielded the best predictions.
-The testing phase showed that, due to the class imbalance (only 2\% of pixels being ditch pixels), the models were not being punished for mislabelling ditches as non-ditches.
+The testing phase showed that, due to the class imbalance (only 2\% of pixels being ditch pixels), the model was not being punished for mislabelling ditches as non-ditches.
 To combat this, we evened out the amount of ditch and non-ditch instances in the training data by first extracting all pixels labelled as ditches as well as pixels within close proximity of ditches.
-Second, pixels were sampled randomly from the entire subsection to still contain most of the geographical features of each area (\hyperref[fig:balancedmasks]{Figure} \ref{fig:balancedmasks}).
+Second, pixels were sampled randomly from the entire subsection to still capture most of the geographical features of each area (\hyperref[fig:balancedmasks]{Figure} \ref{fig:balancedmasks}).
 
@@
 \centering
-    {\begin{tabular}{l|l}
+    {\begin{tabular}{ll}
       \textbf{Terrain Index/Algorithm}\textsuperscript{a} & \textbf{Circular radii}\textsuperscript{b} \\ 
@@
     \end{tabular}}
-    \caption{ The 40 features used to train the models.
+    \caption{\textbf{Feature set.} The 40 features used to train the models.
     \newline \textsuperscript{a} The algorithm used to produce the feature. \newline
@@
     \centering
-    a{
+    \subfigure[]{
         \resizebox*{5.6cm}{!}{\includegraphics{./images/publ_balanced_masks_A_lo.jpg}}}\hspace{5pt}
-    b{
+    \subfigure[]{
         \resizebox*{5.6cm}{!}{\includegraphics{./images/publ_balanced_masks_B_lo.jpg}}}
-    \caption{Black pixels here indicate the balanced masks used to determine what pixels were used when training the models. \textbf{a, b: }Examples of two subsection from the training dataset.}
+    \caption{\textbf{Balancing training data.} Black pixels here indicate the balanced masks used to determine what pixels were used when training the models.
+\textbf{a, b: }Examples of two subsection from the training dataset.}
     \label{fig:balancedmasks}
@@
 
+\newpage
 \subsection{Post-processing}
@@
     \centering
-    a{
+    \subfigure[]{
         \resizebox*{5.6cm}{!}{\includegraphics{./images/publ_post_process_step_1_lo.jpg}}}\hspace{5pt}
-    b{
+    \subfigure[]{
         \resizebox*{5.6cm}{!}{\includegraphics{./images/publ_post_process_step_2_lo.jpg}}}
-    \caption{Step one and two of the post-process. Darker pixels indicate a higher ditch probability. \textbf{a: }Random Forests probability prediction. \textbf{b: }Bilateral de-noising.}
+    \caption{\textbf{Post-processing step one and two.} Darker pixels indicate a higher ditch probability. \textbf{a: }Random Forests probability prediction. \textbf{b: }Bilateral de-noising.}
     \label{fig:postprocessing1}
@@
     \centering
-    a{
+    \subfigure[]{
         \resizebox*{5.6cm}{!}{\includegraphics{./images/publ_post_process_step_3_lo.jpg}}}\hspace{5pt}
-    b{
+    \subfigure[]{
         \resizebox*{5.6cm}{!}{\includegraphics{./images/publ_post_process_step_4_lo.jpg}}}
-    \caption{Step three and four of the post-process. Darker pixels indicate a higher ditch probability. \textbf{a: }Custom de-noising. \textbf{b: }Gap filling.}
+    \caption{\textbf{Post-processing step three and four.} Darker pixels indicate a higher ditch probability. \textbf{a: }Custom de-noising. \textbf{b: }Gap filling.}
     \label{fig:postprocessing2}
@@
 A cluster detection algorithm was developed to remove noise from the final ditch prediction.
-By finding the number of connected pixels with a true value and removing those whose cluster size were below a given threshold, minor noise in the prediction could be removed while still retaining most of the ditch pixels.
+By finding the number of connected pixels with a true value and removing those whose cluster size were below a threshold, minor noise in the prediction could be removed while still retaining most of the ditch pixels.
 A distance calculation was also performed in tandem with this method to find the largest distance of pixels inside each given cluster.
@@
     \centering
-    a{
+    \subfigure[]{
         \resizebox*{5.6cm}{!}{\includegraphics{./images/publ_post_process_step_5_lo.jpg}}}\hspace{5pt}
-    b{
+    \subfigure[]{
         \resizebox*{5.6cm}{!}{\includegraphics{./images/publ_post_process_step_6_lo.jpg}}}
-    \caption{Step five and six of the post-process. Black pixels indicate a ditch prediction. \textbf{a: }Grid zone binarisation \textbf{b: }Cluster removal.}
+    \caption{\textbf{Post-processing step five and six.} Black pixels indicate a ditch prediction. \textbf{a: }Grid zone binarisation \textbf{b: }Cluster removal (final ditch prediction).}
     \label{fig:postprocessing3}
@@
 The Precision-Recall curve and the Area under Precision-Recall curve (AUPRC) are additional metrics that can be used when evaluating datasets with a largely imbalanced class distribution \citep{precision_recall_curve}.
-The weighting causes the Precision-Recall curve to not place an equal value on true negatives and true positives \citep{precision_recall_curve}.
+The weighting causes the Precision-Recall curve to place an unequal weight on true negatives and true positives \citep{precision_recall_curve}.
 For our ditch detection problem, this means that the AUPRC evaluation metric favours accurately classifying ditch pixels over accurately classifying non-ditch pixels.
@@
     \centering
-    a{
-        \resizebox*{6.5cm}{!}{\includegraphics{./images/re_evaluation_1_lo.jpg}}}\hspace{5pt}
-    b{
-        \resizebox*{6.5cm}{!}{\includegraphics{./images/re_evaluation_2_lo.jpg}}}
-    \caption{Illustration of the modified evaluation labels.
-Green marks true positives, red marks false negatives, and blue marks false positives.
+    \subfigure[]{
+        \resizebox*{6.65cm}{!}{\includegraphics{./images/re_evaluation_1_lo.jpg}}}\hspace{5pt}
+    \subfigure[]{
+        \resizebox*{6.65cm}{!}{\includegraphics{./images/re_evaluation_2_lo.jpg}}}
+    \caption{\textbf{Modifying evaluation labels.} Green marks true positives, red marks false negatives, and blue marks false positives.
 False positives and false negatives that lay within one grid zone (9 $m^2$) of a ditch label were evaluated as true positives and true negatives.
@@
 \hyperref[recreatedpredictionperformance]{Table} \ref{recreatedpredictionperformance} shows the evaluation metrics from the experiments where the digital terrain indices were thresholded separately.
-Of the four, the Impoundment Index and the HPMF index performed relatively close to each other, and outperformed the other indices for most of the metrics.
+Of the four, the \hyperref[impoundment]{Impoundment Index} and \hyperref[hpmf]{HPMF} performed relatively close to each other, and outperformed the other indices for most of the metrics.
 
@@
 \centering
-    {\begin{tabular}{l|cccc}
-        \textbf{Metric} & \textbf{Sky View Factor} & \textbf{Impoundment} & \textbf{HPMF} & \textbf{Slope}\\
+    {\begin{tabular}{lcccc}
+        \textbf{Metric} & \textbf{\hyperref[skyviewfactor]{Sky View Factor}} & \textbf{\hyperref[impoundment]{Impoundment}} & \textbf{\hyperref[hpmf]{HPMF}} & \textbf{\hyperref[slope]{Slope}}\\
         \hline
@@
     \end{tabular}}
-    \caption{Metrics from the total results of the four digital terrain indices experiments, where only a sole terrain index is used with a threshold to detect ditches.}
+    \caption{\textbf{Index thresholding metrics.} Metrics from the total results of the four digital terrain indices experiments, where only a sole terrain index is used with a threshold to detect ditches.}
     \label{recreatedpredictionperformance}
@@
 
-\hyperref[predictionperformance]{Table} \ref{predictionperformance} displays all the evaluation metrics for the prediction of our method.
-The confidence intervals were calculated from the results from 11 cross-validation subsections.
+\hyperref[predictionperformance]{Table} \ref{predictionperformance} displays all the evaluation metrics for the prediction of our ditch detector.
+The confidence intervals were calculated from the results from the 11 cross-validation subsections.
 Because the subsections have varying amounts of ditches in them, the confidence intervals will not be completely conclusive, but will produce a close estimation.
@@
 \centering
-    {\begin{tabular}{l|ccc}
+    {\begin{tabular}{lccc}
         \textbf{Metric} & \textbf{Total}\textsuperscript{a} & \textbf{Subsection Average}\textsuperscript{b}& \textbf{CI 95\%}\textsuperscript{c} \\ 
@@
     \end{tabular}}
-    \caption{Metrics for the prediction performance of our ditch detector. \newline
+    \caption{\textbf{Ditch detector performance metrics.} \newline
     \textsuperscript{a} The result of all 11 subsection experiments when combined. \newline
@@
 \begin{figure} [!htb]
-    a{
-        \resizebox*{5.3cm}{!}{\includegraphics{./images/results_illustration_1_lo.jpg}}}\hspace{5pt}
-    b{
-        \resizebox*{5.3cm}{!}{\includegraphics{./images/results_illustration_2_lo.jpg}}}
-    \newline c{
-        \resizebox*{5.3cm}{!}{\includegraphics{./images/results_illustration_3_lo.jpg}}}\hspace{5pt}
-    d{
-        \resizebox*{5.3cm}{!}{\includegraphics{./images/results_illustration_4_lo.jpg}}}
-    \newline e{
-        \resizebox*{5.3cm}{!}{\includegraphics{./images/results_illustration_5_lo.jpg}}}\hspace{5pt}
-    f{
-        \resizebox*{5.3cm}{!}{\includegraphics{./images/results_illustration_6_lo.jpg}}}
-    \caption{The left panel shows the hillshade for a number of subsections, and the right panel shows the hillshade with the extracted ditches superimposed on top.
+\centering
+    \subfigure[]{
+        \resizebox*{4cm}{!}{\includegraphics{./images/results_illustration_1_lo.jpg}}}\hspace{5pt}
+    \subfigure[]{
+        \resizebox*{4cm}{!}{\includegraphics{./images/results_illustration_2_lo.jpg}}}
+    \newline \subfigure[]{
+        \resizebox*{4cm}{!}{\includegraphics{./images/results_illustration_3_lo.jpg}}}\hspace{5pt}
+    \subfigure[]{
+        \resizebox*{4cm}{!}{\includegraphics{./images/results_illustration_4_lo.jpg}}}
+    \newline \subfigure[]{
+        \resizebox*{4cm}{!}{\includegraphics{./images/results_illustration_5_lo.jpg}}}\hspace{5pt}
+    \subfigure[]{
+        \resizebox*{4cm}{!}{\includegraphics{./images/results_illustration_6_lo.jpg}}}
+    \newline
+    \caption{\textbf{Visual analysis of ditch detector.} The left panel shows the hillshade for a number of subsections, and the right panel shows the hillshade with the extracted ditches superimposed on top.
 Green marks correctly classified ditches (true positives), red marks missed ditches (false negatives), blue marks incorrectly classified ditches (false positives), and transparent areas mark correctly classified non-ditches (true negatives).
@@
 \textbf{e} and \textbf{f} illustrate that some of the deepest and more prominent ditches were also missed.
-This occurred as a result of attempting to get rid of more false positives in the form of streams by using the features, as explained in \ref{impoundmentstreamremoval}.}
+This occurred as a result of attempting to get rid of more false positives in the form of streams by using the features explained in \ref{impoundmentstreamremoval}.}
     \label{fig:resultsillustrations}
 \end{figure}
-\clearpage
 
-In \hyperref[featureimportancetable]{Table} \ref{featureimportancetable}, the top 20 (out of 40) features from the Random Forests models are presented with their Gini importances \citep{gini}.
+In \hyperref[featureimportancetable]{Table} \ref{featureimportancetable}, the top 20 (out of 40) features from the Random Forests model are presented with their Gini importances \citep{gini}.
 Features derived from the \hyperref[hpmf]{HPMF} and \hyperref[impoundment]{Impoundment} terrain indices contributed the most to a successful prediction.
@@
     \end{tabular}}
-    \caption{The top 20 features (out of 40) by Gini importance.
+    \caption{\textbf{Feature importances.} The top 20 features (out of 40) by Gini importance.
 \newline \textsuperscript{a} The number next to some of the variables indicates the circular radius used to select what neighbouring pixels to use in the statistical aggregation method.
@@
 
-Our study has shown that it is possible to locate ditches automatically in high-resolution DEMs ($0.5  * 0.5 $ metres, in our case), and that more of the ditches can be detected if several terrain indices (\hyperref[recreatedpredictionperformance]{Table} \ref{recreatedpredictionperformance}) are combined through machine learning than if indices are used separately (\hyperref[predictionperformance]{Table} \ref{predictionperformance}).
+Our study has shown that it is possible to locate ditches automatically in high-resolution DEMs ($0.5  * 0.5 $ metres, in our case), and that more of the ditches can be detected if several terrain indices (\hyperref[predictionperformance]{Table} \ref{predictionperformance}) are combined through machine learning than if indices are used separately (\hyperref[recreatedpredictionperformance]{Table} \ref{recreatedpredictionperformance}).
 The ditch detection performance varied depending on the type of ditches, this is illustrated in \hyperref[fig:ditchpictures]{Figure} \ref{fig:ditchpictures}.
@@
 \begin{figure} [!htb]
-    a{
-        \resizebox*{6.65cm}{!}{\includegraphics{./images/road_ditch_lo.jpg}}}\hspace{7pt}
-    b{
-        \resizebox*{6.65cm}{!}{\includegraphics{./images/mire_ditch_lo.jpg}}}
-    \newline c{
-        \resizebox*{6.65cm}{!}{\includegraphics{./images/forest_ditch_lo.jpg}}}\hspace{7pt}
-    d{
-        \resizebox*{6.65cm}{!}{\includegraphics{./images/overgrown_ditch_lo.jpg}}}
+\centering
+    \subfigure[]{
+        \resizebox*{6.8cm}{!}{\includegraphics{./images/road_ditch_lo.jpg}}}
+    \subfigure[]{
+        \resizebox*{6.8cm}{!}{\includegraphics{./images/mire_ditch_lo.jpg}}}
+    \subfigure[]{
+        \resizebox*{6.8cm}{!}{\includegraphics{./images/forest_ditch_lo.jpg}}}
+    \subfigure[]{
+        \resizebox*{6.8cm}{!}{\includegraphics{./images/overgrown_ditch_lo.jpg}}}
     \caption{\textbf{Ditches in the study catchment.} These photos illustrate the difficulty of capturing the ditches based on their form (i.e.
@@
 \begin{figure}[!htb]
-    a{
+\centering
+    \subfigure[]{
         \resizebox*{5.5cm}{!}{\includegraphics{./images/results_trees_bushes_1_lo.jpg}}}\hspace{7pt}
-    b{
+    \subfigure[]{
         \resizebox*{5.5cm}{!}{\includegraphics{./images/results_trees_bushes_2_lo.jpg}}}
-    \newline c{
+    \subfigure[]{
         \resizebox*{5.5cm}{!}{\includegraphics{./images/results_trees_bushes_3_lo.jpg}}}\hspace{7pt}
-    d{
+    \subfigure[]{
         \resizebox*{5.5cm}{!}{\includegraphics{./images/results_trees_bushes_4_lo.jpg}}}
-    \caption{\textbf{a} and \textbf{b} illustrate that despite a LiDAR point density of 20 points $m^2$, not all points reach the ground in places with dense vegetation.
+    \caption{\textbf{Weak LiDAR scan illustration.} \textbf{a} and \textbf{b} illustrate that despite a LiDAR point density of 20 points $m^2$, not all points reach the ground in places with dense vegetation.
 In \textbf{a} there is a clear difference between the smooth surface of the DEM in the open area (black arrow), and in the areas covered by trees and shrubs (red arrows), where the DEM is much coarser due to interpolation with a lower number of LiDAR points here.
@@
 Had we been able to use labels that were correctly labelled on a pixel basis, i.e.
-each ditch had its actual width recorded, the models could have learned with a better accuracy in the training phase, and the classification results would also most likely have benefitted from this.
+each ditch had its actual width recorded, the model could have learned with a better accuracy in the training phase, and the classification results would also most likely have benefitted from this.
 We managed to bridge gaps and exclude streams to some extent in our predictions.
@@
 A future improvement to the ditch detector could be to use a shape index to remove small natural streams, as natural streams often form complex channels, whereas ditches are often straight.
-A potential issue with generalising our ditch detector for use in other geographical areas is that all the post-processing steps and features were developed based on occurrences in the Krycklan area.
+A potential issue with generalising our ditch detector for use in other geographical areas is that many of the post-processing steps and features were developed based on occurrences in the Krycklan area.
 Different geographical compositions in other areas may require the adjustment of the thresholds used to fill gaps in ditches and to remove noise.
@@
 Ditches represent elongated depressions in a DEM; however, the exact shape and size differs between ditches.
-We combined recently developed digital terrain indices (the most important being Impoundment Index and High Pass Medium Filter) but, unlike previous studies, did not apply generic thresholding on them.
+We combined recently developed digital terrain indices (the most important being \hyperref[impoundment]{Impoundment} Index and High Pass Medium Filter) but, unlike previous studies, did not apply generic thresholding on them.
 Instead, we included local variability in the indices by using different radii to capture ditches with varied shapes.
@@
 
-\textbf{Jag tolkar det som att denna sektion är frivillig.
-De flesta artiklar jag har hittat listar här vad alla enskilda författare har bidragit med.
-Kategorier som andra har med här är: Conceptualisation, Data curation, Formal analysis, Investigation, Methodology, Software, Validation, Visualisation, Writing - original draft, Writing - review & editing, Supervision, Project administration}
+\hl{ Jag tolkar det som att denna sektion \"ar frivillig, men ESWA skriver i guiden att de uppmuntrar att man fyller i denna.
+De flesta artiklar jag har hittat listar h\"ar vad alla enskilda f\"orfattare har bidragit med.
+Kategorier: Conceptualization; Data curation; Formal analysis; Funding acquisition; Investigation; Methodology; Project administration; Resources; Software; Supervision; Validation; Visualization; Roles/Writing - original draft; Writing - review & editing.}
 
@@
 This work was supported by VINNOVA, EU InterReg Baltic Sea project WAMBAF tools and Formas.
-\textbf{ Här ska varje grant definieras med "grant number[xxxx] efter varje aktör"}
+\hl{ H\"ar ska varje grant definieras med "grant number[xxxx] efter varje akt\"or", jag tolkar det ocks\aa som att man ska skriva vilken roll varje finansi\"ar har.}
 

### 2021-01-24 LaTeX ESWA/version-21-01-17.tex
Makes some adjustments to article after meeting. Adds article highlights.

--- 2021-01-17 LaTeX ESWA/article.tex
@@
 \small  
-$^a$ J\"onk\"oping Artificial Intelligence Laboratory, J\"onk\"oping University, Gjuterigatan 5, 551 11, J\"onk\"oping, Sweden \\
+$^a$ Department of Computer Science, School of Engineering, J\"onk\"oping University, Gjuterigatan 5, 553 18, J\"onk\"oping, Sweden \\
 $^b$ Forest Department, Swedish Forest Agency, Skeppargatan 17, 931 32 Skellefte\aa, Sweden \\
@@
 Niklas Lavesson \\
-J\"onk\"oping Artificial Intelligence Laboratory, J\"onk\"oping University, Gjuterigatan 5, 551 11, J\"onk\"oping, Sweden \\
-Tel: (+46) 70-5383338 \hl{Ska detta nummer st\aa  h\"ar? (corresponding author m\aa ste ha telefonnummer)} \\
+Department of Computer Science\newline School of Engineering, J\"onk\"oping University \newline Gjuterigatan 5, 553 18, J\"onk\"oping, Sweden \\
+Tel: (+46) 70-5383338 \\
 Email: niklas.lavesson@ju.se
@@
 \cortext[cor1]{Corresponding author.}
-\address[ju]{J\"onk\"oping Artificial Intelligence Laboratory, J\"onk\"oping University, Gjuterigatan 5, 551 11, J\"onk\"oping, Sweden}
+\address[ju]{Department of Computer Science, School of Engineering, J\"onk\"oping University, Gjuterigatan 5, 553 18, J\"onk\"oping, Sweden}
 \address[fa]{Forest Department, Swedish Forest Agency, Skeppargatan 17, 931 32 Skellefte\aa, Sweden}

### 2021-01-24 LaTeX ESWA/article.tex
Adds a read-through with grammar corrections and more concise writing.

@@
+\UseRawInputEncoding
+
 \documentclass[11pt, review]{elsarticle} % first argument is font-size for document
@@
 We develop a method for detecting ditches in high resolution digital elevation models derived from LiDAR scans.
-Thresholding methods using digital terrain indices can be used to detect ditches, however, a single threshold generally does not capture the variability in the landscape, and generates many false positives and negatives.
+Thresholding methods using digital terrain indices can be used to detect ditches.
+However, a single threshold generally does not capture the variability in the landscape, and generates many false positives and negatives.
 We hypothesise that, by combining the digital terrain indices using supervised learning, we can improve ditch detection at a landscape-scale.
@@
 
-Over the last century, the wetlands in North Europe and parts of North America have been substantially modified and ditched.
+This study presents a forestry decision support system through the mapping of ditch networks, enabling more efficient forest management and better climate change mitigation.
+The wetlands in North Europe and parts of North America have been substantially modified and ditched over the last century.
 In many places both the landscape and the channel network have also been heavily altered  by humans, especially in the agricultural landscapes, for example in mid-western United States \citep{passalacqua}.
-In North Europe, however, ditches have mainly been dug to turn flat peatlands into productive forests, resulting in profound impacts on the hydrology of this region.
-\citep{peatlands}.
+In North Europe, however, ditches have mainly been dug to turn flat peatlands into productive forests, resulting in profound impacts on the hydrology of this region \citep{peatlands}.
 In Sweden, which is one of the North European countries with the most extensive ditch network, over five million hectares of peatland have been drained for forestry in the 20th century.
-As a result, almost one million km of ditches have been introduced, which has doubled the length of watercourses across Sweden  \citep{hasselquist}.
+As a result, almost one million km of ditches have been introduced, doubling the length of watercourses across Sweden \citep{hasselquist}.
 
-The forest industry and agencies have a demand for better maps of the ditch networks in the forest landscape.
-Drained afforested peatlands contribute strongly to Swedish wood production, and consequently to the Swedish economy.
+The forest industry and agencies have a demand for better maps of ditch networks in the forest landscape.
+Drained afforested peatlands contribute greatly to Swedish wood production, and consequently to the Swedish economy.
 At the same time, Swedish authorities have identified drained wetlands as a major target for greenhouse gas mitigation \citep{greenhouse_skogsstyrelsen, greenhouse_jordbruksverket}.
 However, research on the effects of afforested drained wetlands on greenhouse gas emission is inconclusive, and they can constitute both a source and a sink for greenhouse gases \citep{greenhouse_yes, greenhouse_no}.
-There is a need for science-based tools to optimise both forest production and decision support systems for mitigation of greenhouse gas emissions from drained forested wetlands.
+Consequently, there is a need for science-based tools to optimise both forest production and decision support systems for mitigation of greenhouse gas emissions from drained forested wetlands.
 A key step to accomplish this is to map the ditch networks in the forest landscape.
@@
 The study  catchment  is 68 $km^2$, representing diverse landscape features and geomorphological characteristics.
-Additionally, the forest landscape introduces many challenges that are not present when detecting ditches along roads or in agricultural areas.
-The canopy cover in forest areas prevents the use of satellite imagery or aerial photos, as well as lowers the quality of the digital elevation models (DEMs), as fewer LiDAR points reach the ground, instead frequently catching in trees, bushes, or grass.
+Additionally, the forest landscape introduces many challenges that are not present when detecting ditches along roads or in agricultural areas: The canopy cover in forest areas prevents the use of satellite imagery or aerial photos, as well as lowers the quality of the digital elevation models (DEMs), as fewer LiDAR points reach the ground, instead frequently catching in trees, bushes, or grass.
 Forest ditches are also generally not maintained to the same extent as agricultural or road ditches, often being overgrown or covered by windthrows or windsnaps.
-Using raw LiDAR point clouds to detect ditches, as \citet{roelens} and \citet{bailly} did in their studies can thus be problematic in a forest landscape, where a DEM constructed to map the points on the ground should be more suitable.
+Using raw LiDAR point clouds to detect ditches, as \citet{roelens} and \citet{bailly} did in their studies can thus be problematic in a forest landscape.
+Instead, a DEM constructed to map the points on the ground should be more suitable.
+
+Traditionally, machine learning is used in the remote sensing field to identify objects in images.
+These images can be satellite images \citep{biomass_soil_moisture} or LiDAR derived DEMs \citep{integrated_airborne_lidar}.
 \citet{cazorzi} used a DEM with image processing techniques aimed at detecting the elongated ditch shapes and excluding large landscape features.
 This approach worked well for their open agricultural areas, but is difficult to generalise to a diverse forest landscape.
-Instead, we will focus on the topographical incision of the ditches in the DEM, which will be highlighted with a set of digital terrain indices.
+Another way of detecting ditches is to calculate certain digital terrain indices in Geographic Information Systems (GIS), and threshold them to highlight the ditches.
+In this study we combine the machine learning techniques normally used on images in the remote sensing field with the traditional GIS methods of calculating digital terrain indices, focusing on the topographical incision of the ditches in the DEM.
 Previous studies have found that terrain indices such as \hyperref[skyviewfactor]{Sky View Factor} \citep{zaksek}, \hyperref[impoundment]{Impoundment Index} \citep{whiteboxtools} and High Pass Median Filter \citep{whiteboxtools} can detect ditches in forest landscapes in some cases \citep{uppsala}.
-However, different false positives and negatives can be observed when examining each terrain index manually.
+However, different false positives and false negatives can be observed when examining each terrain index manually.
 These inaccuracies can stem either from the gaps in the LiDAR scans, or from a poorly set threshold for the terrain index parameters.
-Thus, it would be useful to combine these terrain indices to prune the weaknesses and amplify the strengths of each index to produce a highly accurate detection of ditch networks.
+Thus, it would be useful to combine these terrain indices to prune the weaknesses, and amplify the strengths of each index to produce a highly accurate detection of ditch networks.
 
-In this study, we introduce a new method for the automated mapping of ditches by integrating multiple high resolution terrain indices using a supervised learning approach.
+We introduce a new method for the automated mapping of ditches by integrating multiple high resolution terrain indices using a supervised learning approach.
 Several features and post-processing steps, specifically tailored for ditch detection are also developed and employed.
-These features aim to both highlight elongated shapes similarly to \citet{cazorzi}, as well as include a map of the neighbouring area as \citet{roelens} did in their study.
+The features aim to both highlight elongated shapes similarly to \citet{cazorzi}, as well as to include a map of the neighbouring area as \citet{roelens} did in their study.
 We hypothesise that by combining the information from all the digital terrain indices using a machine learner (Random Forests), we can improve the detection of the ditches, compared to if a single terrain index is used.
@@
 
-The LiDAR dataset used in this study was produced by TerraTec Sweden AB in August 2015, on demand of the Department of Forest Resource Management at the Swedish University of Agricultural Sciences, and is freely available online \citep{dataset}.
+The LiDAR dataset used in this study was produced by TerraTec Sweden AB in August 2015, on demand of the Department of Forest Resource Management at the Swedish University of Agricultural Sciences, and is freely available online \citep{dataset_raw}.
 The average density of the dataset is 20 pulses per $m^2$.
@@
 To ensure that all ditch pixels were labelled correctly for the model in the training phase, we labelled all pixels within three pixels (1.5 metres) of the vectors as ditch pixels, making the labels 3.5 metres wide (\hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess}).
-Due to all ditches varying in width, it was impossible to produce a perfect representation of each ditch, but this widening helped ensure that almost all ditch pixels were covered by the labels.
-To match the labels with our predictions (where we downscale the predictions into grids to remove outliers), the ground-truth was also downscaled into grids of $6*6$ pixels (9 $m^2$) each by labelling each grid cell containing at least 25\% ditch pixels as a ditch (\hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{c}).
+Due to all ditches varying in width, it was not possible to produce a perfect representation of each ditch, but this widening helped ensure that almost all ditch pixels were covered by the labels.
+To match the labels with our predictions (where we downscale the predictions into grids to remove outliers), the ground-truth was also downscaled into grids of $6*6$ pixels (9 $m^2$) by labelling each grid cell containing at least 25\% ditch pixels as a ditch (\hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{c}).
 
@@
 \textbf{b: }Ditches after a widening process, seven pixels (3.5 metres) wide.
-Used as labels when training the models.
+Used as labels when training the model.
 \textbf{c: }Ditches after a grid zone conversion.
@@
   \item \textbf{Impoundment Index} \label{impoundment} \newline
-    The Impoundment Index in Whitebox Tools \citep{whiteboxtools} can be used as a measure of flow constriction, which is useful for mapping channels such as ditches.
-When tuned to output the flooded volume, it effectively maps the area and average depth of the reservoir that would be created by placing a dam of a user-specified length at each grid cell in the DEM.
-Here, we used dam height calculated with a dam length of 3 metres to attempt to cover the entire ditches.
+    The Impoundment Size Index in Whitebox Tools \citep{whiteboxtools} inserts a digital dam with a user-defined length of cross-sectional profile for each cell in the DEM, calculating the size of the impoundment (area, volume, or depth) that is created as a result of the “digital damming” operation.
+The Impoundment Size Index is calculated in three main steps.
+For detecting the ditches we focus on the first part which calculates the height of a potential dam for each grid cell based on the topographic profile perpendicular to the flow line at the site.
+The algorithm rotates the artificial dam in four different orientations (N-S, NE-SW, E-W, SE-NW) to find the maximum cross sectional damming for each cell across the landscape.
+The dam height is thus a measure of the topographic incision and can be used to detect the small scale ditch channels.
+Here, we used the dam height calculated with a dam length of 3 metres to attempt to cover the entire ditches.
 Typical false positives with this index are natural stream channels with a clear incised channel.
@@
 \end{itemize}
-
+\newpage
 To benchmark our method against a non-machine learning method, we binarised the original four digital terrain indices using thresholds.
-Different threshold values were tested and the value that yielded the highest Cohen's Kappa score for each respective index was chosen.
+Different threshold values were tested, and the value that yielded the highest Cohen's Kappa score for each respective index was chosen.
 This test was conducted for the 11 subsections described in \hyperref[trainingvalidationdatasets]{section} \ref{trainingvalidationdatasets}.
@@
 
-The digital terrain indices provided a satisfactory foundation for the models, but could not always generalise well to outlier situations.
+The digital terrain indices provided a satisfactory foundation for the model, but were not always able to generalise well to the diverse landscape features across the study catchment.
 To give a representation of the area surrounding a specific pixel, we extracted more diverse features using simple statistical aggregates such as mean, median, min, max, and standard deviation in different circular radii around the pixel (similar to the approach of \citet{roelens}).
@@
 \label{impoundmentstreamremoval}
-In an attempt to exclude streams from the predictions, the \hyperref[impoundment]{Impoundment Index} was used to mask out areas with a very large dam height (which would indicate that the area most likely contained a stream).
-This mask was applied on several features to smooth out the values under the mask, generating one new feature from each of them.
+In an attempt to exclude ravines from the predictions, the \hyperref[impoundment]{Impoundment Index} was used to mask out areas with a very large dam height (which would indicate that the area most likely contained a ravine).
+This mask was applied on several features to smooth out the values under the mask, generating one new feature from each original feature.
+Although ravines are uncommon in the Swedish forest landscape, they are overrepresented in the Krycklan catchment.
+Approximately 56 \% of the catchment area is below the highest postglacial coastline, meaning that the postglacial sediments form a thick layer through which the larger streams have deeply incised channels, forming ravines and bluffs of up to 30 metres in height (\hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan}).
 The features marked with \textit{streams removed} in \hyperref[featuretable]{Table} \ref{featuretable} make use of this filter (\hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{i}).
@@
 The 11 hold-out subsections were used as folds in an 11-fold cross validation in the final experiment.
+The digital terrain indices and labels for the 21 subsections are available for download \citep{dataset_processed};
 
@@
 The testing phase showed that, due to the class imbalance (only 2\% of pixels being ditch pixels), the model was not being punished for mislabelling ditches as non-ditches.
-To combat this, we evened out the amount of ditch and non-ditch instances in the training data by first extracting all pixels labelled as ditches as well as pixels within close proximity of ditches.
+To combat this, we evened out the amount of ditch and non-ditch instances in the training data by undersampling non-ditch pixels: First, we extracted all pixels labelled as ditches as well as pixels within close proximity of ditches.
 Second, pixels were sampled randomly from the entire subsection to still capture most of the geographical features of each area (\hyperref[fig:balancedmasks]{Figure} \ref{fig:balancedmasks}).
@@
     \end{tabular}}
-    \caption{\textbf{Feature set.} The 40 features used to train the models.
+    \caption{\textbf{Feature set.} The 40 features used to train the model.
     \newline \textsuperscript{a} The algorithm used to produce the feature. \newline
@@
         \resizebox*{5.6cm}{!}{\includegraphics{./images/publ_balanced_masks_B_lo.jpg}}}
-    \caption{\textbf{Balancing training data.} Black pixels here indicate the balanced masks used to determine what pixels were used when training the models.
+    \caption{\textbf{Balancing training data.} Black pixels here indicate the balanced masks used to determine what pixels were used when training the model.
 \textbf{a, b: }Examples of two subsection from the training dataset.}
@@
 
-To correct errors where lone pixels inside ditches had been incorrectly classified, the resolution was lowered by binary classifying each $6*6$ pixel grid as a ditch or a non-ditch if the mean probability of the pixels inside the grid exceeded 40 \% (\hyperref[fig:postprocessing3]{Figure} \ref{fig:postprocessing3} \hyperref[fig:postprocessing3]{a}).
+To correct errors stemming from lone, incorrectly classified pixels inside ditches, the resolution was lowered by binary classifying each $6*6$ pixel grid as a ditch or a non-ditch if the mean probability of the pixels inside the grid exceeded 40 \% (\hyperref[fig:postprocessing3]{Figure} \ref{fig:postprocessing3} \hyperref[fig:postprocessing3]{a}).
 The same zone partitioning was performed on the labels (\hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{c}).
@@
 A cluster detection algorithm was developed to remove noise from the final ditch prediction.
-By finding the number of connected pixels with a true value and removing those whose cluster size were below a threshold, minor noise in the prediction could be removed while still retaining most of the ditch pixels.
-A distance calculation was also performed in tandem with this method to find the largest distance of pixels inside each given cluster.
+By finding the number of connected pixels with a true value and removing those whose cluster size were below a threshold, minor noise in the prediction could be removed, while still retaining most of the ditch pixels.
+A distance calculation was also performed in tandem with this method to find the largest distance between any two pixels inside each given cluster.
 This helped to remove sinks and hollows that were not removed by the initial small cluster removal, but that did not have a linear directional characteristic, indicating that they did not represent a ditch (\hyperref[fig:postprocessing3]{Figure} \ref{fig:postprocessing3} \hyperref[fig:postprocessing3]{b}).
@@
 Using only accuracy as an evaluation metric when dealing with an imbalanced dataset (roughly 98 \% of all pixels are non-ditch) would produce a poor performance assessment \citep{balanced}; by simply classifying all pixels as non-ditches, we would by default attain 98 \% accuracy.
-For this reason, the results were mainly evaluated using Cohen's Kappa (Cohen's $\kappa$) index, and the Area under Precision-Recall curve.
+For this reason, the results were mainly evaluated using Cohen's Kappa (Cohen's $\kappa$) index, and the Area under Precision-Recall curve (AUPRC).
 Cohen's $\kappa$ index measures how much better a prediction is compared to a completely random prediction, where predicting 2 \% of the occurrences as ditch pixels completely at random would yield a value of zero \citep{kappa123}.
@@
 
-The Precision-Recall curve and the Area under Precision-Recall curve (AUPRC) are additional metrics that can be used when evaluating datasets with a largely imbalanced class distribution \citep{precision_recall_curve}.
+The Precision-Recall curve and the AUPRC  are additional metrics that can be used when evaluating datasets with a largely imbalanced class distribution \citep{precision_recall_curve}.
 The weighting causes the Precision-Recall curve to place an unequal weight on true negatives and true positives \citep{precision_recall_curve}.
@@
 
+\newpage
 \hyperref[predictionperformance]{Table} \ref{predictionperformance} displays all the evaluation metrics for the prediction of our ditch detector.
@@
     \textsuperscript{a} The result of all 11 subsection experiments when combined. \newline
-    \textsuperscript{b} An average score from the 11 different subsections that the experiment was performed on. \newline
+    \textsuperscript{b} An average score from the 11 different experiment subsections. \newline
     \textsuperscript{c} Confidence intervals at 95 \% confidence level.}
@@
     \newline
-    \caption{\textbf{Visual analysis of ditch detector.} The left panel shows the hillshade for a number of subsections, and the right panel shows the hillshade with the extracted ditches superimposed on top.
+    \caption{\textbf{Visual analysis of the ditch detector.} The left panel shows the hillshade for a number of subsections, and the right panel shows the hillshade with the extracted ditches superimposed on top.
 Green marks correctly classified ditches (true positives), red marks missed ditches (false negatives), blue marks incorrectly classified ditches (false positives), and transparent areas mark correctly classified non-ditches (true negatives).
@@
 
+\newpage
 In \hyperref[featureimportancetable]{Table} \ref{featureimportancetable}, the top 20 (out of 40) features from the Random Forests model are presented with their Gini importances \citep{gini}.
@@
 For example, the retrieval rate was slightly higher for the road ditches compared to forest or agricultural ditches.
-This is likely due to the fact that  road ditches are found in open areas alongside roads, and also,   they are generally well maintained for optimum functioning (\hyperref[fig:ditchpictures]{Figure} \ref{fig:ditchpictures} \hyperref[fig:ditchpictures]{a}).
+This is likely due to the fact that  road ditches are found in open areas alongside roads, and also that they are generally well maintained for optimum functioning (\hyperref[fig:ditchpictures]{Figure} \ref{fig:ditchpictures} \hyperref[fig:ditchpictures]{a}).
 However, most of the ditches in Krycklan today are found below the canopy (e.g.
@@
 This, in combination with the surrounding peat having subsided \citep{heikurainen}, have rendered them almost impossible to detect in the DEM today (\hyperref[fig:ditchpictures]{Figure} \ref{fig:ditchpictures} \hyperref[fig:ditchpictures]{d}).
-Despite these issues we had a recall rate of 70.28\%, showing that the majority of the ditches could still be retrieved automatically.
+Despite these issues we achieved a recall rate of 70.28\%, showing that the majority of the ditches could still be retrieved automatically.
 
@@
 These are generally easily detected as they are usually found in open areas along the roads, and have often been maintained.
-\textbf{b: }A ditch in a mire; this is still easily detected, as there is no canopy and there is a clear elevation difference between the ditch and surrounding area.
+\textbf{b: }A ditch in a mire.
+This is still easily detected, as there is no canopy and there is a clear elevation difference between the ditch and surrounding area.
 \textbf{c: }A narrow, more or less  overgrown ditch in a dense forest stand.
@@
 They retrieved 54.8 and 63.8 \% of the ditches on the Aucey and Boucey marshes in France, compared to 70.28 \% in our study.
-However, in our study we had a LiDAR point density of 20 points $m^{2}$, so the quality of the DEM in our study ought to be robust, which may have contributed to our higher retrieval rate.
+However, in our study we had a LiDAR point density of 20 points $m^{2}$, so the quality of the DEM in our study ought to be robust, possibly contributing to our higher retrieval rate.
 \hyperref[fig:resultstreesbushes]{Figure} \ref{fig:resultstreesbushes} \hyperref[fig:resultstreesbushes]{(a, b)} illustrates the difference in the quality of the ground DEM when it is generated from LiDAR where the laser pulses have caught in trees or shrubs, and fewer points have reached the ground.
@@
 
-Detection of artificial ditches over large areas with substantial natural stream coverage is challenging, especially under tree canopy, but highly desired for effective forest management.
+The detection of artificial ditches over large areas with substantial natural stream coverage is challenging, especially under tree canopy, but highly desired for effective forest management.
 Ditches represent elongated depressions in a DEM; however, the exact shape and size differs between ditches.
@@
 Our approach enables accurate ditch detection in complex forested landscapes with tree canopy and many natural streams.
-Future work could introduce more features and post-processing steps, such as using pathfinding algorithms to fill gaps in the ditch model or shape indices to remove stream channels from the prediction.
+Future work could introduce more features and post-processing steps, such as using pathfinding algorithms to fill gaps in the ditch model, or shape indices to remove stream channels from the prediction.
 Moving towards image segmentation instead of pixel classification of tabular pixel values could also be a suitable future approach.
@@
 
-\hl{ Jag tolkar det som att denna sektion \"ar frivillig, men ESWA skriver i guiden att de uppmuntrar att man fyller i denna.
-De flesta artiklar jag har hittat listar h\"ar vad alla enskilda f\"orfattare har bidragit med.
-Kategorier: Conceptualization; Data curation; Formal analysis; Funding acquisition; Investigation; Methodology; Project administration; Resources; Software; Supervision; Validation; Visualization; Roles/Writing - original draft; Writing - review & editing.}
+\textbf{Jonatan Flyckt:} Methodology, Software, Validation, Formal analysis, Investigation, Writing - Original Draft, Visualization.
+\textbf{Filip Andersson:} Methodology, Software, Validation, Formal analysis, Investigation, Writing - Original Draft, Visualization.
+\textbf{Niklas Lavesson:} ToDo.
+\textbf{Liselott Nilsson:} Conceptualization, Resources, Data Curation.
+\textbf{Anneli M.
+Ågren:} Conceptualization, Methodology, Investigation, Writing - Review \& Editing, Visualization, Supervision, Funding acquisition.
 
@@
 This work was supported by VINNOVA, EU InterReg Baltic Sea project WAMBAF tools and Formas.
-\hl{ H\"ar ska varje grant definieras med "grant number[xxxx] efter varje akt\"or", jag tolkar det ocks\aa som att man ska skriva vilken roll varje finansi\"ar har.}
 
